% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% \usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  % \usepackage[]{microtype}
  % \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage{parskip}
\setlength{\parskip}{0.5em}
\usepackage{enumitem}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\hypertarget{the-incompleteness-of-observation}{%
\section{The Incompleteness of
Observation}\label{the-incompleteness-of-observation}}

\hypertarget{why-physics-biggest-contradiction-might-not-be-a-contradiction-at-all}{%
\subsubsection{Why Physics' Biggest Contradiction Might Not Be a
Contradiction at
All}\label{why-physics-biggest-contradiction-might-not-be-a-contradiction-at-all}}

\hypertarget{with-complete-mathematical-detail}{%
\subsubsection{With Complete Mathematical
Detail}\label{with-complete-mathematical-detail}}

\textbf{Alex Maybaum --- March 2026}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-worst-prediction-in-physics}{%
\subsection{The Worst Prediction in
Physics}\label{the-worst-prediction-in-physics}}

Physics has two spectacularly successful theories. Quantum mechanics
describes the behavior of atoms, particles, and light. General
relativity describes gravity, space, and time. Each has been confirmed
to extraordinary precision. They have never disagreed with any
experiment.

They disagree with each other.

Ask quantum mechanics how much energy empty space contains and it gives
you a staggering number: roughly \(10^{113}\) joules per cubic meter.
Ask general relativity the same question --- read the answer off the
expansion rate of the universe --- and you get about
\(6 \times 10^{-10}\) joules per cubic meter. The ratio is \(10^{122}\).
For scale, the number of atoms in the observable universe is about
\(10^{80}\). This is not a close call.

For decades, the assumption has been that something is deeply broken ---
that one or both calculations contain an error, and that finding the
mistake will point the way to a unified theory of everything.

This paper argues the opposite. Neither calculation is wrong. They
disagree because they \emph{must}. In fact, that massive \(10^{122}\)
discrepancy isn't a failure at all. It is the strongest piece of
existing evidence we have that quantum mechanics is not the fundamental
bedrock of reality, but an emergent description forced upon us by our
limited vantage point.

The argument is built from a chain of mathematical proofs, each feeding
into the next. This document explains what the paper claims, walks
through the logic of every major proof, and shows how they connect.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-starting-point-four-axioms-and-three-conditions}{%
\subsection{The Starting Point: Four Axioms and Three
Conditions}\label{the-starting-point-four-axioms-and-three-conditions}}

Every mathematical proof starts from assumptions, and the paper is
explicit about its four. None of them mention quantum mechanics. None of
them mention general relativity. They are:

\textbf{Axiom 1: Deterministic dynamics.} The universe evolves according
to definite rules. Given the complete state at one time, the state at
any other time is uniquely determined. Mathematically, the state lives
in a phase space \ensuremath{\Gamma}, and there's a Hamiltonian H\_tot that governs how
the state changes:

\[\frac{\partial \rho}{\partial t} = \{H_{\text{tot}}, \rho\}\]

The curly braces \{,\} are Poisson brackets --- the classical mechanics
version of ``how things change.'' The function \ensuremath{\rho} represents a
probability distribution, but that probability reflects the observer's
ignorance, not any fundamental randomness. Think of it like a billiard
table: if you knew the exact position and velocity of every ball, you
could predict the future perfectly.

\textbf{Axiom 2: Finiteness.} The system has finitely many
distinguishable states. There's a smallest meaningful size \ensuremath{\varepsilon} --- you
can't resolve anything smaller. This means the configuration space is
finite, not continuous. This matters because finite systems have a
property infinite systems don't --- they must eventually return to their
starting state (Poincaré recurrence). In Part I, \ensuremath{\varepsilon} is left unspecified.
In Part II, self-consistency forces \ensuremath{\varepsilon} = 2 l\_p (twice the Planck
length).

\textbf{Axiom 3: Causal partition.} The total phase space splits into
two pieces:

\[\Gamma = \Gamma_V \times \Gamma_H\]

\ensuremath{\Gamma}\_V is the visible sector (what the observer can access). \ensuremath{\Gamma}\_H is the
hidden sector (what they cannot). The total Hamiltonian splits
correspondingly:

\[H_{\text{tot}} = H_V + H_H + H_{\text{int}}\]

H\_V governs the visible sector alone. H\_H governs the hidden sector
alone. H\_int couples them --- it's how the two sectors talk to each
other. Without H\_int, the two sectors would evolve independently and
the observer would never feel the hidden sector's influence.

\textbf{Axiom 4: Classical probability.} The observer uses standard
Kolmogorov probability theory. No exotic probability theories, no
negative probabilities, no quantum probability --- just ordinary
probability. This is the axiom that makes the result surprising: we're
putting in classical probability and getting out quantum mechanics.

That's it. The claim is that quantum mechanics --- the Schrödinger
equation, the Born rule, superposition, entanglement, Bell inequality
violations --- follows from these four premises alone, given the right
conditions on the hidden sector:

\textbf{C1: Non-zero coupling (H\_int \ensuremath{\neq} 0).} The visible and hidden
sectors interact. Information flows between them. Without this, the
observer's room is perfectly isolated --- nothing interesting happens.

\textbf{C2: Slow bath (\ensuremath{\tau}\_S \ensuremath{\ll} \ensuremath{\tau}\_B).} The hidden sector evolves much
more slowly than the visible sector. \ensuremath{\tau}\_S is the timescale of
visible-sector processes; \ensuremath{\tau}\_B is the timescale of hidden-sector
processes. This is the \emph{opposite} of the usual assumption in
physics. Normally, people assume the environment is fast and chaotic (a
``heat bath'' that quickly forgets everything). Here, the environment is
slow and has a long memory. This is what makes the dynamics
non-Markovian.

\textbf{C3: Sufficient capacity (N\_H \ensuremath{\gg} N\_V).} The hidden sector has
many more degrees of freedom than the visible sector. There's enough
``room'' to store information about the visible sector's history without
running out of space.

The axioms set the stage. The conditions determine what kind of show
plays on it. The next section explains why the cosmological horizon
satisfies all three.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-observers-blind-spot}{%
\subsection{The Observer's Blind Spot}\label{the-observers-blind-spot}}

Light travels at a finite speed, and the universe has a finite age. Put
those together and every observer has a horizon --- a boundary beyond
which no signal has had time to arrive. Everything beyond that boundary
is ordinary physics: fields, particles, radiation. But it is
structurally inaccessible. Not because our telescopes aren't good
enough, but because the geometry of spacetime forbids it. No technology
that obeys the speed of light can reach past it.

This means every observer in the universe is in the same epistemic
situation: there are degrees of freedom --- a vast number of them ---
that influence what you measure but that you can never track. When you
write down the laws of physics for the things you \emph{can} see, you're
forced to average over everything you can't. You have to ``trace out''
the hidden sector.

Here's what that looks like concretely. The total system --- visible
plus hidden --- is deterministic. If you knew the complete state, you
could predict the future exactly. But you don't know the hidden part.
You know the visible state is \(x\), but there are many possible hidden
states compatible with \(x\), and each one sends \(x\) to a different
visible future. Hidden state \(h_1\) might send the particle left;
hidden state \(h_2\) might send it right. Since you can't tell which
hidden state you're in, the best you can do is assign probabilities:
average over all the possible hidden states, weighted by how likely each
one is. The result is a set of \emph{transition probabilities} --- the
chance that visible state \(x\) at time \(t_1\) becomes visible state
\(y\) at time \(t_2\). You've gone from a deterministic system you can't
fully see to a probabilistic one you can. That's a stochastic process,
and it's the only description available to any observer who can't access
the hidden sector.

The standard expectation is that this should produce something boring
--- classical, memoryless noise. And it would, if the hidden sector were
fast and forgettable, like air molecules bouncing off a grain of pollen.
Each kick is independent of the last. Physicists call this
\emph{Markovian} behavior.

But the hidden sector beyond the cosmological horizon is not like that.
It differs in three specific ways, and the paper proves that their
conjunction changes everything.

\textbf{It's coupled.} The horizon is not a static wall. Stress-energy
conservation enforces continuous dynamical correlations across it.
Matter crosses the horizon, and the horizon area adjusts in response to
interior energy density. Information flows in both directions.
(Condition C1.)

\textbf{It's slow.} The hidden sector's correlation time is set by the
Hubble timescale --- roughly \(10^{17}\) seconds, the age of the
universe. Any laboratory experiment operates on timescales of
\(10^{-15}\) seconds or shorter. The ratio is \(10^{-32}\). The hidden
sector cannot ``reset'' between your measurements. Every correlation it
picks up from one experiment is still there when the next one begins.
This is the \emph{opposite} of the standard Markovian regime, where the
environment decorrelates fast. Here, it never decorrelates at all.
(Condition C2.)

\textbf{It's vast.} The hidden sector has roughly \(10^{122}\)
independent degrees of freedom --- the Bekenstein-Hawking entropy of the
cosmological horizon. No experiment you could ever perform would
appreciably disturb its state. Its memory never saturates. (Condition
C3.)

A fast environment with vast capacity would wash out correlations
(Markovian noise). A slow environment with limited capacity would
eventually fill up and stop recording. Only an environment that is
simultaneously coupled, slow, and vast sustains the kind of persistent,
non-decomposable correlations that the paper calls
\emph{P-indivisibility} --- a technical term meaning the system's
transition probabilities at different times cannot be broken into
independent steps.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{partition-relativity-1.4}{%
\subsection{Partition-Relativity
(§1.4)}\label{partition-relativity-1.4}}

This is the first real proof in the paper, and it's beautifully simple.

\textbf{What it proves:} The emergent description (what the observer
sees) depends \emph{only} on the partition --- on which degrees of
freedom are visible and which are hidden. Nothing else.

\textbf{The formula:}

\[T_{ij}(t_2, t_1) = \int_{\Gamma_H} \delta_{x_j}[\pi_V(\phi_{t_2-t_1}(x_i, h))] \, d\mu(h)\]

Unpacking each symbol:

\begin{itemize}
\tightlist
\item
  \textbf{T\_ij}: The probability of transitioning from visible state
  x\_i to visible state x\_j in the time interval from t\_1 to t\_2.
  This is what the observer measures.
\item
  \textbf{(x\_i, h)}: The complete state --- visible part x\_i, hidden
  part h.
\item
  \textbf{\ensuremath{\varphi}\_\{t2-t1\}}: The deterministic evolution. Takes the complete
  state at time t\_1 and returns the complete state at time t\_2.
  Uniquely determined by Axiom 1.
\item
  \textbf{\ensuremath{\pi}\_V}: Projection onto the visible sector. Takes a complete
  state (x, h) and returns just x.
\item
  \textbf{\ensuremath{\delta}\_\{xj\}{[}\ldots{]}}: The Kronecker delta. Equals 1 if the
  visible part ended up at x\_j, equals 0 otherwise.
\item
  \textbf{d\ensuremath{\mu}(h)}: Integration over all possible hidden states, weighted
  by the Liouville measure.
\end{itemize}

\textbf{In plain English:} For each possible hidden state h, check
whether starting at (x\_i, h) and evolving forward lands the visible
part on x\_j. Count up all the hidden states where this happens,
weighted by how likely each hidden state is. The result is the
probability of the transition x\_i \ensuremath{\to} x\_j.

\textbf{The proof:} The formula has exactly three inputs: (1) the
dynamics \ensuremath{\varphi}\_t --- fixed by Axiom 1, (2) the partition (\ensuremath{\Gamma}\_V, \ensuremath{\Gamma}\_H) and
projection \ensuremath{\pi}\_V --- fixed by Axiom 3, and (3) the measure \ensuremath{\mu} --- fixed by
Axiom 4 (Liouville measure is the unique choice). Since inputs 1 and 3
are determined by the axioms, the only free input is the partition.
Therefore: everything about the emergent description depends only on the
partition. QED.

\textbf{Why the Liouville measure is unique:} The observer needs a
``prior'' --- a way to weight the hidden states. Liouville measure is
the unique measure on phase space that is absolutely continuous (no
point masses) and invariant under Hamiltonian flow. Any smooth initial
distribution evolves toward it. Singular measures are excluded by Axiom
4's requirement of standard probability theory. The observer has no
choice.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{emergent-stochasticity-and-the-slow-bath-regime-2.12.2}{%
\subsection{Emergent Stochasticity and the Slow-Bath Regime
(§§2.1--2.2)}\label{emergent-stochasticity-and-the-slow-bath-regime-2.12.2}}

The total system is deterministic. If you knew both x and h, you'd know
the future with certainty. But the observer knows only x. Different
hidden states h send the same visible state x to different futures.

Example: visible state is ``Heads.'' Hidden state could be any die value
1--6. If the die is 1 or 2, the dynamics flip the coin to Tails. If the
die is 3--6, the coin stays at Heads. The observer doesn't know the die,
so they see: P(Heads \ensuremath{\to} Tails) = 2/6 = 1/3. The randomness is epistemic
(from ignorance) not ontological (from fundamental indeterminacy).

In a normal ``heat bath'' scenario, the environment is fast and chaotic.
It scrambles any information you write into it before you can read it
back. This produces Markovian (memoryless) dynamics --- each step is
independent of previous steps.

C2 inverts this. The hidden sector is slow. When the visible sector
interacts with it (writing information through H\_int), the information
stays there. At the next interaction, the hidden sector reads back what
was written before. The observer sees history-dependent transition
probabilities --- what happens next depends on what happened before.

This is non-Markovian dynamics. It's the key ingredient that separates
quantum mechanics from classical stochastic processes.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-p-indivisibility-theorem-2.3}{%
\subsection{The P-Indivisibility Theorem
(§2.3)}\label{the-p-indivisibility-theorem-2.3}}

\textbf{What it claims.} If a deterministic system is split into a
visible and hidden sector, and these sectors are genuinely coupled, then
the visible sector's behavior \emph{cannot} be a simple memoryless
random process. It must exhibit P-indivisibility --- a specific kind of
built-in memory.

\textbf{What ``P-indivisible'' means.} A stochastic process is
``P-divisible'' if you can always find a valid transition matrix
connecting any two time points. Mathematically: for any times t\_1
\textless{} t\_2 \textless{} t\_3, there exists a stochastic matrix \ensuremath{\Lambda}
such that:

\[T(t_3, t_1) = \Lambda(t_3, t_2) \cdot T(t_2, t_1)\]

where \ensuremath{\Lambda} has non-negative entries and rows summing to 1.
``P-indivisible'' means this fails --- the ``intermediate propagator''
would need negative entries, which means it's not a valid probability
matrix.

Breuer, Laine, and Piilo proved that P-indivisibility is equivalent to
``information backflow'' --- the system's distinguishability can
\emph{increase} over time. In a classical Markov process, you can only
lose information (mixing). In a P-indivisible process, information comes
back. This is exactly what quantum systems do --- interference,
revivals, and non-classical correlations all involve information
returning from where it was stored.

\textbf{The setup.} We work on finite sets (Axiom 2). The visible sector
has states C\_V = \{x\_1, x\_2, \ldots\} with \textbar C\_V\textbar{} \ensuremath{\geq}
2. The hidden sector has states C\_H = \{h\_1, h\_2, \ldots\}. The total
dynamics is a bijection \ensuremath{\varphi} on C\_V \ensuremath{\times} C\_H. The transition matrix is:

\[T_{ij} = \frac{|\{h \in \mathcal{C}_H : \pi_V(\varphi(x_i, h)) = x_j\}|}{|\mathcal{C}_H|}\]

\textbf{The key tool --- total variation distance:}

\[d(p, q) = \frac{1}{2}\sum_k |p_k - q_k|\]

This measures how distinguishable two probability distributions are. If
d = 1, they're perfectly distinguishable. If d = 0, they're identical.
For P-divisible processes, d can only decrease or stay constant.

\textbf{Step 1 --- Recurrence.} \ensuremath{\varphi} is a bijection on a finite set. Keep
applying \ensuremath{\varphi} and you must eventually return to where you started --- there
are only finitely many states to visit. Formally: there exists N such
that \ensuremath{\varphi}\^{}N = id. So T\^{}(N) = I, and:

\[d(\delta_i T^{(N)}, \delta_j T^{(N)}) = d(\delta_i, \delta_j) = 1\]

After N steps, states that started distinguishable are still perfectly
distinguishable.

\textbf{Step 2 --- Strict contraction.} T is not a permutation matrix
(this follows from C1 --- the coupling mixes things). So there exist
states i, j, l where both T\_il \textgreater{} 0 and T\_jl
\textgreater{} 0. The total variation distance after one step:

\[d(\delta_i T, \delta_j T) = \frac{1}{2}\sum_k |T_{ik} - T_{jk}| < 1\]

The inequality is strict because the distributions overlap.
Distinguishability has decreased.

\textbf{Step 3 --- The punchline.} At t = 1: d \textless{} 1
(distinguishability decreased). At t = N: d = 1 (distinguishability
restored). The distinguishability went down then came back up ---
non-monotonic behavior. A P-divisible process can only have
non-increasing distinguishability. Therefore the process is
P-indivisible. QED.

The proof uses almost nothing --- just that the dynamics is a bijection
on a finite set (Axioms 1 and 2) and that the coupling is non-trivial
(C1). It is purely combinatorial.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-accessible-timescale-lemma-2.3-continued}{%
\subsection{The Accessible-Timescale Lemma (§2.3
continued)}\label{the-accessible-timescale-lemma-2.3-continued}}

The recurrence proof shows P-indivisibility exists as a mathematical
property. But the recurrence time is absurdly long --- for the
cosmological case, it's e\textsuperscript{(10}122) years. Nobody will
ever observe it.

The accessible-timescale lemma shows that information backflow happens
on \emph{laboratory timescales}, independently of recurrence.

\textbf{The mechanism:} At each interaction (timescale \ensuremath{\tau}\_S), the
coupling H\_int transfers some information from the visible sector to
the hidden sector. Call the amount I\_0. Between interactions, the
hidden sector's correlations decay with a rate set by its spectral gap \ensuremath{\Delta}
\textasciitilde{} 1/\ensuremath{\tau}\_B.

The decay per visible-sector step is:

\[e^{-\Delta \tau_S} \approx 1 - \frac{\tau_S}{\tau_B}\]

When \ensuremath{\tau}\_S \ensuremath{\ll} \ensuremath{\tau}\_B (C2), this is very close to 1 --- almost no decay. The
hidden sector remembers almost perfectly between steps. After k steps,
the cumulative decay is:

\[e^{-k\Delta\tau_S} \approx 1 - \frac{k\tau_S}{\tau_B}\]

As long as k·\ensuremath{\tau}\_S \ensuremath{\ll} \ensuremath{\tau}\_B, the hidden sector retains \textasciitilde k
bits of visible-sector history. The mutual information satisfies:

\[I(X_{<t}; X_{>t} \mid X_t) \geq I_0\left(1 - \frac{k\tau_S}{\tau_B}\right)\]

For the cosmological case: \ensuremath{\tau}\_S \textasciitilde{} 10\^{}\{-15\} s, \ensuremath{\tau}\_B
\textasciitilde{} 10\^{}\{17\} s. Even after k = 10\^{}\{20\} steps,
k·\ensuremath{\tau}\_S/\ensuremath{\tau}\_B \textasciitilde{} 10\^{}\{-12\} --- negligible. The hidden
sector remembers everything.

\textbf{The role of C3:} The hidden sector's memory capacity is
log\_2(\textbar C\_H\textbar) bits. If k bits of history are written but
the capacity is only m \textless{} k bits, old data gets overwritten. C3
ensures m is large enough that the memory never saturates on observable
timescales.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-coin-and-die-model-2.4}{%
\subsection{The Coin-and-Die Model
(§2.4)}\label{the-coin-and-die-model-2.4}}

The paper builds a concrete toy model to make the mechanism tangible.

\textbf{Setup:} - Visible: x \ensuremath{\in} \{0, 1\} (a coin: 0 = Heads, 1 = Tails) -
Hidden: h \ensuremath{\in} \{1, 2, 3, 4, 5, 6\} (a die) - Total: 12 states

\textbf{The permutation \ensuremath{\sigma}:}

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
Input state & Output state & What happens \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
(0, 1) & (1, 1) & Coin flips, die stays \\
(1, 1) & (0, 1) & Coin flips, die stays \\
(0, 2) & (1, 2) & Coin flips, die stays \\
(1, 2) & (0, 2) & Coin flips, die stays \\
(0, 3) & (0, 4) & Coin stays, die changes \\
(0, 4) & (0, 3) & Coin stays, die changes \\
(0, 5) & (0, 6) & Coin stays, die changes \\
(0, 6) & (0, 5) & Coin stays, die changes \\
(1, 3) & (1, 4) & Coin stays, die changes \\
(1, 4) & (1, 3) & Coin stays, die changes \\
(1, 5) & (1, 6) & Coin stays, die changes \\
(1, 6) & (1, 5) & Coin stays, die changes \\
\end{longtable}

Every swap is a transposition (a \ensuremath{\leftrightarrow} b), so \ensuremath{\sigma}\ensuremath{^2} = id (apply twice,
everything returns).

\textbf{Checking the conditions:} C1 (coupling): die values 1 and 2 flip
the coin \checkmark{}. C2 (slow bath): \ensuremath{\sigma}\ensuremath{^2} = id means recurrence time is 2 steps,
giving \ensuremath{\tau}\_S/\ensuremath{\tau}\_B = 1/2 \checkmark{}. C3 (sufficient capacity): 6 hidden states vs 2
visible states \checkmark{}.

\textbf{Computing T(1,0).} Start at x = 0 (Heads). All 6 die values are
equally likely.

\begin{itemize}
\tightlist
\item
  h = 1: \ensuremath{\sigma}(0,1) = (1,1) \ensuremath{\to} Tails
\item
  h = 2: \ensuremath{\sigma}(0,2) = (1,2) \ensuremath{\to} Tails
\item
  h = 3: \ensuremath{\sigma}(0,3) = (0,4) \ensuremath{\to} Heads
\item
  h = 4: \ensuremath{\sigma}(0,4) = (0,3) \ensuremath{\to} Heads
\item
  h = 5: \ensuremath{\sigma}(0,5) = (0,6) \ensuremath{\to} Heads
\item
  h = 6: \ensuremath{\sigma}(0,6) = (0,5) \ensuremath{\to} Heads
\end{itemize}

P(0 \ensuremath{\to} 0) = 4/6 = 2/3, P(0 \ensuremath{\to} 1) = 2/6 = 1/3. By the same logic for x = 1:

\[T(1,0) = \begin{pmatrix} 2/3 & 1/3 \\ 1/3 & 2/3 \end{pmatrix}\]

\textbf{Distinguishability at t = 1:}

\[d(\delta_0 T, \delta_1 T) = \frac{1}{2}(|2/3 - 1/3| + |1/3 - 2/3|) = 1/3\]

Started at d = 1. Now d = 1/3. Distinguishability decreased.

\textbf{What Markov would predict at t = 2:} Apply the same transition
matrix again:

\[T(1,0)^2 = \begin{pmatrix} 5/9 & 4/9 \\ 4/9 & 5/9 \end{pmatrix}\]

Distinguishability would drop to d = 1/9. More mixing.

\textbf{What actually happens at t = 2:} \ensuremath{\sigma}\ensuremath{^2} = id. Every state returns to
its starting point.

\[T(2,0) = I = \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}\]

Distinguishability is back to d = 1. Complete un-mixing. Impossible for
a Markov process.

\textbf{The smoking gun --- negative entries.} If there were a valid
stochastic matrix \ensuremath{\Lambda}(2,1) connecting steps 1 and 2:

\[\Lambda(2,1) = T(2,0) \cdot [T(1,0)]^{-1} = I \cdot \begin{pmatrix} 2 & -1 \\ -1 & 2 \end{pmatrix} = \begin{pmatrix} 2 & -1 \\ -1 & 2 \end{pmatrix}\]

The entries \ensuremath{-}1 are negative. No valid stochastic matrix exists.
\textbf{This is P-indivisibility.}

\textbf{The mechanism in detail.} The die works as a memory register. At
step 1, if the coin was at 0 and the die was at 1, the coin flips to 1
but the die stays at 1. The die value 1 now encodes the information
``the coin was at 0 and I flipped it.'' At step 2, \ensuremath{\sigma} sees (1, 1) and
flips it back to (0, 1). The die read its own memory and reversed the
flip. C1 (coupling) allows writing to the memory. C2 (slow bath) ensures
it isn't erased between reads. C3 (sufficient capacity) ensures there's
enough room. Together, they produce the information backflow that makes
the process P-indivisible --- and therefore, by the stochastic-quantum
correspondence, equivalent to quantum mechanics.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{why-conditions-c2-and-c3-matter-physically}{%
\subsubsection{Why conditions C2 and C3 matter
physically}\label{why-conditions-c2-and-c3-matter-physically}}

The P-indivisibility theorem needs only coupling (C1) and finiteness. So
why does the paper insist on slow memory (C2) and vast capacity (C3)?

Because P-indivisibility without C2 and C3 might only show up at absurd
timescales or might self-destruct. C2 ensures the memory persists on
timescales accessible to actual experiments, not just at cosmic
recurrence times. C3 ensures the hidden sector never runs out of room to
store information --- if it saturates, later imprints overwrite earlier
ones, and the process becomes effectively memoryless. Together, C2 and
C3 guarantee that P-indivisibility is strong, persistent, and
observationally relevant.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-stochastic-quantum-correspondence-3.1-and-appendix-a}{%
\subsection{The Stochastic-Quantum Correspondence (§3.1 and Appendix
A)}\label{the-stochastic-quantum-correspondence-3.1-and-appendix-a}}

This is the key link. Section 2 proved that the embedded observer's
dynamics are P-indivisible. Section 3 shows this is mathematically
equivalent to quantum mechanics.

\textbf{The core statement.} Any P-indivisible stochastic process on a
finite configuration space of size n can be embedded into a unitarily
evolving quantum system. Specifically, there exists a Hilbert space H
(dimension \ensuremath{\leq} n\ensuremath{^3}) and a unitary operator U(t) such that:

\[T_{ij}(t) = |U_{ij}(t)|^2\]

This is the Born rule. The left side is the transition probability
computed by averaging over hidden states (the classical formula from
partition-relativity). The right side is the quantum mechanical
probability --- the squared modulus of a matrix element of the unitary
evolution operator. The equivalence is not approximate. It is not an
analogy. It is a mathematical identity.

\textbf{Two independent routes to the same conclusion.} The primary
route uses Barandes' stochastic-quantum correspondence (2023--2025):
P-indivisibility means transition probabilities can't be factored
through intermediate times --- try it and you get ``negative
probabilities.'' In quantum mechanics, this is \emph{exactly what
happens}: probability amplitudes combine to produce interference
patterns that don't factorize classically. What Barandes proved is that
these are the same mathematical object, written in different notation.

The secondary route, given in Appendix A, uses Stinespring's dilation
theorem (1955): a deterministic bijection on a finite product space
defines a permutation unitary; tracing out the hidden sector with the
Liouville measure produces a completely positive quantum channel whose
diagonal elements recover the classical transition probabilities
exactly. This second route requires only textbook results. Either route
alone suffices; together they ensure the bridge rests on no single
recent result.

\textbf{Where the quantum features come from:}

\begin{itemize}
\item
  \textbf{The Schrödinger equation} arises because U(t) is
  differentiable. Any smooth family of unitary matrices can be written
  as U(t) = exp(-iHt/\ensuremath{\hbar}) for some Hermitian matrix H.
\item
  \textbf{The Born rule} T\_ij = \textbar U\_ij\textbar\ensuremath{^2} is not an
  additional postulate --- it's the definition of how the stochastic
  process maps onto the unitary one.
\item
  \textbf{The action scale \ensuremath{\hbar}} enters when converting from the
  dimensionless unitary to a dimensionful Hamiltonian: Ĥ = i\ensuremath{\hbar} (\ensuremath{\partial}U/\ensuremath{\partial}t)
  U†. The value of \ensuremath{\hbar} cannot be determined from the dimensionless
  transition data alone --- it requires additional physical input from
  the partition geometry (§5).
\item
  \textbf{Bell inequality violations.} Since the transition matrices for
  composite systems don't factorize, entangled systems naturally produce
  correlations that violate Bell inequalities, up to exactly Tsirelson's
  bound.
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-phase-locking-lemma-3.1-continued}{%
\subsection{The Phase-Locking Lemma (§3.1
continued)}\label{the-phase-locking-lemma-3.1-continued}}

A potential objection: the relation T\_ij = \textbar U\_ij\textbar\ensuremath{^2}
throws away phase information. Different unitaries could give the same
transition probabilities. Does this make the quantum description
ambiguous?

The phase-locking lemma shows: no.

\textbf{Setup:} The transition probability at time t is:

\[T_{ij}(t) = \left|\sum_k V_{ik} \, e^{-iE_k t} \, V_{jk}^*\right|^2\]

where V\_ik = \ensuremath{\langle}i\textbar k\ensuremath{\rangle} are the overlaps between the configuration
basis and the energy eigenbasis, and E\_k are the energy eigenvalues.
Expanding the square:

\[T_{ij}(t) = \sum_{k,l} V_{ik}\, V_{jk}^*\, V_{jl}\, V_{il}^*\; e^{-i(E_k - E_l)t}\]

\textbf{The Fourier trick:} This is a sum of oscillating terms at
frequencies \ensuremath{\omega}\_kl = E\_k - E\_l. If all these frequencies are distinct
(condition G2: non-degenerate energy gaps), you can extract each
coefficient by Fourier transform:

\[a_{ij}^{kl} = V_{ik}\, V_{jk}^*\, V_{jl}\, V_{il}^*\]

\textbf{Extracting the moduli:} Setting i = j:
\(a_{ii}^{kl} = |V_{ik}|^2 |V_{il}|^2\). If none of the overlaps are
zero (condition G3), all moduli \textbar V\_ik\textbar{} are determined.

\textbf{Extracting the phases:} Write V\_ik = \textbar V\_ik\textbar{}
e\^{}\{i\ensuremath{\varphi}\_ik\}. The argument of the Fourier coefficient gives:

\[\arg(a_{ij}^{kl}) = (\varphi_{ik} - \varphi_{il}) - (\varphi_{jk} - \varphi_{jl})\]

The only transformation preserving all double differences is \ensuremath{\varphi}\_ik \ensuremath{\to}
\ensuremath{\varphi}\_ik + \ensuremath{\alpha}\_i + \ensuremath{\beta}\_k --- just relabeling (choosing a different phase
convention for the basis states). Once you fix these conventions, all
remaining phases are uniquely determined.

\textbf{Bottom line:} Continuous-time transition probability data
uniquely determines the Hamiltonian up to physically irrelevant
relabeling.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{bell-inequality-violations-3.2}{%
\subsection{Bell Inequality Violations
(§3.2)}\label{bell-inequality-violations-3.2}}

This is the question everyone asks: isn't this ruled out by Bell's
theorem?

\textbf{What Bell's theorem actually requires.} Bell's theorem proves
that no hidden variable theory can reproduce quantum correlations if it
satisfies three conditions simultaneously:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Locality:} The outcome at detector A doesn't depend on the
  setting at detector B
\item
  \textbf{Measurement independence:} The experimenters' choices are
  independent of the hidden variables
\item
  \textbf{Factorizability:} P(a,b \textbar{} x,y,\ensuremath{\lambda}) = P(a\textbar x,\ensuremath{\lambda}) ·
  P(b\textbar y,\ensuremath{\lambda})
\end{enumerate}

The framework satisfies conditions 1 and 2. It violates condition 3.

\textbf{Why factorizability fails.} Factorizability requires that,
conditioned on the hidden variable \ensuremath{\lambda}, the outcomes at the two detectors
are independent --- that \ensuremath{\lambda} carries all the relevant information as a
snapshot at a single moment.

P-indivisible processes don't work this way. The transition
probabilities for a joint system can't be factored:

\[T_{QR} \neq T_Q \otimes T_R\]

Two subsystems that interacted during preparation carry a joint
transition matrix that doesn't decompose into a product. This
non-factorizability IS entanglement.

\textbf{The Jarrett decomposition.} Factorizability splits into
parameter independence (outcome at A doesn't depend on setting at B ---
preserved \checkmark{}) and outcome independence (outcome at A doesn't depend on
outcome at B --- violated X). Parameter independence prevents
faster-than-light signaling. Fine's theorem shows that violating outcome
independence while preserving parameter independence is exactly the
class of theories consistent with quantum correlations.

\textbf{The maximum violation.} Barandes, Hasan, and Kagan prove the
maximum CHSH violation from P-indivisible processes is exactly
Tsirelson's bound: 2\ensuremath{\sqrt{}}2 --- the quantum maximum.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-characterization-theorem-3.3}{%
\subsection{The Characterization Theorem
(§3.3)}\label{the-characterization-theorem-3.3}}

It's not enough to show that embedded observation \emph{produces} QM
(sufficiency). The paper shows QM \emph{requires} embedded observation
under C1--C3 (necessity). The full logical chain:

\begin{itemize}
\tightlist
\item
  Barandes proved: QM \ensuremath{\iff} P-indivisibility
\item
  Section 2.3 proved: C1--C3 \ensuremath{\implies} P-indivisibility (sufficiency)
\item
  Section 3.3 proves: P-indivisibility \ensuremath{\implies} C1--C3 (necessity)
\item
  Combined: \textbf{QM \ensuremath{\iff} P-indivisibility \ensuremath{\iff} embedded observation under
  C1--C3}
\end{itemize}

\textbf{Necessity of C1 (coupling).} If T is a permutation (no
coupling), then T\^{}k is also a permutation for all k. The intermediate
propagator \ensuremath{\Lambda}(k\ensuremath{_2},k\ensuremath{_1}) = T\^{}\{k\ensuremath{_2}-k\ensuremath{_1}\} is always a valid stochastic
matrix. So the process is P-divisible. Contrapositive: P-indivisibility
requires non-trivial coupling.

\textbf{Necessity of C2 (slow bath).} Between coupling events (separated
by \ensuremath{\tau}\_S), the hidden sector evolves under its own Hamiltonian. The
convergence to equilibrium is:

\[\| e^{\mathcal{L}_H \tau_S} \mu_H(\cdot | x_i) - \mu_{\text{eq}} \|_{\text{TV}} \leq C \, e^{-\Delta \tau_S}\]

In the fast-bath regime (\ensuremath{\Delta}·\ensuremath{\tau}\_S \ensuremath{\gg} 1), this is exponentially small. The
hidden sector forgets everything between interactions. Each transition
is computed against the same equilibrium distribution, so T\^{}(k) =
T\^{}k --- a Markov chain, hence P-divisible. Contrapositive:
P-indivisibility requires \ensuremath{\tau}\_S \ensuremath{\ll} \ensuremath{\tau}\_B.

\textbf{Necessity of C3 (sufficient capacity).} The non-Markovian mutual
information is bounded by the hidden sector's size:

\[I(X_{<t} ; X_{>t} \mid X_t) \leq \log_2 m\]

where m = \textbar C\_H\textbar. Proof: the total system is
deterministic, so X\_\{\textgreater t\} is a function of (X\_t, H\_t).
Given X\_t, the chain X\_\{\textless t\} \ensuremath{\to} H\_t \ensuremath{\to} X\_\{\textgreater t\}
is Markov. By the data processing inequality:

\[I(X_{<t} ; X_{>t} \mid X_t) \leq I(X_{<t} ; H_t \mid X_t) \leq H(H_t \mid X_t) \leq H(H_t) \leq \log_2 m\]

Each step uses a standard information-theoretic inequality. The result:
if you want K bits of history dependence, you need m \ensuremath{\geq} 2\^{}K hidden
states.

\textbf{The complete characterization.} For \textbar C\_V\textbar{} \ensuremath{\geq} 2,
the following are equivalent: (1) the process is mathematically
equivalent to unitarily evolving QM, (2) the process is P-indivisible,
(3) the process arises from marginalizing a deterministic bijection with
C1, C2, C3. This is the biconditional: \textbf{QM \ensuremath{\iff} embedded observation
under C1--C3.}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-cosmological-application-4}{%
\subsection{The Cosmological Application
(§4)}\label{the-cosmological-application-4}}

\hypertarget{the-partition}{%
\subsubsection{4.1 The Partition}\label{the-partition}}

The cosmological horizon is the boundary beyond which no signal
traveling at or below c can ever reach the observer. In a universe with
a positive cosmological constant (de Sitter space), this horizon exists
for every observer and has a definite, finite area:

\[A = \frac{4\pi c^2}{H^2}\]

where H is the Hubble parameter. This implements Axiom 3 naturally: \ensuremath{\Gamma}\_V
= everything inside the horizon, \ensuremath{\Gamma}\_H = everything beyond.

\textbf{Why \ensuremath{\hbar} is the same for all observers:} Different observers have
slightly different horizons. But the gap equation \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) depends
only on local geometric quantities (c, G, \ensuremath{\varepsilon}) --- not on the horizon area
or the observer's worldline. So all observers derive the same \ensuremath{\hbar}.

\hypertarget{verification-of-the-conditions}{%
\subsubsection{4.2 Verification of the
Conditions}\label{verification-of-the-conditions}}

\textbf{C1 (coupling).} In general relativity's ADM formulation, the
bulk Hamiltonian is a sum of constraints that vanish on-shell ---
meaning the ``real'' dynamics happens at the boundary. The Hamiltonian
and momentum constraints correlate interior and exterior data. This is
stronger than just H\_int \ensuremath{\neq} 0: the constraints enforce correlations that
persist on all timescales.

\textbf{C2 (slow bath).} \ensuremath{\tau}\_B \textasciitilde{} 1/H \textasciitilde{}
10\ensuremath{^1}\ensuremath{^7} seconds (the Hubble time). \ensuremath{\tau}\_S \textasciitilde{} 10\ensuremath{^{-}}\ensuremath{^1}\ensuremath{^5} seconds (a
typical atomic process). The ratio: \ensuremath{\tau}\_S/\ensuremath{\tau}\_B \textasciitilde{} 10\ensuremath{^{-}}\ensuremath{^3}\ensuremath{^2}.

\textbf{C3 (sufficient capacity).} The hidden sector has A/\ensuremath{\varepsilon}\ensuremath{^2}
\textasciitilde{} 10\ensuremath{^1}\ensuremath{^2}\ensuremath{^2} modes (the de Sitter entropy). The visible
sector has \textasciitilde{} 10\ensuremath{^8}\ensuremath{^0} baryons. No experiment comes close to
saturating the hidden sector.

\hypertarget{application}{%
\subsubsection{4.3 Application}\label{application}}

With the cosmological horizon satisfying all axioms and conditions, the
characterization theorem applies. The observer's reduced description is
P-indivisible and therefore equivalent to unitary quantum mechanics. The
value of \ensuremath{\hbar} is determined by the partition geometry --- which is what
Section 5 derives.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{where-plancks-constant-comes-from-5}{%
\subsection{Where Planck's Constant Comes From
(§5)}\label{where-plancks-constant-comes-from-5}}

Partition-relativity proved that the emergent quantum description is
completely and uniquely determined by the partition. This means
\(\hbar\) cannot be a free parameter --- it must be fixed by the
geometry of the boundary.

\hypertarget{the-classical-horizon-temperature-5.1}{%
\subsubsection{The Classical Horizon Temperature
(§5.1)}\label{the-classical-horizon-temperature-5.1}}

\textbf{Starting point: Jacobson's identity.} Jacobson (1995) showed
that Einstein's field equations follow from applying the first law of
thermodynamics \ensuremath{\delta}Q = TdS to local causal horizons:

\[dE = \frac{c^2 \kappa}{8\pi G} \, dA\]

where \ensuremath{\kappa} is the surface gravity and dA is the change in horizon area.
This is a classical gravitational identity --- no quantum mechanics
involved.

The entropy density is \ensuremath{\eta} = 1/\ensuremath{\varepsilon}\ensuremath{^2} (one degree of freedom per minimal cell
of area \ensuremath{\varepsilon}\ensuremath{^2}). So dS = dA/\ensuremath{\varepsilon}\ensuremath{^2}. From dE = TdS:

\[k_B T_{\text{cl}} = \frac{c^2 \epsilon^2 \kappa}{8\pi G}\]

For the de Sitter horizon, \ensuremath{\kappa} = cH:

\[k_B T_{\text{cl}} = \frac{c^3 \epsilon^2 H}{8\pi G}\]

\textbf{Critical point: no \ensuremath{\hbar} appears anywhere.} This temperature is
computed from purely classical quantities.

\hypertarget{the-four-step-derivation-5.2}{%
\subsubsection{The Four-Step Derivation
(§5.2)}\label{the-four-step-derivation-5.2}}

\textbf{Step 1 (Uniqueness).} Partition-relativity guarantees \ensuremath{\hbar} is
determined by the partition geometry. It's not a free parameter.

\textbf{Step 2 (Boundary-only dependence).} A substantial lemma showing
that \ensuremath{\hbar} depends only on the boundary modes, not on the deep interior of
the hidden sector. Decompose the hidden sector into boundary modes C\_B
(near the horizon) and deep modes C\_D (far from the horizon):

\[\mathcal{C}_H = \mathcal{C}_B \times \mathcal{C}_D\]

The proof has three parts: (i) spatial locality means V talks to B, and
B talks to D, but V doesn't talk directly to D; (ii) on timescales t \ensuremath{\ll}
\ensuremath{\tau}\_B, the deep sector barely moves (displacement \textasciitilde{}
10\ensuremath{^{-}}\ensuremath{^3}\ensuremath{^2}); (iii) because the deep modes are frozen, the sum over them
produces a trivial factor:

\[T_{ij}(t) = \underbrace{\frac{1}{|\mathcal{C}_B|} \sum_{b} \delta_{x_j}[\pi_V(\varphi_t^{VB}(x_i, b))]}_{T^{(B)}_{ij}(t)} + \mathcal{O}(t/\tau_B)\]

The transition probabilities depend only on boundary dynamics. Since \ensuremath{\hbar}
is determined by transition probabilities, \ensuremath{\hbar} depends only on boundary
quantities: c, G, and \ensuremath{\varepsilon}. This excludes dependence on H --- if \ensuremath{\hbar} depended
on H, observers at different cosmic epochs would have different quantum
mechanics.

\textbf{Step 3 (Dimensional analysis).} There's a unique combination of
c, G, and \ensuremath{\varepsilon} with dimensions of action:

\[[\hbar] = \frac{[c]^3 \, [\epsilon]^2}{[G]} = \text{kg·m}^2/\text{s} \quad \checkmark\]

So \ensuremath{\hbar} = \ensuremath{\beta} c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/G, where \ensuremath{\beta} is a dimensionless constant that dimensional
analysis alone can't fix.

\textbf{Step 4 (Thermal self-consistency).} We have two independent
descriptions of the horizon temperature:

\emph{Classical:} T\_cl = c\ensuremath{^2}\ensuremath{\varepsilon}\ensuremath{^2}\ensuremath{\kappa}/(8\ensuremath{\pi}Gk\_B) --- computed from the
geometric substratum, no \ensuremath{\hbar}.

\emph{Quantum:} The emergent QFT lives on a spacetime with a bifurcate
Killing horizon. Standard QFT on curved spacetime gives a KMS thermal
state at temperature:

\[T_Q = \frac{\hbar \kappa}{2\pi c k_B}\]

A thermometer at the horizon must read the same temperature either way.
Setting T\_cl = T\_Q:

\[\frac{c^2 \epsilon^2 \kappa}{8\pi G} = \frac{\hbar \kappa}{2\pi c}\]

The surface gravity \ensuremath{\kappa} cancels from both sides. Solving:

\[\boxed{\hbar = \frac{c^3 \epsilon^2}{4G}}\]

This fixes \ensuremath{\beta} = 1/4.

\textbf{Why this is not circular.} The KMS temperature T\_Q contains \ensuremath{\hbar}
as an \emph{unknown}. The classical temperature T\_cl contains no \ensuremath{\hbar} at
all. The logic is sequential: Part I proves quantum mechanics emerges
with some \ensuremath{\hbar} \ensuremath{\to} the emergent QM predicts T\_GH(\ensuremath{\hbar}) \ensuremath{\to} the classical
temperature T\_cl was calculated independently \ensuremath{\to} matching them fixes \ensuremath{\hbar}.
This is a ``gap equation,'' the same structure used routinely in
condensed matter physics. If T\_cl had depended on the deep
hidden-sector volume (it doesn't), or if T\_Q had been state-dependent
(it isn't), the matching wouldn't work. The fact that it works is a
non-trivial self-consistency check.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-d-gauge-completeness-theorem-5.3}{%
\subsection{The D-Gauge Completeness Theorem
(§5.3)}\label{the-d-gauge-completeness-theorem-5.3}}

\textbf{The problem.} The relation T\_ij = \textbar U\_ij\textbar\ensuremath{^2}
discards phase information. Could different Hamiltonians give the same
transition probabilities?

\textbf{The theorem.} If \textbar U'\_ij(t)\textbar\ensuremath{^2} =
\textbar U\_ij(t)\textbar\ensuremath{^2} for all i, j, t, then H' = DHD† where D is a
diagonal unitary (a physically meaningless relabeling of basis phases).

\textbf{The proof in three steps:}

\emph{Step 1 (eigenvalue recovery):} Fourier analysis of T\_ij(t)
extracts the energy differences E\_k - E\_l. Non-degeneracy of energy
gaps means E'\_k = E\_k + E\ensuremath{_0} (same eigenvalues up to a global shift).

\emph{Step 2 (modulus recovery):} The diagonal Fourier coefficients give
\textbar V\_ik\textbar\ensuremath{^2} directly. So \textbar V'\_ik\textbar{} =
\textbar V\_ik\textbar.

\emph{Step 3 (phase structure):} Writing V'\_ik = V\_ik e\^{}\{i\ensuremath{\delta}\_ik\}
and requiring all Fourier coefficients to match gives the
double-difference condition:

\[\delta_{ik} - \delta_{il} - \delta_{jk} + \delta_{jl} = 0 \pmod{2\pi}\]

The general solution: \ensuremath{\delta}\_ik = \ensuremath{\alpha}\_i + \ensuremath{\beta}\_k --- a sum of a row phase and a
column phase. This is just basis rephasing.

\textbf{The dimensional obstruction.} The unitary U(t) is dimensionless.
The Hamiltonian Ĥ = i\ensuremath{\hbar} \ensuremath{\partial}\_tU · U† contains \ensuremath{\hbar}, which is dimensionful. No
amount of dimensionless data can fix a dimensionful constant. This is
why Step 4 (thermal matching) is not just a convenient check but the
\emph{mathematically obligatory} step: it's the only place where
dimensionful physical input enters the framework.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-discreteness-scale-6}{%
\subsection{The Discreteness Scale
(§6)}\label{the-discreteness-scale-6}}

\hypertarget{what-ux3b5-2-l_p-means}{%
\subsubsection{What \ensuremath{\varepsilon} = 2 l\_p means}\label{what-ux3b5-2-l_p-means}}

Rearranging the gap equation \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G):

\[\epsilon^2 = \frac{4\hbar G}{c^3} = 4 \, l_p^2\]

where l\_p = \ensuremath{\sqrt{}}(\ensuremath{\hbar}G/c\ensuremath{^3}) is the Planck length. Therefore \ensuremath{\varepsilon} = 2 l\_p.

\textbf{What this is:} The framework has one free geometric parameter
--- the discreteness scale \ensuremath{\varepsilon}. The gap equation fixes its relationship to
\ensuremath{\hbar}. Given the measured value of \ensuremath{\hbar}, the discreteness scale is determined:
\ensuremath{\varepsilon} = 2 l\_p \ensuremath{\approx} 3.2 \ensuremath{\times} 10\ensuremath{^{-}}\ensuremath{^3}\ensuremath{^5} meters.

\textbf{What this is NOT:} This is not an independent prediction of \ensuremath{\varepsilon}.
The framework contains one free parameter and one equation relating it
to known constants. The self-consistency condition pins \ensuremath{\varepsilon} to the Planck
regime but doesn't predict a number that wasn't already implicitly
known.

\hypertarget{the-bekenstein-hawking-entropy-why-the-14-factor-is-derived}{%
\subsubsection{The Bekenstein-Hawking Entropy --- Why the 1/4 factor is
derived}\label{the-bekenstein-hawking-entropy-why-the-14-factor-is-derived}}

The number of independent modes on the cosmological horizon is:

\[N_{\text{modes}} = \frac{A}{\epsilon^2} = \frac{A}{4 \, l_p^2}\]

This is the Bekenstein-Hawking entropy:

\[S_{\text{BH}} = \frac{A}{4 \, l_p^2}\]

The factor of 1/4 --- which Bekenstein and Hawking introduced as a
proportionality constant in 1973 --- is here derived: each minimal cell
of area \ensuremath{\varepsilon}\ensuremath{^2} = 4 l\_p\ensuremath{^2} contributes one unit of entropy. The 4 in the
denominator comes from \ensuremath{\varepsilon} = 2 l\_p.

\textbf{Why this is significant:} The 1/4 factor has been a mystery for
50 years. Most frameworks either assume it or derive it within
constructions specifically designed to produce it. Here it follows from
the gap equation with no additional input.

\hypertarget{self-consistency-bounds-on-ux3b5}{%
\subsubsection{Self-consistency bounds on
\ensuremath{\varepsilon}}\label{self-consistency-bounds-on-ux3b5}}

\textbf{If \ensuremath{\varepsilon}\ensuremath{^2} \ensuremath{\ll} l\_p\ensuremath{^2}:} Sub-Planckian cells would need further
subdivision, creating a second trace-out within the first. This would
make \ensuremath{\hbar} multi-valued --- contradicting the observed universality of \ensuremath{\hbar}.

\textbf{If \ensuremath{\varepsilon}\ensuremath{^2} \ensuremath{\gg} l\_p\ensuremath{^2}:} Super-Planckian cells would be too coarse. The
emergent quantum description would assign distinct quantum states to
configurations that are physically identical.

The self-consistency condition \ensuremath{\varepsilon} = 2 l\_p is the unique value where
neither pathology occurs.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-two-levels}{%
\subsection{The Two Levels}\label{the-two-levels}}

Now the cosmological constant problem dissolves. But first, a crucial
link: how does finite-dimensional quantum mechanics become quantum
\emph{field} theory --- with independent modes, each carrying zero-point
energy?

The answer is spatial locality. The classical substratum has it ---
neighboring cells interact, distant cells don't. The paper proves that
the emergent quantum description inherits it. The argument is direct:
the Barandes correspondence maps each classical configuration
\((x_1, x_2, \ldots, x_N)\) to a quantum basis state
\(|x_1, x_2, \ldots, x_N\rangle\). That's already a tensor product. The
only question is whether the emergent Hamiltonian respects the
structure. And it must, because if two configurations differ only at
sites that aren't neighbors, the classical dynamics can't connect them
in an infinitesimal time step --- so the transition probability between
them is zero --- so the corresponding quantum Hamiltonian matrix element
is zero. The emergent Hamiltonian couples only neighboring cells,
exactly like the classical one. You get a lattice quantum field theory
with a built-in ultraviolet cutoff at \(\epsilon = 2\,l_p\).

This is what makes the CC dissolution work. The QFT has independent
modes. Each mode gets a zero-point energy. The sum diverges. And the
``worst prediction in physics'' follows --- but only within the emergent
description.

The critical insight is that general relativity and quantum field theory
are not competing answers to the same question. They are answers to
\emph{different questions}, asked at different levels of the same
reality.

\textbf{Level 1: The classical substratum.} Spacetime geometry is part
of the fundamental layer. The metric tensor evolves via Einstein's field
equations. The stress-energy tensor that sources gravity is the
classical stress-energy of the total microstate. At this level, the
vacuum energy density sits at the critical scale:
\(\rho \sim H^2/G \sim 10^{-9}\) J/m\ensuremath{^3}. No zero-point energy. No
discrepancy.

\textbf{Level 2: The emergent quantum description.} For an embedded
observer tracing out the hidden sector, the mandatory quantum
description assigns a zero-point energy of \(\frac{1}{2}\hbar\omega\)
per mode. Sum to the Planck cutoff and you get
\(\rho_{\text{QFT}} \sim 10^{113}\) J/m\ensuremath{^3}. This number is real
\emph{within the quantum description} --- it reflects the magnitude of
the trace-out noise --- but it is not a source term in Einstein's
equations, because those equations operate at the classical level, which
is logically prior to the quantum description.

This ordering is not a choice but is forced by three independent
requirements. First, the partition must be definite --- not in
superposition --- for the trace-out to be well-defined; a partition in
superposition would yield an incoherent mixture of inequivalent quantum
theories. Second, the partition is defined by the causal structure,
which is determined by null geodesics of the metric; if the metric were
derived from QM, the derivation would be circular (QM \ensuremath{\to} metric \ensuremath{\to} causal
structure \ensuremath{\to} partition \ensuremath{\to} QM). Third, \(\hbar\) is determined by the
boundary geometry; if the geometry were itself quantum-mechanical,
\(\hbar\) would depend on a quantum state, contradicting its observed
universality.

The \(10^{122}\) ratio between the two answers is not a discrepancy. It
equals \(S_{\text{dS}}\) --- the Bekenstein-Hawking entropy of the
cosmological horizon --- which is the number of hidden-sector degrees of
freedom the trace-out compresses into the emergent quantum state. The
``worst prediction in physics'' is the information compression ratio of
the observer's blind spot. A category error, not a fine-tuning failure.

This is not a prediction awaiting future data. The observed vacuum
energy has been measured since 1998, and it sits exactly at the
classical geometric scale --- the value the framework expects.
Meanwhile, every attempt to search for a cancellation mechanism within
quantum-first frameworks have found nothing. The framework explains why:
there is nothing to cancel.

This track record is itself evidence for the ordering. The question of
whether geometry is prior to quantum mechanics or quantum mechanics is
prior to geometry is usually treated as a philosophical preference. But
it has an empirical signature sitting in plain sight: one ordering
produces the worst prediction in physics and has no solution; the other
predicts the observed value and explains the discrepancy as a derived
quantity. That doesn't prove the ordering is correct --- the
gravitational wave echo prediction provides a more direct test --- but
it's existing evidence, not a future hope.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{what-quantum-weirdness-looks-like-from-here}{%
\subsection{What Quantum Weirdness Looks Like From
Here}\label{what-quantum-weirdness-looks-like-from-here}}

If quantum mechanics is an emergent description forced on embedded
observers, the standard quantum puzzles acquire straightforward
readings.

\textbf{The double-slit experiment.} The particle goes through one slit.
The interference pattern arises because the transition probabilities are
computed by marginalizing over the hidden sector, and the hidden sector
includes the field configuration near both slits. Opening or closing the
second slit changes the boundary conditions, which changes the
marginalization, which changes the pattern. The ``wave-like'' behavior
is the hidden sector's influence shifting when the geometry changes.

\textbf{Entanglement.} Two particles prepared together share a joint
transition matrix inherited from the trace-out. The correlations are
encoded in the structure of the dynamics itself, not in a hidden
variable you could integrate over. This is why Bell inequality
violations occur: the standard factorization assumption fails for
indivisible processes. The framework reproduces quantum correlations
exactly up to Tsirelson's bound, without faster-than-light signaling and
without superdeterminism.

\textbf{The measurement problem.} Measurement produces definite outcomes
through the indivisible dynamics. When Wigner can't access his Friend's
lab, he traces out its internal degrees of freedom and assigns a
superposition. The superposition reflects Wigner's epistemic situation
--- what \emph{he} can infer --- not the Friend's physical state.
Branching in the Many-Worlds sense is a feature of the compressed
description, not of the underlying reality.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{predictions}{%
\subsection{Predictions}\label{predictions}}

The framework is not just a reinterpretation. It makes specific,
falsifiable predictions --- and, crucially, it makes predictions in
domains it was not designed to address.

\textbf{Dark energy evolution.} Because the hidden sector's
dimensionality changes as the Hubble parameter \(H\) evolves (the
horizon area is \(A = 4\pi c^2/H^2\)), the emergent vacuum energy
inherits a dependence on \(H\). The predicted form matches the Running
Vacuum Model: \(\Lambda_{\text{eff}} = \Lambda_0 + \nu H^2\). The
coefficient \(\nu\) is computed from the spectral structure of the
horizon: the trace-out compression noise is distributed over
\(\mathcal{N} = \ln(c/\epsilon H)\) spectral decades, giving a conformal
benchmark of \(\nu \approx 2.45 \times 10^{-3}\) with a systematic range
of \(\sim 1\)--\(5 \times 10^{-3}\) (the dominant uncertainty is the
spectral assumption, not the cosmological parameters). DESI's 2024--2025
data releases report evidence for evolving dark energy at
\(2.8\sigma\)--\(4.2\sigma\), with phenomenological fits consistently
finding \(\nu \sim 1\)--\(4 \times 10^{-3}\) --- the range in which the
framework's prediction sits.

\textbf{Gravitational wave echoes.} At a proper distance of about one
discreteness scale (\(\epsilon = 2\,l_p\)) outside a black hole horizon,
an infalling mode's wavelength hits the discreteness floor and must
scatter back. For a 30 solar-mass remnant, the predicted echo delay is
on the order of 50 ms.

\textbf{The dark sector as corroboration.} The trace-out that produces
quantum mechanics has a gravitational consequence that the paper did not
set out to find. The boundary entropy --- the \(S_{\text{dS}}\) modes
traced out to produce the quantum description --- has thermal energy
that, distributed over the Hubble volume, equals the critical density
exactly. A crucial subtlety: this thermal energy is computed entirely
from pre-trace-out (classical) quantities --- the number of boundary
modes, the classical horizon temperature, and the Hubble volume --- with
no reference to \(\hbar\) or the emergent quantum description. This is
what distinguishes the boundary entropy's gravitational contribution
from the QFT zero-point energy (\(\rho_{\text{QFT}} \sim 10^{113}\)
J/m\ensuremath{^3}), which exists only \emph{after} the trace-out and is an artifact
of the emergent description. The framework denies that the zero-point
energy gravitates (this is the CC dissolution of Part II); the boundary
entropy's classical thermal energy \emph{does} gravitate, at the scale
\(\rho_{\text{crit}} \sim 10^{-9}\) J/m\ensuremath{^3}.

The paper proves that this entropy has no operator in the emergent QFT.
The baryonic sector --- what QFT can account for --- is
\textasciitilde5\% of \(\rho_{\text{crit}}\). The remaining
\textasciitilde95\% is the boundary entropy: gravitationally active,
invisible to the emergent description, and persistent through
P-indivisibility (condition C2). This matches the observed composition
of the universe, in which \textasciitilde95\% of the gravitational
content has no source in particle physics. The uniform component
corresponds to dark energy (handled by Part II's CC dissolution); the
structured component (dark matter) is a candidate consequence of
matter-induced entropy redistribution, though the redistribution profile
requires further work. That axioms designed to derive quantum mechanics
also account for the dark sector's total budget is independent
corroboration that observational incompleteness is capturing real
structure.

The conjunction is distinctive: dark energy evolution in RVM form,
gravitational wave echoes, and a natural account of the dark sector's
total gravitational budget --- all from four axioms about embedded
observation. No competing framework produces all three. The parallel
with the cosmological constant dissolution is exact. Part II identifies
the \(10^{122}\) discrepancy as the \emph{information compression ratio}
of the trace-out --- how much the emergent description distorts the
vacuum energy. Part III identifies the \textasciitilde95\% dark sector
as the \emph{gravitational occlusion fraction} --- how much of the
gravitational universe the trace-out hides from the emergent
description. Together, they account for the two largest anomalies in
modern cosmology as two aspects of a single phenomenon: the cost of
observing the universe from within.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{philosophical-lineage}{%
\subsection{Philosophical Lineage}\label{philosophical-lineage}}

The paper is a physics paper, but its core claims --- that observers
face irreducible limits, that two irreconcilable descriptions can both
be correct, that incompleteness is a structural feature rather than a
deficiency --- sit at the intersection of some of the oldest debates in
philosophy. A systematic mapping against the major traditions reveals a
striking pattern: broad support for most of the framework, and
near-universal resistance to one specific thesis.

\hypertarget{the-seven-claims}{%
\subsubsection{The seven claims}\label{the-seven-claims}}

The framework rests on seven implicit philosophical commitments:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Embedded observers face irreducible limits.} No observer
  inside a system can access the complete state.
\item
  \textbf{QM and GR are both correct} within their domains.
\item
  \textbf{The hidden sector is permanently inaccessible} --- not due to
  technological limitations, but structural ones.
\item
  \textbf{The underlying reality is local and definite.} Indeterminacy
  belongs to the observer's description, not to the world.
\item
  \textbf{Incompleteness is structural, not deficient.} The limitation
  arises because the observer is made of the same elements as the
  universe it is trying to describe --- a physical form of
  self-reference. This is analogous to Gödel's incompleteness theorem,
  not to ignorance that better instruments could cure.
\item
  \textbf{The description is observer-relative.} Different partitions
  yield different emergent physics.
\item
  \textbf{The two descriptions are irreconcilable} --- not because one
  is wrong, but because they are complementary projections of a single
  reality that no embedded observer can access directly.
\end{enumerate}

Claims 1, 5, 6, and 7 enjoy broad philosophical support across nearly
every tradition examined. Claim 4 --- that the underlying reality is
definite --- is the paper's most philosophically isolated thesis.

\hypertarget{the-guxf6del-connection}{%
\subsubsection{The Gödel connection}\label{the-guxf6del-connection}}

The analogy between this framework and Gödel's incompleteness theorem is
not merely metaphorical. Gödel proved that a formal system rich enough
to encode arithmetic cannot prove all true statements about itself from
within --- the limitation arises because the system is self-referential,
capable of constructing sentences that refer to its own provability. The
observer in this framework faces a structurally parallel situation: the
reason a cosmological horizon exists at all is that the observer is a
physical subsystem made of the same fields, obeying the same
speed-of-light constraint, as the universe it is trying to describe. An
observer not made of the universe's own elements would not face a
horizon and would not be forced into a quantum description. The
incompleteness is a consequence of self-inclusion.

The connection can be made precise through Wolpert's limits of inference
{[}19{]}, which the paper cites. Wolpert proved, using diagonal
self-referential arguments directly descended from Gödel's, that any
inference device embedded in the universe it is trying to predict faces
fundamental limits --- not because of noise or finite resources, but
because complete self-prediction is logically impossible. The present
framework provides the concrete physical mechanism by which Wolpert's
logical limitation manifests: the causal partition enforces the
trace-out, the trace-out produces P-indivisibility, and P-indivisibility
is quantum mechanics. The \(10^{122}\) compression ratio --- the
Bekenstein-Hawking entropy of the cosmological horizon --- is the
quantitative measure of how much information self-inclusion forces the
observer to lose.

The key difference from Gödel is in the \emph{form} of self-reference.
Gödel's is syntactic: the system encodes a sentence that says ``I am not
provable.'' The framework's is physical: the observer is made of the
described, so complete description would require the observer to fully
encode its own state plus everything causally connected to it, which the
causal structure forbids. Wolpert sits between the two, using Gödelian
logic applied to physical systems. Together they form a chain: Gödel
(formal systems cannot completely describe themselves) \ensuremath{\to} Wolpert
(physical inference devices cannot completely predict the systems they
inhabit) \ensuremath{\to} this framework (the specific mechanism is the causal
partition, and the specific cost is quantum mechanics).

Hofstadter's \emph{Gödel, Escher, Bach} argues that self-referential
systems produce genuinely emergent higher levels --- ``strange loops''
where a system's hierarchical levels fold back on themselves, generating
properties invisible at the lower level. This framework agrees: quantum
mechanics is the real, irreducible description available to any embedded
observer, just as Hofstadter's ``I'' is real even though it emerges from
neurons. The strange loop is intact --- the observer, made of the
universe's own elements, generates through self-inclusion an emergent
description that governs everything the observer can access. The
disagreement with Hofstadter is narrow but consequential: he argues that
once the emergent level is established, the substrate beneath it is
explanatorily inert. The cosmological constant problem suggests
otherwise. The \(10^{122}\) discrepancy exists precisely because the
emergent quantum description and the classical substrate assign
different vacuum energies, and only the substrate's value matches
observation. The loop has a floor, and the floor matters.

\hypertarget{closest-ancestor-nicholas-of-cusa}{%
\subsubsection{Closest ancestor: Nicholas of
Cusa}\label{closest-ancestor-nicholas-of-cusa}}

Of all thinkers surveyed, the fifteenth-century cardinal Nicholas of
Cusa provides the most precise structural alignment. His \emph{docta
ignorantia} (learned ignorance) is essentially Claim 5: the highest
knowledge is knowing what we cannot know, and this is an intellectual
achievement, not a failure. His \emph{coincidentia oppositorum}
(coincidence of opposites) maps onto Claim 7 with remarkable precision:
contradictions irreconcilable in the finite realm dissolve in infinity
--- the infinitely large circle's circumference becomes a straight line,
the infinite polygon becomes a circle. QM and GR, irreconcilable within
any finite observational framework, would be coincident in the infinite
ground that generates both.

Most strikingly, Cusa's ``wall of Paradise'' from \emph{De Visione Dei}
maps onto Claim 3: an insurmountable boundary beyond which finite
intellect cannot pass. Scholars like Emmanuel Falque interpret reaching
this wall not as escaping through it but as \emph{inhabiting the
boundary} --- profoundly compatible with a framework where understanding
the limit is itself the deepest available insight.

\hypertarget{deepest-ontological-parallel-spinoza}{%
\subsubsection{Deepest ontological parallel:
Spinoza}\label{deepest-ontological-parallel-spinoza}}

Spinoza's attribute theory is arguably the single closest ontological
parallel. One substance (God/Nature) expresses itself through infinite
attributes, of which humans know only two: Thought and Extension.
Jonathan Bennett's ``barrier doctrine'' captures the key feature: each
attribute must be conceived through itself --- no explanatory flow
crosses between them. This is precisely Claim 7: two complete, correct
descriptions that are structurally irreconcilable, yet both describe the
same underlying reality. Spinoza's substance is fully determinate,
aligning with Claim 4. His parallelism doctrine --- the order of ideas
is the same as the order of things --- means the two descriptions track
the same structure through incommensurable vocabularies.

The divergence is epistemological: Spinoza believes reason achieves
adequate knowledge of reality through \emph{scientia intuitiva}
(intuitive knowledge). The paper's permanent limits directly contradict
this ambition.

\hypertarget{the-recurring-fault-line-claim-4}{%
\subsubsection{The recurring fault line: Claim
4}\label{the-recurring-fault-line-claim-4}}

Across every tradition examined, a striking pattern emerges. The claim
that underlying reality is ``local and definite'' faces resistance from
virtually every direction:

\textbf{Kant} prohibits positive characterization of the noumenal realm
as dogmatic metaphysics --- you can know \emph{that}
things-in-themselves exist but never \emph{what} they are.
\textbf{Hegel} diagnoses a performative contradiction: to posit a hidden
sector and characterize it as containing standard physics is already to
have crossed the boundary you claim is uncrossable. \textbf{Nietzsche}
attacks it as residual Platonism --- having correctly shown that all
observation is perspectival, the paper reinstates the very ``true
world'' his career was dedicated to destroying. \textbf{Wittgenstein}
rejects it as nonsensical: asserting what lies beyond the limits of the
sayable transgresses exactly the limits the paper identifies.

\textbf{Nagarjuna} identifies it as \emph{svabhāva}-reification ---
attributing inherent existence to what Buddhist emptiness
(\emph{śūnyatā}) says lacks it. \textbf{Daoism} warns against naming the
unnameable: the Dao that can be spoken is not the eternal Dao, and
calling reality ``definite'' is itself an act of conceptual carving.
\textbf{Whitehead} insists reality is fundamentally processual and
creative, involving genuine indeterminacy --- removing that
indeterminacy robs the cosmos of its creative character.

\textbf{Advaita Vedanta} comes closest to full support --- Brahman is
indeed a definite underlying reality --- but insists it is
\emph{accessible}: the observer IS the underlying reality, and
liberation (\emph{mokṣa}) consists in recognizing this identity. Every
Hindu and Buddhist soteriology rejects permanent inaccessibility.

Only Spinoza (whose fully determinate substance is naturally definite)
and Worrall's epistemic structural realism (which posits real but
unknowable natures behind structures) provide genuine philosophical
support for Claim 4.

\hypertarget{whats-genuinely-new}{%
\subsubsection{What's genuinely new}\label{whats-genuinely-new}}

The paper's philosophical lineage is not a single line of descent but a
mosaic. Its structural epistemology draws from Kant through Wittgenstein
to Wolpert. Its complementarity thesis combines Bohr's physics with
Cusa's coincidence of opposites and Daoist yin-yang. Its projection
metaphysics reaches through Plato's cave and Plotinus's emanation to
Advaita Vedanta's maya-Brahman distinction.

What is genuinely new is the \emph{combination}: accepting Bohr's
complementarity while insisting on Einstein's realism, grounding Kantian
limits in physical structure while claiming knowledge of noumenal
character, embracing Cusanian learned ignorance while grounding it in a
concrete physical mechanism --- the Gödel \ensuremath{\to} Wolpert \ensuremath{\to} causal partition
chain described above. The traditions reveal that this combination
creates a productive philosophical tension --- the paper claims to know
the character of what it proves unknowable. Whether this tension is a
contradiction (as Hegel would insist), a residual Platonism (as
Nietzsche would charge), or a legitimate achievement of learned
ignorance (as Cusa would celebrate) may be the framework's deepest
philosophical question.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{frequently-asked-questions}{%
\subsection{Frequently Asked
Questions}\label{frequently-asked-questions}}

\textbf{``Isn't this just another interpretation of quantum
mechanics?''} No.~Interpretations (Copenhagen, Many-Worlds, Bohmian
mechanics) accept the quantum formalism and disagree about what it
\emph{means}. This framework \emph{derives} the quantum formalism from
non-quantum premises and proves the derivation is the only possible one.
It makes quantitative predictions --- the value of \ensuremath{\hbar}, dark energy
evolution, and gravitational wave echoes --- that interpretations do
not, and it accounts for the dark-sector concordance as an automatic
consequence.

\textbf{``Doesn't Bell's theorem rule out hidden variable theories?''}
Bell's theorem rules out \emph{local} hidden variable theories
satisfying a specific factorizability condition. This framework violates
that factorizability --- not through faster-than-light signals, but
because P-indivisible joint dynamics don't permit the decomposition
Bell's theorem assumes. The framework reproduces exactly Tsirelson's
bound (the maximum quantum violation), no more and no less.

\textbf{``If everything is deterministic underneath, where does
randomness come from?''} From ignorance. The total system is
deterministic, but the observer can't access the hidden sector.
Different hidden states compatible with the same visible state lead to
different outcomes. The observer must assign probabilities --- not
because the universe is random, but because their information is
incomplete.

\textbf{``What about dark matter?''} The framework doesn't set out to
explain dark matter, but it produces a natural account of why
\textasciitilde95\% of the universe's gravitational budget has no source
in particle physics. The boundary entropy --- the modes traced out to
produce QM --- has thermal energy equal to the critical density.
Crucially, this thermal energy is computed from purely classical
quantities (the number of boundary modes and the classical horizon
temperature) --- it exists before the trace-out, unlike the QFT
zero-point energy, which is an artifact of the emergent description. The
boundary entropy is therefore gravitationally active (it curves
spacetime at the classical level) but invisible to the emergent QFT (the
boundary modes were traced out to produce it). The baryonic sector
accounts for \textasciitilde5\% of \(\rho_{\text{crit}}\); the rest is
boundary entropy. The uniform component is dark energy (Part II).
Whether the matter-induced \emph{redistribution} of this entropy
produces local effects resembling dark matter at galaxy scales is an
open question requiring a quantitative force law beyond the axioms. What
the framework proves is that the total dark sector budget is an
automatic consequence of the trace-out --- not an additional input.

\textbf{``Why does gravity `see' the classical vacuum energy and not the
quantum one?''} Because the spacetime metric exists at the classical
level, \emph{before} the quantum description emerges. The quantum
zero-point energy is a feature of the observer's compressed description.
It's real for quantum experiments (the Casimir effect, the Lamb shift)
but doesn't appear in the stress-energy tensor that governs curvature.
The \(10^{122}\) discrepancy is the information compression ratio ---
the entropy of the observer's blind spot.

\textbf{``How can the paper claim reality is `definite' if it's
permanently inaccessible?''} This is the paper's most philosophically
contested thesis (see \emph{Philosophical Lineage} above). The paper's
defense is that the claim follows from the derivation's own logic: the
axioms posit deterministic dynamics on a phase space, and the theorem
shows that quantum indeterminacy arises from tracing out part of that
phase space --- not from any indeterminacy in the underlying evolution.
The ``definiteness'' is a consequence of the starting premises, not a
speculative addition. Whether those premises are the right ones to start
from is, of course, an open question --- but within the framework, Claim
4 is a theorem, not an assumption.

\textbf{``Doesn't holographic physics show that spacetime comes from
entanglement?''} This is probably the strongest objection to the
framework's ordering --- classical spacetime first, quantum mechanics
second. The Ryu-Takayanagi formula says entanglement entropy equals
boundary area divided by \(4G\). Van Raamsdonk argued that reducing
entanglement disconnects spacetime. Programs like ER=EPR and ``it from
qubit'' read these results as evidence that quantum entanglement is
prior to geometry.

The framework offers an alternative reading. If the quantum description
is produced by tracing out over a geometric boundary, then \emph{of
course} its entanglement entropy is proportional to the boundary's area
--- the information content of the trace-out is set by the number of
modes crossing the boundary, which scales with area. The Ryu-Takayanagi
formula, on this account, isn't a hint that entanglement builds
geometry; it's a consequence of the fact that geometry built the
entanglement. The Bekenstein-Hawking entropy \(S = A/(4\,l_p^2)\), which
the paper derives, is exactly this statement.

The correlation between entanglement and geometry is real either way.
The question is which direction the arrow of explanation points. The two
orderings make different empirical predictions: if geometry emerges from
entanglement, spacetime structure should break down at high energy; if
quantum mechanics emerges from geometry, the quantum description should
break down near the discreteness scale while the geometric substratum
persists. The gravitational wave echo prediction provides one test. But
the cumulative case is already substantial: the geometry-first ordering
produces the observed vacuum energy, the Bekenstein-Hawking \(1/4\)
factor, the value of \(\hbar\), the dark-sector concordance, and dark
energy evolution consistent with DESI data --- six independent
consequences matching observation. The quantum-first ordering produces
the \(10^{122}\) discrepancy, treats \(\hbar\) and the \(1/4\) factor as
unexplained inputs, and has no natural account of the dark sector.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-decompression-algorithm}{%
\subsection{The Decompression
Algorithm}\label{the-decompression-algorithm}}

There is a way to state what the framework means that is sharper than
anything in the formal paper.

An observer inside the universe receives incomplete data --- the visible
sector only. The characterization theorem proves that there is exactly
one self-consistent algorithm for making predictions from this
incomplete data. That algorithm is quantum mechanics.

Everything in QM is a feature of the algorithm, not a feature of the
underlying reality. The wave function is the algorithm's internal state
variable --- the bookkeeping device it uses to track what it knows and
what it doesn't. Complex amplitudes are the algorithm's arithmetic ---
the specific number system the reconstruction requires. Interference is
what happens when the algorithm combines two incomplete pathways and
their bookkeeping entries partially cancel. Entanglement is the
algorithm's encoding of correlations that were written into the hidden
sector during preparation and haven't been read back yet. The Born rule
is the algorithm's output format --- the way it converts its internal
state into predictions the observer can check.

None of these are properties of the deterministic substratum. In the
substratum, there are no wave functions, no complex numbers, no
interference, no entanglement. There are configurations evolving under a
Hamiltonian flow. That's all. The entire apparatus of quantum mechanics
--- every textbook, every equation, every experiment --- is what the
decompression algorithm looks like from the inside.

\hypertarget{antimatter-as-algorithmic-artifact}{%
\subsubsection{Antimatter as algorithmic
artifact}\label{antimatter-as-algorithmic-artifact}}

This reframing changes the understanding of specific phenomena. Consider
antimatter.

The standard account: the Dirac equation, which combines quantum
mechanics with special relativity, has solutions with both positive and
negative energy. The negative-energy solutions correspond to
antiparticles --- particles with opposite charge and quantum numbers.
Every particle must have an antiparticle, because CPT invariance (a
theorem of any local Lorentz-invariant quantum field theory) requires
it.

The framework's account: the trace-out over the hidden sector forces the
reconstruction algorithm to use two-signed amplitudes. When the
algorithm operates in a relativistic context --- which it must, because
the substratum's causal structure is relativistic --- the two-signed
amplitude structure becomes the two-signed energy structure of the Dirac
equation. Negative-energy solutions aren't additional features of
reality. They're what the algorithm requires for self-consistency when
the incomplete data comes from a relativistic system.

The parallel to the coin-and-die model is direct. The intermediate
propagator:

\[\Lambda(2,1) = \begin{pmatrix} 2 & -1 \\ -1 & 2 \end{pmatrix}\]

has entries of \(-1\) --- ``anti-probabilities'' that don't exist in the
substratum (where every transition probability is between 0 and 1).
These negative entries are the minimal-model ancestors of antimatter.
They arise for the same reason: the algorithm can't reconstruct the
observed dynamics without them. In the toy model, the \(-1\) entries
encode the fact that the die remembers and reverses the coin flip ---
information backflow that no positive-entry matrix can describe. In the
relativistic case, the negative-energy solutions encode the fact that
the relativistic trace-out requires a doubled Hilbert space to remain
self-consistent.

In both cases: the substratum has no doubling. The algorithm demands
one.

\hypertarget{nothing-in-quantum-mechanics-explains-a-fundamental-phenomenon}{%
\subsubsection{Nothing in quantum mechanics explains a fundamental
phenomenon}\label{nothing-in-quantum-mechanics-explains-a-fundamental-phenomenon}}

This is the deepest implication of the framework, and it is worth
stating plainly.

The wave function does not explain what an electron is doing between
measurements. It encodes what the algorithm computes given the
observer's data. Interference does not explain why particles behave like
waves. It reflects the algorithm's method of combining incomplete
information. Entanglement does not explain a mysterious connection
between distant particles. It reflects correlations stored in the hidden
sector that the algorithm must track but cannot directly access.
Antimatter does not explain a second kind of fundamental stuff. It
reflects the algorithm's need for two-signed amplitudes in a
relativistic context.

Every quantum phenomenon is an output of the decompression algorithm.
The only fundamental phenomenon is the compression itself: the observer
cannot see past the horizon, and the characterization theorem dictates
the unique form of the resulting reconstruction.

This doesn't make quantum phenomena less real. Temperature is also an
emergent phenomenon --- a single molecule has no temperature. But
temperature boils water, drives engines, and burns skin. Its emergence
from statistical mechanics doesn't diminish its causal power within the
emergent description. The same holds for every quantum phenomenon: they
are causally potent, experimentally real, and the only physics
accessible to an embedded observer. But they are features of the
observer's necessary algorithm, not features of the universe's
fundamental structure.

This resolves what is arguably the oldest open question in quantum
foundations: \emph{is the wave function real?} Since 1926, physics has
been split between epistemic interpretations (the wave function is just
a bookkeeping device tracking the observer's knowledge --- it doesn't
correspond to anything physical) and ontic interpretations (the wave
function is a real physical entity --- a field on configuration space,
or a branching structure, or a guiding wave). The framework shows that
both sides share a hidden assumption: that ``real'' means
``fundamental.'' The epistemic camp says the wave function isn't
fundamental, therefore it isn't real. The ontic camp says it's real,
therefore it must be fundamental. Both are wrong. The wave function is
real \emph{and} not fundamental --- in exactly the same way as
antimatter, temperature, and every other emergent phenomenon. Within the
bounded projection --- which is all an embedded observer will ever have
access to --- the wave function, the positron, the interference pattern,
and the electromagnetic field all have exactly the same ontological
status. They are all outputs of the decompression algorithm. They are
all experimentally real. And none of them exist in the substratum.

The analogy to data compression is precise. Lossy compression (like JPEG
or MP3) discards information and reconstructs an approximation. The
reconstruction has artifacts --- ringing near edges in JPEG, pre-echo
before transients in MP3. These artifacts are not in the original
signal. They are features of the reconstruction algorithm applied to
incomplete data.

Quantum mechanics is the reconstruction. The \(10^{122}\) discarded bits
(the hidden sector) are the lost information. Wave functions,
interference, entanglement, and antimatter are the artifacts. They are
real --- as real as JPEG ringing is visible --- but they are properties
of the reconstruction, not of the original.

The difference from ordinary compression: with JPEG, you can access the
original file. With the universe, you cannot. The reconstruction is all
we will ever have. Its artifacts are our physics. And the
characterization theorem proves that no other reconstruction is possible
--- this is the unique algorithm, and its artifacts are the unique
artifacts. Quantum mechanics is not one possible decompression of
incomplete cosmological data. It is the only one.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{what-this-means}{%
\subsection{What This Means}\label{what-this-means}}

The search for a ``theory of everything'' that unifies quantum mechanics
and general relativity has assumed that the two theories describe the
same level of reality and must be reconciled there. This paper argues
that assumption is wrong. The two theories operate at different levels
--- one fundamental, one emergent --- and their apparent contradiction
is the information-theoretic cost of being an observer trapped inside
the system you're trying to describe.

The apparent conflict between QM and GR, the \(10^{122}\) vacuum energy
discrepancy, and the dark sector are three faces of the same fact. The
first is the existence of the trace-out. The second is its information
compression ratio. The third is its gravitational occlusion fraction.
All three are mandatory consequences of observational incompleteness ---
and all three match what we observe.

The universe is not broken. We are observing it from within.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{corroboration-the-rigidity-test}{%
\subsection{Corroboration: The Rigidity
Test}\label{corroboration-the-rigidity-test}}

A natural objection to any framework this sweeping is: maybe it only
works because it was built to work. Maybe the axioms were chosen to
produce QM, and the cosmological application was chosen because it fits.
A flexible framework that can accommodate anything predicts nothing.

A companion paper (``Dynamics Selection and Emergent General Relativity
in the Observational Incompleteness Framework'') tests this by asking a
question the main paper doesn't address: if you build a concrete system
satisfying the axioms, does the framework constrain the dynamics? And if
so, does the constrained dynamics produce anything beyond QM ---
something the framework wasn't designed to deliver?

It does, on both counts.

\textbf{The dynamics is unique.} Among all second-order reversible
nearest-neighbor dynamics on a lattice, the requirements of center
independence (necessary for emergent QM), spatial isotropy, and
linearity select exactly one: the discrete wave equation. This holds for
any alphabet size and any dimension. Center independence is necessary
because center-dependent rules allow the visible sector to partially
predict itself, suppressing the information backflow that produces QM
--- an effect proven analytically on the full lattice via an
information-screening mechanism. Linearity is selected by three
independent criteria: it gives maximum propagation speed, it uniquely
maximizes P-indivisibility among linear rules, and it is the only choice
for which horizons are in thermal equilibrium (nonlinear dispersion
breaks the Unruh effect).

\textbf{The dynamics passes seven independent checks.} The wave equation
--- selected solely by the QM requirement plus symmetry --- turns out to
produce: (1) non-Markovian reduced dynamics (emergent QM), (2) a causal
structure with the correct spacetime dimension, (3) the gap equation for
\ensuremath{\hbar}, (4) entanglement entropy proportional to boundary area, (5) a
Lorentz-invariant dispersion relation, (6) the correct Unruh temperature
at horizons, and (7) all inputs to Jacobson's thermodynamic derivation
of Einstein's field equations. Each of these is an independent check
that could have come out wrong --- the entropy could have scaled with
volume, the dispersion could have been non-relativistic, the horizon
state could have been non-thermal. None of these failures occurs.

\textbf{No free parameters.} The wave equation is selected, not chosen.
The lattice spacing is fixed by the gap equation. The entropy
coefficient is determined by thermal matching. The dispersion relation
is an algebraic identity. There is nothing to tune.

This is what rigidity looks like. A framework that produces QM from one
set of arguments and then, without modification, produces the inputs for
GR from a completely different set of arguments --- lattice dynamics,
dispersion relations, entanglement Hamiltonians --- is making a
structural claim that goes well beyond the original derivation. Each
passed check is a point where the framework could have been falsified
and wasn't. Seven passed checks with no free parameters is not proof,
but it is the kind of evidence that distinguishes a framework describing
something real from one that was engineered to fit.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{what-is-the-lattice}{%
\subsection{What Is the Lattice?}\label{what-is-the-lattice}}

The rigidity test proves that the wave equation on a lattice produces
both QM and GR. But what \emph{is} the lattice? Is the universe
literally a grid of cells at the Planck scale? What would the grid be
made of? What would it sit in?

A third companion paper (``The Substrate Problem: Structural Foundations
of the Observational Incompleteness Framework'') addresses these
questions head-on. Its answer: the lattice is not a physical object. It
is the \emph{coupling structure} of the dynamics --- the pattern of
which degrees of freedom affect which others.

\textbf{The minimal structure.} The paper audits every assumption in the
framework and identifies which are necessary for the theorems and which
are artifacts of the particular construction. The result: only six
structural properties matter --- deterministic bijectivity, finite
boundary entropy, bounded coupling degree, statistical isotropy,
non-trivial partition coupling, and slow-bath capacity. The regular
cubic lattice, the specific alphabet size q, the dimensionality d, and
even the wave equation are all either derived from these six properties
or irrelevant to the predictions. The minimal object is not a lattice.
It is a triple: a finite set S, a bijection \ensuremath{\varphi}, and a partition V.

\textbf{Space as coupling graph.} Given any bijection on a finite set,
you can ask: which components of the state affect which others? The
answer defines a graph --- sites are vertices, and two sites are
connected if the dynamics couples them. This graph is not drawn on a
pre-existing space. It IS space, at the fundamental level. Distance
means ``how many coupling steps apart.'' Area means ``how many edges
cross the boundary.'' The area law, the dispersion relation, the
Myrheim-Meyer dimension --- every spatial property used in the
derivation chain is a property of this coupling graph.

This dissolves the container problem. The lattice doesn't sit in
anything. There is no ambient manifold. The coupling graph is an
abstract mathematical structure --- like a number or a group --- and any
spatial embedding we draw is a representation for our convenience, not a
physical fact. Asking ``what does the lattice sit in?'' is like asking
what the number 7 sits in. It's a category error.

\textbf{The alphabet is a gauge freedom.} The companion papers prove
every result for any alphabet size q \ensuremath{\geq} 2. The gap equation contains no
q. The Bekenstein-Hawking formula contains no q. The dispersion relation
is valid for any q. No experiment could measure q, even in principle.
This makes q a gauge freedom --- a choice of mathematical description,
like choosing a coordinate system or a gauge for the electromagnetic
potential. The physical content is the coupling structure, not the
microscopic state space.

\textbf{Connection to causal set theory.} The bijection's coupling
graph, extended in time, generates a causal partial order: event A
precedes event B if B is within A's future coupling light cone. This
partial order is a causal set in the sense of Bombelli, Lee, Meyer, and
Sorkin --- the starting point of causal set theory, one of the
established approaches to quantum gravity. Causal set theory has always
had a specific gap: it postulates a causal order but lacks a
deterministic dynamics that produces QM and GR. The OI framework
provides exactly this. Sorkin's slogan is ``Order + Number = Geometry.''
The OI version is ``Bijection + Locality = QM + GR.''

\textbf{The hierarchy of physics.} If the fundamental object is (S, \ensuremath{\varphi})
--- a finite set and a bijection --- then every concept in physics is a
different aspect of this pair:

\begin{itemize}
\tightlist
\item
  \textbf{\ensuremath{\varphi}} is the dynamical law --- the complete rule mapping states
  to states.
\item
  \textbf{Space} is the coupling structure of \ensuremath{\varphi} --- which degrees of
  freedom affect which others. It is determined by \ensuremath{\varphi} as the
  factorization minimizing coupling degree. Space is not a container; it
  is a relationship.
\item
  \textbf{Matter} is the state --- the values assigned to the degrees of
  freedom. A particle is a localized pattern that propagates through the
  coupling graph. Space and matter are the graph topology and the graph
  coloring; both come from (S, \ensuremath{\varphi}).
\item
  \textbf{Energy} is the rate of change of the state under iteration of
  \ensuremath{\varphi}. A high-energy excitation changes rapidly from step to step; the
  vacuum changes least. Energy is not a substance --- it is a measure of
  how fast a pattern evolves.
\item
  \textbf{Time} is the iteration itself --- the stepping from one state
  to the next. There is no continuous time at the fundamental level.
\item
  \textbf{Quantum mechanics} is the observer's compressed view of V. It
  exists because the observer must marginalize over the hidden sector.
\item
  \textbf{Gravity} is the thermodynamic limit of the coupling structure
  --- the macroscopic behavior of bounded-degree graphs with area-law
  entropy.
\item
  \textbf{Conservation laws} are emergent. Bijectivity of \ensuremath{\varphi} preserves
  state-space volume (information conservation). Energy conservation is
  what this looks like in the emergent quantum description (Noether's
  theorem). Momentum conservation reflects the symmetry of the coupling
  graph.
\end{itemize}

None of these are independent substances. They are the same structure
viewed at different scales or from different angles. The framework does
not unify them by reducing them to a common material. It unifies them by
showing they were never separate.

\textbf{What remains.} The structural reading leaves remarkably little
genuinely open. The spatial dimensionality \(d = 3\) is derived by four
independent self-consistency filters (the dark sector concordance,
propagating gravity, stable matter, and renormalizability). Background
independence is achieved through state-dependent bijections, with the
discrete Einstein equation identified as the Ollivier-Ricci curvature
condition. The observer is proved to be generic: any small subgraph of
any large bounded-degree bijection satisfies the conditions for emergent
QM. The fundamental object is (S, \ensuremath{\varphi}) --- a finite set and a bijection.
The partition V, the dimension, and the laws of physics are all derived.
What remains undetermined are the 18+ parameter values of the Standard
Model, which depend on the specific bijection \ensuremath{\varphi} --- analogous to how
Einstein's equations don't determine the mass of Jupiter.

But the central claim stands: the fundamental object is (S, \ensuremath{\varphi}) --- a
finite set and a bijection. The observer, the dimension, and the laws of
physics are all emergent.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{why-these-particles}{%
\subsection{Why These Particles?}\label{why-these-particles}}

The first three papers establish that the framework produces quantum
mechanics and general relativity. But quantum mechanics is a
\emph{framework}, not a \emph{theory}. It tells the observer to use
Hilbert spaces, unitary evolution, and the Born rule --- but it's
compatible with infinitely many different quantum field theories. You
could have quantum mechanics with an SU(7) gauge group, or with 15
generations of fermions, or with no gauge fields at all. Deriving QM
answers the question ``what kind of probability does the observer use?''
It doesn't answer ``what particles exist?'' or ``what forces act between
them?''

The Standard Model of particle physics --- quarks, leptons, the strong
and weak nuclear forces, electromagnetism, the Higgs boson --- has been
confirmed to extraordinary precision. But its \emph{structure} has
always been taken as empirical. We observe three generations, SU(3) \ensuremath{\times}
SU(2) \ensuremath{\times} U(1), specific hypercharge assignments, and we accept them as
given. Nobody has explained \emph{why} these particles and not others.

A fourth companion paper (``Why These Particles: Standard Model
Structure from Observational Incompleteness'') asks whether the specific
lattice dynamics selected by the QM and GR requirements --- the wave
equation on a d = 3 hypercubic lattice with checkerboard partition ---
determines which quantum field theory the observer sees. The answer is
yes --- almost completely.

\textbf{Fermions from the wave equation.} The wave equation selected by
the QM requirement (the discrete lattice Klein-Gordon equation) has a
well-known mathematical property: it factors into first-order operators
called staggered Dirac operators. This factorization, discovered by
Susskind in 1977, means the \emph{same dynamics} that produces bosonic
waves also describes fermionic matter. Fermions are not added to the
framework. They \emph{are} the framework, seen from a different
mathematical angle.

The factorization produces a specific structure: on a three-dimensional
lattice, the staggered construction yields exactly 4 ``tastes'' ---
independent species of fermions. These decompose as 1 + 3 under the
cubic symmetry of the lattice: one singlet and one triplet. The triplet
count equals the spatial dimension \(d = 3\). Since \(d = 3\) is derived
by the Ontology paper, the three-generation structure of the Standard
Model --- the fact that there are three copies of each type of quark and
lepton --- is traced back to the dimensionality of space.

Numerically, the singlet and triplet are not just distinguished by
symmetry --- they have quantitatively different coupling strengths. The
singlet has coupling \(|\mu| = 1\) (the maximum), while the three
triplet members each have \(|\mu| = 1/3\). The triplet members are
exactly degenerate by cubic symmetry, providing a lattice-level origin
for ``generation symmetry.''

\textbf{The Higgs mechanism from one condition.} The same center
independence condition (\(\alpha = 0\)) that produces quantum mechanics
also enforces exact chiral symmetry in the staggered fermions. This
means fermion mass terms are forbidden in the fundamental Lagrangian.
The \emph{only} way to give fermions mass while preserving unitarity and
renormalizability is the Higgs mechanism. One algebraic condition ---
center independence --- simultaneously produces QM, chiral fermions, and
the necessity of a Higgs boson.

\textbf{Gauge structure from a matrix.} The wave equation naturally
generalizes to multiple components. Each lattice site carries a
\(K\)-component vector instead of a single number. The selection
criteria (center independence, isotropy, reversibility, linearity)
uniquely produce the \emph{matrix wave equation}, where a coupling
matrix \(M\) is the sole new parameter.

This matrix \(M\) determines everything about the gauge structure. Its
eigenvalue multiplicities give the gauge group --- the group of
transformations that leave \(M\) invariant (technically, its commutant
in U(K)). Its eigenvalues give the mass spectrum. The gauge group and
the mass spectrum are dual descriptions of the same matrix. This is a
theorem, not an approximation.

\textbf{Why K = 6.} In three dimensions, each lattice site has 6 nearest
neighbors (\ensuremath{\pm}x, \ensuremath{\pm}y, \ensuremath{\pm}z). If the internal components correspond to these
link directions --- the natural choice from the factorization principle,
which says to decompose the per-site state space into its independent
dynamical channels --- then \(K = 2d = 6\).

These 6 link directions form a representation of the cubic rotation
group \(O\). Standard character theory decomposes this representation
into irreducibles:

\[6 = T_1(3) \oplus E(2) \oplus A_1(1)\]

The dimensions are 3, 2, and 1. By Schur's lemma, the coupling matrix
\(M\) acts as an independent scalar on each irreducible subspace:
\(M = \text{diag}(\mu_c I_3, \mu_w I_2, \mu_y)\). The gauge group ---
the commutant of \(M\) --- is therefore:

\[U(3) \times U(2) \times U(1) \supset SU(3) \times SU(2) \times U(1)\]

This is the Standard Model gauge group. It comes from the representation
theory of the cubic lattice in three spatial dimensions. The three
factors correspond to the three irreducible representations of \(O\):
the vector (\(T_1\), dimension 3) gives color SU(3), the quadrupole
(\(E\), dimension 2) gives weak SU(2), and the scalar (\(A_1\),
dimension 1) gives hypercharge U(1).

\textbf{Chiral coupling.} The Standard Model has a peculiar feature: the
weak force treats left-handed and right-handed particles differently.
Left-handed particles form SU(2) doublets; right-handed particles are
SU(2) singlets. No other known force has this asymmetry. Where does it
come from?

In the OI framework, the answer is the partition itself. The
checkerboard partition that produces quantum mechanics --- even sites
visible, odd sites hidden --- is \emph{identical} to the chirality
decomposition of staggered fermions. Even sites carry left-handed spinor
components; odd sites carry right-handed. The trace-out that produces QM
simultaneously integrates out the right-handed sector. The staggered
Dirac operator has the property \(D_{LL} = D_{RR} = 0\): it couples
\emph{only} across sublattices, never within them. After the trace-out,
the effective Lagrangian has only left-handed external legs ---
precisely the chiral coupling structure of SU(2)\(_L\).

Meanwhile, color SU(3) acts on the \emph{internal} components (the
\(K\)-component index at each site), which are preserved by the spatial
trace-out. Both left and right fields carry color. SU(3) remains
vector-like. The distinction: chirality lives in the spatial structure
(affected by the trace-out); color lives in the internal structure
(unaffected).

\textbf{The strong CP problem solved.} QCD has a parameter
\(\bar{\theta}\) that violates CP symmetry. Experimentally,
\(|\bar{\theta}| < 10^{-10}\). Why is it so small? This is the strong CP
problem. The standard solution invokes a new particle (the axion).

The OI framework solves it without new particles. The wave equation is
exactly invariant under time reversal \(T\). The trace-out is a spatial
operation --- it doesn't break \(T\). Therefore the emergent QFT
inherits \(T\)-invariance. The \(\theta\)-term in QCD is \(T\)-odd, so
\(T\)-invariance forces \(\theta = 0\). The physical parameter
\(\bar{\theta} = \theta + \arg\det(Y_u Y_d)\) also vanishes because the
visible-sector transition matrix satisfies detailed balance
(\(T_{ij} = T_{ji}\), proved and verified numerically), which constrains
the Yukawa matrices to be real. The prediction: \(\bar{\theta} = 0\)
exactly, no axion, and the neutron electric dipole moment is exactly
zero.

\textbf{No empirical inputs.} The hypercharge U(1)\(_Y\) ---
traditionally an empirical input --- is automatic: any U(\(n\)) gauge
group contains a U(1) factor, and anomaly cancellation uniquely
determines the hypercharge assignments. The parity violation --- also
traditionally empirical --- follows from the partition's chirality
structure. The filter chain from (S, \ensuremath{\varphi}) to the Standard Model contains
no empirical inputs beyond the framework's two structural properties
(bounded coupling and statistical isotropy).

\textbf{Where the matter representations come from.} A subtle point: the
K = 6 internal space decomposes as a \emph{direct sum} 3 + 2 + 1, but
the SM quark doublet Q\(_L\) = (3, 2, +1/6) is a \emph{tensor product}
representation --- it transforms simultaneously under both SU(3) and
SU(2). This representation doesn't come from the lattice geometry. It
comes from the \emph{observer's existence}: the observer genericity
theorem (proved in the Ontology paper) guarantees that any observer has
a self-consistent quantum description. A self-consistent chiral QFT
requires anomaly cancellation. Anomaly cancellation with SU(3) \ensuremath{\times} SU(2) \ensuremath{\times}
U(1) uniquely determines the matter content per generation: Q\(_L\),
u\(_R\), d\(_R\), L\(_L\), e\(_R\) with specific hypercharges. The
lattice gives the gauge group and the generation count; the observer's
self-consistency gives the representations. Three degenerate tastes then
give three copies --- three generations.

\textbf{What the framework does not determine.} The coupling matrix
\(M\) has three independent eigenvalues (\(\mu_c\), \(\mu_w\),
\(\mu_y\)), constrained by the max-speed condition to have
\(\mu_y = 1\). The remaining two eigenvalues set the strong and weak
coupling strengths at the lattice scale. These, along with the Yukawa
couplings (which determine fermion masses and mixing angles) and the
Higgs potential parameters, are properties of the specific bijection
\(\varphi\) --- the specific ``universe'' the observer inhabits. The
framework determines the \emph{form} of the laws of physics but not the
\emph{constants}.

\textbf{Predictions.} The SM paper makes seven predictions: (1) no
additional gauge groups below the Planck scale; (2) no fundamental
scalars beyond the Higgs; (3) no supersymmetric partners; (4) no fourth
generation; (5) \(\bar{\theta} = 0\) exactly, no axion; (6) the CKM
CP-violating phase arises from parity violation, not time-reversal
violation; (7) neutrinos are Majorana --- the singlet taste is the Higgs
sector (not a right-handed neutrino), so neutrinoless double beta decay
should be observed.

Combined with the earlier papers' predictions --- dark energy evolution
with \(\nu_{\text{OI}} \approx 2.45 \times 10^{-3}\), gravitational wave
echoes, and dark gravity --- the framework makes a suite of falsifiable
predictions that no competing theory makes jointly.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-full-picture}{%
\subsection{The Full Picture}\label{the-full-picture}}

The framework now covers the three pillars of modern physics:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2105}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4211}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3684}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Pillar
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What is derived
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source paper
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Quantum mechanics} & Emerges from embedded observation & Main
paper {[}1{]} \\
\textbf{General relativity} & Emerges from lattice dynamics passing 7
checks & GR paper {[}2{]} \\
\textbf{Standard Model structure} & SU(3) \ensuremath{\times} SU(2) \ensuremath{\times} U(1), 3 generations,
Higgs, chiral coupling, \ensuremath{\bar{\theta}} = 0 & SM paper {[}5{]} \\
\textbf{Structural foundations} & \(d = 3\), coupling graph ontology,
\(q\)-gauge, background independence & Ontology paper {[}3{]} \\
\end{longtable}

The starting point for all of it: a finite set, a bijection, and a
partition.

The framework does not claim to be a complete theory of everything. It
does not determine the 18+ parameters of the Standard Model --- these
depend on the specific bijection \ensuremath{\varphi}, just as the mass distribution of the
universe depends on initial conditions, not on Einstein's equations.
What it does claim --- and what the derivation chain supports --- is
that the \emph{structure} of the known laws of physics is not arbitrary.
It is what mathematical consistency looks like when a finite
deterministic system is observed from within. The fundamental object is
(S, \ensuremath{\varphi}): a finite set and a bijection. Everything else --- the observer,
the dimension, quantum mechanics, general relativity, and the Standard
Model --- is emergent.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{what-is-s-ux3c6}{%
\subsection{What Is (S, \ensuremath{\varphi})?}\label{what-is-s-ux3c6}}

The papers derive what (S, \ensuremath{\varphi}) \emph{produces}. But what kind of thing
\emph{is} it?

\hypertarget{storage-and-memory}{%
\subsubsection{Storage and memory}\label{storage-and-memory}}

S is the set of all distinguishable states. Its physical meaning is
\emph{finite capacity}: there are \textbar S\textbar{} configurations
that can be told apart, carrying log\ensuremath{_2}\textbar S\textbar{} bits of
information. The finiteness is not a regularization --- it's
load-bearing. Finite sets recur (\ensuremath{\varphi}\^{}N = id), recurrence produces
P-indivisibility, and P-indivisibility produces quantum mechanics. An
infinite S wouldn't recur, and the whole mechanism breaks.

\ensuremath{\varphi} is a bijection: every state has exactly one predecessor and exactly
one successor. Its physical meaning is \emph{perfect memory}.
Information is never created, never destroyed. The past is always
recoverable from the present, because \ensuremath{\varphi}\ensuremath{^{-}}\ensuremath{^1} exists. A non-bijective map
would send two states to the same successor, erasing the distinction
between them. \ensuremath{\varphi} preserves all distinctions, permanently.

Together, (S, \ensuremath{\varphi}) is a \emph{finite lossless memory}: a register with
finite capacity whose contents are updated without loss at each step.
The partition V defines the observer's \emph{read access} --- which
portion of the register they can see. The hidden sector H is the rest:
storage that the dynamics writes and reads but the observer cannot.

The framework's results all have a clean interpretation in this
language. The conditions C1--C3 are access conditions on the memory: C1
means the readable and unreadable parts are linked (writes to one affect
the other). C2 means the unreadable part retains its contents between
reads. C3 means the unreadable part is large enough to store the full
interaction history without overflow.

Quantum mechanics is the \emph{read statistics} of a lossless memory
through a partial interface. The Schrödinger equation governs how read
probabilities evolve. The Born rule is the equilibrium distribution over
unread states. Interference is the signature of information written to
hidden storage and read back later. The 10\ensuremath{^1}\ensuremath{^2}\ensuremath{^2} cosmological constant
discrepancy is the compression ratio between total storage and readable
storage. The dark sector is the gravitational weight of the unreadable
storage. The Bekenstein-Hawking entropy is the storage capacity of the
partition boundary. \ensuremath{\hbar} is the conversion factor between storage geometry
and read statistics.

An immediate objection: ``Memory made of \emph{what}?'' In everyday
life, memory requires a physical substrate --- silicon, magnetic
domains, neurons. The answer: (S, \ensuremath{\varphi}) is a \emph{complete description} of
reality --- it determines every observable, with nothing left over that
any measurement could detect. S is the totality of distinguishable
configurations; \ensuremath{\varphi} is the structure of dynamical dependence. Space, time,
and matter are all derived from (S, \ensuremath{\varphi}), so they can't appear in its
definition without circularity. Whether (S, \ensuremath{\varphi}) \emph{is} reality or
\emph{describes} reality is a question the framework proves to be gauge
--- undecidable by any experiment. The ``memory'' is a complete
mathematical description from which everything observable is derived.
Asking what it's made of presupposes a deeper level that no measurement
can access --- like asking what the number 7 is made of.

\hypertarget{not-a-computer}{%
\subsubsection{Not a computer}\label{not-a-computer}}

A Turing machine has a tape (storage), a head (read/write access), and a
transition function (update rule). That maps suggestively onto (S, \ensuremath{\varphi},
V). But three differences matter.

A Turing machine's tape is potentially infinite. S is finite --- and
finiteness is what produces QM.

A Turing machine is generally irreversible: it can erase, overwrite,
halt. \ensuremath{\varphi} is a bijection: nothing is ever erased, nothing is ever created,
there is no halting. This is what makes (S, \ensuremath{\varphi}) a lossless memory rather
than a computation.

Most fundamentally: a Turing machine computes an \emph{extrinsic
function} --- it maps input to output for an external user who feeds in
the input and reads the result. (S, \ensuremath{\varphi}) has no external user. There is no
input, no output, no halting, no answer. It's a closed permutation that
cycles through states and returns.

An information theorist might object: any reversible finite automaton
\emph{is} a computer --- the initial state is the input, the final state
is the output, and the cycle is the computation. Formally, that's
correct. But the distinction isn't formal --- it's about \emph{purpose}.
A Turing machine computes something \emph{for someone}. (S, \ensuremath{\varphi}) computes
nothing for nobody. It's a closed system; there's no recipient of the
``answer.'' What the framework shows is that the appearance of
purposeful physical law --- evolution equations, conserved quantities,
gauge symmetries --- arises entirely from the \emph{internal}
perspective of the partition. The complexity is in the observer, not in
the map.

People ask: ``Is the universe a computer?'' The framework's answer: the
universe is something \emph{less} than a computer. It's a finite
lossless cycle. And physics is the \emph{appearance of complexity}
created by observing a simple thing through a keyhole.

\hypertarget{the-arrow-of-time}{%
\subsubsection{The arrow of time}\label{the-arrow-of-time}}

If the universe is a finite permutation, it must eventually cycle back:
\ensuremath{\varphi}\^{}N = id for some N. So where does the arrow of time come from? Why
does entropy increase?

The answer: entropy increase is a property of the observer's
coarse-grained description, not of the substratum. The observer sees
transition probabilities, not the full state. From their perspective,
the visible sector evolves stochastically, and their probability
distribution spreads over accessible states --- entropy increases. This
is the standard Boltzmann mechanism: coarse-graining a reversible
microscopic dynamics produces irreversible macroscopic behavior.

The substratum itself has no arrow of time. \ensuremath{\varphi} and \ensuremath{\varphi}\ensuremath{^{-}}\ensuremath{^1} are equally valid.
The total fine-grained entropy is exactly conserved. The Second Law is a
theorem about the read statistics, not about the memory. Like quantum
mechanics itself, the direction of time is emergent --- a consequence of
observing a lossless cycle from within.

\hypertarget{the-incompleteness-family}{%
\subsubsection{The incompleteness
family}\label{the-incompleteness-family}}

This places the framework in a family of results where self-reference
under finite resources produces rigid structure.

\textbf{Gödel (1931):} A sufficiently powerful formal system cannot
prove all truths about itself. The unprovable statements aren't random
--- they have a precise structure (the Gödel sentence, the consistency
statement, the arithmetical hierarchy).

\textbf{Turing (1936):} A universal computer cannot decide all questions
about its own behavior. The undecidable problems aren't random --- they
have a precise structure (the halting problem, Turing degrees, the
oracle hierarchy).

\textbf{OI:} An observer embedded in (S, \ensuremath{\varphi}) cannot access the complete
state. The emergent description isn't random --- it has a precise
structure (unitary quantum mechanics with the Schrödinger equation, Born
rule, and Bell violations).

The common thread: a system with finite resources tries to completely
model something it's part of, and the structural impossibility of doing
so determines the \emph{form} of what it produces instead. The
limitation and the law are the same object viewed from two sides.

Gödel and Turing are usually read as negative results --- limits on
knowledge. The framework recasts the physical instance as
\emph{generative}. Quantum mechanics is not what we're stuck with
because we can't see the full state. It's the unique, mathematically
rigid consequence of finite lossless memory observed from within.

The universe is completely described by a finite memory that never
forgets. Physics is what that memory looks like from inside.

\hypertarget{why-does-mathematics-describe-physics}{%
\subsubsection{Why does mathematics describe
physics?}\label{why-does-mathematics-describe-physics}}

This reframes one of the oldest questions in the philosophy of science.

The standard view: mathematics \emph{describes} physics. The physical
world exists independently, and we discover that mathematical equations
capture its behavior. Wigner called this the ``unreasonable
effectiveness of mathematics'' --- why should abstract structures
developed by pure mathematicians turn out to describe the physical
world?

Tegmark's answer: the physical world \emph{is} a mathematical structure.
But this is a metaphysical claim that can't be tested, and it doesn't
explain \emph{which} mathematical structure or \emph{why} this
particular one.

The OI framework provides a different answer, with a mechanism.
Mathematics doesn't describe physics. Physics is what mathematics
\emph{looks like from inside}.

The mathematical structure is (S, \ensuremath{\varphi}) --- a finite set and a bijection.
An observer embedded in that structure, seeing only part of it through
the partition V, produces a description of what it can see. That
description is quantum mechanics. Not because QM was designed to
describe nature. Not because nature is mysteriously mathematical. But
because the structural constraints of embedded observation --- finite
capacity, lossless dynamics, partial access --- uniquely determine the
form of what the observer produces.

The trace-out is the mechanism. It takes the complete mathematical
structure --- the full evolution matrix with all its eigenvalues, Jordan
blocks, and nilpotent monodromy --- and projects it onto what the
observer can see: the semisimple spectral data, organized by the
representation theory of the partition. Physics is this projection.

What gets erased? The nilpotent part --- the Jordan blocks that exist
over finite fields but vanish over the reals, the monodromy operator
that the observer can't detect, the arithmetic data about eigenvalue
orders in finite field extensions. This is genuine mathematical
structure. It's \emph{there} in the dynamics. But the observer, looking
through the keyhole of the partition, sees only the diagonalizable
shadow.

So Wigner's puzzle dissolves. Mathematics isn't ``unreasonably
effective'' at describing physics. Mathematics is the most complete
description. Physics is the observable part of that description ---
drawn by someone who can't see the whole thing. The observable part is
effective because it's a faithful projection --- the semisimple part
genuinely captures the eigenvalue structure. But it's not the whole
description. The full mathematical description has structure (the
nilpotent monodromy) that no physical measurement can reveal.

But there's a stronger result. The reconstruction theorem proves the
converse: starting from the observed physics, mathematical consistency
forces you uniquely back to the equivalence class {[}(S, \ensuremath{\varphi}){]}. The two
descriptions determine each other.

The forward direction: start from (S, \ensuremath{\varphi}) and derive physics. That's the
program's main result. The converse: start from the observed physics and
reconstruct (S, \ensuremath{\varphi}). And the pieces are already in hand.

If you observe quantum mechanics with Bell violations on a finite
system, the characterization theorem forces you into some (S, \ensuremath{\varphi}, V) ---
there must be a deterministic system with a partition. If your physics
includes Einstein gravity via the thermodynamic route, the dynamics
selection theorem pins the wave equation --- it's the unique dynamics
consistent with QM emergence, isotropy, and linearity. If you see SU(3)
\ensuremath{\times} SU(2) \ensuremath{\times} U(1) with three generations, the gauge structure pins K = 6
with multiplicities (3, 2, 1) --- the cubic group decomposition is
unique. If \ensuremath{\bar{\theta}} = 0, the dynamics is T-invariant. If \ensuremath{\hbar} has the specific
value c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G), the gap equation pins \ensuremath{\varepsilon} = 2l\_p.~And d = 3 is the
unique self-consistent dimension.

Chain all of these together and you get: the observed physics determines
a unique equivalence class {[}(S, \ensuremath{\varphi}){]} --- a finite set with a
bijection, pinned up to the framework's gauge freedoms (the alphabet
size q, the choice of specific isotropic bounded-degree bijection, and
the specific partition V within the generic class).

This is exactly how physics always works. In general relativity,
measurements determine an equivalence class of spacetime metrics modulo
diffeomorphisms --- not a unique coordinate system. In gauge theory,
experiments determine an equivalence class of connections modulo gauge
transformations --- not a unique field configuration. Here, the observed
physics determines an equivalence class of finite bijections modulo the
OI gauge freedoms --- not a unique (S, \ensuremath{\varphi}).

So the question ``is mathematics reality, or does it just describe
reality?'' is not unanswered --- it's provably unanswerable by any
measurement. The observed physics and the mathematical structure
determine each other uniquely up to gauge. Any experiment that could
distinguish ``the universe is (S, \ensuremath{\varphi})'' from ``the universe is described
by (S, \ensuremath{\varphi})'' would need to detect something outside the equivalence
class. But the equivalence class already accounts for everything
observable.

This is the same structure as gauge invariance in electromagnetism. The
potential A is underdetermined --- only the field strength F is
measurable. We don't agonize over whether A is ``real.'' We identify the
exact boundary between the physical (F) and the gauge (A). The OI
framework identifies the same kind of boundary for the math-physics
relationship: the physics is the equivalence class {[}(S, \ensuremath{\varphi}){]}; the
ontological status of (S, \ensuremath{\varphi}) is gauge. The boundary is a theorem.

This suggests a precise way to understand what physics is. The
mathematical structure (S, \ensuremath{\varphi}) has two parts: the part accessible to an
embedded observer (the equivalence class, the gauge-invariant content,
the semisimple spectral data), and the part that isn't (the specific
representative, the nilpotent monodromy, the choice of q). The first
part is what we call physics. A better name for it might be
\emph{observable mathematics} --- the portion of mathematical structure
that an embedded subsystem with partial read access can detect.

Under this definition, Wigner's puzzle dissolves to the point of
tautology. ``Why is mathematics effective at describing physics?''
becomes ``Why is mathematics effective at describing observable
mathematics?'' --- which isn't a mystery. It's a definition. Physics is
the observable part of mathematics. Of course mathematics describes it.
The unreasonable effectiveness was never unreasonable; it was a category
error, treating mathematics and physics as separate domains when one is
a subset of the other.

And the incompleteness family extends one step further: Gödel showed
that proof can't capture all truth. Turing showed that computation can't
predict all behavior. The OI framework shows that observation can't
capture all dynamics. In each case, what \emph{can't} be captured has
rigid mathematical structure, and what \emph{is} captured is determined
by the incompleteness itself. Physics --- all of it, from quantum
mechanics to the Standard Model --- is the rigid structure of what
embedded observation captures. Mathematics is larger. Physics is the
view from inside.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\emph{This is a companion overview to ``The Incompleteness of
Observation'' (Maybaum, March 2026), which presents the formal arguments
with detailed derivations; ``Dynamics Selection and Emergent General
Relativity in the Observational Incompleteness Framework'' (Maybaum,
March 2026), which presents the rigidity test; ``The Substrate Problem:
Structural Foundations of the Observational Incompleteness Framework''
(Maybaum, 2026), which addresses the ontological status of the lattice;
and ``Why These Particles: Standard Model Structure from Observational
Incompleteness'' (Maybaum, 2026), which derives the Standard Model's
gauge group, matter content, and chiral structure. The philosophical
lineage section draws on a systematic analysis mapping the paper's
claims against the major traditions in Western, Eastern, and
contemporary philosophy of science.}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-mathematical-summary-tables}{%
\subsection{Appendix: Mathematical Summary
Tables}\label{appendix-mathematical-summary-tables}}

\hypertarget{sections-12-what-each-piece-does}{%
\subsubsection{Sections 1--2: What Each Piece
Does}\label{sections-12-what-each-piece-does}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Component
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What it establishes
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What it uses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Axioms 1--4 & The physical setup (no QM assumed) & Nothing --- these are
starting points \\
Partition-relativity (§1.4) & Emergent description depends only on
partition & Axioms 1, 3, 4 \\
Emergent stochasticity (§2.1) & Determinism looks random to embedded
observer & Axioms 1, 3, 4 \\
P-indivisibility proof (§2.3) & The stochastic process is non-Markovian
& Axioms 1, 2 + C1 \\
Accessible-timescale lemma (§2.3) & Non-Markovianity is observable, not
just formal & C1 + C2 + C3 \\
Coin-and-die model (§2.4) & Concrete demonstration of all mechanisms &
All axioms + all conditions \\
Stochastic-quantum correspondence (§3.1) & P-indivisibility = quantum
mechanics & Barandes {[}10,11{]} or Stinespring (Appendix A) \\
Necessity proof (§3.3) & QM \emph{requires} C1, C2, C3 & Contrapositives
of sufficiency \\
\end{longtable}

\hypertarget{sections-36-the-logical-flow}{%
\subsubsection{Sections 3--6: The Logical
Flow}\label{sections-36-the-logical-flow}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Step
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What it establishes
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Key equation
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
§3.1: Stochastic-quantum correspondence & P-indivisibility = QM & T\_ij
= \textbar U\_ij\textbar\ensuremath{^2} \\
§3.1: Phase-locking & Quantum Hamiltonian is unique & Fourier analysis
of T\_ij(t) \\
§3.2: Bell analysis & Framework evades Bell's theorem & Outcome
independence violated, parameter independence preserved \\
§3.3: Necessity of C1 & QM requires coupling & No coupling \ensuremath{\to} permutation
\ensuremath{\to} P-divisible \\
§3.3: Necessity of C2 & QM requires slow bath & Fast bath \ensuremath{\to} Markov \ensuremath{\to}
P-divisible \\
§3.3: Necessity of C3 & QM requires capacity & I \ensuremath{\leq} log\ensuremath{_2} m bounds
memory \\
§3.3: Characterization theorem & QM \ensuremath{\iff} C1+C2+C3 & Biconditional \\
§4: Cosmological horizon & The universe satisfies the axioms & C1 \checkmark{}, C2
\checkmark{}, C3 \checkmark{} \\
§5.1: Classical temperature & T\_cl = c\ensuremath{^2}\ensuremath{\varepsilon}\ensuremath{^2}\ensuremath{\kappa}/(8\ensuremath{\pi}Gk\_B) & No \ensuremath{\hbar} \\
§5.2 Step 1: Uniqueness & \ensuremath{\hbar} is determined, not free &
Partition-relativity \\
§5.2 Step 2: Boundary-only & \ensuremath{\hbar} depends only on c, G, \ensuremath{\varepsilon} & Deep sector
frozen \\
§5.2 Step 3: Dimensional analysis & \ensuremath{\hbar} = \ensuremath{\beta} c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/G & Unique combination \\
§5.2 Step 4: Thermal matching & T\_cl = T\_Q fixes \ensuremath{\beta} = 1/4 & \ensuremath{\hbar} =
c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) \\
§5.3: D-gauge theorem & No phase ambiguity & Double-difference
condition \\
§6: Self-consistency & \ensuremath{\varepsilon} = 2 l\_p, S\_BH = A/(4l\_p\ensuremath{^2}) & \ensuremath{\varepsilon}\ensuremath{^2} = 4\ensuremath{\hbar}G/c\ensuremath{^3} \\
\end{longtable}

The chain from axioms to \ensuremath{\hbar}:

Axioms \ensuremath{\to} P-indivisibility \ensuremath{\to} QM \ensuremath{\to} Hamiltonian with unknown \ensuremath{\hbar} \ensuremath{\to} thermal
matching at the horizon \ensuremath{\to} \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) \ensuremath{\to} \ensuremath{\varepsilon} = 2 l\_p \ensuremath{\to} S\_BH = A/(4
l\_p\ensuremath{^2})

Every link is either a proof or a calculation. No link requires quantum
mechanics as input --- QM appears as output at step 3 and its parameters
are determined by steps 4--6.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-key-mathematical-subtleties}{%
\subsection{Appendix: Key Mathematical
Subtleties}\label{appendix-key-mathematical-subtleties}}

\textbf{1. The gap equation is a gap equation, not a derivation from
nothing.}

The framework has one free parameter (\ensuremath{\varepsilon}). The gap equation relates \ensuremath{\varepsilon} to
\ensuremath{\hbar}. You need to know one to determine the other. The predictive content
is: - The \emph{relationship} \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) --- which could have been
any function of c, G, \ensuremath{\varepsilon} - The \emph{consequences} --- Bekenstein-Hawking
entropy, CC dissolution, dark-sector fraction --- which follow from this
specific relationship

\textbf{2. The KMS condition on the lattice.}

The thermal matching in Step 4 uses the KMS condition, which is proved
for continuum QFT on curved spacetime. The emergent QFT from Part I is
lattice-regularized with cutoff \ensuremath{\varepsilon}. Thermal periodicity on a lattice has
corrections of order (\ensuremath{\varepsilon}\ensuremath{\kappa}/c)\ensuremath{^2}. For the cosmological horizon: \ensuremath{\varepsilon}\ensuremath{\kappa}/c = \ensuremath{\varepsilon}H/c
\textasciitilde{} 10\ensuremath{^{-}}\ensuremath{^6}\ensuremath{^1}, so corrections are \textasciitilde{} 10\ensuremath{^{-}}\ensuremath{^1}\ensuremath{^2}\ensuremath{^2}.
Negligible, but a referee could ask for a lattice-native proof.

\textbf{3. The D-gauge theorem requires genericity.}

The phase-locking and D-gauge results assume non-degenerate spectrum,
non-degenerate energy gaps, and non-vanishing overlaps (conditions
G1--G3). These fail on a measure-zero set of Hamiltonians. For ``almost
all'' Hamiltonians, the results hold. A referee might ask what happens
on the measure-zero set --- the answer is that the Hamiltonian is still
determined up to a finite discrete ambiguity, which doesn't affect the
physics.

\textbf{4. Why doesn't the deep sector matter?}

The boundary-only dependence lemma is crucial because without it, \ensuremath{\hbar}
might depend on the vast interior of the hidden sector --- which we
can't observe or constrain. The lemma shows the deep sector is frozen on
observable timescales, so its details are irrelevant. \ensuremath{\hbar} is a property of
the boundary geometry, not of the bulk.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Acknowledgment of AI-Assisted Technologies:} The author
acknowledges the use of \textbf{Claude Opus 4.6} and \textbf{Gemini 3.1
Pro} to assist in synthesizing technical concepts and refining clarity.
The final text and all scientific claims were reviewed and verified by
the author.

\end{document}
