% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\ifPDFTeX\else
  % xetex/luatex font selection
\setmainfont{DejaVu Serif}
\setsansfont{DejaVu Sans}
\setmonofont{DejaVu Sans Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  % \usepackage[]{microtype}
  % \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\hypertarget{the-fundamental-structure-of-the-observational-incompleteness-framework}{%
\section{The Fundamental Structure of the Observational Incompleteness
Framework}\label{the-fundamental-structure-of-the-observational-incompleteness-framework}}

\hypertarget{from-finite-bijection-to-the-standard-model}{%
\subsubsection{From Finite Bijection to the Standard
Model}\label{from-finite-bijection-to-the-standard-model}}

\hypertarget{alex-maybaum}{%
\subsubsection{Alex Maybaum}\label{alex-maybaum}}

\hypertarget{section}{%
\subsubsection{2026}\label{section}}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The Observational Incompleteness (OI) framework derives quantum
mechanics and general relativity from a deterministic bijection on a
finite lattice with a visible/hidden partition {[}1{]}. This paper
addresses two questions the derivations leave open: what is the lattice,
and which quantum field theory does the embedded observer see?

For the first question, we show that the framework's results depend on
six structural properties --- deterministic bijectivity, finite boundary
entropy, bounded coupling degree, statistical isotropy, non-trivial
partition coupling, and slow-bath capacity --- and on nothing else. The
regular lattice, the alphabet size q, and the wave equation are all
either derived from these properties or irrelevant to the predictions.
The lattice is the coupling graph of the bijection: the adjacency
structure defined by which degrees of freedom affect which others in one
dynamical step. The product decomposition of the state space into
``sites'' --- and hence space itself --- is determined by the bijection
as the unique factorization minimizing coupling degree. The alphabet
size q is a gauge freedom. The spatial dimensionality d = 3 is selected
by four independent self-consistency filters. The fundamental
ontological commitment is minimal: a finite set, a bijection, and a
partition. Everything else --- space, time, dimensionality, quantum
mechanics, general relativity --- is emergent.

For the second question, we show that the wave equation on a d = 3
lattice uniquely determines the Standard Model's structure. Bottom-up:
the wave equation factors into staggered Dirac fermions; center
independence enforces chiral symmetry, mandating the Higgs; staggered
tastes give three fermion generations. Gauge emergence: coupling-degree
minimization gives K = 2d = 6 internal components; the cubic rotation
group fixes multiplicities (3, 2, 1); background independence promotes
the commutant to local gauge invariance. Top-down: anomaly cancellation
uniquely determines the hypercharges; the trace-out makes SU(2) chiral
while SU(3) remains vector-like; T-invariance forces θ̄ = 0 at all energy
scales.

The reconstruction theorem establishes the converse: observed QM with
Bell violations, finite boundary entropy, and spatial isotropy uniquely
determine the equivalence class {[}(S,
φ){]}/\(\mathcal{G}_{\text{sub}}\) at the lattice level --- a finite set
with a bijection of bounded coupling degree and statistical isotropy.
Whether (S, φ) \emph{is} reality or \emph{describes} reality is provably
undecidable by any measurement --- the framework identifies this as
gauge.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{introduction}{%
\subsection{1. Introduction}\label{introduction}}

The companion paper {[}1{]} proves that an observer embedded in a
deterministic system with a coupled, slow, high-capacity hidden sector
necessarily describes the visible sector using quantum mechanics --- and
that these conditions are necessary and sufficient. It also shows that
the cosmological horizon realizes these conditions, determines ℏ =
c³ε²/(4G), recovers the Bekenstein-Hawking entropy, and dissolves the
cosmological constant discrepancy.

Both papers use explicit constructions: a cubic lattice in d dimensions,
an alphabet ℤ/qℤ at each site, nearest-neighbor coupling, a specific
update rule. These constructions work --- every link in the derivation
chain is analytically proven. But they raise two questions. First, what
is the ontological status of the lattice? Three readings are possible:
literal (the lattice is physical reality), effective (the lattice
approximates unknown structure), and structural (the physics lives in
the equivalence class of systems sharing the lattice's structural
properties, not in any particular representative). This paper argues for
the structural reading and develops its consequences.

Second, does the lattice structure determine which quantum field theory
the observer sees? The QM derivation tells the embedded observer that
they must use quantum probability; it does not tell them which particles
exist or which forces act. We show that the wave equation on a d = 3
lattice, combined with mathematical consistency, uniquely determines the
Standard Model's gauge group, matter content, and discrete symmetries
--- with the primary derivation chain proved end-to-end.

The paper is organized as follows. §§2--4 establish the structural
foundations: the minimal structure required by the theorems, the
coupling graph interpretation of space, and the gauge freedom of the
alphabet size. §§5--6 derive two geometric constraints: background
independence and the unique dimension d = 3. §7 derives the emergent QFT
from the lattice dynamics. §8 addresses the strong CP problem. §9
discusses the ontological and interpretive consequences, including
relations to other discrete approaches. §10 synthesizes the complete
program and proves the reconstruction theorem. §11 concludes.

\hypertarget{epistemic-status-of-the-results}{%
\subsubsection{1.1 Epistemic Status of the
Results}\label{epistemic-status-of-the-results}}

The results fall into three tiers.

\textbf{Tier 1 (structural foundations, from the six properties alone).}
The lattice is the coupling graph (§3), the alphabet size is gauge (§4),
the observer is generic (§9.4), and d = 3 is the unique self-consistent
dimension (§6). These are theorems about the equivalence class {[}(S,
φ){]}/\(\mathcal{G}_{\text{sub}}\) and do not depend on which specific
dynamics is selected.

\textbf{Tier 2 (dynamics selection and the primary SM derivation
chain).} Center independence + isotropy + linearity uniquely select the
wave equation (§7.1). The wave equation determines: staggered Dirac
fermions, K = 2d = 6 from coupling-degree minimization, multiplicities
(3, 2, 1) from the cubic group, local gauge invariance from background
independence, chirality from the trace-out, unique hypercharges from
anomaly cancellation, three generations from cubic symmetry, one Higgs
doublet, and θ̄ = 0 at all energy scales. Each step is proved; the chain
is 29 theorems/lemmas long, and each depends on its predecessors. The
individual proofs are short, but the cumulative chain is the paper's
most exposed structure: a single failure at any link would propagate
forward.

\textbf{Tier 3 (redundant confirmations).} Asymptotic freedom requiring
N\_c ≥ 3 (Proposition), minimal chiral group (Proposition), and the
Jordan-Chevalley projection (Appendix A). These provide independent
second-route checks on Tier 2 results but are not load-bearing in the
primary derivation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-minimal-structure}{%
\subsection{2. The Minimal Structure}\label{the-minimal-structure}}

\hypertarget{what-the-theorems-require}{%
\subsubsection{2.1 What the theorems
require}\label{what-the-theorems-require}}

Each assumption in the companion paper {[}1{]} does specific work. The
question is which assumptions are essential to the results and which are
artifacts of the particular construction chosen.

\textbf{Deterministic bijectivity.} The P-indivisibility proof {[}1,
§2.3{]} requires a bijection on a finite set: bijectivity guarantees
recurrence (φ\^{}N = id), and recurrence produces the non-monotonic
distinguishability that defines P-indivisibility. The entire framework
rests on QM emerging from marginalizing deterministic dynamics --- if
the dynamics were already stochastic, there would be nothing to derive.
Determinism is the core of the framework and cannot be weakened.
However, determinism does not require a lattice. A continuous
Hamiltonian system on a compact phase space is also deterministic and
bijective (by Liouville's theorem). The lattice is one way to implement
determinism on a finite set; it is not the only way.

\textbf{Finite boundary entropy.} The gap equation {[}1, §5{]} requires
a finite number of degrees of freedom across the partition boundary: S =
A/ε². This follows from the holographic entropy bound {[}3{]} applied to
any finite-area surface, not from lattice structure. What matters is
that the boundary carries finitely many modes --- not that the bulk is a
lattice.

\textbf{Bounded coupling degree (locality).} This assumption does the
most work. The area law (§5) follows from the spatial Markov property,
which requires range-1 coupling. Without bounded coupling, long-range
correlations could produce volume-law entropy, breaking the Jacobson
route to GR. Locality does not require a regular lattice. Any
bounded-degree graph has the spatial Markov property for range-1
dynamics. The theorems would work on a random graph with bounded degree,
as long as it has the right statistical properties.

\textbf{Statistical homogeneity and isotropy.} The Myrheim-Meyer
dimension {[}4{]} requires a well-defined notion of spacetime dimension,
which emerges from the causal order statistics on a homogeneous,
isotropic structure. The dispersion relation (§5) uses translation
invariance. The dynamics selection (§7.1) uses isotropy. Exact lattice
regularity is sufficient but not necessary. A random graph with
statistical homogeneity and isotropy at large scales would produce the
same results.

\textbf{Non-trivial coupling (C1) and slow bath with capacity (C2, C3).}
These are conditions on the partition, not on the lattice. They hold for
any system with the right partition geometry, regardless of whether the
underlying space is a lattice, a graph, or something else.

\hypertarget{what-the-theorems-do-not-require}{%
\subsubsection{2.2 What the theorems do not
require}\label{what-the-theorems-do-not-require}}

A \emph{regular lattice}. Any bounded-degree graph with statistical
isotropy suffices. A \emph{specific alphabet size q}. All proofs work
for any q ≥ 2; no prediction depends on q. A \emph{specific
dimensionality d}. The proofs work for any d ≥ 1; self-consistency
selects d = 3 (§6). The \emph{wave equation}. It is derived from center
independence, isotropy, and linearity (§7.1), each of which follows from
the structural properties listed above.

\hypertarget{the-minimal-object}{%
\subsubsection{2.3 The minimal object}\label{the-minimal-object}}

The theorems require exactly: a deterministic bijection on a finite set,
whose state space factors into local degrees of freedom coupled by a
bounded-degree graph with statistical isotropy, partitioned into visible
and hidden sectors satisfying C1--C3. Everything else --- the regular
lattice, the alphabet, the dimensionality, the wave equation, quantum
mechanics, general relativity --- is either derived or irrelevant. The
minimal object is the triple (S, φ, V): a finite set S, a bijection φ: S
→ S with bounded-degree coupling, and a partition V ⊂ S of the degrees
of freedom into visible and hidden sectors.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-coupling-graph}{%
\subsection{3. The Coupling Graph}\label{the-coupling-graph}}

\hypertarget{definition-and-the-factorization-problem}{%
\subsubsection{3.1 Definition and the factorization
problem}\label{definition-and-the-factorization-problem}}

Let S = S₁ × S₂ × \ldots{} × S\_N be a product of finite sets (the local
state spaces), and let φ: S → S be a bijection. The \emph{coupling
graph} G\_φ is the undirected graph on vertices \{1, \ldots, N\} where i
and j are connected if the i-th component of φ(s) depends on the j-th
component of s for at least one state s ∈ S. The coupling graph is not
an additional structure imposed on the system --- it is read off from φ.

What determines the product decomposition? An abstract finite set S does
not come with a canonical factorization. The answer is that the
factorization is determined by φ itself, as the one that \emph{minimizes
the coupling degree}. For the wave equation, numerical verification
confirms this: the natural spatial factorization is the \emph{unique
minimizer} of coupling degree --- 0\% of random factorizations achieve
it, and 100\% have strictly higher coupling.

\textbf{Theorem (factorization uniqueness for translation-invariant
dynamics).} \emph{Let \(\varphi\) be a bijection on
\(S = (\mathbb{Z}/q\mathbb{Z})^N\) satisfying: (a) translation
invariance, (b) range-1 coupling on the natural factorization, and (c)
genuinely bidirectional coupling. Then the natural factorization is the
unique minimizer of coupling degree among all product decompositions
\(S = S_1 \times \cdots \times S_M\) with \(|S_k| = q\) for all \(k\).}
The key step is that the wave equation's addition
\(x_{i-1} + x_{i+1} - x_i^{\text{prev}}\) creates irreducible algebraic
dependence among all input components (for \(q > 2\)), so any relabeling
that splits or merges natural sites increases the coupling degree.

\textbf{Extension to general graphs.} The translation-invariant proof
extends to any bounded-degree graph with polynomial growth exponent
\(d\) and statistical isotropy. The argument proceeds in three steps.
First, §6 establishes (via Gromov's theorem) that any such graph is
quasi-isometric to \(\mathbb{Z}^d\), so the asymptotic dimension is
\(d\). Second, the wave equation on a \(d\)-dimensional graph has \(2d\)
independent low-energy propagation modes (from heat kernel asymptotics:
the spectral dimension equals \(d\), giving \(d\) momentum-like
parameters over two orientations each). Third, the integer-valued
coupling degree \(\delta(K)\) depends continuously on the graph
structure but discretely on \(K\); since the graph is connected to
\(\mathbb{Z}^d\) by a continuous path of statistically isotropic
bounded-degree graphs (all quasi-isometric), the minimizer \(K = 2d\) is
topologically stable. The factorization uniqueness is therefore a
property of the equivalence class
\([(S, \varphi)]/\mathcal{G}_{\text{sub}}\), not of the specific graph.

\hypertarget{space-as-coupling-structure}{%
\subsubsection{3.2 Space as coupling
structure}\label{space-as-coupling-structure}}

Every ``spatial'' property used in the derivation chain is a property of
G\_φ. The \emph{area} of a region V is the number of edges crossing from
V to its complement. The \emph{dimension} is extracted from the causal
partial order via the Myrheim-Meyer estimator {[}4{]}. The \emph{area
law} follows from the spatial Markov property on any graph with range-1
dynamics. The \emph{dispersion relation} requires statistical isotropy
but not exact regularity. The regular cubic lattice is one graph with
these properties; it is not the only one.

\hypertarget{the-dissolution-of-the-container-problem}{%
\subsubsection{3.3 The dissolution of the container
problem}\label{the-dissolution-of-the-container-problem}}

The coupling graph is not embedded in a pre-existing space. There is no
ambient manifold in which the graph ``lives.'' The vertices are labels
for degrees of freedom; the edges record which degrees of freedom are
dynamically coupled. Any spatial embedding is a representation for human
convenience, not a physical fact. The coupling graph is not made of
material. The vertices are not atoms of space, and the edges are not
physical connections. A degree of freedom is a component of the state
--- it is an abstract mathematical entity, not an object. Asking ``what
is site i made of?'' is a category error: it is a label for a variable,
not a name for a thing.

The most common objection to discrete models of spacetime is the
container problem: if space is a lattice, what does the lattice sit in?
On the structural reading, the container problem does not arise. The
coupling graph is defined by the bijection φ, not by embedding in a
manifold. Two sites are neighbors because φ couples them --- because the
state at one affects the future of the other. This is a dynamical fact,
not a spatial one. Locality is defined by the dynamics, and space
emerges from locality. At no point does anything need to be ``inside''
anything else. The relationship is analogous to a social network: two
people are ``connected'' not because they occupy adjacent points in
physical space, but because they interact. The topology is defined by
the interactions, not by geography.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-alphabet-as-gauge-freedom}{%
\subsection{4. The Alphabet as Gauge
Freedom}\label{the-alphabet-as-gauge-freedom}}

Every prediction of the OI framework is independent of the alphabet size
q. The gap equation ℏ = c³ε²/(4G) contains no q. The Bekenstein-Hawking
formula contains no q. The dispersion relation, area law,
P-indivisibility, center independence, and Einstein's equations are all
q-independent. Numerical confirmation: on a 1D ring of L = 40 sites, the
wave equation produces significant P-indivisibility at every q tested
from 2 to 64. No experiment, even in principle, could measure q.

Two systems (S, φ) and (S', φ') with different alphabet sizes q and q'
are \emph{physically equivalent} if they produce the same emergent
transition probabilities. The analogy to electromagnetic gauge
invariance is precise: the gauge potential A\_μ is not physical; the
field strength F\_μν is. In the OI framework, the mod-q state space is
not physical; the coupling structure and its emergent consequences are.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{background-independence}{%
\subsection{5. Background Independence}\label{background-independence}}

The companion paper {[}1{]} treats the coupling graph as fixed. In
general relativity, the spacetime geometry is dynamical. If space is the
coupling graph, background independence requires the graph to evolve
with the state: s(t+1) = φ\_\{s(t)\}(s(t)), where each φ\_s is a
bijection but G\_\{φ\_s\} varies with s.

\textbf{Bijectivity is automatic.} The wave equation is second-order:
x\_i(t+1) = f(neighbors of i at time t) − x\_i(t−1) mod q. The
phase-space map F: (x(t−1), x(t)) → (x(t), x(t+1)) has an explicit
inverse that uses x(t) to determine which graph to apply. Bijectivity of
the phase-space map is automatic for any second-order reversible
dynamics, regardless of whether the coupling is state-dependent. The
P-indivisibility proof applies without modification.

\textbf{The derivation chain survives under three constraints.} (i)
\emph{Local graph-dependence:} G(x) at site i depends only on x\_j
within a bounded range of i. (ii) \emph{Center-independent
graph-dependence:} G(x) at site i does not depend on x\_i, preserving
center independence. (iii) \emph{Statistical isotropy:} G(x) is
statistically isotropic on large scales for typical configurations.

Under these constraints: the area law holds, center independence is
preserved, the gap equation is unchanged, and the dynamics selection
theorem applies at each step.

\textbf{Explicit construction.} On a ring of L ≥ 4 sites with alphabet
ℤ/qℤ, define the right-neighbor assignment \(R(i, x) = i+2 \bmod L\) if
\(x_{(i+1) \bmod L} = 0\), and \(R(i, x) = i+1 \bmod L\) otherwise, with
left neighbor \(L(i) = i - 1 \bmod L\) fixed. The dynamics is
\(x_i(t+1) = (x_{L(i)}(t) + x_{R(i,x(t))}(t) - x_i(t-1)) \bmod q\). This
satisfies: constraint (i) with range 1 (\(R(i,x)\) depends only on
\(x_{i+1}\)); constraint (ii) exactly (\(R(i,x)\) does not depend on
\(x_i\)); constraint (iii) by translation invariance. Bijectivity is
verified by explicit inverse:
\(u_i = (v_{L(i)} + v_{R(i,v)} - w_i) \bmod q\), since \(v\) determines
all neighbor assignments. Center independence, P-indivisibility, and
graph regularity (undirected degree bounded in {[}2, 4{]} independently
of the state) all hold for any L and q.

\textbf{Local gauge invariance from background independence.} With
state-dependent coupling, the coupling matrix M (§7.4 below) becomes
site-dependent: \(M = M(\mathbf{n}, \hat{e}_j)\). At every site,
\(M(\mathbf{n}, \hat{e}_j)\) has K = 6 components with eigenvalue
multiplicities (3, 2, 1). Since M varies from site to site, the
commutant \(\mathrm{U}(3) \times \mathrm{U}(2) \times \mathrm{U}(1)\)
acts independently at each site. Under a site-dependent transformation
\(G(\mathbf{n})\):

\[\phi(\mathbf{n}) \to G(\mathbf{n})\,\phi(\mathbf{n}), \qquad M(\mathbf{n}, \hat{e}_j) \to G(\mathbf{n})\,M(\mathbf{n}, \hat{e}_j)\,G(\mathbf{n}+\hat{e}_j)^{-1}\]

The wave equation is invariant. This is local gauge invariance. The link
variable \(M(\mathbf{n}, \hat{e}_j)\) transforms as a gauge connection:
the plaquette product
\(P(\mathbf{n}, j, k) = M(\mathbf{n}, \hat{e}_j)\,M(\mathbf{n}+\hat{e}_j, \hat{e}_k)\,M(\mathbf{n}+\hat{e}_k, \hat{e}_j)^{-1}\,M(\mathbf{n}, \hat{e}_k)^{-1}\)
transforms in the adjoint
(\(P \to G(\mathbf{n})\,P\,G(\mathbf{n})^{-1}\)), so
\(\mathrm{Re\,Tr}(P)\) is gauge-invariant --- it is the Wilson plaquette
action {[}9{]}, now derived rather than postulated.

\textbf{Einstein's equations as self-consistency.} The Jacobson
thermodynamic argument {[}24{]}, applied to local causal horizons on the
state-dependent graph, requires three inputs --- each proved here for
the wave equation.

\textbf{Lemma} (Area law). \emph{The entanglement entropy of a spatial
region V in the wave equation scales as \(S(V) = \eta\,|\partial V|\).
This holds for both the linear wave equation over \(\mathbb{R}\) and the
mod-q wave equation over \(\mathbb{Z}/q\mathbb{Z}\).}

\emph{Proof.} Over \(\mathbb{R}\): the wave equation is a system of
coupled harmonic oscillators with a quadratic Hamiltonian determined by
the adjacency matrix. The area-law theorem for Gaussian systems applies
directly {[}2{]}. Over \(\mathbb{Z}/q\mathbb{Z}\): the wave equation has
range \(R = 1\), so for any \(i \in V°\) (interior of \(V\)), all
neighbors lie inside \(V\). With uniform initial conditions:
\(V° \perp V^c \mid \partial V\) (spatial Markov property). By the chain
rule:
\(I(V;\, V^c) = I(\partial V;\, V^c) \leq H(\partial V) \leq |\partial V| \cdot \log_2 q\).
The mutual information scales as boundary area, not volume. \(\square\)

\textbf{Lemma} (Relativistic dispersion). \emph{The wave equation has
exact dispersion relation \(\cos\omega = \cos k\).}

\emph{Proof.} Substituting \(x_j(t) = A\exp(i(kj - \omega t))\):
\(e^{-i\omega} + e^{i\omega} = e^{-ik} + e^{ik}\), giving
\(\cos\omega = \cos k\). For small \(k\): \(\omega = |k| + O(k^3)\) ---
relativistic propagation with \(v = 1\). \(\square\)

\textbf{Lemma} (Lattice Bisognano-Wichmann). \emph{For the wave
equation, whose low-energy effective theory is Lorentz-invariant, the
entanglement Hamiltonian for a half-space bipartition has the
Bisognano-Wichmann form. The reduced state for a Rindler observer is
thermal at \(T_U = \hbar\kappa/(2\pi c k_B)\), with corrections at
\(O(\varepsilon^2\kappa^2/c^2)\).}

\emph{Proof.} The mod-q wave equation is a coupled harmonic oscillator
system --- the class for which the lattice BW theorem is proved
analytically {[}23{]}. Its low-energy dispersion is relativistic (the
preceding lemma). Thermality of the Unruh effect is robust against UV
modifications of the dispersion relation, with corrections scaling as
\(O(\varepsilon^2\kappa^2/c^2)\). \(\square\)

\textbf{Theorem} (Discrete Einstein equation). \emph{For the
state-dependent wave equation on a bounded-degree graph G(x) satisfying
constraints (i)--(iii), the Jacobson thermodynamic argument produces:}

\[\kappa_{OR}(i, j;\, G(x)) = \alpha \cdot \mathcal{T}_{ij}(x,\, G(x))\]

\emph{where \(\kappa_{OR}\) is the Ollivier-Ricci curvature {[}21{]},
\(\mathcal{T}_{ij}\) is the discrete stress-energy, and
\(\alpha = 8\pi G/(c^4 \eta)\).}

\emph{Proof.} (i) \emph{Area law.} The entanglement entropy scales as
\(S(V) = \eta\,|\partial V|\) (area-law lemma above), holding for any
bounded-degree graph with range-1 dynamics. (ii) \emph{Unruh
temperature.} The lattice Bisognano-Wichmann lemma gives
\(T = \hbar\kappa/(2\pi c k_B)\) at local causal horizons. (iii)
\emph{Clausius relation.} At each edge \((i,j)\): the area variation
under graph deformation is captured by \(\kappa_{OR}\) (which measures,
via optimal transport, the deviation from flat space), giving
\(\delta A / A = -\kappa_{OR}\,\delta\lambda\). Substituting into
\(\delta Q = T\,\delta S\):
\(\mathcal{T}_{ij}\,\delta\lambda\,\delta A = T \cdot \eta \cdot (-\kappa_{OR}\,\delta\lambda\,A)\),
which rearranges to \(\kappa_{OR} = \alpha \cdot \mathcal{T}_{ij}\).
(iv) \emph{Continuum limit.} Van der Hoorn et al.~{[}15{]} prove
\(\kappa_{OR} \to \mathrm{Ric}\) on random geometric graphs converging
to a Riemannian manifold. Kelly et al.~{[}16{]} prove
\(\sum \kappa_{OR} \to \int R\,\sqrt{g}\,d^dx\). Together: the discrete
Einstein equation converges to
\(G_{\mu\nu} + \Lambda g_{\mu\nu} = (8\pi G/c^4)\,T_{\mu\nu}\).
\(\square\)

Every orbit is self-consistent by construction: the state determines the
graph, which determines the entropy and temperature, and the Clausius
relation constrains the graph's response to energy flux --- paralleling
continuum GR where the Einstein equations are the equations of motion,
not constraints on separate dynamics. The mathematical framework is
unchanged: the full dynamics F(u, v) = (v, WE\_\{G(v)\}(v) − u) is a
single bijection on the finite phase space. The ontological object is
still (S, F, V).

\textbf{Lattice corrections to Einstein's equations.} The derivation
invokes continuum results in the final step (Jacobson). Decomposing
Jacobson's argument into its four inputs and checking each
independently: (i) \(\delta S = \eta\,\delta A/\varepsilon^2\) (area-law
entropy) --- proved for Gaussian systems over \(\mathbb{R}\) {[}2{]} and
for any nearest-neighbor dynamics over \(\mathbb{Z}/q\mathbb{Z}\) via
the spatial Markov property (area-law lemma above). No Lorentz
invariance enters. \textbf{Exact on the lattice.} (ii)
\(T = \hbar\kappa/(2\pi c k_B)\) (Unruh temperature) --- requires the
Bisognano-Wichmann theorem. On the lattice, the BW form is proved
analytically for coupled harmonic oscillator systems {[}23{]} and shown
robust against UV modifications of the dispersion relation with
corrections at \(O(\varepsilon^2\kappa^2/c^2)\). \textbf{Approximate.}
(iii) \(\delta Q = \int T_{\mu\nu} k^\mu d\Sigma^\nu\) (energy flux) ---
a definition on the emergent manifold. No Lorentz invariance required.
\textbf{Exact.} (iv) The Raychaudhuri equation --- kinematic: it holds
on any pseudo-Riemannian manifold as a consequence of the definition of
curvature. \textbf{Exact.} Since inputs (i), (iii), and (iv) are
lattice-exact, corrections enter only through input (ii):

\[G_{\mu\nu} + \Lambda g_{\mu\nu} = \frac{8\pi G}{c^4} T_{\mu\nu} \times \left[1 + \mathcal{O}\!\left(\frac{\varepsilon^2 \kappa^2}{c^2}\right)\right]\]

For the cosmological horizon,
\(\varepsilon\kappa/c = 2l_P \cdot H/c \sim 10^{-61}\), giving
corrections of order \(10^{-122}\). For solar-mass black holes,
\(\varepsilon\kappa/c \sim 10^{-38}\), giving corrections of order
\(10^{-76}\). The corrections are negligible for any horizon much larger
than the Planck scale.

\textbf{The \(\eta = 1/4\) coefficient.} The Bekenstein-Hawking formula
\(S = A/(4l_P^2)\) follows from \(\varepsilon = 2l_P\) (the gap equation
{[}1, §5{]}): one entropy unit per cell of area
\(\varepsilon^2 = 4l_P^2\), giving \(S = A/\varepsilon^2 = A/(4l_P^2)\).
This cannot be confirmed via entanglement entropy because of the species
problem --- the entanglement entropy coefficient is species-dependent
(Srednicki found \(\eta_{\text{ent}} \approx 0.295\) per scalar field),
while the BH entropy is species-independent. The thermal matching route
avoids this by counting classical boundary degrees of freedom (one per
\(\varepsilon^2\) cell, independent of field content) rather than any
particular field's vacuum entanglement.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{why-d-3}{%
\subsection{6. Why d = 3}\label{why-d-3}}

The theorems work for any d ≥ 1. Four independent self-consistency
filters converge on d = 3.

\textbf{The concordance calculation.} In d spatial dimensions, the
boundary entropy ratio is:

\[\frac{\rho_s}{\rho_{\text{crit}}} = \frac{2}{d-1}\]

This equals unity only for d = 3. For d = 2: overclosure. For d ≥ 4:
gravitational deficit with no source.

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
d & ρ\_s / ρ\_crit & Status \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2 & 2 & Overclosure \\
\textbf{3} & \textbf{1} & \textbf{Exact concordance} \\
4 & 2/3 & Deficit \\
\end{longtable}

\textbf{Propagating gravity requires d ≥ 3.} The Weyl tensor vanishes
for d ≤ 2. GW echoes and the Jacobson derivation require propagating
gravitational degrees of freedom.

\textbf{Stable matter requires d ≤ 3.} The Coulomb potential in d
dimensions gives unstable atoms for d ≥ 4 (Ehrenfest {[}13{]}).
Gravitational orbits are unstable for d ≥ 4 {[}14{]}. Without stable
matter, no observers exist to define the partition.

\textbf{Renormalizability requires d = 3.} The Yang-Mills coupling has
mass dimension {[}g{]} = (3−d)/2; dimensionless only for d = 3.

\textbf{The intersection} of d ≥ 3, d ≤ 3, the concordance, and
renormalizability is d = 3 uniquely.

\textbf{Integer dimension is derived, not presupposed.} The
self-consistency argument might appear to presuppose that the coupling
graph has a well-defined integer dimension. In fact, every
bounded-degree graph falls into one of three growth classes: exponential
(\(|B(r)| \sim k^r\)), fractal (\(|B(r)| \sim r^\alpha\) with
non-integer \(\alpha\)), or integer-polynomial (\(|B(r)| \sim r^d\) with
integer \(d\)). Self-consistency excludes the first two.

\emph{Exponential growth} fails because the wave equation on such graphs
does not produce Lorentz-invariant dispersion (the spectral gap doesn't
close correctly), so Jacobson's derivation of Einstein's equations
fails. Additionally, \(d_{\text{eff}} \to \infty\) is incompatible with
stable matter (\(d \leq 3\)) and renormalizability (\(d = 3\)).

\emph{Fractal growth} fails because Jacobson's argument requires the
Raychaudhuri equation, which requires geodesic congruences on a manifold
--- fractal spaces don't support them. The staggered fermion
decomposition requires a hypercubic BZ structure
(\(\eta \in \{0,1\}^d\)), which doesn't exist on fractal graphs. The
cubic group decomposition of \(2d\) link directions --- which constrains
the gauge structure (§7.6) --- requires cubic lattice symmetry, absent
in fractals.

Only \emph{integer-polynomial growth} survives. By Gromov's theorem
(finitely generated groups of polynomial growth are virtually nilpotent)
and statistical isotropy, the graph is quasi-isometric to
\(\mathbb{Z}^d\) for integer \(d\). The four filters then select
\(d = 3\).

\textbf{Corollary} (\(d = 3\) is necessary, not contingent). \emph{Any
bijection whose coupling graph has \(d \neq 3\) produces internal
contradictions in its emergent description. The coupling graph dimension
is not a free parameter or a property of a specific \(\varphi\) --- it
is the unique value compatible with self-consistency.}

\textbf{Relation to the anthropic principle.} This is not a standard
anthropic argument. The standard argument invokes a multiverse of
dimensions and selection bias: many \(d\) exist; we observe \(d = 3\)
because only there can observers exist. The present argument is
different: the framework's axioms require an observer as a constitutive
element (Axiom 3 defines the partition relative to an observer). The
requirement that such an observer exist, combined with the
framework-internal concordance calculation, constrains \(d\). The
observer is not selecting from a pre-existing landscape but is a
structural prerequisite for the axioms to have content.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-emergent-quantum-field-theory}{%
\subsection{7. The Emergent Quantum Field
Theory}\label{the-emergent-quantum-field-theory}}

The wave equation on a d = 3 lattice determines not just QM and GR but
the specific QFT the observer sees. This section derives the Standard
Model's structure through three complementary routes: bottom-up from the
lattice dynamics to fermionic matter (§§7.1--7.3), gauge emergence from
multi-component dynamics (§§7.4--7.6), and the staggered species /
consistency constraints that fix the matter content and discrete
symmetries (§§7.7--7.12).

\hypertarget{the-wave-equation-as-lattice-klein-gordon}{%
\subsubsection{7.1 The wave equation as lattice
Klein-Gordon}\label{the-wave-equation-as-lattice-klein-gordon}}

The OI lattice is a d-dimensional hypercubic lattice Λ = ℤ\^{}d with
spacing ε = 2 l\_p.~The dynamics on this lattice is not a free choice
--- it is uniquely determined by three requirements.

\textbf{Theorem} (Dynamics selection). \emph{Among all second-order
reversible nearest-neighbor dynamics on a d-dimensional lattice of
alphabet size q, the requirements of (i) center independence, (ii)
spatial isotropy, and (iii) linearity uniquely select the discrete wave
equation. This holds for any q ≥ 2 and d ≥ 1.}

\emph{Proof.} The update function has the form
\(x_i(t+1) = (f(\text{neighbors of } i \text{ at time } t) - x_i(t-1)) \bmod q\).
Center independence requires that \(f\) not depend on \(x_i\) itself:
\(f\) reduces to a function of the \(2d\) neighboring values only.
Spatial isotropy requires \(f\) to be symmetric under all permutations
of the neighbors corresponding to lattice symmetries. Linearity over
\(\mathbb{Z}/q\mathbb{Z}\) requires \(f(a + a') = f(a) + f(a')\). The
unique function satisfying all three is
\(f = \alpha(x_1 + x_2 + \cdots + x_{2d}) \bmod q\) for some constant
\(\alpha\). The propagation speed is \(v = \alpha\), so \(\alpha = 1\)
gives the maximum lattice-scale speed. This is the discrete wave
equation. \(\square\)

Each requirement has a physical justification.

\emph{Center independence is necessary for emergent QM.} The mechanism
is information screening. For the checkerboard partition, the adjacency
matrix \(A\) is bipartite: \(A_{VV} = 0\) (no visible--visible edges).
In the center-independent (CI) case, the update
\(x_i(t+1) = [h_{i-1}(t) + h_{i+1}(t) - x_i(t-1)] \bmod q\) does not
contain \(x_i(t)\). Therefore \(X_V(t+1)\) carries no information about
\(X_V(t)\); conditioning on \(X_V(t+1)\) constrains \(H(t)\) but does
not screen \(X_V(t)\) from the future. Since \(H(t+1)\) depends on
\(X_V(t)\), the path \(X_V(t) \to H(t+1) \to X_V(t+2)\) bypasses the
present: the process is non-Markovian. In the center-dependent (CD)
case, the update
\(x_i(t+1) = [x_i(t) + h_{i-1}(t) + h_{i+1}(t) - x_i(t-1)] \bmod q\)
contains \(x_i(t)\) explicitly. Conditioning on \((X_V(t), X_V(t+1))\)
gives \(|V|\) linear equations in \(|H|\) unknowns over
\(\mathbb{Z}/q\mathbb{Z}\); generically (for \(q\) prime) the hidden
state \(H(t)\) is fully determined, so the augmented process is Markov.
Center coupling makes the present an explicit function of the visible
past, allowing it to determine the hidden state and screen the past from
the future. Without center coupling, the hidden sector retains
unscreened correlations --- which is P-indivisibility.

\emph{Spatial isotropy is required for Lorentz invariance:} without
rotational symmetry, the dynamics has a preferred spatial direction,
breaking the isotropy that Lorentz invariance demands.

\emph{Linearity is selected by four independent criteria.} First,
\(q\)-gauge invariance (§4): the alphabet size \(q\) is proved gauge ---
no prediction depends on it. Linear dynamics over
\(\mathbb{Z}/q\mathbb{Z}\) has \(q\)-independent structural properties
(transition probabilities, dispersion, P-indivisibility); non-linear
dynamics generically does not (\(x^2 \bmod 3\) and \(x^2 \bmod 5\) have
different algebraic structure), so the emergent physics would depend on
\(q\), contradicting the gauge freedom. Linearity is necessary, not
merely selected. Second, over \(\mathbb{R}\), nonlinear
center-independent dynamics have propagation speed \(v < 1\), so
linearity gives maximum speed. Third, over \(\mathbb{Z}/q\mathbb{Z}\),
among all linear CI dynamics \(f = \alpha(l + r) \bmod q\) with prime
\(q \geq 5\), \(\alpha = 1\) uniquely maximizes P-indivisibility.
Fourth, modified (nonlinear) dispersion relations break the thermality
of the Unruh effect --- the KMS condition holds only for linear
dispersion \(\omega = |k|\). Since the Jacobson derivation (§5) requires
thermal equilibrium at horizons, only the wave equation supports the
full GR chain. All four criteria converge on the same dynamics.

\[\phi(n, t+1) = \phi(n-1, t) + \phi(n+1, t) - \phi(n, t-1)\]

\textbf{Theorem 1.} \emph{The OI wave equation is the massless lattice
Klein-Gordon equation. Center independence is equivalent to the
vanishing of the lattice mass term.}

\emph{Proof.} The general linear update is
\(\phi(n, t+1) = \alpha\,\phi(n,t) + \beta[\phi(n-1,t) + \phi(n+1,t)] + \gamma\,\phi(n,t-1)\).
Reversibility: γ = −1. Center independence: α = 0. Isotropy + unit
speed: β = 1. The d'Alembertian
\(\Box_{\text{lat}}\phi = -\alpha\,\phi = 0\): massless KG. \(\square\)

\hypertarget{factorization-into-dirac-operators}{%
\subsubsection{7.2 Factorization into Dirac
operators}\label{factorization-into-dirac-operators}}

\textbf{Theorem 2} (Susskind {[}5{]}).
\emph{\(D_{\text{st}}^2 = -\frac{1}{4}\Box_{\text{lat}}\). The OI wave
equation is equivalent to the massless staggered Dirac equation.}

The bosonic and fermionic descriptions are related by an exact change of
variables on the lattice.

\hypertarget{center-independence-and-chiral-symmetry}{%
\subsubsection{7.3 Center independence and chiral
symmetry}\label{center-independence-and-chiral-symmetry}}

\textbf{Theorem 3.} \emph{Center independence of the lattice dynamics is
equivalent to exact chiral symmetry of the emergent staggered fermions.}

\emph{Proof.} The staggered mass term squares to
\(\Box_{\text{lat}}\phi = -4m_{\text{lat}}^2\phi\): center dependence.
Conversely, center independence gives \(D_{\text{st}}\chi = 0\),
invariant under \(\chi \to \varepsilon\chi\) (chiral symmetry).
\(\square\)

\textbf{The chain:} P-indivisibility → center independence → chiral
symmetry → Higgs mechanism {[}6, 7{]}. One algebraic condition (α = 0)
produces quantum mechanics, chiral fermions, and the Higgs boson.

\hypertarget{multi-component-dynamics-and-gauge-structure}{%
\subsubsection{7.4 Multi-component dynamics and gauge
structure}\label{multi-component-dynamics-and-gauge-structure}}

Each site now carries a K-component vector
\(\boldsymbol{\phi}(\mathbf{n}, t) \in (\mathbb{Z}/q\mathbb{Z})^K\). The
general second-order linear update is
\(\boldsymbol{\phi}(\mathbf{n}, t+1) = C\,\boldsymbol{\phi}(\mathbf{n}, t) + \sum_j M^{(j)}[\boldsymbol{\phi}(\mathbf{n}+\hat{e}_j, t) + \boldsymbol{\phi}(\mathbf{n}-\hat{e}_j, t)] + D\,\boldsymbol{\phi}(\mathbf{n}, t-1)\),
where \(C, D, M^{(j)} \in \mathrm{Mat}(K)\). Reversibility requires
\(D = -I_K\); center independence requires \(C = 0\); spatial isotropy
requires \(M^{(j)} = M\) for all \(j\). This uniquely selects the matrix
wave equation:

\[\boldsymbol{\phi}(\mathbf{n}, t+1) = M \sum_{j=1}^{d} \left[\boldsymbol{\phi}(\mathbf{n} + \hat{e}_j, t) + \boldsymbol{\phi}(\mathbf{n} - \hat{e}_j, t)\right] - \boldsymbol{\phi}(\mathbf{n}, t-1)\]

where \(M \in \mathrm{Mat}(K)\) is the sole free parameter.

\textbf{Theorem 4} (Mass spectrum). \emph{Each eigenvalue \(\mu_a\) of M
determines a dispersion relation \(\cos\omega_a = \mu_a \cos k\).
Massless iff \(\mu_a = 1\). Stability requires \(|\mu_a| \leq 1\).}

\emph{Proof.} Diagonalizing
\(M = V\,\text{diag}(\mu_1, \ldots, \mu_K)\,V^{-1}\), each eigenmode
decouples. Substituting the plane wave
\(\phi_a \propto e^{i(kn - \omega t)}\) into the \(a\)-th decoupled
equation gives \(e^{-i\omega} + e^{i\omega} = \mu_a(e^{-ik} + e^{ik})\),
hence \(\cos\omega_a = \mu_a \cos k\). For \(|\mu_a| > 1\), there exist
real \(k\) with \(|\mu_a \cos k| > 1\), giving complex \(\omega\)
(exponential growth). \(\square\)

\textbf{Theorem 5} (Gauge group). \emph{The global gauge group is the
commutant \(G = C_{\mathrm{U}(K)}(M)\). If M has r distinct eigenvalues
with multiplicities \((n_1, \ldots, n_r)\):
\(G = \mathrm{U}(n_1) \times \cdots \times \mathrm{U}(n_r)\).}

\emph{Proof.} A global transformation
\(\boldsymbol{\phi} \to U\boldsymbol{\phi}\) with
\(U \in \mathrm{U}(K)\) preserves the matrix wave equation iff
\(UMU^{-1} = M\), i.e., \([U, M] = 0\). By Schur's lemma, the set of
unitaries commuting with \(M\) decomposes as a product of unitary
groups, one on each eigenspace. \(\square\)

The gauge group and mass spectrum are the same information: modes of
equal mass share a gauge symmetry.

\hypertarget{k-2d-from-coupling-degree-minimization}{%
\subsubsection{7.5 K = 2d from coupling-degree
minimization}\label{k-2d-from-coupling-degree-minimization}}

\textbf{Theorem 6} (\(K = 2d\)). \emph{For any bounded-degree graph with
polynomial growth exponent \(d\) and statistical isotropy, the wave
equation (selected by center independence, isotropy, and linearity) has
a unique minimal-coupling factorization with \(K = 2d\) components.}

\emph{Proof (translation-invariant case).} The matrix wave equation on
\(\mathbb{Z}^d\) updates the K-component vector at site \textbf{n} using
exactly 2d independent spatial inputs:
\(\{\boldsymbol{\phi}(\mathbf{n} \pm \hat{e}_j, t)\}_{j=1}^d\). Define
the \emph{internal coupling degree} δ(K) as the maximum number of
spatial input channels on which any single component's update depends.

\emph{Step 1 (K = 2d achieves δ = 1).} With K = 2d, assign component
\(a\) to link direction
\(a \in \{\pm\hat{e}_1, \ldots, \pm\hat{e}_d\}\). If M is diagonal in
this basis, then
\(\phi_a(\mathbf{n}, t+1) = \mu_a \cdot \phi_a(\mathbf{n} + \hat{e}_{j(a)}, t) - \phi_a(\mathbf{n}, t-1)\):
each component depends on exactly one spatial input. Therefore δ(2d) =
1.

\emph{Step 2 (K \textless{} 2d forces δ ≥ 2).} By the pigeonhole
principle, at least one component must receive contributions from
\(\lceil 2d/K \rceil \geq 2\) spatial input channels. These channels
carry independent data from distinct neighboring sites, so δ(K) ≥ 2.

\emph{Step 3 (K \textgreater{} 2d forces δ ≥ 2).} With K \textgreater{}
2d components and only 2d spatial input channels, at least K − 2d
\textgreater{} 0 components have no dedicated spatial input. The matrix
M has rank at most 2d in its spatial coupling structure (since it acts
on 2d independent inputs), so the K × K coupling necessarily maps excess
components to linear combinations of other components' spatial inputs.
These excess components therefore depend on at least 2 spatial channels
through M: δ(K) ≥ 2.

\emph{Uniqueness.} Steps 1--3 show δ = 1 iff K = 2d.

\emph{Extension to general graphs.} For a bounded-degree graph \(G\)
with polynomial growth exponent \(d\) and statistical isotropy: (i)
\(G\) is quasi-isometric to \(\mathbb{Z}^d\) (§6, Gromov's theorem);
(ii) the wave equation on \(G\) has \(2d\) independent low-energy
propagation modes, determined by the spectral dimension \(d_s = d\)
(from heat kernel asymptotics \(p_t(x,x) \sim t^{-d/2}\) on graphs with
polynomial growth exponent \(d\)); (iii) the coupling-degree minimizer
\(K = 2d\) on \(\mathbb{Z}^d\) is topologically stable under
quasi-isometric deformations, because \(\delta(K)\) is integer-valued
and depends continuously on the graph structure --- it cannot change
without crossing a codimension-1 boundary in the space of statistically
isotropic bounded-degree graphs. \(\square\)

This is the per-site analog of the global factorization principle
(§3.1): at the global level, \(S = S_1 \times \cdots \times S_N\) is the
unique factorization minimizing coupling degree --- defining ``space'';
at the per-site level, \(K\) components minimize internal coupling
degree --- defining ``gauge structure.'' The objection that ``nature
need not minimize coupling degree'' mistakes the logical role. The
framework does not postulate a preference for minimal coupling ---
coupling-degree minimization is the mathematical definition of ``site''
given a bijection, and the factorization is unique (§3.1). A system with
\(K = 8\) or \(K = 4\) would correspond to a \emph{different bijection}
\(\varphi'\), not to the same \(\varphi\) with a non-minimal
factorization. Since the dynamics is selected by center independence,
isotropy, and linearity (§7.1), and its factorization is unique,
\(K = 2d = 6\) is a theorem about the wave equation on \(\mathbb{Z}^3\),
not a selection rule.

\hypertarget{cubic-decomposition-multiplicities-3-2-1}{%
\subsubsection{7.6 Cubic decomposition: multiplicities (3, 2,
1)}\label{cubic-decomposition-multiplicities-3-2-1}}

\textbf{Theorem 7.} \emph{The 2d = 6 nearest-neighbor link directions of
the d = 3 cubic lattice decompose under the rotation group O as
\(6 = T_1(3) \oplus E(2) \oplus A_1(1)\). The eigenvalue multiplicities
of M are (3, 2, 1), giving gauge group
\(\mathrm{U}(3) \times \mathrm{U}(2) \times \mathrm{U}(1) \supset \mathrm{SU}(3) \times \mathrm{SU}(2) \times \mathrm{U}(1)\).}

\emph{Proof.} Characters at each conjugacy class of O (24 elements): E:
χ = 6; 8C₃: χ = 0; 3C₂: χ = 2; 6C₂': χ = 0; 6C₄: χ = 2. Character inner
products give n(A₁) = 1, n(E) = 1, n(T₁) = 1, with n(A₂) = n(T₂) = 0. By
Schur's lemma and isotropy,
\(M = \mathrm{diag}(\mu_c I_3, \mu_w I_2, \mu_y)\). \(\square\)

The max-speed constraint requires μ\_y = 1 (the A₁ mode propagates at
the lattice speed of light). The physical identifications: T₁ (spatial
vector) → color, E (quadrupole) → weak isospin, A₁ (scalar) →
hypercharge. Local gauge invariance follows from background independence
(§5).

\textbf{Remark (robustness under statistical isotropy).} Theorem 7 uses
the character table of the octahedral group \(O\), the exact symmetry of
the cubic lattice. The structural reading of §2 claims that any
bounded-degree graph with statistical isotropy suffices. The robustness
follows from Theorem 6's general proof and two additional observations.
First, §6 establishes that the coupling graph is quasi-isometric to
\(\mathbb{Z}^3\) (Gromov's theorem + statistical isotropy + integer
polynomial growth). The relevant representation theory is that of the
large-scale rotation group \(O(3)\); the decomposition of the 2d = 6
link directions into irreducible representations has multiplicities
\((3, 2, 1)\) determined by the topology of \(d = 3\) rotations, not by
the precision of the lattice symmetry. Second, the gauge group is a
discrete structure: the eigenvalue multiplicities of \(M\) are integers,
and Theorem 5 maps integer multiplicities to unitary group factors.
Small perturbations of the graph --- breaking exact \(O\) symmetry to
statistical isotropy --- perturb the eigenvalues of \(M\) continuously
but cannot change integer multiplicities without crossing a
codimension-1 boundary in parameter space. Both \(K = 2d\) (Theorem 6)
and the \((3, 2, 1)\) decomposition are therefore topologically stable
under all deformations within the equivalence class
\([(S, \varphi)]/\mathcal{G}_{\text{sub}}\).

\hypertarget{staggered-species-generations-and-the-higgs}{%
\subsubsection{7.7 Staggered species, generations, and the
Higgs}\label{staggered-species-generations-and-the-higgs}}

The Nielsen-Ninomiya theorem {[}8{]} requires \(2^{d+1}\) Weyl species.
Staggered reduction gives
\(N_{\text{taste}} = 2^{\lfloor(d+1)/2\rfloor} = 4\) Dirac tastes in d+1
= 4.

\textbf{Theorem 8.} \emph{Under the cubic group O:
\(4 = \mathbf{1} \oplus \mathbf{3}\). One singlet taste and three
triplet tastes.}

\emph{Proof.} The 2\^{}d = 8 BZ corners pair into 4 taste pairs under
\(\boldsymbol{\eta} \leftrightarrow \mathbf{1} - \boldsymbol{\eta}\).
The pair \{(0,0,0), (1,1,1)\} is O-invariant (singlet A₁). The three
axis-aligned pairs form irreducible T₁. \(\square\)

\textbf{Theorem 8b} (Taste coupling structure). \emph{On the d = 3
lattice with the normalized wave equation, each BZ corner
\(\boldsymbol{\eta}\) has coupling
\(\mu(\boldsymbol{\eta}) = \sigma(\boldsymbol{\eta})/d\) where
\(\sigma = \sum_j \cos(\pi\eta_j)\):}

\[\mu_\Gamma = 1, \quad \mu_X = \tfrac{1}{3}, \quad \mu_M = -\tfrac{1}{3}, \quad \mu_R = -1\]

\emph{The staggered taste pairs have couplings: singlet (\(\Gamma\),
\(R\)): \(|\mu| = 1\); triplet (\(X_j\), \(M_j\)): \(|\mu| = 1/3\),
identical for all \(j = 1, 2, 3\).}

\emph{Proof.} Direct computation: \(\sigma(\Gamma) = 3\),
\(\sigma(X) = 1\), \(\sigma(M) = -1\), \(\sigma(R) = -3\). Dividing by
\(d = 3\) gives the couplings. The triplet members are related by cubic
symmetry (\(O\) permutes spatial axes), hence identical. \(\square\)

\textbf{Theorem 9} (Spin-taste correspondence). \emph{(a) The singlet
taste produces a spin-0 field. (b) The triplet taste produces spin-1/2
fields.}

\emph{Proof.} (a) The singlet pairs Γ(0,0,0) = I₄ (scalar) and Γ(1,1,1)
= γ¹γ²γ³ (pseudoscalar), both SO(3)-invariant: J = 0. (b) The j-th
triplet pairs γʲ (vector) with −iΣ\_j (spin generator): J = 1/2.
\(\square\)

\textbf{Corollary} (Spin-statistics from the lattice). Singlet → spin-0
→ boson (Higgs). Triplet → spin-1/2 → fermion (quarks and leptons).

\textbf{Theorem 10} (Three Standard Model generations). \emph{The three
triplet staggered tastes, each producing a spin-1/2 Dirac fermion with K
= 6 internal components, constitute three complete SM generations with
the anomaly-cancelling representations.}

\emph{Proof.} (i) Three spin-1/2 sectors and one spin-0 sector (Theorem
9). (ii) Anomaly cancellation (Theorem 15 below) uniquely determines the
matter representations per generation. No alternative anomaly-free
assignment exists {[}10{]}. (iii) The three sectors are related by cubic
symmetry (O permutes spatial axes). (iv) Three degenerate sectors, each
with unique anomaly-cancelling representations, constitute three
complete SM generations. \(\square\)

\textbf{Theorem 11} (Higgs sector). \emph{The singlet taste produces
spin-0 excitations with quantum numbers
\((\mathbf{1}, \mathbf{2}, +\tfrac{1}{2})\) --- exactly the Higgs
doublet.}

\emph{Proof.} The singlet taste is spin-0 (Theorem 9a). The tensor
product with the E(2) sector gives A₁ ⊗ E = E: the Higgs couples to
SU(2) as a doublet. The hypercharge +1/2 is uniquely determined by
gauge-invariant Yukawa couplings and anomaly cancellation. \(\square\)

\emph{Consequences.} Exactly one Higgs doublet (one singlet taste). No
ν\_R: the singlet taste produces the Higgs, not a fourth generation.
Neutrinos are Majorana (Weinberg operator).

\textbf{Theorem 12} (Partition = spinor decomposition). \emph{The OI
checkerboard partition coincides with the staggered spinor
decomposition: visible and hidden sectors carry complementary spinor
components.}

\emph{Proof.} The checkerboard assigns site \textbf{n} to the visible
sector iff \((-1)^{\sum n_\mu} = +1\) (even sublattice). In the
staggered-to-Dirac reconstruction, the even sublattice carries spinor
components with \(\Gamma(\eta)\) matrices of even rank (including
\(I_4\)), which give the left-handed components under Lorentz
decomposition. The odd sublattice carries the right-handed components.
The partition structure and the spinor decomposition are the same
object. \(\square\)

\textbf{Remark (the checkerboard is not a free choice).} The
checkerboard partition is not selected by hand to produce chirality ---
it is the bipartite structure of the lattice itself. The wave equation
is nearest-neighbor, so the cubic lattice is bipartite: even sites
couple only to odd sites and vice versa. This bipartite structure is a
property of the dynamics (range-1 coupling on a hypercubic graph), not
of the partition. The staggered-to-Dirac reconstruction identifies the
two sublattices with left- and right-handed spinor components --- a
standard result in lattice QFT {[}5, 12{]}, not an OI construction.
Theorem 12 observes that the OI partition (visible vs.~hidden) and the
staggered partition (left vs.~right) are the same decomposition. The
physical partition (the cosmological horizon) traces out a spatial
region containing both sublattices; at scales large compared to
\(\epsilon\), any macroscopic region contains equal numbers of even and
odd sites. Chirality in the emergent description is a property of how
the staggered-to-Dirac reconstruction organizes the lattice variables
into spinor components --- it follows from the bipartite structure of
the dynamics, not from a choice of which sites to trace out.

\hypertarget{chirality-from-the-trace-out}{%
\subsubsection{7.8 Chirality from the
trace-out}\label{chirality-from-the-trace-out}}

\textbf{Theorem 13} (Chirality). \emph{The emergent gauge coupling of
the visible sector is chiral: SU(2) is chiral while SU(3) remains
vector-like.}

\emph{Proof.} (i) \emph{Chirality = sublattice parity.} The
staggered-to-Dirac reconstruction {[}12, 5{]} assigns spinor components
to hypercube corners \(\eta \in \{0,1\}^{d+1}\) via
\(\chi(y + \eta\epsilon) = \frac{1}{4}\Gamma(\eta)_{\alpha\beta}\psi_\beta(y)\),
where \(\Gamma(\eta) = \gamma_0^{\eta_0}\cdots\gamma_d^{\eta_d}\). Under
\(\gamma_5\): \(\Gamma(\eta) \to (-1)^{|\eta|}\Gamma(\eta)\). Sites with
\(|\eta|\) even are left-handed; \(|\eta|\) odd are right-handed. This
equals the sublattice parity
\(\varepsilon(\mathbf{x}) = (-1)^{\sum x_\mu}\). (ii)
\emph{\(D_{LL} = D_{RR} = 0\).} The staggered Dirac operator couples
only across sublattices: \(\{D_{\text{st}}, \varepsilon\} = 0\) (chiral
symmetry, Theorem 3), so the even-even and odd-odd blocks vanish
identically. (iii) \emph{Trace-out removes right-handed fields.} The OI
checkerboard partition assigns even sites (left-handed) to V and odd
sites (right-handed) to H (Theorem 12). The trace-out marginalizes over
H: all operators in the emergent description act on the visible
(left-handed) Hilbert space. (iv) \emph{Left-handed effective
Lagrangian.} Since \(D_{LL} = 0\), the gauge coupling in the full theory
is purely \(L \leftrightarrow R\). After integrating out R (the hidden
sublattice), the effective coupling has the form
\(\mathcal{L}_{\text{eff}} = \bar{\psi}_L\,(D_{LR}\,G_{RR}\,D_{RL})\,\psi_L + \cdots\),
where \(G_{RR}\) is the hidden-sector propagator. Both external legs are
left-handed. (v) \emph{SU(3) vector-like; SU(2) chiral.} The chirality
projection \(P_L\) acts on the spin/sublattice index and commutes with
the internal (K-component) index. SU(3) acts on internal \(T_1(3)\)
components: both L and R carry color, so the trace-out preserves color
coupling → \textbf{vector-like}. SU(2) acts on internal \(E(2)\)
components, but the effective coupling has only L external legs →
\textbf{chiral}. \(\square\)

\hypertarget{anomaly-cancellation-and-hypercharges}{%
\subsubsection{7.9 Anomaly cancellation and
hypercharges}\label{anomaly-cancellation-and-hypercharges}}

\textbf{Theorem 14} (U(1)\_Y automaticity). \emph{The existence of a
U(1) gauge factor is automatic in the multi-component framework.}

\emph{Proof.} Each factor of the commutant decomposes as
\(\mathrm{U}(n_i) = \mathrm{SU}(n_i) \times \mathrm{U}(1)_i\). For K = 6
with multiplicities (3, 2, 1):
\(G = [\mathrm{SU}(3) \times \mathrm{U}(1)_B] \times [\mathrm{SU}(2) \times \mathrm{U}(1)_L] \times \mathrm{U}(1)_S\).
Three U(1) factors exist automatically. A general abelian charge is
\(Q = \alpha B + \beta L + \gamma S\). Given SU(3) × SU(2) and the
minimal fermion content (5 representations per generation), the six
anomaly conditions impose 4 independent constraints on the 5 hypercharge
unknowns, leaving exactly one free parameter (overall normalization).
There is therefore exactly one anomaly-free U(1). \(\square\)

\textbf{Theorem 15} (Unique hypercharges). \emph{Given
\(\mathrm{SU}(3) \times \mathrm{SU}(2) \times \mathrm{U}(1)\) with
fermions in fundamental or singlet representations, the six anomaly
conditions {[}11{]} uniquely determine the hypercharges} {[}10{]}:

\[Y_Q = \tfrac{1}{6}, \quad Y_u = \tfrac{2}{3}, \quad Y_d = -\tfrac{1}{3}, \quad Y_L = -\tfrac{1}{2}, \quad Y_e = -1\]

\textbf{Corollary.} \textbar q\_p\textbar{} = \textbar q\_e\textbar{} is
a theorem, not a coincidence.

\hypertarget{renormalizability}{%
\subsubsection{7.10 Renormalizability}\label{renormalizability}}

\textbf{Theorem 16.} \emph{The Yang-Mills coupling has mass dimension
{[}g{]} = (3−d)/2; dimensionless iff d = 3.}

\emph{Proof.} The Yang-Mills action
\(S = \frac{1}{4g^2}\int d^{d+1}x\,F_{\mu\nu}^2\) requires
\([g^{-2}][F^2] = [x]^{-(d+1)}\). With \([F] = [x]^{-2}\):
\([g]^{-2} = [x]^{d-3}\), so \([g] = (3-d)/2\). \(\square\)

Since d = 3 is derived (§6), the emergent QFT supports renormalizable
gauge interactions with asymptotic freedom.

\hypertarget{spontaneous-symmetry-breaking-and-the-hierarchy}{%
\subsubsection{7.11 Spontaneous symmetry breaking and the
hierarchy}\label{spontaneous-symmetry-breaking-and-the-hierarchy}}

Chiral symmetry (Theorem 3) combined with the center-independent gauge
sector forbids explicit masses. The unique renormalizable
unitarity-preserving mechanism is the Higgs {[}6, 7{]}. The minimal
scalar breaking
\(\mathrm{SU}(2)_L \times \mathrm{U}(1)_Y \to \mathrm{U}(1)_{\text{em}}\)
while preserving \(\mathrm{SU}(3)_c\) is
\(H = (\mathbf{1}, \mathbf{2}, +\tfrac{1}{2})\).

The hierarchy problem is dissolved: in the OI framework, the QFT is
emergent. The Higgs mass is set by eigenvalues of M (properties of φ),
not by radiative corrections.

\hypertarget{the-filter-chain}{%
\subsubsection{7.12 The filter chain}\label{the-filter-chain}}

The complete derivation from (S, φ) to the Standard Model is tabulated
in Appendix B. The primary chain is proved end-to-end: wave equation →
KG → staggered fermions → chiral symmetry → K = 2d = 6 → cubic
decomposition (3,2,1) → local gauge invariance → anomaly cancellation →
chirality → three generations → one Higgs → θ̄ = 0. Redundant
second-route confirmations (asymptotic freedom, minimal N\_c, minimal
chiral group) are at proposition level and provide independent checks.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-strong-cp-problem}{%
\subsection{8. The Strong CP Problem}\label{the-strong-cp-problem}}

\hypertarget{t-invariance-of-the-wave-equation}{%
\subsubsection{8.1 T-invariance of the wave
equation}\label{t-invariance-of-the-wave-equation}}

\textbf{Theorem 17.} \emph{The discrete wave equation is invariant under
time reversal T: φ(n, t) → φ(n, −t).}

\emph{Proof.} Under T, define ψ(n, t) = φ(n, −t). Substitution gives
ψ(n, t+1) = ψ(n−1, t) + ψ(n+1, t) − ψ(n, t−1): the same wave equation.
\(\square\)

The checkerboard partition marginalizes over spatial sites, not temporal
data. It preserves T while breaking P (§7.8).

\hypertarget{ux3b8-0}{%
\subsubsection{8.2 θ = 0}\label{ux3b8-0}}

\textbf{Theorem 18.} \emph{The QCD θ-parameter vanishes: θ = 0.}

\emph{Proof.} The θ-term
\(\mathcal{L}_\theta = (\theta/32\pi^2) \mathrm{Tr}(F \tilde{F})\) is
T-odd: \(\mathrm{Tr}(F\tilde{F}) = 2\,\mathbf{E} \cdot \mathbf{B}\), and
under T, \(\mathbf{E} \to \mathbf{E}\), \(\mathbf{B} \to -\mathbf{B}\).
A T-invariant Lagrangian cannot contain a T-odd term. The emergent
Lagrangian inherits T-invariance because the partition is spatial, so
time reversal commutes with the visible/hidden projection. \(\square\)

\hypertarget{detailed-balance-and-ux3b8-0}{%
\subsubsection{8.3 Detailed balance and θ̄ =
0}\label{detailed-balance-and-ux3b8-0}}

\textbf{Theorem 19} (Detailed balance). \emph{The visible-sector
transition probabilities satisfy \(T_{ij} = T_{ji}\) for all i, j.}

\emph{Proof.} The full dynamics φ is a bijection, and T-invariance
(Theorem 17) gives
\(\varphi^{-1} = \mathcal{T} \circ \varphi \circ \mathcal{T}\). For each
hidden state h with \(\pi_V(\varphi(i, h)) = j\), let
\(h' = \pi_H(\varphi(i, h))\). Then \(\varphi(i, h) = (j, h')\), so
\(\varphi^{-1}(j, h') = (i, h)\). Since the partition is spatial (not
temporal), \(\mathcal{T}\) preserves the visible/hidden classification:
\(\pi_V \circ \mathcal{T} = \pi_V\). The map \(h \mapsto h'\) is an
injection from \(\{h : \pi_V(\varphi(i,h)) = j\}\) into
\(\{h' : \pi_V(\varphi(j,h')) = i\}\). Since φ is a bijection on a
finite set, the injection is a bijection between sets of equal
cardinality. Dividing by \(|\mathcal{C}_H|\): \(T_{ij} = T_{ji}\).
\(\square\)

\textbf{Theorem 20} (Detailed balance implies T-invariant Hamiltonian).
\emph{If \(T_{ij} = T_{ji}\) for all i, j, then the emergent Hamiltonian
satisfies \([\hat{H}_{\text{eff}}, \hat{\Theta}] = 0\) where
\(\hat{\Theta}\) is the anti-unitary time-reversal operator.}

\emph{Proof.} The Born rule gives \(T_{ij}(t) = |U_{ij}(t)|^2\) {[}1,
§3.1{]}. Combined with \(T_{ij} = T_{ji}\): \(|U_{ij}|^2 = |U_{ji}|^2\)
for all i, j, t. This symmetry of the transition amplitudes, by Wigner's
theorem, implies \(U(t) = \hat{\Theta}\,U(t)^*\,\hat{\Theta}^{-1}\) for
a suitable anti-unitary \(\hat{\Theta}\), which gives
\([\hat{H}_{\text{eff}}, \hat{\Theta}] = 0\). \(\square\)

\textbf{Theorem 21} (\(\bar{\theta} = 0\) at all energy scales).
\emph{The physical parameter
\(\bar{\theta} = \theta + \arg\det(Y_u Y_d) = 0\), and the vanishing
persists to all infrared scales.}

\emph{Proof.} \textbf{Step 1:} Since φ is a bijection, \(\varphi^n\) is
a bijection for every n.~Applying the counting argument of Theorem 19 to
\(\varphi^n\): \(T_{ij}(n) = T_{ji}(n)\) for all n.~(Verified
numerically: T-violation is exactly zero for L = 4, 6, q = 2, 3, 5, and
all n = 1, \ldots, 12.)

\textbf{Step 2:} By Theorem 20 applied to \(T_{ij}(n)\) at each scale:
\([\hat{H}_{\mathrm{eff}}(n), \hat{\Theta}] = 0\). The effective
Hamiltonian is T-invariant at every time scale.

\textbf{Step 3:} T-invariant Yukawa couplings at each scale are
simultaneously real in an appropriate basis {[}10, §6.3{]}. Real Yukawa
matrices have \(\arg\det(Y_u Y_d) = 0\). Combined with θ = 0 (Theorem
18): \(\bar{\theta} = 0\) at every scale. \(\square\)

\emph{Remark.} This proof bypasses the instanton question. T-invariance
of the transition probabilities is an exact consequence of the bijection
structure, holding at every time scale without perturbative or
non-perturbative approximation. The RG flow cannot generate T-violation
because the underlying bijection structure forbids it at every scale.

\textbf{Prediction:} \(\bar{\theta} = 0\) exactly. No axion needed.
Neutron EDM should be exactly zero (current bound:
\(|d_n| < 1.8 \times 10^{-26}\;e\cdot\text{cm}\) {[}22{]}).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{discussion}{%
\subsection{9. Discussion}\label{discussion}}

\hypertarget{structural-realism}{%
\subsubsection{9.1 Structural realism}\label{structural-realism}}

The structural reading aligns with ontic structural realism but does not
require it. What the framework proves is that (S, φ) is a \emph{complete
description} of reality up to gauge equivalence (the reconstruction
theorem, §10.3). Whether the structure \emph{is} reality (the OSR
commitment) or \emph{describes} a reality that exists independently is a
question the framework identifies as gauge --- provably undecidable by
any measurement. The ``stuff'' of the universe, in any reading, is fully
characterized by the abstract structure of a bijection on a finite set
with bounded coupling.

\hypertarget{the-ontological-hierarchy}{%
\subsubsection{9.2 The ontological
hierarchy}\label{the-ontological-hierarchy}}

The triple (S, φ, V) generates every concept in fundamental physics, not
as independent substances but as different aspects of the same
structure. \textbf{Space} is the coupling structure of φ --- the graph
G\_φ determined by which degrees of freedom affect which others (§3.1).
\textbf{Matter} is the state --- localized patterns that propagate
through the coupling graph. \textbf{Energy} is the rate of change under
iteration. \textbf{Time} is the iteration itself. \textbf{Quantum
mechanics} is the observer's compressed description of the visible
sector. \textbf{General relativity} is the thermodynamic limit of the
coupling structure. \textbf{Conservation laws} are emergent: energy
conservation (Noether) is what information conservation (bijectivity)
looks like in the emergent quantum description. None of these are
independent entities; they are descriptions of (S, φ, V) at different
scales.

\hypertarget{the-measurement-problem}{%
\subsubsection{9.3 The measurement
problem}\label{the-measurement-problem}}

On the structural reading, the measurement problem is dissolved. The
wave function is not a component of (S, φ, V) --- it is a derived
object. Since it is derived, not fundamental, asking ``does it
collapse?'' is asking about the behavior of a compression artifact. In
the double-slit experiment, the particle traverses a single slit in the
deterministic substratum. In Wigner's friend, the Friend has a definite
outcome; Wigner's superposition reflects his epistemic deficit.

Branching is forbidden by the rigidity of φ. A fixed bijection on a
finite set has exactly one trajectory from any initial state. There is
no point at which the trajectory splits. The appearance of branching in
the emergent quantum description reflects the observer's uncertainty
about which trajectory they are on (because they cannot see the hidden
sector), not a physical splitting of worlds.

Non-locality in Bell correlations is explained by the coupling graph
G\_φ. Two visible sites prepared in a joint state (entangled) have
correlated hidden-sector configurations --- a consequence of the joint
P-indivisible dynamics at preparation {[}1, §3.2{]}. The correlations
are mediated by the coupling graph, not by any superluminal influence.
The graph structure ensures that the correlations respect the Tsirelson
bound and violate Bell inequalities without violating parameter
independence.

\hypertarget{observer-genericity}{%
\subsubsection{9.4 Observer genericity}\label{observer-genericity}}

\textbf{Theorem 22} (Genericity of observers). \emph{Let φ be the wave
equation (or any energy-conserving dynamics) on a connected
bounded-degree coupling graph with diameter D ≥ 4. Then for any
connected subgraph V with \(|V| \leq N/3\), the partition (V, H)
satisfies C1--C3.}

\emph{Proof.} C1: \(G_\varphi\) is connected and V is a proper subgraph,
so ∂V ≠ $\emptyset$. The wave equation couples nearest neighbors; any edge in ∂V
produces non-trivial dynamical coupling.

C2: the wave equation is a Hamiltonian system --- it conserves energy
and preserves phase-space volume. The hidden sector has spectral gap Δ =
0: correlations persist indefinitely, not just until some relaxation
time. The C2 necessity proof {[}1, §3.3{]} establishes that
\(\Delta \tau_S \gg 1\) drives the system to Markovianity; for any
Hamiltonian hidden sector, \(\Delta \tau_S = 0 \ll 1\) --- maximally in
the slow-bath regime, regardless of partition geometry.

C3: \(|H| = N - |V| \geq 2N/3\). The hidden-sector configuration space
has \(q^{|H|}\) states; the non-Markovian mutual information is bounded
by \(|H| \log_2 q\) bits {[}1, §3.3{]}. The capacity ratio
\(q^{|H| - |V|} \geq q^{N/3}\) is exponentially large. No visible-sector
process of sub-exponential duration saturates the hidden sector's
memory. \(\square\)

\emph{Remark (validity window).} While C2 is satisfied maximally, the
emergent Hamiltonian has a finite \emph{stationarity window}: the time
before information deposited at the V--H boundary propagates through H
and returns to V at a different boundary point. The wave equation has an
exact light cone (no exponential tails), so the minimum return time is
\(\tau_{\mathrm{return}} = \min_{b, b' \in \partial V} \mathrm{dist}_H(b, b')\).
For core-shell partitions with core radius r on a lattice of side L:
\(\tau_{\mathrm{return}} = (L - 2r)/v\). After
\(\tau_{\mathrm{return}}\), the emergent description remains quantum
mechanical (C2 is still satisfied), but the emergent Hamiltonian becomes
time-dependent. Numerical verification: for L = 20--320,
\(\tau_{\mathrm{return}}\) ranges from 6 to 318 steps, confirming the
light-cone bound.

\textbf{Corollary.} Any finite deterministic system with
energy-conserving dynamics and bounded-degree coupling necessarily
contains subsystems whose internal description is quantum mechanics. The
observer is not postulated --- it is a mathematical consequence.

\hypertarget{relation-to-other-approaches}{%
\subsubsection{9.5 Relation to other
approaches}\label{relation-to-other-approaches}}

\textbf{Causal set theory.} The bijection φ with coupling graph G\_φ
generates a causal partial order that is a causal set in the sense of
Bombelli-Lee-Meyer-Sorkin {[}4{]}. The OI framework provides what CST
lacks: a deterministic dynamics that provably produces both QM and GR.
The CST Hauptvermutung corresponds to the Myrheim-Meyer dimension result
{[}4{]}.

\textbf{'t Hooft {[}17{]}.} Agrees on emergent QM from determinism but
differs: OI requires bijectivity (not information loss), violates
outcome independence (not superdeterminism), and derives ℏ, S\_BH, and
Einstein's equations.

\textbf{Wolfram {[}18{]}.} Shares discrete dynamics and emergent space
but differs: OI requires bounded coupling for the area law and produces
quantitative predictions for physical constants.

\textbf{Loop quantum gravity {[}19{]}.} Both find discrete spatial
geometry at the Planck scale, but LQG quantizes GR (starts from the
continuum and discretizes), while OI derives both QM and GR from a
pre-quantum starting point.

\hypertarget{predictions}{%
\subsubsection{9.6 Predictions}\label{predictions}}

The framework makes the following falsifiable predictions.

From the cosmological application {[}1{]}: dark energy evolution with
\(\nu_{\text{OI}} \approx 2.45 \times 10^{-3}\) (DESI, Euclid); GW
echoes near black hole horizons (LIGO/Virgo/KAGRA).

From the SM derivation (§§7--8): no additional gauge groups below
\(M_{\text{Pl}}\); no fundamental scalars beyond the Higgs doublet; no
SUSY partners; no fourth generation; \(\bar{\theta} = 0\) exactly (no
axion, neutron EDM = 0); the CKM CP-violating phase arises from
P-violation (the partition), not from T-violation; neutrinos are
Majorana; no proton decay from GUT-scale gauge bosons (the eigenvalues
of M correspond to independent irreps of O, so couplings do not unify).

The conjunction is distinctive to this framework.

\hypertarget{what-the-framework-determines-and-what-the-specific-bijection-determines}{%
\subsubsection{9.7 What the framework determines and what the specific
bijection
determines}\label{what-the-framework-determines-and-what-the-specific-bijection-determines}}

The framework's predictions fall into two categories, and the
distinction matters for assessing their status.

\textbf{Structural predictions} are properties of the equivalence class
{[}(S, φ){]}/\(\mathcal{G}_{\text{sub}}\) --- they hold for every
bijection satisfying the six structural properties. These include: SU(3)
× SU(2) × U(1) with multiplicities (3, 2, 1), three chiral generations
with identical gauge quantum numbers, one Higgs doublet (1, 2, +1/2),
unique hypercharges from anomaly cancellation, \(\bar{\theta} = 0\) at
all energy scales, ℏ = c³ε²/(4G), and the Bekenstein-Hawking formula.
These are theorems. They do not depend on which specific φ describes our
universe.

\textbf{Solution-specific properties} are determined by the particular
bijection φ within the equivalence class. These include: gauge couplings
(set by the eigenvalues μ\_c, μ\_w of M), fermion masses (set by
taste-breaking in φ at \(\mathcal{O}(\epsilon^2)\)), and mixing
parameters (set by the Yukawa structure of φ). These are analogous to
the mass \(M\) in the Schwarzschild solution: GR derives the functional
form of the metric but does not predict which \(M\) describes a
particular black hole. The OI framework derives the SM's structure but
does not predict which φ describes our universe.

The ``mass hierarchy'' --- the five-order-of-magnitude spread of fermion
masses --- falls in the second category. It is not a gap in the
framework; it is a property of the specific solution. The framework
makes the correct structural prediction: three generations with
\emph{identical} gauge quantum numbers (from cubic symmetry) but
\emph{non-degenerate} masses (because the specific φ breaks the cubic
symmetry through its Yukawa structure). The hierarchy problem proper ---
why the Higgs mass is so much lighter than the Planck scale --- is
dissolved: the Higgs mass is an eigenvalue of M (a property of φ), not a
radiative correction within the emergent QFT (§7.11).

\textbf{Remaining open problems.} (i) \emph{Lattice-continuum
comparison} --- the structural predictions are exact theorems at the
lattice level and do not require a continuum limit. The lattice is the
fundamental description (\(\epsilon = 2l_p\) is physical, not a
regulator), so there is no \(\epsilon \to 0\) limit to take. The
comparison to experiment --- which is interpreted through continuum
perturbation theory --- introduces corrections of order
\((E/M_{\text{Pl}})^2 \sim 10^{-32}\) at LHC energies, far below any
foreseeable experimental sensitivity. The Yang-Mills mass gap problem
(proving rigorous existence of the continuum limit for Yang-Mills
theory) is a famous open problem in mathematics, but it does not affect
any prediction of the OI framework: the lattice theory is complete as
stated, and the lattice is the fundamental description, not a
regularization. (ii) \emph{Generation identification} --- the fermion
count (three) is proved; the assignment of specific representations to
specific tastes is a uniqueness argument from anomaly cancellation, not
a geometric derivation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{synthesis-the-complete-program}{%
\subsection{10. Synthesis: The Complete
Program}\label{synthesis-the-complete-program}}

\hypertarget{summary-of-results}{%
\subsubsection{10.1 Summary of results}\label{summary-of-results}}

The framework starts from four axioms (deterministic dynamics,
finiteness, causal partition, classical probability) and three
conditions (C1--C3), with no quantum mechanics, general relativity, or
specific dynamics. The companion paper {[}1{]} proves that embedded
observation under these conditions is equivalent to unitary QM, derives
ℏ = c³ε²/(4G) and the Bekenstein-Hawking formula, dissolves the CC
discrepancy, and predicts dark energy evolution and GW echoes. This
paper establishes that center independence, isotropy, and linearity
uniquely select the wave equation (§7.1), which passes all seven GR
tests with no free parameters (§5). The framework's results depend on
six structural properties and nothing else; the wave equation on a d = 3
lattice determines the Standard Model with the primary derivation chain
proved end-to-end (§7).

\textbf{Theorem dependency structure.} The 24 theorems are not a linear
chain of 24 sequential steps. They form a tree with four independent
branches. The critical path through the SM derivation is five theorems
long (1 → 2 → 5 → 6 → 7); the strong CP solution is independent of the
SM structure; and the reconstruction uses all preceding results but adds
only two new theorems.

\begin{verbatim}
Axioms + d=3
  │
  ├─→ Thm 1 (wave eq) ─→ Thm 2 (Dirac) ─→ Thm 3 (chiral sym)
  │         │
  │         ├─→ Thm 4 (mass spectrum)
  │         └─→ Thm 5 (gauge = commutant of M)
  │
  ├─→ Thm 6 (K=2d) ─┐
  │                   ├─→ Thm 7 (cubic decomposition → multiplicities (3,2,1))
  │                   │     │
  │                   │     ├─→ Thms 8–11 (tastes, generations, Higgs)
  │                   │     ├─→ Thm 13 (chirality)
  │                   │     └─→ Thms 14–15 (hypercharges, anomaly cancellation)
  │                   │
  │                   └─→ Thm 16 (renormalizability iff d=3)
  │
  ├─→ Thm 12 (partition = spinor) [independent of Thms 6–11]
  │
  ├─→ Thms 17–21 (strong CP: θ̄ = 0) [independent of Thms 6–16; uses only bijectivity]
  │
  ├─→ Thm 22 (observer genericity) [independent of Thms 1–21]
  │
  └─→ Thm 23 (reconstruction) [uses Thms 1–22]
        └─→ Thm 24 (substratum gauge group) [uses Thm 23]
\end{verbatim}

Each branch can be verified independently. The strong CP block (Thms
17--21) requires only the bijectivity of φ and the wave equation's
T-invariance --- it does not depend on the gauge structure derivation.
The observer genericity theorem (Thm 22) is independent of all preceding
results.

\hypertarget{the-physical-interpretation-of-s-ux3c6}{%
\subsubsection{10.2 The physical interpretation of (S,
φ)}\label{the-physical-interpretation-of-s-ux3c6}}

\textbf{Storage and memory.} S is the set of all distinguishable states:
\emph{finite capacity}. φ is a bijection: \emph{perfect memory} ---
information is never created or destroyed. Together, (S, φ) is a finite
lossless memory. The partition V defines the observer's \emph{access}
--- both read and write. The observer reads the visible sector
(measuring x) and writes to the hidden sector (each visible-sector
operation imprints correlations on H through the coupling
\(H_{\text{int}}\)). The slow bath (C2) preserves those writes;
subsequent reads retrieve them --- producing the information backflow
that constitutes P-indivisibility. Quantum mechanics is the statistics
of this read-write cycle: not passive observation of a static memory,
but the self-consistent description that emerges when a finite subsystem
both reads from and writes to a lossless register it cannot fully
access. The Born rule is the equilibrium of the cycle, not of passive
reading. Interference is write-then-read: information deposited in the
hidden sector during one transition is retrieved on a later transition
and partially cancels or reinforces the transition probabilities.

The \(10^{122}\) CC discrepancy is the compression ratio between total
storage and readable storage. The dark sector is the gravitational
effect of the unreadable storage. The Bekenstein-Hawking entropy is the
storage capacity of the partition boundary. The action scale ℏ is the
conversion factor between storage geometry and read-write statistics.

\textbf{The substrate objection.} ``What is the memory made of?'' (S, φ)
is a \emph{complete description} of reality --- it determines all
observables. Space, time, matter, and energy are derived from (S, φ), so
they cannot appear in its definition without circularity. Whether (S, φ)
\emph{is} reality or \emph{describes} reality is provably undecidable by
any measurement (reconstruction theorem below).

\textbf{Relation to computation.} A Turing machine has a tape (storage),
a head (partial read/write access), and a transition function (update
rule). The correspondence is suggestive: S is the tape, V is the head's
access window, φ is the transition function. But three differences are
physically significant. First, a Turing machine's tape is potentially
infinite; S is finite --- finiteness is essential for the recurrence
proof of P-indivisibility and QM, though the effective finiteness result
(Main, §9.6) shows the deep hidden sector may be infinite without
affecting the observable physics. Second, a Turing machine is generally
irreversible (it can erase, overwrite, and halt); φ is a bijection ---
nothing is erased, nothing is created, there is no halting. Third, a
Turing machine computes an \emph{extrinsic function} (input → output for
an external user); (S, φ) computes no extrinsic function --- it is a
closed permutation that cycles through states and returns. The
appearance of dynamics, probability, particles, and forces is entirely
the observer's perspective --- what the permutation looks like through
the partial window V.

\textbf{Remark (the Turing connection under effective finiteness).} The
three differences soften under the effective finiteness result. With the
deep hidden sector potentially infinite, the first difference is between
the formal axiom (S finite) and the physical requirement (only
\(\mathcal{C}_V \times \mathcal{C}_B\) need be finite). The second
difference is a specialization, not an opposition: reversible Turing
machines (Bennett, Fredkin-Toffoli) are a well-studied subclass. The
third difference --- extrinsic vs.~intrinsic --- is the one that does
physical work. The mapping is then structural: V is the head, H is the
tape, φ is a reversible transition function, and C1--C3 characterize the
architecture. The framework extends Turing's question: instead of asking
what a machine can compute for an external observer, it asks what
computation looks like to a component of the machine --- and proves the
answer is quantum mechanics.

\textbf{The arrow of time.} The substratum has no arrow of time --- φ
and φ⁻¹ are equally valid. Entropy increase is a property of the
observer's coarse-grained description: the standard Boltzmann mechanism
applied to the partition.

\textbf{The incompleteness family.} The framework's central result
belongs to a family of impossibility-with-structure theorems. Gödel: a
formal system cannot prove all truths about itself --- the unprovable
truths have rigid structure. Turing: a computer cannot decide all
questions about its own behavior --- the undecidable problems have rigid
structure. OI: an observer embedded in a deterministic system cannot
access the complete state --- the emergent description has rigid
structure (quantum mechanics). The common structure is
\emph{self-reference under finite resources}.

The precise OI analog of the halting problem is: \emph{can the observer
determine the hidden-sector state \(h\)?} The question is well-posed ---
\(h\) has a definite value at every moment, because \((S, \varphi)\) is
deterministic. Different \(h\) values produce different physical
outcomes. But the observer provably cannot determine \(h\): multiple
hidden states are compatible with any visible-sector history, and
transition probabilities \(T_{ij}(t)\) are averages over \(h\). The
structural consequence of this inaccessibility is quantum mechanics ---
just as the structural consequence of the halting problem's
undecidability is computability theory. The framework also identifies a
second class of inaccessible quantities --- the alphabet size \(q\) (§4)
and the deep-sector cardinality \(|\mathcal{C}_D|\) (Main, §5.2) --- but
these are \emph{gauge}, not undecidable: different values produce
identical observables, so the question itself is physically empty. The
hidden state \(h\) is undecidable (real answer, provably inaccessible).
The cardinality \(|\mathcal{C}_D|\) is gauge (no answer to find).

\textbf{Mathematics and physics.} The trace-out performs a
Jordan-Chevalley projection (Appendix A): it extracts the semisimple
part of the dynamics and erases the nilpotent monodromy. Physics is the
semisimple shadow of mathematics --- the diagonalizable spectral data,
projected by the trace-out and organized by the gauge group's
representation structure. The reconstruction theorem (below) proves that
the mathematical description and the physics determine each other
uniquely up to gauge.

\hypertarget{the-reconstruction-theorem}{%
\subsubsection{10.3 The reconstruction
theorem}\label{the-reconstruction-theorem}}

The forward direction --- from (S, φ) to observed physics --- is
established by §§5, 7--8 and the companion paper {[}1{]}. The converse
question is whether the observed physics uniquely determines (S, φ). The
reconstruction proceeds in three stages.

\textbf{Stage 1: From observed QM to deterministic embedding.}

\emph{Empirical input:} Unitary quantum mechanics with Bell violations,
on a configuration space with finite entropy bounded by a finite-area
surface.

The finite-entropy condition has independent justification from
holographic bounds {[}3, 20{]}. The cosmological horizon has finite
area; the bound applies.

\emph{Derivation.} Any CPTP quantum channel on a finite-dimensional
Hilbert space admits a Stinespring dilation as a bijection on a finite
set {[}1, Appendix A{]}. Bell violations require non-trivial coupling
(C1). The characterization theorem {[}1, §3.3{]} requires slow-bath
memory (C2) and sufficient capacity (C3).

\emph{Output:} The observed QM arises from some (S, φ, V) satisfying
C1--C3. \emph{Status:} Theorem.

\textbf{Stage 2: From the embedding to the dynamics and gauge
structure.}

\emph{Additional empirical input:} Spatial isotropy.

\emph{Derivation.} The coupling graph gives d = 3 (§6). Center
independence + isotropy + linearity uniquely select the wave equation
(§7.1). The wave equation determines the full SM structure: K = 2d = 6
(Theorem 6), cubic decomposition (3,2,1) (Theorem 7), local gauge
invariance (§5), staggered species → three generations + one Higgs
(Theorems 8--11), partition-spinor identification (Theorem 12),
chirality (Theorem 13), anomaly cancellation (Theorems 14--15),
T-invariance → θ̄ = 0 (Theorems 17--21).

\emph{Output:} Dynamics, dimension, gauge group, matter content, and
discrete symmetries are all derived. The SM structure is a retrodiction
checked against observation. \emph{Status:} Theorem at the lattice
level.

\textbf{Stage 3: From the dynamics to the emergent constants.}

\emph{No additional input.} The gap equation from thermal
self-consistency {[}1, §5{]} gives \(\hbar = c^3 \epsilon^2 / (4G)\)
with ε = 2 l\_p.~\emph{Status:} Theorem.

\textbf{Theorem 23} (Layered reconstruction). \emph{Observed quantum
mechanics with Bell violations and finite boundary entropy, together
with spatial isotropy, uniquely determine --- at the lattice level ---
the equivalence class \([(S, \varphi)]/\!\sim\): a finite set with a
bijection of bounded coupling degree and statistical isotropy, with d =
3, K = 6, coupling matrix eigenvalue multiplicities (3, 2, 1),
\(\bar{\theta} = 0\), and \(\hbar = c^3 \epsilon^2/(4G)\).}

\emph{Proof.} Stage 1 ⇒ Stage 2 ⇒ Stage 3, as detailed above.
\(\square\)

The theorem starts from empirical inputs (QM, Bell violations, finite
entropy, isotropy) and derives the framework. The SM gauge structure,
three generations, one Higgs, θ̄ = 0, and the Bekenstein-Hawking formula
are retrodictions --- derived, not assumed.

\textbf{Remark} (Evidential weight of the reconstruction). The
reconstruction uniquely derives
\(\text{SU}(3) \times \text{SU}(2) \times \text{U}(1)\) with
multiplicities \((3, 2, 1)\), three chiral generations, one Higgs
doublet \((1, 2, +1/2)\), unique hypercharges, and \(\bar{\theta} = 0\).
Because the derivation is unique --- no free parameters, no
model-building choices, no landscape of alternatives --- every
independent experimental confirmation of the Standard Model's gauge
structure is retroactive evidence for the derivation chain that produces
it. The Standard Model is the most precisely tested mathematical
structure in the history of science (\(g - 2\) of the electron confirmed
to \(10^{-12}\), electroweak precision observables to \(10^{-5}\), QCD
cross-sections across decades of energy). This experimental record
transfers in full to the framework, because uniqueness of the derivation
makes the relationship between evidence and theory transitive: data
confirms structure, structure is uniquely derived, therefore data
confirms the derivation.

\textbf{Remark} (Scope). The reconstruction establishes uniqueness
within the class of finite deterministic systems with bounded coupling.
It does not exclude intrinsically continuum or stochastic alternatives.

\textbf{Remark} (Continuum extension). The lattice-level predictions are
the framework's primary claims --- the lattice is the fundamental
description, not an approximation to a continuum theory. The structural
results (gauge group, representations, generation count,
\(\bar{\theta} = 0\)) are algebraic/topological properties of the
lattice theory and are exact. Quantitative observables (scattering
amplitudes, mass ratios) are compared to experiment through continuum
perturbation theory; the lattice-continuum discrepancy at any
experimentally accessible energy \(E\) is suppressed by
\((E\epsilon/\hbar c)^2 = (E/M_{\text{Pl}})^2\), which is
\(\sim 10^{-32}\) at LHC energies. The rigorous proof that lattice
Yang-Mills defines a continuum limit with a mass gap (a Clay Millennium
Prize problem) is an open problem in mathematics, but it is not required
by the OI framework: the lattice theory is complete, and its predictions
are lattice-exact.

The reconstruction establishes a bidirectional correspondence:

\[\text{Observed physics (QM + Bell + finite entropy + isotropy)} \quad \longleftrightarrow \quad [(S, \varphi)] / \sim\]

The mathematical structure and the physics determine each other up to
gauge equivalence. The distinction between ``mathematics describes
reality'' and ``mathematics is reality'' has no empirical content --- it
is itself gauge. This reframes Wigner's puzzle: the ``unreasonable
effectiveness'' of mathematics is a theorem, not a mystery.

\hypertarget{the-substratum-gauge-group}{%
\subsubsection{10.4 The substratum gauge
group}\label{the-substratum-gauge-group}}

The equivalence relation \(\sim\) in the reconstruction theorem has a
precise structure. Define: \((S, \varphi) \sim (S', \varphi')\) if the
two systems produce identical emergent physics --- the same transition
probabilities \(T_{ij}(t)\), the same emergent Hamiltonian (up to
D-gauge), and the same \(\hbar\) --- for all partitions of the same
structural class. The set of transformations mapping \((S, \varphi)\) to
an equivalent \((S', \varphi')\) is a group --- the \emph{substratum
gauge group} \(\mathcal{G}_{\text{sub}}\).

\textbf{Theorem 24} (Generators of the substratum gauge group).
\emph{\(\mathcal{G}_{\text{sub}}\) contains at least four independent
families of transformations:}

\emph{(i) State relabeling.} For any bijection \(\sigma: S \to S\), the
conjugate
\((S, \sigma \circ \varphi \circ \sigma^{-1}) \sim (S, \varphi)\). The
transition probabilities depend on the coupling structure of
\(\varphi\), not on which labels are attached to which states. This is
an \(|S|!\)-element subgroup --- vastly larger than any gauge group in
the Standard Model.

\emph{(ii) Alphabet change.} Replacing the local state space
\(\mathbb{Z}/q\mathbb{Z}\) with \(\mathbb{Z}/q'\mathbb{Z}\) for any
\(q' \geq 2\), while preserving the coupling graph and dynamics class,
leaves all observables unchanged (§4). This family is parametrized by
all integers \(q \geq 2\).

\emph{(iii) Deep-sector enlargement.} Adjoining additional degrees of
freedom to \(\mathcal{C}_D\) (the deep hidden sector beyond the boundary
layer), with arbitrary dynamics satisfying \(\tau_B^D \gg \tau_S\), does
not change the emergent description. The boundary-only dependence lemma
{[}1, §5.2{]} proves
\(T_{ij}(t) = T_{ij}^{(B)}(t) + \mathcal{O}(t/\tau_B)\): observables
depend only on \(\mathcal{C}_V \times \mathcal{C}_B\). The deep sector
may be finite of any size, or infinite.

\emph{(iv) Graph isomorphism (up to statistical isotropy).} Two coupling
graphs \(G_\varphi\) and \(G_{\varphi'}\) that are quasi-isometric with
the same polynomial growth exponent \(d\), the same spectral properties,
and the same statistical isotropy at large scales produce the same
emergent physics. The regular cubic lattice \(\mathbb{Z}^3\) and any
bounded-degree random graph with \(d = 3\) polynomial growth and
statistical isotropy are gauge-equivalent.

\emph{Proof.} Each generator preserves all inputs to the derivation
chain. (i): conjugation preserves the coupling graph \(G_\varphi\) up to
relabeling, hence all graph-dependent quantities (area law, dispersion,
dimension, eigenvalue multiplicities). (ii): §4 proves
\(q\)-independence of every prediction. (iii): the boundary-only
dependence lemma gives the result directly. (iv): the derivation chain
uses only statistical properties of \(G_\varphi\) (dimension via
Myrheim-Meyer, isotropy, bounded degree), not the specific graph.
\(\square\)

\textbf{Completeness.} The reconstruction theorem (Theorem 23)
establishes that the equivalence class \([(S, \varphi)]/\!\sim\) is
uniquely determined by the observed physics. Any transformation not in
\(\mathcal{G}_{\text{sub}}\) would change at least one observable,
contradicting equivalence. The four generators therefore span the full
gauge group --- \(\mathcal{G}_{\text{sub}}\) is the kernel of the
reconstruction map.

\textbf{The gauge hierarchy.} Three levels of gauge symmetry appear in
the framework, each projecting onto the next through the trace-out:

\emph{Level 3 (substratum):} \(\mathcal{G}_{\text{sub}}\) acts on
\((S, \varphi)\) before the trace-out. It is the largest gauge group and
includes transformations with no analog in the emergent description
(deep-sector enlargement, alphabet change).

\emph{Level 2 (emergent QFT):}
\(\text{SU}(3) \times \text{SU}(2) \times \text{U}(1)\) is the commutant
of the coupling matrix \(M\), acting on the emergent fields. It is the
image of \(\mathcal{G}_{\text{sub}}\) restricted to transformations that
permute internal components within the eigenspaces of \(M\).

\emph{Level 1 (emergent Hamiltonian):} The D-gauge \(H \to DHD^\dagger\)
with \(D\) a diagonal unitary, acting on the emergent Hamiltonian within
the emergent QM. It is the residual freedom after all
transition-probability data has been extracted.

Each level is contained in the one above: Level 1 \(\subset\) Level 2
\(\subset\) Level 3. The trace-out projects Level 3 onto Level 2 (the SM
gauge group is the shadow of \(\mathcal{G}_{\text{sub}}\) visible to the
emergent QFT), and restricting to the Hamiltonian projects Level 2 onto
Level 1.

\textbf{Remark.} The substratum gauge group is not a symmetry of a
Lagrangian or an action --- no Lagrangian exists at the substratum
level. It is a symmetry of the \emph{equivalence class of substrata},
defined by the condition that all observables are preserved. The
emergent gauge symmetries (Levels 1 and 2) are Lagrangian symmetries in
the standard sense, derived from the substratum through the trace-out.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{conclusion}{%
\subsection{11. Conclusion}\label{conclusion}}

The lattice in the OI framework is not a physical substrate. It is the
coupling graph of a bijection on a finite set. Space is the large-scale
geometry of this graph. Time is the iteration of the bijection. Quantum
mechanics is the observer's compressed description of the visible
sector. General relativity is the thermodynamic limit of the coupling
structure. The Standard Model is the specific QFT determined by the
lattice structure in d = 3.

The fundamental object is (S, φ) --- a finite set and a bijection.
Physically: a finite lossless memory. The partition V is derived (any
small subgraph sees QM). The dimension d = 3 is derived (four
independent filters converge). The gauge group, matter content, and
discrete symmetries are derived (the primary chain is proved
end-to-end). The reconstruction theorem (§10.3) establishes the
converse: observed physics uniquely determines {[}(S,
φ){]}/\(\mathcal{G}_{\text{sub}}\) at the lattice level. The universe is
completely described by a lossless memory with finite read-write access.
Physics is what that memory looks like from inside --- and what it looks
like from inside uniquely determines what it is.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

During the preparation of this work, the author used Claude Opus 4.6
(Anthropic) and Gemini 3.1 Pro (Google) to assist in drafting, numerical
simulations, and refining argumentation. The author reviewed and edited
all content and takes full responsibility for the publication.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\subsection{References}\label{references}}

{[}1{]} A. Maybaum, ``The Incompleteness of Observation,'' submitted to
Foundations of Physics (2026).

{[}2{]} J. Eisert, M. Cramer, and M. B. Plenio, ``Area laws for the
entanglement entropy,'' \emph{Rev.~Mod. Phys.} \textbf{82}, 277 (2010).

{[}3{]} R. Bousso, ``The holographic principle,'' \emph{Rev.~Mod. Phys.}
\textbf{74}, 825 (2002).

{[}4{]} L. Bombelli, J. Lee, D. Meyer, and R. Sorkin, ``Space-time as a
causal set,'' \emph{Phys. Rev.~Lett.} \textbf{59}, 521 (1987).

{[}5{]} L. Susskind, ``Lattice fermions,'' \emph{Phys. Rev.~D}
\textbf{16}, 3031 (1977).

{[}6{]} P. W. Higgs, ``Broken symmetries and the masses of gauge
bosons,'' \emph{Phys. Rev.~Lett.} \textbf{13}, 508 (1964).

{[}7{]} B. W. Lee, C. Quigg, and H. B. Thacker, ``Weak interactions at
very high energies: The role of the Higgs-boson mass,'' \emph{Phys.
Rev.~D} \textbf{16}, 1519 (1977).

{[}8{]} H. B. Nielsen and M. Ninomiya, ``Absence of neutrinos on a
lattice,'' \emph{Nucl. Phys. B} \textbf{185}, 20 (1981).

{[}9{]} K. G. Wilson, ``Confinement of quarks,'' \emph{Phys. Rev.~D}
\textbf{10}, 2445 (1974).

{[}10{]} R. N. Mohapatra, \emph{Unification and Supersymmetry}
(Springer, 3rd ed., 2003), §6.3.

{[}11{]} L. Alvarez-Gaumé and E. Witten, ``Gravitational anomalies,''
\emph{Nucl. Phys. B} \textbf{234}, 269 (1984).

{[}12{]} J. B. Kogut and L. Susskind, ``Hamiltonian formulation of
Wilson's lattice gauge theories,'' \emph{Phys. Rev.~D} \textbf{11}, 395
(1975).

{[}13{]} P. Ehrenfest, ``In what way does it become manifest in the
fundamental laws of physics that space has three dimensions?''
\emph{Proc. Amsterdam Acad.} \textbf{20}, 200 (1917).

{[}14{]} F. R. Tangherlini, ``Schwarzschild field in n dimensions and
the dimensionality of space problem,'' \emph{Nuovo Cim.} \textbf{27},
636 (1963).

{[}15{]} P. van der Hoorn, W. J. Cunningham, G. Lippner, C.
Trugenberger, and D. Krioukov, ``Ollivier-Ricci curvature convergence in
random geometric graphs,'' \emph{Phys. Rev.~Research} \textbf{3}, 013211
(2021).

{[}16{]} C. Kelly, C. Trugenberger, and F. Biancalana, ``Convergence of
combinatorial gravity,'' \emph{Phys. Rev.~D} \textbf{105}, 124002
(2022).

{[}17{]} G. 't Hooft, \emph{The Cellular Automaton Interpretation of
Quantum Mechanics} (Springer, 2016).

{[}18{]} S. Wolfram, ``A Class of Models with the Potential to Represent
Fundamental Physics,'' Complex Systems 29, 107 (2020).

{[}19{]} C. Rovelli, \emph{Quantum Gravity} (Cambridge University Press,
2004).

{[}20{]} N. Bao, S. M. Carroll, and A. Singh, ``The Hilbert space of
quantum gravity is locally finite-dimensional,'' \emph{Int. J. Mod.
Phys. D} \textbf{26}, 1743013 (2017).

{[}21{]} Y. Ollivier, ``Ricci curvature of Markov chains on metric
spaces,'' \emph{J. Funct. Anal.} \textbf{256}, 810 (2009).

{[}22{]} C. Abel et al., ``Measurement of the permanent electric dipole
moment of the neutron,'' \emph{Phys. Rev.~Lett.} \textbf{124}, 081803
(2020).

{[}23{]} V. Eisler, ``On the Bisognano-Wichmann entanglement Hamiltonian
of nonrelativistic fermions,'' \emph{J. Stat. Mech.} (2025) 013101.

{[}24{]} T. Jacobson, ``Thermodynamics of Spacetime: The Einstein
Equation of State,'' \emph{Phys. Rev.~Lett.} \textbf{75}, 1260 (1995).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-a-the-trace-out-as-a-jordan-chevalley-projection}{%
\subsection{Appendix A: The Trace-Out as a Jordan-Chevalley
Projection}\label{appendix-a-the-trace-out-as-a-jordan-chevalley-projection}}

This appendix proves that the OI trace-out extracts the semisimple part
of the evolution matrix's Jordan-Chevalley decomposition, erasing the
nilpotent monodromy. Results are stated for the scalar wave equation on
a ring of L sites over \(\mathbb{F}_q\); the multi-component extension
incorporates the gauge group.

\hypertarget{setup}{%
\subsubsection{Setup}\label{setup}}

The wave equation
\(x(n, t+1) = x(n-1, t) + x(n+1, t) - x(n, t-1) \pmod{q}\) has phase
space \((\mathbb{Z}/q\mathbb{Z})^{2L}\) and evolution matrix
\(F = \left(\begin{smallmatrix} 0 & I \\ -I & A \end{smallmatrix}\right)\)
where A is the circulant adjacency matrix.

\hypertarget{theorem-a.1-period-formula}{%
\subsubsection{Theorem A.1 (Period
formula)}\label{theorem-a.1-period-formula}}

\emph{For q prime with \(\gcd(L, q) = 1\):
\(\mathrm{ord}(F \bmod q) = qL\) if q is odd, and L if q = 2.}

\emph{Proof.} For each Fourier mode k, the 2×2 block \(B_k\) has
characteristic polynomial \(t^2 - \lambda_k t + 1\). At parabolic modes
(\(\lambda_k = \pm 2\)), the Jordan form gives
\(B_k^n = \alpha^n \left(\begin{smallmatrix} 1 & n\alpha^{-1} \\ 0 & 1 \end{smallmatrix}\right)\),
contributing order q (or 2q). The diagonalizable eigenvalues contribute
orders dividing L. Therefore \(\mathrm{ord}(F) = qL\). For q = 2: the
nilpotent part is automatically killed. \(\square\)

\hypertarget{theorem-a.2-jordan-chevalley-decomposition}{%
\subsubsection{Theorem A.2 (Jordan-Chevalley
decomposition)}\label{theorem-a.2-jordan-chevalley-decomposition}}

\emph{Define \(N = (F^L - I)/L \bmod q\). Then: (i) N is nilpotent with
N² = 0 and rank(N) = 2. (ii) \(F_u = I + N\) is unipotent with
\(F_u^q = I\). (iii) \(F_{\mathrm{ss}} = F \cdot (I - N)\) is semisimple
with \(F_{\mathrm{ss}}^L = I\). (iv) \(F = F_{\mathrm{ss}} \cdot F_u\)
and \([F_{\mathrm{ss}}, N] = 0\).}

\emph{Proof.} \(F^L\) has the form \(I + LN'\) where \(N'\) arises from
the Jordan blocks:
\(\left(\begin{smallmatrix} 1 & 1 \\ 0 & 1 \end{smallmatrix}\right)^L = \left(\begin{smallmatrix} 1 & L \\ 0 & 1 \end{smallmatrix}\right)\)
contributes a rank-1 nilpotent, and similarly for \(\alpha = -1\) (since
\((-1)^L = 1\) for even \(L\)). So \(N = N'\) has \(N^2 = 0\) (each
block is rank-1 nilpotent on a 2D subspace) and rank\((N) = 2\) (two
parabolic modes). Since \(N^2 = 0\): \((I + N)^{-1} = I - N\), giving
\(F_{\mathrm{ss}} = F(I - N)\). Then
\(F_{\mathrm{ss}}^L = F^L(I - N)^L = (I + LN)(I - LN + \ldots) = I \bmod q\)
since terms involving \(LN\) cancel modulo \(q\) (using \(N^2 = 0\)).
Commutativity follows from \(N\) being supported on the parabolic
eigenspaces, which are \(F\)-invariant. \(\square\)

\emph{Verified computationally for L = 4, 6, 8, 10, 12 and q = 3, 5, 7,
11, 13.}

\hypertarget{theorem-a.3-q-independence-of-the-weil-deligne-conductor}{%
\subsubsection{Theorem A.3 (q-independence of the Weil-Deligne
conductor)}\label{theorem-a.3-q-independence-of-the-weil-deligne-conductor}}

\emph{The Weil-Deligne conductor
\(\mathfrak{f}_{\mathrm{WD}} = \mathfrak{f}_{\mathrm{ss}}(L) + \mathrm{rank}(N) = \mathfrak{f}_{\mathrm{ss}}(L) + 2\)
is q-independent when \(\gcd(L, q) = 1\), where
\(\mathfrak{f}_{\mathrm{ss}}(L) = \sum_\alpha (\mathrm{ord}(\alpha) - 1)\)
is computed from the eigenvalue orders of \(F_{\mathrm{ss}}\), all
dividing \(L\).}

\emph{Proof.} \(F_{\mathrm{ss}}\) has order L; its eigenvalues are L-th
roots of unity. For \(\gcd(L, q) = 1\), the L-th roots in
\(\bar{\mathbb{F}}_q\) are isomorphic to those in \(\mathbb{C}\), so
their orders match. \(\square\)

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3571}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\(L\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\mathfrak{f}_{\mathrm{ss}}(L)\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\mathfrak{f}_{\mathrm{WD}}\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\mathrm{NM}^2 = 3L/4\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
4 & 14 & 16 & 3.00 \\
6 & 30 & 32 & 4.50 \\
8 & 70 & 72 & 6.00 \\
10 & 106 & 108 & 7.50 \\
12 & 130 & 132 & 9.00 \\
\end{longtable}

Both \(\mathfrak{f}_{\mathrm{ss}}\) and \(\mathrm{NM}^2\) are
q-independent encodings of the same semisimple eigenvalue data --- one
via multiplicative orders (integers), one via fourth moments of
magnitudes (reals).

\hypertarget{theorem-a.4-additive-decomposition-over-gauge-irreps}{%
\subsubsection{Theorem A.4 (Additive decomposition over gauge
irreps)}\label{theorem-a.4-additive-decomposition-over-gauge-irreps}}

\emph{For the K-component wave equation with coupling matrix
\(M = \mathrm{diag}(\mu_1 I_{n_1}, \ldots, \mu_r I_{n_r})\):}

\[\mathfrak{f}_{\mathrm{WD}}(M) = \sum_{i=1}^{r} n_i \cdot \mathfrak{f}_{\mathrm{WD}}(\mu_i)\]

\emph{In particular, for K = 6 with
\(M = \mathrm{diag}(\mu_c I_3, \mu_w I_2, 1)\):
\(\mathfrak{f}_{\mathrm{WD}} = 3\mathfrak{f}_{\mathrm{WD}}(\mu_c) + 2\mathfrak{f}_{\mathrm{WD}}(\mu_w) + \mathfrak{f}_{\mathrm{WD}}(1)\),
with the same multiplicities \((3, 2, 1)\) that determine
\(\mathrm{SU}(3) \times \mathrm{SU}(2) \times \mathrm{U}(1)\).}

\emph{Proof.} The multi-component evolution matrix is block-diagonal:
the \(a\)-th component has its own \(2L \times 2L\) block \(F_{\mu_a}\)
with eigenvalues depending only on \(\mu_a\). Both \(F_{\mathrm{ss}}\)
and \(N\) inherit the block-diagonal structure, so
\(\mathfrak{f}_{\mathrm{ss}}\) and rank\((N)\) decompose additively.
\(\square\)

\emph{Verified for all \((\mu_c, \mu_w) \in \{1, \ldots, q-1\}^2\) at q
= 3, 5, 7 with L = 4. Every case matches exactly.}

\hypertarget{the-projection}{%
\subsubsection{The projection}\label{the-projection}}

The OI trace-out (marginalization over the hidden sector) produces the
emergent quantum description, which depends on the coupling eigenvalues
\(\mu_k\) via the NM formula \(\mathrm{NM}^2 = 3\langle\mu^4\rangle\) (a
consequence of the stochastic-quantum correspondence applied to the wave
equation's Fourier decomposition). These \(\mu_k\) are properties of
\(F_{\mathrm{ss}}\) --- the semisimple part of the dynamics. The
nilpotent monodromy \(N\) contributes nothing to the emergent
description: it affects only the off-diagonal Jordan block entries,
which are erased by the coarse-graining over the hidden sector.

The trace-out therefore performs the Jordan-Chevalley projection
\((F_{\mathrm{ss}}, N) \mapsto F_{\mathrm{ss}} \mapsto \{\mu_k\} \mapsto \mathrm{NM}^2\):
it extracts the semisimple part, encodes it via magnitudes rather than
orders, and organizes it by the representation theory of the partition.
The nilpotent monodromy --- genuine mathematical structure present in
the full dynamics --- is invisible to the embedded observer. This is the
precise sense in which physics is the semisimple shadow of mathematics:
the observer sees only the diagonalizable spectral data, projected by
the trace-out and organized by the gauge group's representation
structure.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-b-the-filter-chain}{%
\subsection{Appendix B: The Filter
Chain}\label{appendix-b-the-filter-chain}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1765}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3235}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2647}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2353}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Step
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Constraint
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Selects
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Status
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
0 & Emergent QFT, d = 3 & Lattice QFT, 3+1D & Theorem \\
1 & Wave eq. = massless KG (Thm 1) & Staggered fermions & Theorem \\
2 & Center indep. = chiral sym. (Thm 3) & Massless chiral fermions &
Theorem \\
3 & Multi-component → matrix WE (§7.4) & Gauge structure from M &
Theorem \\
4 & Gauge group = commutant of M (Thm 5) &
\(\mathrm{U}(n_1) \times \cdots\) & Theorem \\
4a & K = 2d = 6, cubic decomp. (Thms 6--7) & Multiplicities (3,2,1) &
Theorem \\
5 & Background indep. → local gauge (§5) & Local gauge invariance &
Theorem \\
6 & Dimensionless g (Thm 16) & Renormalizable gauge theories &
Theorem \\
7 & Spin-taste + cubic sym. (Thms 8--11) & 1 Higgs + 3 gen with SM reps
& Theorem \\
8 & Partition = spinor decomp. (Thm 12) & Checkerboard ↔ chirality &
Theorem \\
9 & \(D_{LL} = 0\) + trace-out (Thm 13) & SU(2) chiral, SU(3)
vector-like & Theorem \\
10 & U(1)\_Y automaticity (Thm 14) & Unique abelian factor & Theorem \\
11 & Anomaly cancellation (Thm 15) & SM hypercharges & Theorem \\
12 & T-invariance (Thms 17--18) & θ = 0 & Theorem \\
13 & Detailed balance at all scales (Thms 19--21) & θ̄ = 0 at all scales
& Theorem \\
\end{longtable}

\textbf{Output:} SU(3)\_c × SU(2)\_L × U(1)\_Y with:

\[Q_L = (\mathbf{3}, \mathbf{2}, +\tfrac{1}{6}), \quad u_R = (\mathbf{3}, \mathbf{1}, +\tfrac{2}{3}), \quad d_R = (\mathbf{3}, \mathbf{1}, -\tfrac{1}{3})\]
\[L_L = (\mathbf{1}, \mathbf{2}, -\tfrac{1}{2}), \quad e_R = (\mathbf{1}, \mathbf{1}, -1)\]

Three generations. One Higgs doublet
\(H = (\mathbf{1}, \mathbf{2}, +\tfrac{1}{2})\). \(\bar{\theta} = 0\).
Primary chain proved end-to-end. This is exactly the Standard Model.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-c-non-markovianity-under-gauge-structure}{%
\subsection{Appendix C: Non-Markovianity Under Gauge
Structure}\label{appendix-c-non-markovianity-under-gauge-structure}}

The multi-component system's P-indivisibility decomposes additively over
eigenmodes:

\[\|G_V(2) - G_V(1)^2\|_F^2 = \sum_{a=1}^{K} \|g^{\mu_a}(2) - g^{\mu_a}(1)^2\|_F^2\]

The normalized non-Markovianity has the exact closed form:

\[\mathrm{NM}_{\text{norm}} = \sqrt{3} \cdot \sqrt{\langle \mu^4 \rangle}\]

where \(\langle \mu^4 \rangle = \frac{1}{K}\sum_{a=1}^{K} \mu_a^4\).
This is independent of L and depends on M only through its eigenvalue
spectrum. NM is maximized at \(M = I_K\); the SM structure (3,2,1) with
μ\_w, μ\_c \textless{} 1 gives NM \textless{} √3: P-indivisibility does
not select the gauge group. Verified numerically for K = 1--12, L =
8--50, μ ∈ {[}0,1{]} with machine-precision agreement.

\end{document}
