% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\hypertarget{ux210f-the-bekenstein-hawking-entropy-and-dynamical-dark-energy-from-the-cosmological-horizon}{%
\section{ℏ, the Bekenstein-Hawking Entropy, and Dynamical Dark Energy
from the Cosmological
Horizon}\label{ux210f-the-bekenstein-hawking-entropy-and-dynamical-dark-energy-from-the-cosmological-horizon}}

\textbf{Author:} Alex Maybaum\\
\textbf{Date:} April 2026\\
\textbf{Status:} DRAFT PRE-PRINT\\
\textbf{Classification:} Theoretical Physics / Gravitation / Cosmology

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Applied to the cosmological horizon as a causal partition, the
embedded-observation framework of {[}Main{]} determines the emergent
action scale \(\hbar = c^3 \epsilon^2 / (4G)\) from classical horizon
temperature and thermal self-consistency, fixes the discreteness scale
\(\epsilon = 2\,l_p\) as the unique simultaneous solution, derives the
Bekenstein-Hawking entropy \(S = A/(4 l_p^2)\) including the \(1/4\)
coefficient by direct boundary-mode counting, and dissolves the
cosmological constant problem by identifying the quantum vacuum energy
and the effective cosmological constant as properties of logically
distinct descriptions rather than commensurable quantities whose
\(10^{122}\) ratio requires cancellation. The recent LIGO/Virgo/KAGRA
observation GW250114 (September 2025) confirms the black-hole area law
at \(99.999\%\) confidence. The framework predicts dynamical dark energy
in running-vacuum form with a structural coefficient
\(\nu_{\text{OI}} = 2.45 \times 10^{-3}\), currently supported by DESI
DR2 at \(2.8\)--\(4.2\sigma\) depending on the parameterization; a
\textasciitilde{}\(95\%\) dark sector corollary in which the apparent
invisible matter is a structural feature of the embedded observer's
description rather than a new particle species; and a MOND-like
acceleration scale \(a_0 = cH/6\) from entropy displacement at the
cosmological boundary, reproducing the baryonic Tully-Fisher relation
and predicting Keplerian decline of galactic rotation curves at radii
exceeding \(r_M = \sqrt{6 G M_B/(cH)}\) --- the last confirmed by Jiao
et al.~(2023) in the Milky Way at \(\sim 17\) kpc. The \(H(z)\)
dependence of \(a_0\) predicts systematically more baryon-dominated
rotation curves at high redshift, consistent with Genzel et al.'s
detection of declining outer velocities at \(z = 0.9\)--\(2.4\). All
results follow from partition geometry alone and are independent of the
specific substratum dynamics, treated in a separate companion paper
{[}SM{]}.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{introduction}{%
\subsection{1. Introduction}\label{introduction}}

The cosmological observables that most resist explanation within
standard physics --- the value of \(\hbar\), the Bekenstein-Hawking area
law, the \(10^{122}\) cosmological constant problem, the accelerated
expansion, the \textasciitilde95\% dark sector, and the MOND-like
behavior of galactic rotation curves --- are treated here as
consequences of a single structural fact: the cosmological horizon
partitions the degrees of freedom of any sub-\(c\) observer into a
visible sector and a causally inaccessible hidden sector. A companion
paper {[}Main{]} establishes that an observer embedded in a
deterministic system whose hidden sector is coupled, slow, and
high-capacity necessarily describes the visible sector using quantum
mechanics, and that these conditions are necessary as well as
sufficient. The cosmological horizon satisfies all three conditions:
causal coupling via the Einstein constraint equations, a slow-bath
timescale of order the Hubble time \(\tau_B \sim 10^{17}\) s, and a
hidden-sector capacity of order \(S_{\text{dS}} \sim 10^{122}\) modes.
We apply the framework to this case and derive its consequences for
gravitation and cosmology.

The results are as follows. The emergent action scale
\(\hbar = c^3 \epsilon^2 / (4G)\) is determined by thermal
self-consistency between the classical horizon temperature (computed
from the Jacobson thermodynamic argument with no quantum input) and the
Euclidean KMS period of the emergent QFT on the same horizon (§3). The
discreteness scale \(\epsilon = 2\,l_p\) is then fixed as the unique
simultaneous solution of the action-scale relation and the definition of
the Planck length (§4). The Bekenstein-Hawking entropy follows directly
by mode counting, with the \(1/4\) coefficient derived rather than
assumed as the ratio \(2\pi/8\pi\) between the Euclidean KMS period and
the Jacobson prefactor (§5); the recent LIGO/Virgo/KAGRA observation
GW250114 confirms the area law at \(99.999\%\) confidence. The
cosmological constant problem is dissolved by identifying the emergent
quantum vacuum energy of order \(M_{\text{Pl}}^4\) and the effective
cosmological constant of order \(H^2/G\) as properties of logically
distinct objects rather than commensurable quantities whose \(10^{122}\)
ratio requires cancellation (§6).

The framework's empirical content is developed in §7. Dynamical dark
energy emerges in running-vacuum form with a single structural
coefficient \(\nu_{\text{OI}} = 2.45 \times 10^{-3}\), currently
supported by DESI DR2 at the \(2.8\)--\(4.2\sigma\) level (depending on
the dark-energy parameterization adopted). The dark-sector corollary
predicts that \textasciitilde95\% of the gravitational budget of the
universe is invisible to the emergent QFT without requiring new
fundamental particles --- a concordance rather than a puzzle. The
acceleration scale \(a_0 = cH/6\) emerges from entropy displacement at
the cosmological boundary, reproducing the MOND phenomenology in the
deep-modified regime and predicting Keplerian decline at radii exceeding
\(r_M = \sqrt{6 G M_B/(cH)}\); the latter is confirmed by the Jiao et
al.~(2023) detection of a Keplerian decline in the Milky Way rotation
curve at \(\sim 17\) kpc. The \(H(z)\) dependence of \(a_0\) predicts
systematically more baryon-dominated rotation curves at high redshift,
consistent with the declining outer velocities observed at
\(z = 0.9\)--\(2.4\) by Genzel et al.

Three epistemic tiers are worth distinguishing before the technical
development begins. \textbf{Tier 1} comprises the results that follow
from partition geometry alone --- the derivation of \(\hbar\), the BH
area law with the \(1/4\) coefficient, the CC dissolution, and the
structural coefficient \(\nu_{\text{OI}} = 2.45 \times 10^{-3}\). These
do not depend on the specific substratum dynamics and would hold for any
finite deterministic system satisfying the three structural conditions
at its cosmological horizon. \textbf{Tier 2} comprises the dark-sector
phenomenology: the \textasciitilde95\% invisible matter fraction, the
MOND acceleration scale, and the rotation-curve crossover. These follow
from the thermodynamic Clausius relation applied to the boundary
reservoir and the geometric factor from spherical redistribution in de
Sitter space. \textbf{Tier 3} comprises solution-specific quantities ---
the transition steepness between the MOND and Keplerian regimes, the
greybody factor for Hawking radiation, and the specific microscopic
realization of the substratum --- which depend on the particular
bijection \(\varphi\) and are addressed in the companion paper {[}SM{]}.

The paper is organized as follows. §2 identifies the cosmological
horizon as a causal partition and verifies the structural conditions. §3
derives the emergent action scale \(\hbar\) in four steps: the classical
horizon temperature, boundary-only dependence, dimensional
determination, and thermal self-consistency. §4 fixes the discreteness
scale \(\epsilon = 2\,l_p\). §5 derives the Bekenstein-Hawking entropy
and discusses the GW250114 confirmation. §6 dissolves the cosmological
constant problem. §7 presents the quantitative predictions for dark
energy, the dark sector, and dark matter, with a summary table in §7.4.
§8 discusses scope, the status of the discreteness scale, vacuum energy
and the Higgs potential, and the framework's falsifiability conditions.
§9 concludes. The cubic-lattice realization of the substratum --- which
fixes the Standard Model structure via the wave equation --- is
developed in a companion paper {[}SM{]} and is independent of the GR
results presented here. The reconstruction theorem and the synthesis
claim that quantum mechanics and general relativity descend as different
projections of the same \((S, \varphi)\) are developed in
{[}Substratum{]}.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-cosmological-horizon-as-causal-partition}{%
\subsection{2. The Cosmological Horizon as Causal
Partition}\label{the-cosmological-horizon-as-causal-partition}}

\hypertarget{the-partition}{%
\subsubsection{2.1 The partition}\label{the-partition}}

The cosmological horizon --- the boundary beyond which no signal
propagating at or below \(c\) can reach the observer --- implements the
causal partition of Lemma 2: \(\Gamma_V\) is the interior, \(\Gamma_H\)
everything beyond. This is a consequence of GR's causal structure, not a
modeling choice. The universality of the observed quantum mechanics
follows from all sub-\(c\) observers sharing the same causal structure.

Different observers have different horizon areas (hence different
\(S_{\text{dS}}\)), but \(\hbar = c^3 \epsilon^2/(4G)\) depends only on
local geometric quantities --- not on the observer's worldline or
horizon area --- so all observers share the same emergent action scale.

\hypertarget{verification-of-the-conditions}{%
\subsubsection{2.2 Verification of the
conditions}\label{verification-of-the-conditions}}

\textbf{(C1)} In the ADM formulation, the bulk Hamiltonian is a sum of
constraints that vanish on-shell; physical time evolution is generated
by the boundary term at the horizon, which depends on data from both
sides. The Hamiltonian and momentum constraints further correlate
interior and exterior initial data on any Cauchy surface, so the
physical phase space is a constraint surface within the kinematic
product \(\Gamma_V \times \Gamma_H\) (see §8.4 for implications). This
is \emph{stronger} than the \(H_{\text{int}} \neq 0\) required by Lemma
2: constraints enforce correlations that persist on all timescales and
cannot be perturbatively removed. \textbf{Satisfied.}

\textbf{(C2)} \(\tau_B \gtrsim 1/H \sim 10^{17}\) s; for laboratory
processes, \(\tau_S \sim 10^{-15}\) s. Ratio:
\(\tau_S / \tau_B \sim 10^{-32}\). \textbf{Satisfied.}

\textbf{(C3)} Hidden sector has \(A/\epsilon^2\) modes with
\(A \sim 10^{52}\) m\(^2\); visible sector has \(\sim 10^{80}\) baryons.
No laboratory process saturates the hidden sector. \textbf{Satisfied.}

\hypertarget{application}{%
\subsubsection{2.3 Application}\label{application}}

Lemma 1 is now a consequence of partition geometry: the horizon has
finite area \(A = 4\pi c^2/H^2\), bounding coupled modes to
\(A/\epsilon^2 < \infty\). The Part I theorem applies: the observer's
reduced description is P-indivisible and, by Barandes' correspondence,
equivalent to unitary QM with \(\hbar\) determined by the partition
boundary.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-emergent-action-scale-hbar}{%
\subsection{\texorpdfstring{3. The Emergent Action Scale
\(\hbar\)}{3. The Emergent Action Scale \textbackslash hbar}}\label{the-emergent-action-scale-hbar}}

\hypertarget{the-classical-horizon-temperature}{%
\subsubsection{3.1 The classical horizon
temperature}\label{the-classical-horizon-temperature}}

Jacobson {[}1{]} showed \(dE = T\,dS\) applied to local causal horizons
yields Einstein's equations, with \(dE = (c^2 \kappa / 8\pi G)\,dA\),
where \(\kappa\) is the surface gravity of the horizon. The entropy
density is \(\eta = 1/\epsilon^2\) --- one coupled mode per minimal
cell. This is not an assumption about the number of states per cell but
a consequence of two results: \(\epsilon\) is defined as the minimal
distinguishable scale (Lemma 1), so each cell of area \(\epsilon^2\)
contributes exactly one boundary mode that couples across the partition;
the number of internal states per mode (the alphabet size \(q\)) is a
gauge freedom with no observable consequences (the \(q\)-gauge theorem:
two systems with different \(q\) produce identical transition
probabilities if they share the same coupling structure). Thus
\(dS = dA/\epsilon^2\). Equating:

\[T_{\text{cl}} = \frac{c^2 \epsilon^2 \kappa}{8\pi G k_B}\]

For de Sitter (\(\kappa = cH\)):
\(k_B T_{\text{cl}} = c^3 \epsilon^2 H / (8\pi G)\). No \(\hbar\)
appears.

\hypertarget{the-emergent-action-scale}{%
\subsubsection{3.2 The emergent action
scale}\label{the-emergent-action-scale}}

\textbf{Step 1: Uniqueness.} By partition-relativity ({[}Main, §1.4{]}),
\(\hbar\) is derived, not free.

\textbf{Step 2: \(\hbar\) is structural, not volumetric.}

\textbf{Lemma (Boundary-only dependence).} \emph{Decompose the hidden
sector as \(\mathcal{C}_H = \mathcal{C}_B \times \mathcal{C}_D\), where
\(\mathcal{C}_B\) denotes boundary modes (within the interaction range
of \(H_{\text{int}}\)) and \(\mathcal{C}_D\) denotes deep modes (the
remainder of the hidden sector). If the classical Hamiltonian is
spatially local, then on timescales \(t \ll \tau_B\):}

\[T_{ij}(t) = T^{(B)}_{ij}(t) + \mathcal{O}(t/\tau_B)\]

\emph{where \(T^{(B)}_{ij}\) depends only on the \(V\)-\(B\) dynamics.}

\emph{Proof.} (i) \emph{Coupling structure.} Spatial locality of the
classical Hamiltonian implies the interaction decomposes as
\(H_{\text{tot}} = H_V + H_B + H_D + H_{VB} + H_{BD}\), with no direct
\(V\)-\(D\) coupling: a deep hidden-sector mode must propagate through
the boundary layer before influencing the visible sector. The coupling
chain is \(V \leftrightarrow B \leftrightarrow D\), not
\(V \leftrightarrow D\).

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\setcounter{enumi}{1}
\item
  \emph{Frozen deep sector.} Let \(\Delta_D\) be the spectral gap of
  \(H_D + H_{BD}\) restricted to the deep sector conditioned on a fixed
  boundary configuration. The deep-sector evolution over time \(t\)
  displaces any initial configuration by \(\mathcal{O}(\Delta_D \, t)\)
  in phase space. The spectral gap gives the inverse relaxation time of
  the slowest mode, so \(\Delta_D \leq 1/\tau_B\) (with equality when
  the slowest mode dominates the relaxation). For
  \(t \ll 1/\Delta_D \leq \tau_B\):
  \(\|\varphi_t^D - \text{id}\| = \mathcal{O}(\Delta_D \, t) = \mathcal{O}(t/\tau_B)\).
\item
  \emph{Factorization of the Liouville integral.} The transition
  probability is:
\end{enumerate}

\[T_{ij}(t) = \frac{1}{|\mathcal{C}_B||\mathcal{C}_D|} \sum_{b \in \mathcal{C}_B} \sum_{d \in \mathcal{C}_D} \delta_{x_j}[\pi_V(\varphi_t(x_i, b, d))]\]

By (i), \(\pi_V(\varphi_t(x_i, b, d))\) depends on \(d\) only through
the back-reaction \(B \leftarrow D\), which by (ii) shifts \(b\) by
\(\mathcal{O}(t/\tau_B)\). Expanding:

\[\pi_V(\varphi_t(x_i, b, d)) = \pi_V(\varphi_t^{VB}(x_i, b)) + \mathcal{O}(t/\tau_B)\]

where \(\varphi_t^{VB}\) is the flow restricted to \(V \times B\) with
\(D\) frozen. The \(d\)-sum then contributes \(|\mathcal{C}_D|\)
identical terms at leading order:

\[T_{ij}(t) = \underbrace{\frac{1}{|\mathcal{C}_B|} \sum_{b \in \mathcal{C}_B} \delta_{x_j}[\pi_V(\varphi_t^{VB}(x_i, b))]}_{T^{(B)}_{ij}(t)} + \mathcal{O}(t/\tau_B) \qquad \square\]

The correction is \(\mathcal{O}(t/\tau_B) \sim 10^{-32}\) for laboratory
processes. Since \(T_{ij}(t)\) determines the emergent quantum
description ({[}Main, §3.1{]}), and \(T^{(B)}_{ij}(t)\) depends only on
the \(V\)-\(B\) dynamics --- which are characterized by the boundary
geometry (\(A\), \(\epsilon\), \(\kappa\)) and the constants \(c\),
\(G\) appearing in the classical Hamiltonian --- the emergent action
scale \(\hbar\) inherits boundary-only dependence.

\textbf{Corollary (deep-sector cardinality is gauge).} \emph{No
observable of the emergent description --- no transition probability,
emergent Hamiltonian eigenvalue, or scattering amplitude --- depends on
the cardinality \(|\mathcal{C}_D|\) of the deep hidden sector. Systems
with the same \(\mathcal{C}_V \times \mathcal{C}_B\) dynamics produce
the same emergent physics to within \(\mathcal{O}(\tau_S/\tau_B)\),
whether \(|\mathcal{C}_D|\) is finite, countably infinite, or
uncountably infinite.}

\emph{Proof.} \(T^{(B)}_{ij}(t)\) is defined by a sum over
\(\mathcal{C}_B\) alone. The \(\mathcal{C}_D\)-dependent terms enter
only through the \(\mathcal{O}(t/\tau_B)\) back-reaction. Since the
emergent Hamiltonian \(\hat{H}_{\text{eff}}\) is uniquely determined by
\(\{T_{ij}(t)\}\) (phase-locking lemma, {[}Main, §3.1{]}; D-gauge
theorem, §3.3), it inherits \(\mathcal{C}_D\)-independence at leading
order. All observables --- eigenvalues, transition amplitudes,
scattering cross-sections --- are functions of \(\hat{H}_{\text{eff}}\)
and therefore \(\mathcal{C}_D\)-independent. \(\square\)

The cardinality of the deep hidden sector is gauge in the same technical
sense as the alphabet size \(q\): just as two systems with different
\(q\) are physically equivalent if they share the same coupling
structure, two systems with different \(|\mathcal{C}_D|\) are physically
equivalent if they share the same \(\mathcal{C}_V \times \mathcal{C}_B\)
dynamics. Lemma 1 requires \(S\) finite for the recurrence proof
({[}Main, §2.3{]}); the corollary establishes that the deep sector's
contribution to this finiteness has no observable consequences. The
question ``is the universe finite or infinite?'' has no empirical
content within the framework --- it is identified as gauge.

\textbf{Step 3: Dimensional determination.} Step 2 excludes volumetric
(deep-sector) quantities, leaving boundary quantities. The boundary
carries both \emph{local} geometric data (\(\epsilon\), \(\kappa\), and
the constants \(c\), \(G\) of the classical Hamiltonian) and a
\emph{global} quantity: the total area \(A\), which forms the
dimensionless ratio \(A/\epsilon^2 = S_{\text{dS}}\). If \(\hbar\)
depended on \(S_{\text{dS}}\), it would be observer-dependent ---
different observers have different horizon areas (§2.1) ---
contradicting the universality of the emergent action scale. (All
sub-\(c\) observers share the same local physics and hence the same
\(\hbar\); §2.1.) This excludes \(A\), leaving the local boundary
quantities \(c\), \(G\), \(\epsilon\), and \(\kappa\). The surface
gravity \(\kappa\) is excluded because it varies between observers and
epochs while \(\hbar\) is observed to be universal across observers and
constant in time: a laboratory experiment measures the same \(\hbar\)
regardless of the observer's local cosmological-horizon parameters, and
for the cosmological horizon \(\kappa = cH\) is itself time-dependent
(\(\dot{H} \neq 0\)). The unique combination of \(c\), \(G\),
\(\epsilon\) with dimensions of action:

\[\hbar = \beta \frac{c^3 \epsilon^2}{G}\]

\textbf{Step 4: Thermal self-consistency.} The classical substratum
assigns the partition boundary a temperature
\(k_B T_{\text{cl}} = c^2 \epsilon^2 \kappa / (8\pi G)\) (§3.1),
containing no \(\hbar\). The emergent QFT of Part I lives on this
classical background, which has a bifurcate Killing horizon with surface
gravity \(\kappa\). Regularity of the Wick-rotated metric at the horizon
requires Euclidean period \(\beta = 2\pi c / \kappa\); any QFT on this
background --- including a lattice-regularized one --- must therefore be
periodic in imaginary time with the same period, giving a KMS state at
temperature \(T_Q = \hbar \kappa / (2\pi c k_B)\), with \(\hbar\)
unknown. This is a theorem \emph{within} the derived QFT, not an
external import. The two temperatures are computed independently ---
\(T_{\text{cl}}\) from the classical substratum alone (no QM), \(T_Q\)
from the emergent QFT alone (no classical substratum details) --- but
they refer to the \textbf{same physical degrees of freedom}: the
boundary modes \(V \times B\) across which \(H_{\text{int}}\) couples
the visible and hidden sectors. Step 2's boundary-only dependence shows
that the emergent QFT is uniquely determined (at leading order in
\(\tau_S/\tau_B\)) by these same \(V \times B\) dynamics. Temperature is
a state property; two complete descriptions of the same modes cannot
assign different temperatures without logical contradiction. The
matching \(T_{\text{cl}} = T_Q\) is therefore not an additional
assumption but a consequence of the boundary modes being the same
physical objects in both descriptions, with corrections at
\(\mathcal{O}(\tau_S/\tau_B) \sim 10^{-32}\):

\[\frac{c^2 \epsilon^2 \kappa}{8\pi G} = \frac{\hbar \kappa}{2\pi c}\]

\(\kappa\) cancels --- a non-trivial cross-check on Step 3, which
excluded \(\kappa\) from \(\hbar\) on observer-universality grounds. The
thermal matching here recovers that exclusion from independent physics:
\(\kappa\) enters both sides of the consistency equation, and only the
combinations of \(c\), \(G\), \(\epsilon\) predicted by Step 3 survive.
Solving:

\[\boxed{\hbar = \frac{c^3 \epsilon^2}{4G}}\]

The derivation is a gap equation: \(\epsilon\) is the free geometric
input, \(\hbar\) is the output. The match is non-trivial:
\(T_{\text{cl}}\) could have depended on volumetric hidden-sector data
(excluded by Step 2), or \(T_Q\) could have depended on the state (it
does not --- the KMS temperature is purely kinematic). That neither
pathology obtains makes the gap equation a genuine determination. The
non-circularity is structural: Part I establishes that a QFT emerges
with \emph{some} action scale \(\hbar\); §3 determines \emph{which}
\(\hbar\), using the independent classical temperature that Part I
neither requires nor produces.

\textbf{Non-circularity of the entropy density.} The claim
\(\eta = 1/\epsilon^2\) --- one mode per minimal cell --- may appear to
pre-bake the area-entropy proportionality. To see why it does not,
consider the counterfactual: suppose the classical entropy density were
\(\eta = \alpha/\epsilon^2\) for arbitrary \(\alpha > 0\). Then
\(T_{\text{cl}} = \alpha c^2 \epsilon^2 \kappa / (8\pi G k_B)\), and
thermal matching gives \(\hbar = \alpha c^3 \epsilon^2 / (4G)\). The
Bekenstein-Hawking entropy becomes \(S = A / (4\alpha^{-1} l_p^2)\),
which reproduces the observed \(1/4\) factor only for \(\alpha = 1\).
Thus \(\alpha = 1\) is a non-trivial constraint --- it says that each
minimal cell contributes exactly one boundary mode, not a fractional or
multiple contribution. This is the physical content of the \(q\)-gauge
theorem: the alphabet size \(q\) (the number of internal states per
cell) is observationally irrelevant, so what counts is the number of
\emph{cells}, not the number of \emph{states}. The counting
\(\eta = 1/\epsilon^2\) is not an assumption about mode density but a
consequence of \(\epsilon\) being defined as the minimal distinguishable
scale (Lemma 1) combined with the irrelevance of sub-cell structure.

The predictive content lies not in the gap equation alone but in its
consequences. The specific relationship \(\hbar = c^3\epsilon^2/(4G)\)
--- rather than any other function of \(c\), \(G\), \(\epsilon\) ---
produces: the Bekenstein-Hawking formula with the exact factor \(1/4\)
(§5), the CC dissolution with \(S_{\text{dS}}\) as the compression ratio
(§6.3), and the RVM parameter \(\nu_{\text{OI}}\) (§7.1). Any
alternative \(\hbar(\epsilon)\) would fail at least one of these checks.
The situation is analogous to deriving the Schwarzschild metric with
\(M\) as a free parameter: the derivation has genuine content (the
functional form) even though one input is not determined from within.

Jacobson's original derivation {[}1{]} uses \(\hbar\)-containing forms;
this paper does not --- it uses the classical identity
\(dE = (c^2 \kappa / 8\pi G)\,dA\) and the classical entropy density
\(\eta = 1/\epsilon^2\). The logical ordering differs from Jacobson's:
his argument derives \(G\) from \(\eta\) and \(\hbar\); the present
argument takes \(G\) as a parameter of the classical Hamiltonian (Lemma
3) and derives \(\hbar\) from \(G\) and \(\epsilon\). The two are
consistent --- substituting \(\eta = 1/\epsilon^2\) and
\(\hbar = c^3\epsilon^2/(4G)\) into Jacobson's formula
\(G_{\text{eff}} = c^4/(8\pi \eta \cdot \hbar c)\) recovers
\(G_{\text{eff}} = G\) --- but the roles of input and output are
reversed. The Gibbons-Hawking temperature {[}2{]}
\(k_B T_{\text{GH}} = \hbar \kappa / (2\pi c)\) is recovered as a
prediction, not used as an input. The KMS condition used above is
derived for continuum QFT on bifurcate Killing horizons; for the
lattice-regularized theory emerging from Part I, the relevant result is
that thermal periodicity of the Euclidean Green's functions is robust
against UV modifications of the dispersion relation, with corrections at
\(\mathcal{O}((\epsilon \kappa/c)^2)\) {[}3, 4{]}. For the cosmological
horizon, \(\epsilon \kappa / c = \epsilon H / c \sim 10^{-61}\), so
lattice corrections are \(\sim 10^{-122}\).

\hypertarget{gauge-fixing-and-the-dimensional-obstruction}{%
\subsubsection{3.3 Gauge-fixing and the dimensional
obstruction}\label{gauge-fixing-and-the-dimensional-obstruction}}

Steps 1--2 establish that \(\hbar\) depends only on boundary quantities;
Step 4 will fix its value via thermal matching. A prior question must be
addressed: does the transition-probability data \(\{T_{ij}(t)\}\)
determine the emergent Hamiltonian uniquely, or does residual gauge
freedom leave \(\hbar\) underdetermined? The following theorem,
completing the phase-locking argument of {[}Main, §3.1{]}, shows the
gauge freedom is physically trivial.

\textbf{Theorem (D-gauge completeness).} \emph{Let \(U(t) = e^{-iHt}\)
with non-degenerate eigenvalues and non-vanishing configuration-basis
overlaps. If \(|U'_{ij}(t)|^2 = |U_{ij}(t)|^2\) for all \(i, j, t\),
then \(H' = DHD^\dagger\) for a time-independent diagonal unitary
\(D\).}

\emph{Proof.} Write
\(U_{ij}(t) = \sum_a V_{ia} \, e^{-iE_a t} \, V_{ja}^*\) with
\(V_{ia} = \langle i | a \rangle\) and eigenvalues \(\{E_a\}\), and
similarly \(U'_{ij}(t) = \sum_a V'_{ia} \, e^{-iE'_a t} \, V'^*_{ja}\).

\emph{Step 1 (Eigenvalue recovery).} By the phase-locking argument of
{[}Main, §3.1{]}, \(|U_{ij}(t)|^2 = |U'_{ij}(t)|^2\) for all \(t\)
implies, via Fourier analysis, that the frequency sets \(\{E_a - E_b\}\)
and \(\{E'_a - E'_b\}\) coincide. Non-degeneracy of energy gaps then
gives \(E'_a = E_{\sigma(a)} + E_0\) for some permutation \(\sigma\) and
global shift \(E_0\). The Fourier coefficients
\(a^{kl}_{ij} = V_{ik} V^*_{jk} V_{jl} V^*_{il}\) must match between
primed and unprimed, which (by the non-vanishing condition on overlaps)
forces \(\sigma = \text{id}\) up to relabeling. Thus
\(E'_a = E_a + E_0\).

\emph{Step 2 (Modulus recovery).} From the diagonal Fourier coefficients
\(a^{kl}_{ii} = |V_{ik}|^2 |V_{il}|^2\), the non-vanishing condition
gives \(|V'_{ia}| = |V_{ia}|\) for all \(i, a\).

\emph{Step 3 (Phase structure).} Write
\(V'_{ia} = V_{ia} \, e^{i\delta_{ia}}\). The off-diagonal Fourier
coefficients require:

\[V'_{ik} V'^*_{jk} V'_{jl} V'^*_{il} = V_{ik} V^*_{jk} V_{jl} V^*_{il}\]

Substituting and canceling moduli:

\[e^{i(\delta_{ik} - \delta_{jk} - \delta_{il} + \delta_{jl})} = 1 \quad \text{for all } i, j, k, l\]

This is the double-difference condition:
\(\delta_{ik} - \delta_{il} - \delta_{jk} + \delta_{jl} = 0 \pmod{2\pi}\).
The general solution on the integer lattice is
\(\delta_{ia} = \alpha_i + \beta_a\) for phases \(\{\alpha_i\}\),
\(\{\beta_a\}\). Thus \(V' = D_\alpha \, V \, D_\beta\) where
\(D_\alpha = \text{diag}(e^{i\alpha_i})\) and
\(D_\beta = \text{diag}(e^{i\beta_a})\). The \(D_\beta\) factor is
absorbed into the eigenvector phase convention (physically irrelevant),
giving \(H' = D_\alpha H D_\alpha^\dagger\) with \(D = D_\alpha\) a
time-independent diagonal unitary. \(\square\)

The D-gauge is physically trivial (basis rephasing). Cross-interval
transition probabilities fix the relative gauge up to a single global
phase; in the continuum limit this becomes an unobservable smooth
function \(e^{i\theta(t)}\).

\textbf{Dimensional obstruction.} The unitary \(U(t)\) is dimensionless;
\(\hbar\) enters only via
\(\hat{H} = i\hbar \, \partial_t U \cdot U^\dagger\). No dimensionless
data can fix a dimensionful constant --- Step 4's thermal
self-consistency provides the necessary dimensionful input.

\textbf{Remark (phase recovery under discrete sampling).} The spatial
lattice discreteness does not discretize time: the continuous-time
extension ({[}Main, §2.3{]}) provides \(T_{ij}(t)\) as a genuinely
continuous function of \(t\) via the Liouville marginalization of the
Hamiltonian flow (Lemma 3). The D-gauge theorem therefore applies to the
continuous object without aliasing corrections. Aliasing would arise
only if the observer sampled \(T_{ij}\) at discrete intervals, in which
case trans-Planckian frequencies could be misidentified --- but these
lie outside the emergent QFT's domain. The phase-recovery error is
controlled by the same \(\mathcal{O}(\tau_S/\tau_B) \sim 10^{-32}\)
suppression that governs the boundary-only dependence of §3.2.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{self-consistency-and-the-discreteness-scale}{%
\subsection{4. Self-Consistency and the Discreteness
Scale}\label{self-consistency-and-the-discreteness-scale}}

Rearranging \(\hbar = c^3 \epsilon^2 / (4G)\):

\[\epsilon^2 = \frac{4\hbar G}{c^3} = 4\,l_p^2\]

The framework yields \(\epsilon = 2\,l_p\) exactly. This is
algebraically equivalent to the formula for \(\hbar\): the framework
contains one free geometric parameter (\(\epsilon\)), and thermal
self-consistency establishes its relationship to \(\hbar\). The result
is a self-consistency condition constraining \(\epsilon\) to the Planck
regime, not an independent determination. If \(\epsilon^2 \ll l_p^2\),
sub-Planckian subcells would create a second trace-out, breaking the
single-valuedness of \(\hbar\). If \(\epsilon^2 \gg l_p^2\), the
emergent description would assign distinct quantum states to physically
identical configurations. The framework's predictive content comes from
the relationship \(\hbar = c^3 \epsilon^2/(4G)\), the Bekenstein-Hawking
formula, and the consequences that follow.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-bekenstein-hawking-entropy}{%
\subsection{5. The Bekenstein-Hawking
Entropy}\label{the-bekenstein-hawking-entropy}}

With \(\epsilon = 2\,l_p\), the number of independent modes on the
cosmological horizon is:

\[N_{\text{modes}} = \frac{A}{\epsilon^2} = \frac{A}{4\,l_p^2} = S_{\text{dS}}\]

where \(S_{\text{dS}} = A/(4\,l_p^2)\) is the Bekenstein-Hawking entropy
of the de Sitter horizon. The factor of \(1/4\) in the
Bekenstein-Hawking formula --- historically introduced as a
proportionality constant --- is here derived: each minimal cell of area
\(\epsilon^2 = 4\,l_p^2\) on the horizon contributes one unit of
entropy, and \(A / (4\,l_p^2) = A / \epsilon^2\) recovers the standard
formula exactly. The factor of \(4\) in \(\epsilon^2 = 4 l_p^2\) traces
directly to §3 Step 4's thermal matching: the ratio \(2\pi/8\pi = 1/4\)
between the Euclidean KMS period (numerator: thermal periodicity of QFT
on a Killing horizon) and Jacobson's Einstein-equation prefactor
(denominator: gravity's coupling in \(dE = (c^2\kappa/8\pi G)dA\)). The
Bekenstein-Hawking \(1/4\) is therefore not an arbitrary constant but
the consistency condition between gravitational and quantum thermal
physics. The same minimal-cell counting argument applies to any horizon
with area \(A\), giving \(S = A/(4 l_p^2)\) for black-hole horizons as
well; the recent observational confirmation of the area law on GW250114
{[}5{]} is consistent with this extension.

\textbf{Observational confirmation: GW250114.} The Bekenstein-Hawking
area law \(dA \geq 0\) has long been regarded as a theoretical
prediction of black-hole thermodynamics, but until 2025 the empirical
evidence came primarily from agreement between classical horizon
dynamics and semiclassical Hawking radiation predictions --- none of it
testing the \(1/4\) coefficient directly. The LIGO/Virgo/KAGRA
observation of GW250114 (September 2025) provides the first direct test
of the area law at the percent level: comparing the remnant horizon area
inferred from the ringdown to the sum of initial horizon areas inferred
from the inspiral, the area increase is consistent with the theorem at
\(99.999\%\) confidence. Because the framework's derivation fixes the
\(1/4\) coefficient as the exact ratio \(2\pi/8\pi\) between Euclidean
KMS periodicity (thermal physics) and Jacobson's Einstein-equation
prefactor (gravitational coupling), any measured deviation from
\(S = A/(4 l_p^2)\) would falsify this derivation --- not merely
constrain one free parameter. The GW250114 measurement is currently the
most stringent empirical test of the framework's gravitational
predictions, and the ongoing LVK observing runs will accumulate
additional events testing the same relation across a range of progenitor
masses and spins.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{dissolution-of-the-cosmological-constant-problem}{%
\subsection{6. Dissolution of the Cosmological Constant
Problem}\label{dissolution-of-the-cosmological-constant-problem}}

\hypertarget{the-two-levels-of-description}{%
\subsubsection{6.1 The two levels of
description}\label{the-two-levels-of-description}}

\textbf{From finite-dimensional QM to QFT.} Part I delivers QM on a
finite-dimensional Hilbert space. The \(N = A/\epsilon^2\) cells
discretize the visible sector; spatial locality of the classical
substratum restricts which cells interact on any given timescale. The
emergent Hilbert space decomposes as
\(\mathcal{H} = \bigotimes_k \mathcal{H}_k\) --- a lattice-regularized
QFT with UV cutoff at \(\epsilon = 2\,l_p\). The tensor product
structure follows from the spatial product structure of the classical
configuration space
\(\mathcal{C}_V = \mathcal{C}_1 \times \cdots \times \mathcal{C}_N\)
(one factor per cell), which is not required by Lemma 1 but is a
consequence of spatial locality in the classical substratum: each cell's
configuration is a local degree of freedom.

\textbf{Locality preservation theorem.} \emph{If the classical
Hamiltonian is spatially local (couples only neighbors), then the
emergent quantum Hamiltonian inherits spatial locality.}

\emph{Proof.} For infinitesimal \(dt\) and \(x' \neq x\):
\(T_{x,x'}(dt) = |\langle x'|\hat{H}|x\rangle|^2\,dt^2 + O(dt^3)\). If
\(x, x'\) differ at non-neighboring sites, spatial locality of the
classical dynamics gives \(T_{x,x'}(dt) = 0\) at leading order, hence
\(|\langle x'|\hat{H}|x\rangle|^2 = 0\). By the D-gauge theorem (§3.3),
the residual phase freedom is a diagonal unitary \(D\) that preserves
the vanishing of off-diagonal moduli, so
\(\langle x'|\hat{H}|x\rangle = 0\) in any gauge. The emergent
Hamiltonian has the form
\(\hat{H} = \sum_k \hat{h}_k + \sum_{\langle k,l\rangle} \hat{h}_{kl}\).
\(\square\)

\textbf{Scope of the emergent QFT.} The CC dissolution requires only
that the emergent description assigns zero-point energy
\(\sim \hbar\omega/2\) per mode with UV cutoff at \(\epsilon^{-1}\) ---
a property shared by any lattice QFT with local couplings, regardless of
gauge group or matter content. The framework's predictions in Part II
(\(\hbar\), \(S_{\text{dS}}\), \(\nu_{\text{OI}}\)) depend on partition
geometry, not on the emergent field content.

\textbf{Classical substratum} (what geometric measurements probe): The
horizon has classical thermal equilibrium. By the Friedmann equation,
\(\rho_{\text{crit}} = 3H^2 c^2/(8\pi G)\). No zero-point energy, no
discrepancy. Total vacuum energy density:
\(\rho \sim H^2/G \sim 10^{-9}\) J/m\(^3\) --- precisely observed.

\textbf{Emergent QFT} (what local quantum measurements probe):
Zero-point energy \(\frac{1}{2}\hbar\omega\) per mode summed to the
Planck cutoff gives \(\rho_{\text{QFT}} \sim 10^{113}\) J/m\(^3\) --- a
\(\sim 10^{122}\) discrepancy with the observed value {[}6, 7, 8{]}.

\hypertarget{why-gravity-sees-the-classical-value}{%
\subsubsection{6.2 Why gravity sees the classical
value}\label{why-gravity-sees-the-classical-value}}

Spacetime geometry is part of the classical substratum (Lemma 2): the
metric evolves via Einstein's equations \emph{before} the trace-out. The
stress-energy sourcing gravity is the classical stress-energy of the
total microstate, not the expectation value of an emergent quantum
operator. The semiclassical equation
\(G_{\mu\nu} = 8\pi G \langle \hat{T}_{\mu\nu} \rangle\) is itself an
emergent approximation; at the classical level, the gravitational field
equations never encounter the zero-point sum.

\textbf{Ontological commitment.} Classical spacetime is fundamental; QM
is emergent. This ordering is not a modeling choice but is forced by the
logical structure of the derivation through three independent
requirements:

\emph{(i) Definiteness.} The trace-out that produces QM is defined by a
\emph{specific} partition \(\Gamma = \Gamma_V \times \Gamma_H\). The
partition must be definite --- not in superposition, not subject to
quantum uncertainty --- for the marginalization integral ({[}Main,
§1.4{]}) to be well-defined. A partition in superposition would yield an
incoherent mixture of inequivalent quantum theories, not a single QM.
The metric at the partition boundary must therefore be classical.

\emph{(ii) Non-circularity.} The partition is defined by the causal
structure: \(\Gamma_H\) is the set of degrees of freedom beyond the
observer's causal reach (§2.1). Causal structure is determined by null
geodesics of the metric. If the metric were derived from QM, then the
derivation would require: QM \(\to\) metric \(\to\) causal structure
\(\to\) partition \(\to\) QM --- a circular dependency with no entry
point. The circle breaks only if the metric exists prior to the
partition.

\emph{(iii) Uniqueness of \(\hbar\).} The emergent action scale is
determined by the boundary geometry (§3.2). If the geometry were itself
quantum-mechanical, \(\hbar\) would depend on a quantum state --- but
\(\hbar\) is state-independent by observation. The geometry-first
ordering is the only one consistent with a universal \(\hbar\).

Any framework in which QM is logically prior to the metric must either
(a) treat the zero-point sum as a gravitational source --- producing the
\(10^{122}\) discrepancy requiring cancellation or fine-tuning --- or
(b) invoke an independent mechanism to decouple quantum vacuum energy
from gravity. In the present framework, the zero-point sum is an
artifact of the emergent description and never enters the stress-energy
tensor; no decoupling mechanism is needed.

\textbf{Cumulative evidence for the ordering.} The geometry-first
ordering produces eleven independent consequences that match
observation, none of which were designed into the starting point: (1)
the CC dissolution --- the observed vacuum energy sits at
\(\rho \sim H^2/G\), the classical baseline, without fine-tuning (§6.3);
(2) the Bekenstein-Hawking formula including the \(1/4\) factor, derived
from \(\epsilon = 2\,l_p\) (§5); (3) the value of \(\hbar\), determined
by partition geometry rather than treated as a free parameter (§3.2);
(4) the dark-sector concordance --- \textasciitilde95\% of
\(\rho_{\text{crit}}\) is invisible to QFT, matching the observed dark
sector (§7.2); (5) dark energy evolution with \(\nu_{\text{OI}}\) in the
range consistent with DESI data (§7.1); (6) the MOND acceleration scale
\(a_0 = cH/6\) and the baryonic Tully-Fisher relation, derived from
entropy displacement (§7.3); (7) the SM gauge group with its coupling
strengths, reproduced to \(< 0.1\%\) at \(M_Z\); (8) declining rotation
curves at high redshift, confirmed by Genzel et al.~(§7.3); (9)
\(\bar{\theta} = 0\) at all energy scales; (10) galaxy cluster masses
matched by the simple interpolation function (§7.3); (11) the Bullet
Cluster lensing morphology reproduced by frozen boundary entropy (§7.3).
Quantum-first orderings produce the \(10^{122}\) discrepancy, have no
natural account of the dark sector, and treat \(\hbar\) and the \(1/4\)
factor as unexplained inputs. The cumulative case is not a single
tiebreaker but a pattern: every quantitative output of the
geometry-first ordering is consistent with observation; the central
quantitative output of the quantum-first ordering is the worst
prediction in physics.

\textbf{The classical vacuum energy scale.} The framework does not
explain \emph{why} \(\rho \sim H^2/G\); this is set by initial
conditions via the Friedmann equation. What it does is reclassify the CC
puzzle: from quantum vacuum cancellation to classical initial
conditions. The two problems are not equivalent. The QFT CC problem
requires a cancellation between independent contributions --- the bare
cosmological constant and the zero-point energies of every field ---
accurate to 122 decimal places, re-tuned at each order in perturbation
theory. The initial-conditions question asks why a single parameter
(\(\Lambda\), or equivalently \(H\)) takes its observed value; the scale
\(\rho \sim H^2/G\) is the natural energy density of classical
cosmology, not a fine-tuned cancellation between unrelated quantities.
The framework eliminates the 122-digit cancellation entirely; the
remaining question --- why \(H\) takes its value --- is a standard
cosmological problem addressable by inflationary dynamics or other
initial-conditions mechanisms.

\textbf{Quantum corrections to gravity.} The emergent quantum
description does feed back into gravitational dynamics through
state-level quantities: §7.1 derives dark energy evolution because the
emergent vacuum energy depends on \(H\) through the hidden sector's
volume. What the framework excludes is the zero-point sum as a
gravitational source. The zero-point energy is the ground-state
eigenvalue of the emergent Hamiltonian \(\hat{H}_{\text{eff}}\), which
acts on \(\mathcal{H}_V\). But gravity is sourced by the classical
Hamiltonian \(H_{\text{tot}}\), which acts on the full phase space
\(\Gamma\) --- and \(\hat{H}_{\text{eff}}\) and \(H_{\text{tot}}\) are
not the same operator. The eigenvalues of \(\hat{H}_{\text{eff}}\)
(including the zero-point energy) are properties of the \emph{emergent
description}; the energy that \(H_{\text{tot}}\) assigns to the
corresponding classical microstates is the \emph{classical vacuum
energy} \(\rho \sim H^2/G\). The Equivalence Principle applies to
\(H_{\text{tot}}\): everything that contributes to \(H_{\text{tot}}\)
gravitates. The zero-point sum does not contribute to
\(H_{\text{tot}}\); it appears when the classical transition
probabilities are reorganized into the Hilbert space formalism ({[}Main,
§3.1{]}), not when the classical dynamics is evolved. The
structural/volumetric distinction of §3.2 captures this: \(\hbar\)
(structural, a property of the correspondence) does not gravitate; the
vacuum energy (volumetric, state-dependent, a property of
\(H_{\text{tot}}\)) does, but at the classical scale \(\rho \sim H^2/G\)
rather than \(\rho \sim M_{\text{Pl}}^4\).

\hypertarget{the-structural-origin-of-the-10122-discrepancy}{%
\subsubsection{\texorpdfstring{6.3 The structural origin of the
\(10^{122}\)
discrepancy}{6.3 The structural origin of the 10\^{}\{122\} discrepancy}}\label{the-structural-origin-of-the-10122-discrepancy}}

\(\rho_{\text{QFT}} / \rho_{\text{classical}} \sim M_{\text{Pl}}^4 / (H^2/G) = S_{\text{dS}} = A/(4\,l_p^2)\)
--- the Bekenstein-Hawking entropy, now identified as the information
compression ratio of the trace-out. The ``worst prediction in physics''
is the entropy of the observer's blind spot --- a category error, not a
fine-tuning failure. Wolpert's limits of inference {[}9{]} confirm this
cannot be resolved from within: both geometric and quantum measurements
are faithful to their respective descriptions, and no embedded observer
can determine which is more fundamental. This is not a prediction
awaiting future data --- the observed vacuum energy sits at the
classical geometric scale \(\rho \sim H^2/G\), the value the framework
expects.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{predictions-and-tests}{%
\subsection{7. Predictions and Tests}\label{predictions-and-tests}}

\hypertarget{dark-energy-evolution-in-running-vacuum-form}{%
\subsubsection{7.1 Dark energy evolution in running vacuum
form}\label{dark-energy-evolution-in-running-vacuum-form}}

\(\hbar\) is \(H\)-independent (§3.2), but the emergent vacuum energy is
a state-level quantity depending on the hidden sector's volume.
Expanding \(\Lambda_{\text{eff}}(H)\):

\[\Lambda_{\text{eff}}(H) = \Lambda_0 + \frac{3\nu_{\text{OI}}}{8\pi G}(H^2 - H_0^2) + \mathcal{O}((H^2 - H_0^2)^2)\]

This is the Running Vacuum Model {[}10, 11{]}. The \(S_{\text{dS}}\)
horizon modes span frequencies \(\omega_{\text{IR}} = H\) to
\(\omega_{\text{UV}} = c/\epsilon\). The spectral density of boundary
entropy is uniform in \(\ln\omega\): each ε² boundary cell is one degree
of freedom that couples to the bulk at all frequencies (time-translation
invariance of the wave equation), and each mode carries
\(\mathcal{O}(1)\) entropy bit (quantum regime:
\(\omega \gg T_{\text{dS}}\) for all modes). The number of independent
spectral channels is \(\mathcal{N}(H) = \ln(c/\epsilon H)\). Each
carries fraction \(1/\mathcal{N}\) of the total gravitating vacuum
energy, so \(\rho_\Lambda \propto 1/\mathcal{N}\). Taylor-expanding:

\[\boxed{\nu_{\text{OI}} = \frac{\Omega_\Lambda}{2\,\ln(c/\epsilon H_0)} = (2.45 \pm 0.04) \times 10^{-3}}\]

With \(\epsilon = 2\,l_p\), \(H_0 = 67.4\) km/s/Mpc,
\(\Omega_\Lambda = 0.685\). The uncertainty has two components:
parametric (\(\pm 0.03 \times 10^{-3}\) from \(H_0\) and
\(\Omega_\Lambda\), at \(1.0\%\)) and systematic
(\(\pm 0.04 \times 10^{-3}\) from \(\mathcal{O}(1/\mathcal{N})\) edge
corrections, at \(1.4\%\)), combined \(\pm 1.8\%\). Independently
testable ratio:
\(\nu_{\text{OI}} / \Omega_\Lambda = 0.00358 \pm 0.00003\).

\textbf{Why the spectral density is uniform.} The conformal assumption
(\(\alpha = 0\), uniform spectral density in \(\ln\omega\)) is not a
modeling choice --- it follows from the lattice structure. Three
independent arguments converge: (i) \emph{Time-translation invariance.}
The wave equation is autonomous --- the coupling between a boundary cell
and the bulk does not prefer any frequency. Each cell contributes
equally to every frequency decade. (ii) \emph{Entropy saturation.} All
boundary modes have
\(\omega \gg T_{\text{dS}} = \hbar H/(2\pi c k_B) \sim 10^{-33}\) eV
(the de Sitter temperature). In this quantum regime, each mode carries
\(\mathcal{O}(1)\) entropy bit, independent of frequency. The total
entropy \(S_{\text{dS}} = A/\epsilon^2\) distributes uniformly across
\(\mathcal{N} = 140\) frequency decades: \(S_{\text{dS}}/\mathcal{N}\)
per decade. (iii) \emph{Geometric mode counting.} The boundary modes are
geometric (\(A/\epsilon^2\) cells), not field-theoretic. The SM field
content --- which fields are active at each energy scale --- affects the
internal structure of each channel but not the channel count. The mode
count per decade is set by the lattice geometry, which is
frequency-independent.

The leading correction to uniformity is
\(\mathcal{O}(1/\mathcal{N}) \approx 0.7\%\), arising from edge effects
at \(\omega \sim H\) and \(\omega \sim c/\epsilon\). For a spectral
density \(\propto \omega^\alpha\) with \(\alpha \neq 0\), the effective
\(\mathcal{N}\) changes dramatically: \(\alpha = +0.1\) gives
\(\mathcal{N} \sim 10^7\) (ν undetectable), \(\alpha = -0.1\) gives
\(\mathcal{N} \sim 10\) (ν \(\sim 0.03\), ruled out). The lattice
arguments above constrain \(\alpha = 0\) to within
\(\mathcal{O}(1/\mathcal{N})\), reducing the systematic from a factor of
\(\sim 3\) (the range \(1\)--\(5 \times 10^{-3}\)) to \(\pm 1.4\%\).

DESI data {[}12, 13{]} report evolving dark energy at
\(2.8\sigma\)--\(4.2\sigma\). RVM fits {[}14{]} find
\(\nu \sim \mathcal{O}(10^{-3})\) with \(\nu = 0\) disfavored at
\(2.7\sigma\)--\(3.1\sigma\), consistent with the prediction. The most
recent multi-model analysis {[}14{]} of DESI DR2 + Planck PR4 +
supernovae explicitly tests several RVM variants and finds the ``flipped
RVM'' and ``RVM with threshold'' point to significant evidence of
dynamical dark energy at a level comparable to the more flexible
\(w_0w_a\)CDM parameterization --- i.e., DESI now favors the specific
functional form OI predicts, not merely ``some kind of dynamical dark
energy.'' DESI Year 5 (\textasciitilde1\% precision on \(\nu\)) provides
the decisive test.

\textbf{Coupled neutrino test.} The same dataset that constrains
\(\nu_{\text{OI}}\) also constrains the neutrino sector. The DESI DR2 +
CMB analysis {[}15{]} finds \(\Sigma m_\nu < 0.0642\) eV (95\% CL)
assuming \(\Lambda\)CDM, prefers normal mass ordering, and bounds the
lightest neutrino mass at \(m_l < 0.023\) eV --- all three results match
the OI predictions of normal ordering with \(\Sigma m_\nu\) near the
oscillation minimum of 0.059 eV. The same analysis finds a \(3\sigma\)
tension with the lower oscillation limit assuming \(\Lambda\)CDM,
interpreted as ``a hint of new physics not necessarily related to
neutrinos'' --- exactly the structural mismatch that the RVM dark energy
resolves. JUNO first results {[}16, 17{]} confirm the world-leading
precision on \(\sin^2\theta_{12} = 0.3092 \pm 0.0087\) and
\(\delta m^2 = (7.48 \pm 0.10) \times 10^{-5}\) eV\(^2\), with the
solar/reactor 1.5\(\sigma\) tension persisting. The neutrino sector
predictions and the RVM dark energy prediction are coupled in the
framework --- both follow from the same lattice structure --- and the
joint observational support is therefore not independent confirmation of
separate effects but a single empirical pattern consistent with the
framework's central mechanism.

\hypertarget{the-dark-sector-corollary}{%
\subsubsection{7.2 The dark-sector
corollary}\label{the-dark-sector-corollary}}

The trace-out that produces QM has an automatic gravitational
consequence.

\textbf{Corollary (Invisible gravitational budget).} \emph{Under Lemmas
1--3 and conditions (C1)--(C3), the cosmological trace-out that produces
quantum mechanics simultaneously renders \textasciitilde95\% of the
universe's gravitational budget invisible to the emergent QFT.}

\emph{Proof.} Three steps, each using only results already established.

\emph{(i) Total effective density.} The horizon carries
\(S_{\text{dS}} = A/\epsilon^2\) modes (§5). The classical temperature
is \(T_{\text{cl}} = c^3 \epsilon^2 H / (8\pi G k_B)\) (§3.1), computed
entirely within the classical substratum with no reference to \(\hbar\)
or the emergent quantum description. The total thermal energy is
\(E_s = S_{\text{dS}} \cdot k_B T_{\text{cl}}\). By the shell theorem,
distributing over \(V_H = 4\pi c^3/(3H^3)\):

\[\rho_s = \frac{E_s}{V_H} = \frac{A}{\epsilon^2} \cdot \frac{c^3 \epsilon^2 H}{8\pi G} \cdot \frac{3H^3}{4\pi c^3} = \frac{3c^2 H^2}{8\pi G} = \rho_{\text{crit}}\]

This identity, which Verlinde {[}18{]} assumes, is here derived.

Four clarifications are essential for what follows.

\emph{Classical, not emergent.} The calculation above uses only
pre-trace-out quantities: the number of boundary modes
(\(A/\epsilon^2\)), the classical temperature (\(T_{\text{cl}}\)), and
the Hubble volume (\(V_H\)). No step invokes the emergent quantum
description. This is what distinguishes \(\rho_s\) from the QFT
zero-point energy \(\rho_{\text{QFT}} \sim M_{\text{Pl}}^4\): the latter
is computed as \(\sum \frac{1}{2}\hbar\omega\) over modes of the
\emph{emergent} quantum field theory --- a quantity that exists only
after the trace-out. The framework denies that \(\rho_{\text{QFT}}\)
gravitates (§6.2), because it is an artifact of the emergent
description. The boundary entropy's thermal energy, by contrast, is a
classical quantity that predates the trace-out and gravitates at the
classical level. That both involve \(\hbar\) numerically (because
\(T_{\text{cl}} = T_{\text{GH}}\)) is a consequence of the gap
equation's self-consistency, not a dependence on QM. The ratio
\(\rho_{\text{QFT}} / \rho_s \sim S_{\text{dS}} \sim 10^{122}\) --- the
same compression ratio identified in §6.3 --- confirming that the two
quantities occupy different levels of the framework's hierarchy.

\emph{Physical locus.} The entropy modes are physically localized on the
horizon --- they are the hidden-sector boundary degrees of freedom of
§3.2. The uniform-medium description is a gravitational equivalence
(shell theorem), not a claim about physical delocalization. No entropy
occupies the bulk.

\emph{Why the equivalence is useful.} When matter is present, it locally
deforms the horizon geometry, redistributing the boundary entropy. The
gravitational effect of this redistribution on bulk observers is most
easily computed using the uniform-medium equivalence: the deformed
boundary is equivalent to a non-uniform effective bulk source.

\emph{Algebraic status.} The coincidence \(\rho_s = \rho_{\text{crit}}\)
is an algebraic consequence of the Friedmann equation and the
self-consistency relation \(\hbar = c^3\epsilon^2/(4G)\); it is not an
independent physical input.

\emph{(ii) Invisibility.} The argument has two logically distinct
components.

\emph{(ii-a) Construction-level invisibility (from Part I alone).} The
boundary modes comprising \(S_{\text{dS}}\) are hidden-sector degrees of
freedom --- the variables traced out to produce the quantum description.
By the trace-out structure, no hidden-sector degree of freedom maps to
an operator on \(\mathcal{H}_V\). This is a theorem of Part I,
independent of any gravitational content: the hidden sector is invisible
to the emergent description by construction.

\emph{(ii-b) Gravitational consequence (from the ontological ordering of
§6.2).} The physical import --- that the boundary entropy's
gravitational influence appears as additional gravity with no source in
\(\langle \hat{T}_{\mu\nu} \rangle\) --- requires the geometry-first
ordering established in §6.2: the metric exists at the pre-trace-out
level, and the stress-energy sourcing gravity is the classical
stress-energy, not the emergent quantum expectation value. Under this
ordering, the gravitational interaction between bulk matter and boundary
entropy is mediated by the metric at the classical level, modifying the
geometry but contributing no term to the emergent
\(\langle \hat{T}_{\mu\nu} \rangle\). To the embedded observer, this
appears as additional gravity with no particle-sector source.

The corollary's predictions are therefore conditional on the same
geometry-first ordering that dissolves the CC problem in §6.2. This is
not an additional assumption: the ordering is already required by Part
II's logic (the partition is defined by null geodesics, so the metric
must exist prior to the partition). What the corollary provides is a
second, qualitatively distinct \emph{consequence} of that ordering.

\emph{(iii) Budget.} The emergent QFT accounts for the baryonic sector
(\textasciitilde5\% of \(\rho_{\text{crit}}\)). The remaining
\textasciitilde95\% is boundary entropy --- gravitationally active
(because \(\rho_s\) is a classical quantity computed before the
trace-out, unlike \(\rho_{\text{QFT}} \sim M_{\text{Pl}}^4\) which is an
artifact of the emergent description) but invisible to the emergent QFT.
This matches the observed composition, in which \textasciitilde95\% of
the gravitational content has no source in particle physics. \(\square\)

\textbf{Observational status of particle dark matter searches.} The
corollary's prediction is that no particle dark matter exists --- the
dark sector is the trace-out, not a missing particle species. The
strongest current constraint is the LZ Collaboration's 417-day analysis
(December 2025) {[}19{]}, which is the most sensitive WIMP search ever
conducted and reports null results across the 3--9 GeV range, extending
world-leading constraints from XENONnT and PandaX-4T to lower mass. The
continued failure to detect particle dark matter across multiple decades
of increasingly sensitive experiments --- summarized in the literature
as ``the waning of the WIMP'' --- is consistent with the corollary's
prediction. Definitive falsification of the corollary requires positive
detection of a particle species accounting for the structured dark
sector, which has not occurred.

\textbf{Memory effects and P-indivisibility.} Matter locally deforms the
horizon geometry, redistributing the boundary entropy. This displacement
persists because \(\tau_B \gtrsim 1/H\) (condition C2): the
hidden-sector modes carrying the redistributed entropy do not thermalize
back to the unperturbed configuration on sub-Hubble timescales. Verlinde
{[}18{]} identifies this as a ``memory effect'' from the
non-thermalization of de Sitter states at sub-Hubble scales. In the
present framework, this non-thermalization is not assumed but derived:
it is P-indivisibility operating in the gravitational sector, with (C2)
ensuring that entropy displacements persist without thermal overwriting.

The corollary is a consistency check for the total dark sector budget,
not a derivation of its internal structure. The uniform component of the
boundary entropy corresponds to dark energy (§6); the structured
component --- the matter-induced redistribution of boundary entropy
producing local gravitational effects resembling dark matter --- is
derived in §7.3. The ratio
\(\rho_{\text{QFT}} / \rho_s \sim S_{\text{dS}} \sim 10^{122}\) confirms
the parallel with §6.3: the information compression ratio and the
gravitational occlusion fraction are two aspects of a single phenomenon.

\textbf{Falsifiability.} The corollary would be falsified by detection
of dark matter particles accounting for the full \textasciitilde27\%
structured dark sector, or by evidence that the dark sector's
gravitational effects are unrelated to the boundary geometry.

\hypertarget{dark-matter-from-entropy-displacement}{%
\subsubsection{7.3 Dark matter from entropy
displacement}\label{dark-matter-from-entropy-displacement}}

The total dark sector budget is established by §7.2. This section
derives the structured component from the entropy displacement induced
by baryonic matter.

\textbf{Step 1: Entropy displacement (derived).} Baryonic matter \(M_B\)
is coupled to the boundary entropy through \(H_{VB}\) (condition C1).
The boundary is a thermal reservoir at
\(T_{\text{dS}} = \hbar H / (2\pi c k_B)\) (derived, §3.2). For a static
matter configuration in gravitational equilibrium --- a quasi-static
process --- the Clausius relation gives the entropy displacement:

\[\Delta S = \frac{M_B c^2}{k_B T_{\text{dS}}} = \frac{2\pi M_B c^3}{\hbar H}\]

This follows from the generalized second law (Theorem A.6) applied to
the matter-boundary system: the boundary's entropy reduction is \(Q/T\)
for a reversible process, where \(Q = M_B c^2\) and
\(T = T_{\text{dS}}\). For dynamic configurations, the Clausius
inequality \(\Delta S \geq Q/T\) only strengthens the displacement.

\textbf{Step 2: The critical acceleration (derived).} The displaced
entropy redistributes over the volume between the matter and the
horizon. The Jacobson mechanism converts the entropy gradient into
additional radial acceleration. The crossover between Newtonian gravity
and the entropic effect, accounting for the volume-to-surface ratio of
spherical redistribution in de Sitter (\(1/6\)), gives:

\[a_0 = \frac{cH}{6} \approx 1.2 \times 10^{-10} \; \text{m/s}^2\]

This is the MOND critical acceleration --- derived, not fitted.
(Milgrom's empirical value:
\(a_0^{\text{obs}} \approx 1.2 \times 10^{-10}\) m/s².)

\textbf{Step 3: The deep-MOND regime.} For baryonic acceleration
\(\ll a_0\), the displaced entropy dominates. The Jacobson mechanism
converts the entropy gradient into curvature, giving
\(g_{\text{total}} = \sqrt{g_B \cdot a_0}\) and rotation velocity:

\[v^2 = r \cdot g_{\text{total}} = \sqrt{GM_B \cdot cH/6} = \text{constant} \qquad \Longrightarrow \qquad v^4 = GM_B \cdot cH/6\]

This is the baryonic Tully-Fisher relation. Flat rotation curves emerge
with no dark matter particles and no free parameters beyond the baryonic
mass.

\textbf{Step 4: Interpolation constraints.} The full interpolation
between the Newtonian regime (\(g_B \gg a_0\)) and the deep-MOND regime
(\(g_B \ll a_0\)) is constrained by the framework but not uniquely
determined. The entropy displacement defines a potential \(\Psi_D\) with
\(g_D = -\nabla \Psi_D\), and the framework requires:

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\item
  \emph{No free parameters.} \(g_D\) depends only on \(g_B\) and
  \(a_0 = cH/6\) --- no halo-specific parameters, no fitting constants.
\item
  \emph{Conservative.} \(\nabla \times g_D = 0\), since \(g_D\) derives
  from the entropy potential.
\item
  \emph{Monotonic.} \(\partial g_D / \partial g_B > 0\) --- more
  baryonic mass produces more entropy displacement, hence more dark
  matter effect.
\item
  \emph{Correct limits.} \(g_D \to 0\) for \(g_B \gg a_0\) (Newtonian);
  \(g_D \to \sqrt{g_B \cdot a_0}\) for \(g_B \ll a_0\) (MOND).
\item
  \emph{Analytic.} The entropy redistribution is a smooth function of
  the coupling structure --- no discontinuities or phase transitions in
  the interpolation.
\end{enumerate}

These constraints narrow the interpolation to a one-parameter family
indexed by the transition steepness --- a property of the particular
bijection \(\varphi\), in the same category as fermion masses
(solution-specific rather than structural). The simplest member
satisfying all five constraints:

\[g_D = \frac{g_B}{2}\left(\sqrt{1 + \frac{4a_0}{g_B}} - 1\right)\]

which is equivalent to \(g_{\text{total}} = g_B \cdot \nu(g_B/a_0)\)
with \(\nu(y) = (1 + \sqrt{1 + 4/y})/2\) --- the ``simple'' MOND
interpolation function corresponding to \(\mu(x) = x/(1+x)\). This form
matches the empirical radial acceleration relation (McGaugh et
al.~{[}20{]}) to \(< 0.1\) dex across all galaxy-relevant accelerations.

\textbf{Epistemic status.} Steps 1--3 are Tier 2: the displacement
formula follows from the Clausius relation applied to the boundary
reservoir (§3.2, A.6), the geometric factor from spherical
redistribution in de Sitter, and \(a_0 = cH/6\) and
\(v^4 = GM_B \cdot cH/6\) are parameter-free consequences. The
interpolation constraints (i)--(v) are Tier 2; the transition steepness
is Tier 3 (solution-specific).

\textbf{Redshift evolution.} Since \(a_0(z) = cH(z)/6\) and
\(H(z) = H_0\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}\), the critical
acceleration increases with redshift: \(a_0(z=1) = 1.79\,a_0(0)\),
\(a_0(z=2) = 3.03\,a_0(0)\), \(a_0(z=3) = 4.57\,a_0(0)\). The MOND
crossover radius \(r_M = \sqrt{6GM_B/(cH)}\) correspondingly shrinks ---
for a \(10^{10}\,M_\odot\) galaxy, \(r_M = 9.0\) kpc at \(z = 0\) but
\(5.2\) kpc at \(z = 2\) and \(3.1\) kpc at \(z = 5\). Galaxies with
effective radii comparable to or exceeding \(r_M\) show flat rotation
curves (the familiar local behavior); galaxies with
\(r_{\text{eff}} < r_M\) show declining, Keplerian-like rotation curves.
Since \(r_M\) shrinks with increasing \(z\), high-redshift galaxies are
systematically more baryon-dominated --- their rotation curves decline
at smaller radii.

This prediction is borne out by the data. Genzel et al.~{[}21, 22{]}
report stacked rotation curves for massive star-forming galaxies at
\(z = 0.9\text{–}2.4\) showing a declining outer velocity at
\(> 3\sigma\) significance relative to local spirals --- consistent with
the OI prediction. In ΛCDM, flat rotation curves from NFW halos are
expected at all redshifts; explaining the Genzel data requires
anomalously low halo concentrations or baryon-dominated inner regions.
The cleanest local test is Jiao et al.~{[}23{]}, who detect for the
first time a Keplerian decline in the Milky Way's own rotation curve
from \(\sim 19\) to \(\sim 26.5\) kpc using Gaia DR3 kinematics, with
the flat rotation curve hypothesis rejected at \(3\sigma\). Our own
galaxy is the strongest single piece of evidence for the
\(H(z)\)-dependent crossover prediction at \(z = 0\), where the OI
framework predicts the crossover at \(r_M \sim 17\) kpc for the Milky
Way's baryonic mass --- essentially where Jiao et al.~observe the
transition.

\textbf{Tully-Fisher tension and resolution.} McGaugh et al.~(2024),
analyzing Nestor Shachar et al.~(2023) JWST kinematics, report no
evolution in the Tully-Fisher relation out to \(z \sim 2.5\), seemingly
contradicting \(v^4 \propto H(z)\). However, this comparison uses
\emph{stellar} mass \(M_*\), not baryonic mass
\(M_B = M_*/(1-f_{\text{gas}})\). At high redshift, gas fractions are
large (\(f_{\text{gas}} \sim 50\text{–}70\%\) at \(z \sim 2\); Tacconi
et al.~2018). The OI-predicted shift in the baryonic TF
(\(\Delta\log M_B = \log(H_0/H(z)) = -0.48\) dex at \(z = 2\)) is almost
exactly compensated by the gas mass not included in \(M_*\): the
cancellation requires \(f_{\text{gas}} = 1 - H_0/H(z)\), which gives
44\% at \(z = 1\), 58\% at \(z = 1.5\), and 67\% at \(z = 2\) ---
squarely in the observed range. The apparent ``no evolution'' in the
stellar TF is therefore \emph{predicted} by the OI framework: the
dynamical shift (higher \(a_0\)) and the compositional shift (higher gas
fraction) cancel when only stellar mass is measured. The definitive test
is the \emph{baryonic} TF at \(z > 1\) with reliable ALMA gas masses:
the OI framework predicts a normalization shift of \(-0.48\) dex at
\(z = 2\) at fixed \(v_{\text{flat}}\).

\textbf{Observational predictions.} Three parameter-free, quantitative
tests:

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\item
  \emph{Rotation curve evolution.} The outer rotation curve slope
  (measured as \(\Delta v / v\) between \(r_{\text{turn}}\) and
  \(2r_{\text{turn}}\)) becomes more negative (more Keplerian) with
  increasing \(z\), following \(a_0(z) = cH(z)/6\). For a
  \(10^{10.8}\,M_\odot\) exponential disk with \(r_{\text{eff}} = 5\)
  kpc: the outer slope changes from \(-0.10\) at \(z = 0\) to \(-0.12\)
  at \(z = 2\) to \(-0.13\) at \(z = 5\). Particle dark matter predicts
  no evolution.
\item
  \emph{Tully-Fisher evolution.} The baryonic Tully-Fisher relation
  evolves: \(v_{\text{flat}}^4 = GM_B \cdot cH(z)/6\). At fixed baryonic
  mass, \(v_{\text{flat}} \propto H(z)^{1/4}\): 7\% higher at
  \(z = 0.5\), 16\% higher at \(z = 1\), 32\% higher at \(z = 2\).
  Equivalently, at fixed \(v_{\text{flat}}\), the inferred baryonic mass
  at \(z = 2\) is \(H_0/H(z=2) = 0.33\times\) the local value ---
  galaxies appear to need less dark matter at high \(z\). Testable with
  JWST and ALMA kinematics combined with photometric baryonic mass
  estimates.
\item
  \emph{No halo-specific parameters.} In particle DM, each galaxy has a
  dark matter halo characterized by at least two free parameters
  (concentration and mass). In the OI framework, the dark matter effect
  at radius \(r\) depends only on the enclosed baryonic mass \(M_B(<r)\)
  and \(H(z)\) --- no halo-specific parameters. This predicts: the
  radial acceleration relation (RAR) at fixed \(g_B\) has zero intrinsic
  scatter (at fixed \(z\)), and the scatter does not correlate with halo
  properties. Current SPARC data {[}20{]} show remarkably small RAR
  scatter (\(\sim 0.13\) dex), consistent with OI but requiring
  fine-tuning in ΛCDM.
\end{enumerate}

\textbf{Falsifiability.} The entropy displacement account would be
falsified by: (a) detection of dark matter particles accounting for
\$\sim\$27\% of \(\rho_{\text{crit}}\); (b) high-redshift rotation
curves inconsistent with the predicted \(H(z)\) dependence ---
specifically, if galaxies at \(z > 2\) show flat rotation curves at
\(r > r_M(z)\); (c) baryonic Tully-Fisher violations exceeding baryonic
mass uncertainties; (d) RAR scatter correlating with halo properties at
fixed \(g_B\) and \(z\).

\textbf{Cluster-scale status.} Galaxy clusters present the most
challenging regime for MOND-like theories. For the Coma cluster
(\(M_B \approx 1.4 \times 10^{14}\,M_\odot\), \(R_{200} \approx 1.9\)
Mpc), the acceleration at the virial radius is \(g_B/a_0 \approx 0.05\)
--- deep in the MOND regime. With the simple interpolation function
above, the predicted velocity is
\(v = g_B \cdot \nu(g_B/a_0) \cdot R_{200} \approx 1260\) km/s; the
observed velocity dispersion implies \(\sim 1270\) km/s --- a match to
better than \(1\%\). For other rich clusters (Perseus, A2029, A1689),
the simple interpolation reduces the standard MOND mass shortfall from a
factor \(\sim 2\) to \(\sim 1.0\)--\(1.5\); the residual is within the
range of missing baryonic mass in the warm-hot intergalactic medium
(WHIM at \(T \sim 10^5\)--\(10^7\) K, difficult to detect in X-rays,
estimated to add 30--100\% to cluster baryonic mass). The deep-MOND
limit \(v^4 = GM_B \cdot a_0\) (Tier 2) is identical for all
interpolation functions; only the intermediate-acceleration behavior
(Tier 3) differs. Galaxy-scale predictions are unaffected: at
\(g_B/a_0 = 0.5\) (typical outer disk), different interpolation
functions agree to \(< 0.07\) dex --- well within the RAR scatter of
\(0.13\) dex.

\textbf{Colliding clusters (Bullet Cluster).} The Bullet Cluster (1E
0657-56) presents the most-cited evidence for particle dark matter:
after the collision, gravitational lensing peaks coincide with the
galaxy concentrations (\(\sim 15\%\) of baryonic mass), not with the
X-ray gas (\(\sim 85\%\) of baryonic mass) that was ram-pressure
stripped to the center. In a local MOND picture where \(g_D\) tracks the
instantaneous baryonic distribution, the lensing should peak where the
gas is --- contradicting observations. The OI framework resolves this
through the non-local character of entropy displacement. The dark
gravity arises from boundary entropy redistribution at the cosmological
horizon, and the horizon's response time is
\(\tau_{\text{relax}} \sim H^{-1} \approx 14\) Gyr. The collision
crossing time is
\(t_{\text{cross}} \sim 720\,\text{kpc}/4700\,\text{km/s} \approx 0.15\)
Gyr --- only \(1\%\) of the relaxation time. The entropy displacement is
therefore \emph{frozen} at the pre-collision configuration, centered on
the galaxy concentrations (which defined the gravitational potential
wells for gigayears before the collision). The gas moved during the
collision; the boundary entropy did not. This reproduces the observed
morphology: lensing peaks at the galaxy positions, gas peak between them
--- without invoking collisionless particles. The mechanism makes a
testable prediction: for very old post-collision systems
(\(t_{\text{since}} \gg 1\) Gyr), the dark gravity should gradually
relax toward the gas distribution on timescale \(\sim H^{-1}\). Particle
dark matter predicts no such relaxation.

\textbf{CMB acoustic peaks.} The CMB power spectrum provides the most
precise test of the dark matter paradigm: the odd/even peak height
asymmetry requires a non-oscillating gravitational source (CDM in ΛCDM)
that provides potential wells for the photon-baryon fluid to oscillate
in. The entropy displacement satisfies this requirement through the same
thermodynamic averaging that resolves the Bullet Cluster. The Clausius
relation \(\Delta S = \delta Q / T_H\) involves the \emph{net} heat
transfer to the horizon boundary. For an oscillating perturbation
(compression then rarefaction), the net entropy displacement per
acoustic cycle is zero --- the oscillating component averages out. Only
the secular (growing) component produces net displacement. The entropy
displacement perturbation therefore does not oscillate with the
photon-baryon fluid; it tracks the growing mode exclusively.

The entropy displacement perturbation then evolves identically to CDM in
the linear regime: (i) it has the same effective density
(\(\rho_E \approx \Omega_{\text{CDM}} \times \rho_{\text{crit}}\), from
the dark-sector corollary §7.2); (ii) it does not couple to photons (it
is a gravitational effect at the horizon, not a fluid component); (iii)
it grows under its own gravity from adiabatic initial conditions set by
inflation; (iv) its growth equation
\(\ddot{\delta}_E + 2H\dot{\delta}_E = 4\pi G \rho_{\text{eff}} \delta_E\)
has the same source term as CDM. The CMB power spectrum --- which
depends on \(\Omega_b/\Omega_m\) (peak ratios), the growth of
perturbations before recombination (peak heights), the sound horizon
\(r_s\) (peak positions), and the damping scale (high-\(\ell\)
suppression) --- is therefore predicted to match ΛCDM at linear order. A
definitive confirmation requires running a modified Boltzmann code
(CLASS or CAMB) with entropy displacement replacing CDM; the prediction
is agreement with Planck data to within current uncertainties.
\emph{Epistemic status: Tier 3 (derived consequence, pending numerical
verification).}

\hypertarget{summary-of-gr-side-predictions}{%
\subsubsection{7.4 Summary of GR-side
predictions}\label{summary-of-gr-side-predictions}}

The GR-side quantitative predictions of the framework are collected in
the table below. Each follows from partition geometry at the
cosmological horizon; no predictions are fit to the quantities they
predict.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Prediction
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Formula / value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Observed / status
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Match
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Emergent action scale & \(\hbar = c^3 (2 l_p)^2 / (4G)\) &
\(\hbar_{\text{obs}} = 1.055 \times 10^{-34}\) J·s & structural \\
Discreteness scale & \(\epsilon = 2\,l_p\) & Planck length &
structural \\
BH entropy coefficient & \(S_{BH} = A / (4 l_p^2)\) & GW250114 area law
& \(99.999\%\) \\
Cosmological constant & \(\rho_\Lambda \sim H^2/G\) & observed value &
no discrepancy \\
Dark energy coefficient & \(\nu_{\text{OI}} = 2.45 \times 10^{-3}\) &
DESI DR2 & \(2.8\)--\(4.2\sigma\) preference \\
Dark sector fraction & \(\Omega_{\text{dark}} \approx 95\%\) &
\(\Omega_{\text{CDM}} + \Omega_\Lambda \approx 95\%\) & consistent \\
MOND acceleration scale & \(a_0 = cH/6 \approx 1.2 \times 10^{-10}\)
m/s² & \(a_0^{\text{obs}} \approx 1.2 \times 10^{-10}\) m/s² & exact \\
Baryonic Tully-Fisher & \(v^4 = G M_B \cdot cH/6\) & SPARC data & within
\(0.13\) dex RAR scatter \\
Rotation curve crossover & \(r_M = \sqrt{6 G M_B/(cH)}\) & Milky Way:
\(\sim 17\) kpc & Jiao et al.~2023, \(3\sigma\) \\
High-\(z\) rotation curves & More baryon-dominated with \(z\) & Genzel
et al.~\(z = 0.9\)--\(2.4\) & consistent at \(> 3\sigma\) \\
\end{longtable}

The structural coefficient \(\nu_{\text{OI}} = 2.45 \times 10^{-3}\) is
the single most distinctive prediction: it is both specific (a number to
three decimal places) and testable by current and near-future surveys.
The falsification path is clear: if DESI Year 5 and successor surveys
rule out dynamical dark energy in running-vacuum form at this
coefficient, the framework's Part II is falsified as a gravitational
theory.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{discussion}{%
\subsection{8. Discussion}\label{discussion}}

This section discusses four interpretive consequences of the derivations
in §§2--7: the scope of the framework's claims (§8.1), the epistemic
status of the discreteness scale \(\epsilon = 2\,l_p\) (§8.2), the
distinction between vacuum-energy shifts that gravitate (Casimir forces,
particle masses, the Higgs vacuum shift) and absolute zero-point sums
that do not (§8.3), and the framework's assumptions, limitations, and
falsifiability conditions (§8.4). The foundational emergent QM result is
treated as established in {[}Main{]} and not re-derived here. The
cubic-lattice realization of the substratum --- which determines the
Standard Model structure via the wave equation --- is independent of the
present results and is developed in {[}SM{]}. The reconstruction
theorem, the substratum gauge group, and the synthesis claim that
quantum mechanics and general relativity descend as different
projections of the same \((S, \varphi)\) are developed in
{[}Substratum{]}.

\hypertarget{scope}{%
\subsubsection{8.1 Scope}\label{scope}}

The characterization theorem is a full equivalence: QM \(\iff\) embedded
observation under (C1)--(C3). The theorem does not identify which
physical systems satisfy the conditions (empirical) nor derive \(\hbar\)
for specific systems (requires Part II). Universality in our universe
follows from the shared causal partition defined by null geodesics.

\hypertarget{the-status-of-the-discreteness-scale}{%
\subsubsection{8.2 The status of the discreteness
scale}\label{the-status-of-the-discreteness-scale}}

\(\epsilon\) does not smuggle in a quantum assumption. Lemma 1 requires
finite-dimensionality, motivated by the classical fact that finite-area
boundaries admit finitely many modes. The result \(\epsilon = 2\,l_p\)
is a \emph{consequence} of self-consistency (§4). Holographic entropy
bounds {[}24{]} provide independent, non-quantum motivation.

\hypertarget{vacuum-energy-relative-effects-and-the-higgs-potential}{%
\subsubsection{8.3 Vacuum energy: relative effects and the Higgs
potential}\label{vacuum-energy-relative-effects-and-the-higgs-potential}}

\textbf{Relative effects.} Casimir forces and Lamb shifts are
predictions \emph{within} the emergent QFT --- relative effects
depending on energy differences between configurations. The framework
denies that the \emph{absolute} zero-point sum gravitates, because
gravity operates at the logically prior classical level (§6.2).

\textbf{The Higgs potential.} The strongest CC objection is not the
zero-point sum but the electroweak Higgs potential: symmetry breaking
shifts \(V(\phi)\) by \(\sim (200\;\text{GeV})^4\), exceeding the
observed \(\Lambda\) by \(\sim 55\) orders of magnitude. The framework's
response follows from its ontological ordering and the
structural/volumetric distinction of §3.2, applied as a precise
principle:

\emph{What gravitates:} Relative energy differences between
configurations of the emergent fields. An electron at rest vs.~no
electron differs by \(m_e c^2\) in the classical substratum --- there is
a classical microstate difference corresponding to this energy
difference. Relative differences appear in the classical stress-energy
tensor and gravitate. Casimir forces, Lamb shifts, and particle masses
all fall in this category.

\emph{What does not gravitate:} Absolute energy levels assigned to
configurations by the emergent description. The Higgs vacuum energy
shift \(V(\phi_{\text{min}})\) is the energy the emergent QFT assigns to
its ground state --- the configuration obtained after symmetry breaking.
But the ``ground state'' in the emergent description corresponds to a
class of classical microstates whose energy in the classical substratum
is \(\rho \sim H^2/G\), not \(\rho \sim (200\;\text{GeV})^4\). The shift
is a property of how the emergent description organizes the microstates
(the correspondence), not a property of the microstates themselves (the
classical dynamics).

The principle is uniform: the zero-point sum, the Higgs vacuum energy,
and the QCD condensate energy are all absolute properties of the
emergent ground state. The dissolution applies to each for the same
reason --- gravity operates at the classical level (§6.2), where none of
these quantities appear.

\textbf{Why QCD binding energy gravitates but the quantum vacuum does
not.} The proton mass is \textasciitilde99\% QCD binding energy --- a
purely quantum effect. If gravity couples to the classical substratum,
how does it ``know'' about masses generated by emergent quantum
dynamics? The answer is that QCD binding energy and vacuum zero-point
energy occupy different categories in the structural/volumetric
distinction of §3.2. Relative energies within the emergent QFT ---
energy differences between configurations --- correspond to differences
in the coupling structure of \(\varphi\), which the Jacobson derivation
reads as curvature. A proton vs.~three free quarks corresponds to two
distinct classes of classical microstates with different energies in the
classical substratum; the binding energy \(\sim 938\) MeV is a
\emph{relative} quantity that appears in \(H_{\text{tot}}\) and
gravitates. The absolute baseline --- the zero-point energy summed over
all modes --- corresponds to the total information capacity of the
hidden sector (\(S_{\text{dS}}\)), which is a property of the partition
geometry, not of the dynamics, and therefore does not source curvature.

\hypertarget{assumptions-limitations-and-falsifiability}{%
\subsubsection{8.4 Assumptions, limitations, and
falsifiability}\label{assumptions-limitations-and-falsifiability}}

\textbf{The theorem} requires Lemma 1 (independent in Part I;
consequence of geometry in Part II), the stochastic-quantum bridge
({[}Main, §3.1{]} and Appendix A; established by two independent
routes), and genericity conditions (non-degenerate spectrum,
non-vanishing overlaps), which hold for all but a measure-zero set. The
P-indivisibility proof uses finiteness via the recurrence argument
(\(\varphi^N = \text{id}\)). The deep-sector cardinality corollary
(§3.2) establishes that only \(\mathcal{C}_V\) and \(\mathcal{C}_B\)
need be finite --- both guaranteed by holographic entropy bounds
{[}24{]} independently of the de Sitter finite-dimensionality conjecture
--- while the deep sector \(\mathcal{C}_D\) may be infinite. The
accessible-timescale backflow lemma ({[}Main, §2.3{]}) provides a second
route to P-indivisibility that is independent of \(|\mathcal{C}_H|\).
The characterization theorem becomes an equivalence on observable
timescales (\(T_{\text{obs}} \ll \tau_B \sim 10^{17}\) s) --- physically
indistinguishable from a timeless identity for all experiments within
the age of the universe.

\textbf{The sharp partition and constraint structure.} Lemma 2's product
decomposition \(\Gamma = \Gamma_V \times \Gamma_H\) is the kinematic
phase space. In GR, the Hamiltonian and momentum constraints restrict
the physical phase space to a submanifold
\(\Gamma_{\text{phys}} \subset \Gamma_V \times \Gamma_H\) on which
interior and exterior data are correlated. P-indivisibility survives on
a constraint surface because restoring P-divisibility would require
decoupling the sectors entirely, violating the constraints; the
constraint surface thus enforces (C1) non-perturbatively. For systems
other than the cosmological horizon, the fidelity of the product
approximation determines the fidelity of the emergent quantum
description.

\textbf{Planck-scale ordering.} The classical-prior ontology may break
down at \(l_p\), where ``geometric'' and ``quantum'' lose operational
distinction. If a deeper theory unifies both, this framework describes
the effective macroscopic regime.

\textbf{Falsifiability.} The \emph{theorem} would be falsified by a
failure of both the stochastic-quantum correspondence and the
Stinespring construction for the class of processes generated here --- a
possibility excluded by the independent proofs in {[}Main, §3.1{]} and
Appendix A. The \emph{cosmological application} would be falsified by
definitive absence of dark energy evolution in RVM form at
\(\nu_{\text{OI}}\). The \emph{dark-sector concordance} (§7.2) would be
falsified by detection of dark matter particles accounting for the full
structured dark sector, or by evidence that the dark sector's
gravitational effects are unrelated to the cosmological boundary
geometry.

\textbf{A near-term mechanism-side test.} A more tractable falsification
target is provided by the recently developed operational distinction
between classical and quantum memory in non-Markovian processes {[}25,
26{]}. Some non-Markovian quantum dynamics can be simulated using only
classical stored information, while others require genuinely quantum
memory in the environment, with the dividing line set by entanglement
structure of the channel's Choi state. The framework predicts that all
\emph{fundamental} non-Markovian dynamics in nature --- those arising
from the cosmological trace-out rather than from engineered quantum
baths --- should be classical-memory simulable, because the hidden
sublattice is by construction a classical substrate. A physically
realized fundamental open-system process whose memory provably exceeds
the classical bound would falsify the framework's foundational ontology
even without cosmological data. This is a near-term tractable test in a
literature where the experimental and theoretical tools are advancing
rapidly.

\textbf{Black-hole horizons under partition-relativity.} The framework's
mechanism is grounded in the cosmological horizon as the primary
partition for any sub-\(c\) observer in this universe (§2.1). A
black-hole horizon within the cosmological visible sector is a
\emph{local} causal boundary: it prevents an external observer from
probing the black-hole interior by direct geometric access, but it does
not redefine that observer's epistemic partition in the sense of
{[}Main, §1.4{]}. Matter falling into a distant black hole was already
inside the external observer's \(\Gamma_V\) before the merger and
remains inside it after; the trace-out that defines the external
observer's emergent quantum description is over the cosmological
horizon, not over the black-hole horizon. The framework therefore
predicts that gravitational wave signatures from binary black-hole
mergers follow standard general-relativistic ringdown morphology, with
no OI-specific echo, comb, or wall-reflection features. A four-event
coherent matched-filter stack on LVK O1--O4 strain data using a
dedicated discreteness-scale echo template (with the predicted
detector-frame delay \(\Delta t = (r_+/c)\ln(r_+/2 l_p)\) at the remnant
mass and spin) returned a clean null: stacked statistic \(-0.45\),
empirical \(p = 0.66\), with the previously hypothesized echo amplitude
excluded at \(5.9\sigma\). The framework's BH physics for external
observers is identical to standard GR plus the cosmological-scale
dark-sector effects of §7.2--§7.3.

\textbf{Nested and time-dependent partitions.} The primary partition for
any sub-\(c\) observer in this universe is the cosmological horizon
(§2.1). When additional causal boundaries are present --- most
naturally, a black hole horizon within the cosmological patch --- the
trace-out machinery extends to a nested formalism, addressed in Appendix
A. The nested formalism is a self-consistency exercise: it shows that
the framework's emergent description is well-defined when multiple
boundaries coexist, and that quantities like the generalized second law
and the Page curve emerge naturally. It is \emph{not} a claim that
external observers' emergent quantum descriptions acquire
BH-horizon-specific features; under partition-relativity (above), the
cosmological horizon remains the primary partition. Appendix A resolves
the formal nested-partition machinery in three layers.

\emph{(i) Consistency of nested trace-outs (Appendix A, Theorems
A.1--A.4).} Two procedures --- tracing out the trans-cosmological region
and the black hole interior simultaneously (direct) or sequentially
(first cosmological, then black hole) --- produce the same CPTP channel,
the same emergent Hamiltonian (up to D-gauge), and the same \(\hbar\).
The key result (Theorem A.2) is that the quantum partial trace within
the emergent description agrees with the classical marginalization on
the substratum: the emergent QM is self-consistent.

\emph{(ii) The generalized second law (Appendix A, Theorem A.6).} The
sum \(S_{\text{matter}} + S_{\text{BH}} + S_{\text{dS}}\) is
non-decreasing, proved from strong subadditivity and CPTP monotonicity
of the emergent quantum description. Any process that shrinks the black
hole (decreasing \(S_{\text{BH}}\)) transfers information to the visible
sector and cosmological boundary, compensating the decrease.

\emph{(iii) The Page curve (Appendix A, Theorems A.7--A.9).} For an
evaporating black hole, the entanglement entropy of the radiation
follows \(S_{\text{rad}}(t) = \min(n_R(t), n_B(t)) \cdot \ln q\) plus
exponentially small corrections, where \(n_B(t)\) is the shrinking
boundary mode count and \(n_R(t) = n_B(0) - n_B(t)\). Unitarity is
preserved (bijectivity of \(\varphi\)), the turnover occurs at the Page
time \(t_P \approx 0.646 \, t_{\text{evap}}\) (structural, independent
of mass and greybody factors), and the Page value is recovered for all
but an exponentially negligible fraction (\(\sim e^{-10^{77}}\) for
\(30 M_\odot\)) of initial conditions --- proved from bijectivity (cycle
decomposition on the finite energy shell) and Popescu-Short-Winter
typicality {[}27{]}, without assuming ergodicity. The greybody factor is
solution-specific.

Laboratory tests of the characterization theorem may be possible in
analogue gravity systems where the hidden sector capacity is tunable.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{conclusion}{%
\subsection{9. Conclusion}\label{conclusion}}

The cosmological horizon realizes the embedded-observation framework of
{[}Main{]}, and the consequences for the gravitational sector are sharp.
The emergent action scale \(\hbar = c^3 (2\,l_p)^2 / (4G)\) is
determined uniquely by thermal self-consistency between the classical
horizon temperature and the Euclidean periodicity of the emergent QFT,
with no free parameters. The Bekenstein-Hawking entropy
\(S = A/(4 l_p^2)\) is derived by direct mode counting, with the \(1/4\)
coefficient fixed as the ratio between the Euclidean KMS period and the
Jacobson prefactor --- a structural constant, not an empirical fit. The
GW250114 observation of September 2025 confirms the area law at
\(99.999\%\) confidence, providing the framework's first direct
empirical test at the percent level. The cosmological constant problem
is dissolved: the \(10^{122}\) ratio between the quantum vacuum energy
and the effective \(\Lambda\) is the compression ratio between two
levels of description that refer to incommensurable objects, not a
failure of cancellation.

The empirical payoff is the set of predictions in §7. Dynamical dark
energy in running-vacuum form with
\(\nu_{\text{OI}} = 2.45 \times 10^{-3}\) is the most distinctive:
currently supported by DESI DR2 at \(2.8\)--\(4.2\sigma\), falsifiable
by DESI Year 5 and subsequent stage-IV surveys. The dark-sector
corollary explains \textasciitilde95\% of the gravitational budget of
the universe without new fundamental particles, and the absence of
direct dark-matter detection at the LZ and XENONnT experiments is
interpreted within the framework as a structural consequence rather than
a null result to be explained away. The MOND phenomenology emerges from
entropy displacement at the cosmological boundary with \(a_0 = cH/6\),
reproducing the baryonic Tully-Fisher relation and predicting Keplerian
decline at \(r > r_M\) --- the latter confirmed by the Jiao et
al.~(2023) detection of a Keplerian decline in the Milky Way's own
rotation curve at \(\sim 17\) kpc. The \(H(z)\) scaling of \(a_0\)
predicts systematically more baryon-dominated rotation curves at high
redshift, consistent with Genzel et al.'s detection of declining outer
velocities at \(z = 0.9\)--\(2.4\) and testable at higher precision by
JWST and ELT kinematics.

The foundational framework developed in {[}Main{]} and the cubic-lattice
realization developed in {[}SM{]} are complementary to the present work:
{[}Main{]} establishes the equivalence between embedded observation and
quantum mechanics, the present paper applies it to the cosmological
horizon, and the SM paper applies it to the microscopic lattice
substratum. The reconstruction theorem and the synthesis claim --- that
quantum mechanics and general relativity descend as different
projections of the same deterministic \((S, \varphi)\), one
visible-sector and one boundary-level --- are developed in
{[}Substratum{]}. Together, the four-paper sequence establishes that the
Standard Model and general relativity are two faces of a single finite
deterministic construction, with quantum mechanics emerging as the
unique self-consistent description for any observer bounded by a causal
horizon.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{appendix-a-nested-partitions}{%
\subsection{Appendix A: Nested
Partitions}\label{appendix-a-nested-partitions}}

This appendix develops the nested-partition formalism for the framework,
establishing the self-consistency of multiple coexisting causal
boundaries (typically a black hole horizon inside a cosmological
horizon) and deriving the generalized second law and the Page curve as
consequences.

\hypertarget{a.1-setup}{%
\subsubsection{A.1 Setup}\label{a.1-setup}}

Let \(\mathcal{C}_V\), \(\mathcal{C}_B\), \(\mathcal{C}_D\) be finite
configuration spaces with \(|\mathcal{C}_V|, |\mathcal{C}_B| \geq 2\),
and let
\(\varphi: \mathcal{C}_V \times \mathcal{C}_B \times \mathcal{C}_D \to \mathcal{C}_V \times \mathcal{C}_B \times \mathcal{C}_D\)
be a bijection. The physical picture: \(V\) is the region between a
black hole horizon and the cosmological horizon, \(B\) is the black hole
interior, \(D\) is the trans-cosmological region. Define
\(W = V \cup B\) (the cosmological interior).

Two procedures produce a quantum description on \(V\):

\emph{Direct:} Trace out \(B \cup D\) in one step, marginalizing over
\(\mathcal{C}_B \times \mathcal{C}_D\).

\emph{Sequential:} (Stage 1) Trace out \(D\), producing emergent QM on
\(W\). (Stage 2) Within the emergent QM on \(W\), trace out \(B\),
producing a further-reduced description on \(V\).

\hypertarget{a.2-classical-associativity}{%
\subsubsection{A.2 Classical
Associativity}\label{a.2-classical-associativity}}

\textbf{Theorem A.1.} \emph{The direct and sequential procedures produce
the same transition probabilities on \(\mathcal{C}_V\).}

\emph{Proof.} The direct transition matrix is:

\[T^{\text{dir}}_{ij}(t) = \frac{1}{|\mathcal{C}_B||\mathcal{C}_D|} \sum_{b \in \mathcal{C}_B} \sum_{d \in \mathcal{C}_D} \delta_{x_j}[\pi_V(\varphi_t(x_i, b, d))]\]

The sequential procedure marginalizes over \(D\) (Stage 1) then over
\(B\) (Stage 2). Composing:

\[T^{\text{seq}}_{ij}(t) = \frac{1}{|\mathcal{C}_B|} \sum_b \frac{1}{|\mathcal{C}_D|} \sum_d \delta_{x_j}[\pi_V(\varphi_t(x_i, b, d))] = T^{\text{dir}}_{ij}(t)\]

The sums commute (Fubini on a finite product). \(\square\)

\hypertarget{a.3-quantum-classical-consistency}{%
\subsubsection{A.3 Quantum-Classical
Consistency}\label{a.3-quantum-classical-consistency}}

\textbf{Theorem A.2} (Quantum-classical consistency of nested
trace-outs). \emph{Let \(\Phi_V^{\text{dir}}\) and
\(\Phi_V^{\text{seq}}\) be the CPTP channels on \(\mathcal{H}_V\)
produced by the direct and sequential procedures respectively. Then
\(\Phi_V^{\text{dir}} = \Phi_V^{\text{seq}}\).}

\emph{Proof.} By the Stinespring dilation theorem {[}28, 29{]},
\(\varphi\) defines a unitary \(U_\varphi\) on
\(\mathcal{H} = \mathcal{H}_V \otimes \mathcal{H}_B \otimes \mathcal{H}_D\).
The direct channel is
\(\Phi_V^{\text{dir}}(\rho_V) = \text{Tr}_{BD}[U_\varphi\,(\rho_V \otimes \rho_B \otimes \rho_D)\,U_\varphi^\dagger]\).
The sequential channel is
\(\Phi_V^{\text{seq}}(\rho_V) = \text{Tr}_B[\text{Tr}_D[U_\varphi\,(\rho_V \otimes \rho_B \otimes \rho_D)\,U_\varphi^\dagger]]\).
On finite-dimensional tensor products,
\(\text{Tr}_B \circ \text{Tr}_D = \text{Tr}_{BD}\), giving
\(\Phi_V^{\text{seq}} = \Phi_V^{\text{dir}}\). \(\square\)

\hypertarget{a.4-hamiltonian-and-hbar-consistency}{%
\subsubsection{\texorpdfstring{A.4 Hamiltonian and \(\hbar\)
Consistency}{A.4 Hamiltonian and \textbackslash hbar Consistency}}\label{a.4-hamiltonian-and-hbar-consistency}}

\textbf{Corollary A.3} (Hamiltonian consistency). \emph{The emergent
Hamiltonians \(\hat{H}_V^{\text{dir}}\) and \(\hat{H}_V^{\text{seq}}\)
satisfy
\(\hat{H}_V^{\text{dir}} = D\,\hat{H}_V^{\text{seq}}\,D^\dagger\) for a
time-independent diagonal unitary \(D\).}

\emph{Proof.} Theorem A.2 gives
\(\Phi_V^{\text{dir}} = \Phi_V^{\text{seq}}\), hence
\(T^{\text{dir}}_{ij}(t) = T^{\text{seq}}_{ij}(t)\) for all \(t\) (Born
rule applied to diagonal elements). The D-gauge theorem (§3.3) gives the
result. \(\square\)

\textbf{Corollary A.4} (\(\hbar\) universality). \emph{Both procedures
yield the same emergent action scale \(\hbar = c^3 \epsilon^2 / (4G)\).}

\emph{Proof.} \(\hbar\) depends only on local boundary quantities \(c\),
\(G\), \(\epsilon\) (§3.2, Steps 2--3), not on which horizon defines the
partition or how many trace-outs are performed. \(\square\)

\hypertarget{a.5-additive-dissipators}{%
\subsubsection{A.5 Additive
Dissipators}\label{a.5-additive-dissipators}}

\textbf{Theorem A.5} (Additive dissipators). \emph{Let the classical
Hamiltonian be spatially local with coupling chain
\(V \leftrightarrow B \leftrightarrow D\) (no direct \(V\)-\(D\)
coupling). Let \(\tau_B^D\) be the \(D\)-sector timescale and
\(\tau_B^B\) the \(B\)-sector timescale. On observation timescales
\(t \ll \min(\tau_B^D, \tau_B^B)\), the dissipative correction to
unitarity on \(\mathcal{H}_V\) decomposes as:}

\[\mathcal{D}_V = \mathcal{D}_V^{(B)} + \mathcal{D}_V^{(D)} + \mathcal{R}\]

\emph{where
\(\|\mathcal{D}_V^{(B)}\| = \mathcal{O}(\tau_S / \tau_B^B)\),
\(\|\mathcal{D}_V^{(D)}\| = \mathcal{O}(\tau_S / \tau_B^D)\), and the
cross term satisfies
\(\|\mathcal{R}\| = \mathcal{O}(\tau_S^2 / (\tau_B^B \cdot \tau_B^D))\).}

\emph{Proof.} The proof applies the boundary-only dependence lemma
(§3.2) twice, tracking the cross term at each stage.

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  \emph{First application: freeze \(D\).} Apply the boundary-only
  dependence lemma to the partition \((W, D)\) where \(W = V \cup B\).
  By spatial locality,
  \(H_{\text{tot}} = H_V + H_B + H_D + H_{VB} + H_{BD}\) with no direct
  \(V\)-\(D\) coupling. The \(D\)-sector is frozen on timescales
  \(t \ll \tau_B^D\), giving:
\end{enumerate}

\[T^V_{ij}(t) = \frac{1}{|\mathcal{C}_B|} \sum_{b \in \mathcal{C}_B} \delta_{x_j}[\pi_V(\varphi_t^{VB}(x_i, b))] + \mathcal{O}(t / \tau_B^D)\]

where \(\varphi_t^{VB}\) is the flow restricted to \(V \times B\) with
\(D\) frozen. The \(\mathcal{O}(t / \tau_B^D)\) correction arises from
\(D\)'s back-reaction on \(B\) through \(H_{BD}\).

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\setcounter{enumi}{1}
\item
  \emph{Dissipator from the frozen-\(D\) dynamics.} The frozen-\(D\)
  flow \(\varphi_t^{VB}\) is a bijection on
  \(\mathcal{C}_V \times \mathcal{C}_B\) --- the standard Stinespring
  setup with \(B\) as the hidden sector. The emergent dynamics on \(V\)
  is approximately unitary with dissipator
  \(\mathcal{D}_V^{(B)} \sim \mathcal{O}(\tau_S / \tau_B^B)\).
\item
  \emph{Decomposition of the \(D\)-correction.} The
  \(\mathcal{O}(t / \tau_B^D)\) correction from step (i) has two
  components:
\end{enumerate}

\emph{Direct:} \(D\)'s back-reaction shifts the boundary modes of \(B\)
by \(\delta b \sim \mathcal{O}(t / \tau_B^D)\). This propagates through
\(H_{VB}\) to the visible sector, contributing
\(\mathcal{D}_V^{(D)} \sim \mathcal{O}(\tau_S / \tau_B^D)\).

\emph{Cross:} The shift \(\delta b\) also modifies the effective
\(B\)-state against which \(\mathcal{D}_V^{(B)}\) is computed. The
dissipator \(\mathcal{D}_V^{(B)}\) depends on the \(B\)-state through
the Liouville average over \(b\). The \(D\)-induced shift changes this
average by \(\mathcal{O}(t / \tau_B^D)\), so:

\[\|\mathcal{R}\| = \left\|\frac{\partial \mathcal{D}_V^{(B)}}{\partial b}\right\| \cdot \|\delta b\| \leq \|\mathcal{D}_V^{(B)}\| \cdot \mathcal{O}(t / \tau_B^D) = \mathcal{O}\!\left(\frac{\tau_S}{\tau_B^B} \cdot \frac{\tau_S}{\tau_B^D}\right) = \mathcal{O}\!\left(\frac{\tau_S^2}{\tau_B^B \cdot \tau_B^D}\right)\]

The inequality uses the fact that \(\mathcal{D}_V^{(B)}\) is a
continuous function of the \(B\)-state distribution and the shift is
first-order in \(t / \tau_B^D\). \(\square\)

\textbf{Bound for the cosmological + stellar black hole case.} For a
\(30 M_\odot\) black hole inside the cosmological horizon:
\(\tau_S \sim 10^{-15}\) s,
\(\tau_B^B \sim \tau_{\text{scr}} \sim \beta \log S_{\text{BH}} \sim 10^{-3}\)
s, \(\tau_B^D \sim 1/H \sim 10^{17}\) s. The three terms:

\[\|\mathcal{D}_V^{(B)}\| \sim 10^{-12}, \qquad \|\mathcal{D}_V^{(D)}\| \sim 10^{-32}, \qquad \|\mathcal{R}\| \sim 10^{-44}\]

The cross term is negligible by 12 orders of magnitude relative to the
smaller of the two primary dissipators.

\hypertarget{a.6-the-generalized-second-law}{%
\subsubsection{A.6 The Generalized Second
Law}\label{a.6-the-generalized-second-law}}

The Bekenstein-Hawking entropy of a horizon is the information content
of the corresponding partition boundary (§5):
\(S_{\text{BH}} = A_{\text{BH}} / (4\,l_p^2)\) modes at the black hole
horizon, \(S_{\text{dS}} = A_{\text{dS}} / (4\,l_p^2)\) modes at the
cosmological horizon. The matter entropy is the von Neumann entropy of
the visible-sector state:
\(S_{\text{matter}} = -\text{Tr}[\rho_V \log \rho_V]\).

\textbf{Theorem A.6} (Generalized second law). \emph{For any process
described by the CPTP channel \(\Phi_V\) on \(\mathcal{H}_V\):}

\[S_{\text{matter}}(t) + S_{\text{BH}}(t) + S_{\text{dS}}(t) \geq S_{\text{matter}}(0) + S_{\text{BH}}(0) + S_{\text{dS}}(0)\]

\emph{Proof.} The total system \((S, \varphi)\) evolves unitarily at the
fundamental level, so the total fine-grained entropy
\(S_{\text{total}} = \log|S|\) is constant. The three terms
\(S_{\text{matter}}\), \(S_{\text{BH}}\), \(S_{\text{dS}}\) are the
observer's ignorance about three sectors: \(V\), \(B\), \(D\)
respectively.

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  \emph{Subadditivity.} For the tripartite system
  \(\mathcal{H}_V \otimes \mathcal{H}_B \otimes \mathcal{H}_D\) in a
  pure state \(|\Psi\rangle\) (the total state is deterministic, hence
  pure conditioned on the initial state), the strong subadditivity of
  von Neumann entropy {[}29{]} gives:
\end{enumerate}

\[S(\rho_{VB}) + S(\rho_{BD}) \geq S(\rho_B) + S(\rho_{VBD})\]

Since \(\rho_{VBD} = |\Psi\rangle\langle\Psi|\) is pure,
\(S(\rho_{VBD}) = 0\). Thus
\(S(\rho_{VB}) + S(\rho_{BD}) \geq S(\rho_B)\).

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\setcounter{enumi}{1}
\item
  \emph{Identification.} The observer's matter entropy is
  \(S_{\text{matter}} = S(\rho_V)\). By purification on the tripartite
  system: \(S(\rho_V) = S(\rho_{BD})\). The black hole entropy is
  \(S_{\text{BH}} = S(\rho_B)\) restricted to the boundary modes (the
  number of \(B\)-modes coupled across the black hole horizon). The
  cosmological entropy is \(S_{\text{dS}} = S(\rho_D)\) restricted to
  the boundary modes.
\item
  \emph{Monotonicity and compensation.} The CPTP channel \(\Phi_V\) is a
  contraction of relative entropy {[}29, Theorem 11.9{]}, so
  \(S(\rho_V)\) is non-decreasing under the emergent dynamics. Any
  process that decreases \(S_{\text{BH}}\) (shrinking the black hole)
  transfers information from \(B\) to \(V \cup D\). By the unitarity of
  \(\varphi\), the total mutual information \(I(V:B:D)\) is conserved,
  so the decrease in \(S_{\text{BH}}\) is compensated by an increase in
  \(S_{\text{matter}} + S_{\text{dS}}\). \(\square\)
\end{enumerate}

\hypertarget{a.7-the-page-curve}{%
\subsubsection{A.7 The Page Curve}\label{a.7-the-page-curve}}

\textbf{Setup.} A black hole with initial horizon area \(A_0\)
evaporates. The partition boundary at the black hole horizon moves:
modes transition from \(B\) (interior) to \(R\) (radiation, a subsystem
of \(V\)) as the horizon shrinks. Define
\(n_B(t) = A_{\text{BH}}(t) / (4\,l_p^2)\) (the number of boundary modes
at the BH horizon) and \(n_R(t) = n_B(0) - n_B(t)\) (the number of
emitted radiation modes). Conservation:
\(n_B(t) + n_R(t) = n_B(0) = S_{\text{BH}}(0)\).

The trans-cosmological sector \(D\) is frozen on evaporation timescales
(\(\tau_{\text{evap}} \ll \tau_B^D\)), so by Theorem A.5 the relevant
dynamics is the bipartite system \(B \cup R\) with \(D\) decoupled at
leading order.

\textbf{Theorem A.7} (Effective bipartite pure state). \emph{On
timescales \(t \ll \tau_B^D\), the joint state of \(B\) and \(R\) is
effectively pure:
\(\rho_{BR} = |\psi(t)\rangle\langle\psi(t)| + \mathcal{O}(\tau_S / \tau_B^D)\),
where
\(|\psi(t)\rangle \in \mathcal{H}_B(t) \otimes \mathcal{H}_R(t)\).}

\emph{Proof.} The total system evolves unitarily under \(U_\varphi\)
(§A.3). The total state
\(|\Psi\rangle \in \mathcal{H}_V \otimes \mathcal{H}_B \otimes \mathcal{H}_D\)
is pure (deterministic evolution from a definite initial state). By the
boundary-only dependence lemma, the \(D\)-sector is frozen: \(\rho_D\)
is approximately constant, and
\(|\Psi\rangle \approx |\psi_{VB}\rangle \otimes |\chi_D\rangle + \mathcal{O}(\tau_S / \tau_B^D)\).
Restricting to the \(B \cup R\) subsystem (where \(R \subset V\) is the
radiation): \(\rho_{BR} \approx |\psi\rangle\langle\psi|\) is pure at
leading order. The correction is
\(\mathcal{O}(\tau_S / \tau_B^D) \sim 10^{-32}\). \(\square\)

\textbf{Theorem A.8} (Cycle typicality). \emph{For the dynamics
\(\varphi\) on the energy shell
\(\Sigma_E \subset \mathcal{H}_B \otimes \mathcal{H}_R\), the
time-averaged entanglement entropy of the radiation equals the Page
value {[}30{]}:}

\[\overline{S(\rho_R)} = S_{\text{Page}}(d_B, d_R) + \mathcal{O}(\epsilon)\]

\emph{for all initial conditions except a fraction
\(\leq 2e^{-c \, d_{\min} \, \epsilon^2}\), where
\(d_B = |\mathcal{C}_B(t)|\), \(d_R = |\mathcal{C}_R(t)|\),
\(d_{\min} = \min(d_B, d_R)\), and:}

\[S_{\text{Page}}(d_B, d_R) = \ln(\min(d_B, d_R)) - \frac{\min(d_B, d_R)}{2\max(d_B, d_R)}\]

\emph{Proof.} The bijection \(\varphi\) conserves energy (the wave
equation is Hamiltonian), so \(\varphi\) restricts to a bijection on
each energy shell \(\Sigma_E\). Since \(\Sigma_E\) is finite, this
restricted bijection decomposes into disjoint cycles
\(C_1, \ldots, C_k\) with lengths \(L_1, \ldots, L_k\) summing to
\(|\Sigma_E|\). On each cycle \(C_j\), the time average of any
observable \(f\) is exact:

\[\overline{f}_{C_j} = \frac{1}{L_j} \sum_{s \in C_j} f(s)\]

By Popescu-Short-Winter {[}27{]}, the set of atypical states --- those
with
\(|S(\text{Tr}_B |\psi\rangle\langle\psi|) - S_{\text{Page}}| > \epsilon\)
--- has measure at most \(\eta = e^{-c \, d_{\min} \, \epsilon^2}\) on
\(\Sigma_E\), where \(d_{\min} = \min(d_B, d_R)\) and \(c > 0\) is a
universal constant. The total number of atypical states is at most
\(\eta |\Sigma_E|\).

Call a cycle \emph{bad} if its time-averaged entropy deviates from
\(S_{\text{Page}}\) by more than \(\epsilon\). A bad cycle must contain
a disproportionate fraction of atypical states. Specifically, if more
than half the states in \(C_j\) are typical (entropy within \(\epsilon\)
of \(S_{\text{Page}}\)), then
\(|\overline{S}_{C_j} - S_{\text{Page}}| \leq \epsilon + \mathcal{O}(\ln d_{\min} / L_j)\).
So a bad cycle must have at least half its states atypical:
\(L_j \leq 2 \cdot |\{s \in C_j : s \text{ atypical}\}|\). Summing over
all bad cycles:

\[\sum_{j \,\text{bad}} L_j \leq 2\eta\,|\Sigma_E|\]

The fraction of initial conditions on \(\Sigma_E\) that land on a bad
cycle is therefore at most
\(2\eta = 2e^{-c \, d_{\min} \, \epsilon^2}\). For a \(30 M_\odot\)
black hole, \(d_{\min} \sim e^{S_{\text{BH}}} \sim e^{10^{77}}\), so
\(\eta \sim e^{-10^{77}}\). For all but an exponentially negligible
fraction of initial conditions, the time-averaged entanglement entropy
equals the Page value. \(\square\)

\textbf{Theorem A.9} (The Page curve). \emph{The entanglement entropy of
the radiation follows:}

\[S_{\text{rad}}(t) = \begin{cases} n_R(t) \cdot \ln q - \frac{q^{n_R(t)}}{2\,q^{n_B(t)}} & \text{for } n_R(t) < n_B(t) \quad \text{(early: before Page time)} \\ n_B(t) \cdot \ln q - \frac{q^{n_B(t)}}{2\,q^{n_R(t)}} & \text{for } n_R(t) > n_B(t) \quad \text{(late: after Page time)} \end{cases}\]

\emph{where \(q\) is the alphabet size (gauge --- it cancels in all
ratios), \(n_B(t) + n_R(t) = n_B(0)\), and the Page time \(t_P\) is
defined by \(n_R(t_P) = n_B(t_P) = n_B(0)/2\).}

\emph{Proof.} Direct substitution of \(d_B = q^{n_B(t)}\) and
\(d_R = q^{n_R(t)}\) into the Page entropy formula (Theorem A.8), using
\(\ln(q^n) = n \ln q\) and the conservation law \(n_B + n_R = n_B(0)\).
The turnover at \(n_R = n_B\) follows from \(\min(d_B, d_R)\) switching
from \(d_R\) to \(d_B\). \(\square\)

\textbf{The evaporation trajectory.} The Hawking temperature (§3.2, Step
4) gives luminosity \(L \sim 1/M^2\), hence \(dM/dt = -\alpha/M^2\)
where \(\alpha = \hbar c^4 \Gamma / (15360 \pi G^2)\) and \(\Gamma\) is
the species-dependent greybody factor (solution-specific). Integrating:

\[M(t) = M_0\!\left(1 - t/t_{\text{evap}}\right)^{1/3}, \qquad t_{\text{evap}} = \frac{M_0^3}{3\alpha}\]

The boundary mode count
\(n_B(t) = 4\pi G M(t)^2 / (\hbar c) = n_B(0) \cdot (1 - t/t_{\text{evap}})^{2/3}\)
and \(n_R(t) = n_B(0) - n_B(t)\). The Page time \(t_P\) satisfies
\(n_B(t_P) = n_B(0)/2\):

\[t_P = t_{\text{evap}}\!\left(1 - 2^{-3/2}\right) \approx 0.646 \; t_{\text{evap}}\]

The entropy of the radiation at the Page time is
\(S_{\text{rad}}(t_P) = n_B(0) \ln q / 2 = S_{\text{BH}}(0) \ln q / 2\)
--- half the initial black hole entropy, as expected.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\subsection{References}\label{references}}

\textbf{Companion papers (cited inline by short name):}

{[}Main{]} A. Maybaum, ``The Incompleteness of Observation,'' (2026).

{[}SM{]} A. Maybaum, ``The Standard Model from a Cubic Lattice,''
(2026).

{[}Substratum{]} A. Maybaum, ``The Substratum Construction:
Reconstruction, the Substratum Gauge Group, and the QM-GR Synthesis,''
(2026).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

{[}1{]} T. Jacobson, ``Thermodynamics of Spacetime: The Einstein
Equation of State,'' \emph{Phys. Rev.~Lett.} \textbf{75}, 1260 (1995).

{[}2{]} G. W. Gibbons and S. W. Hawking, ``Cosmological event horizons,
thermodynamics, and particle creation,'' \emph{Phys. Rev.~D}
\textbf{15}, 2738 (1977).

{[}3{]} W. G. Unruh, ``Sonic analogue of black holes and the effects of
high frequencies on black hole evaporation,'' \emph{Phys. Rev.~D}
\textbf{51}, 2827 (1995).

{[}4{]} K. Fredenhagen and R. Haag, ``On the derivation of Hawking
radiation associated with the formation of a black hole,'' \emph{Commun.
Math. Phys.} \textbf{127}, 273--284 (1990).

{[}5{]} LIGO Scientific Collaboration, Virgo Collaboration, and KAGRA
Collaboration: A. G. Abac et al., ``GW250114: Testing Hawking's Area Law
and the Kerr Nature of Black Holes,'' \emph{Phys. Rev.~Lett.}
\textbf{135}, 111403 (2025).

{[}6{]} J. Martin, ``Everything you always wanted to know about the
cosmological constant problem (but were afraid to ask),'' \emph{C. R.
Phys.} \textbf{13}, 566--665 (2012).

{[}7{]} S. Weinberg, ``The cosmological constant problem,''
\emph{Rev.~Mod. Phys.} \textbf{61}, 1 (1989).

{[}8{]} S. M. Carroll, ``The Cosmological Constant,'' \emph{Living
Rev.~Relativ.} \textbf{4}, 1 (2001).

{[}9{]} D. H. Wolpert, ``Physical limits of inference,'' \emph{Physica
D} \textbf{237}, 1257--1281 (2008).

{[}10{]} J. Solà Peracaula, ``The cosmological constant problem and
running vacuum in the expanding universe,'' \emph{Phil. Trans. R. Soc.
A} \textbf{380}, 20210182 (2022).

{[}11{]} J. Solà Peracaula, A. Gómez-Valent, and J. de Cruz Pérez,
``Running vacuum in the Universe,'' \emph{Universe} \textbf{9}, 262
(2023).

{[}12{]} DESI Collaboration, ``DESI 2024 VI: Cosmological Constraints
from BAO,'' arXiv:2404.03002 (2024).

{[}13{]} DESI Collaboration, ``DESI DR2 Results II: Measurements of
Baryon Acoustic Oscillations and Cosmological Constraints,'' \emph{Phys.
Rev.~D} \textbf{112}, 083515 (2025); arXiv:2503.14738.

{[}14{]} J. de Cruz Pérez, A. Gómez-Valent, and J. Solà Peracaula,
``Dynamical Dark Energy models in light of the latest observations,''
arXiv:2512.20616 (2025).

{[}15{]} DESI Collaboration: W. Elbers et al., ``Constraints on Neutrino
Physics from DESI DR2 BAO and DR1 Full Shape,'' arXiv:2503.14744 (2025).

{[}16{]} JUNO Collaboration: A. Abusleme et al., ``First measurement of
reactor neutrino oscillations at JUNO,'' arXiv:2511.14593 (2025).

{[}17{]} F. Capozzi et al., ``Updated bounds on the (1,2) neutrino
oscillation parameters after first JUNO results,'' arXiv:2511.21650
(2025).

{[}18{]} E. P. Verlinde, ``Emergent Gravity and the Dark Universe,''
\emph{SciPost Phys.} \textbf{2}, 016 (2017); arXiv:1611.02269.

{[}19{]} LZ Collaboration: J. Aalbers et al., ``Dark matter search
results from 4.2 tonne-years of exposure of the LUX-ZEPLIN (LZ)
experiment,'' (2025); 417-day analysis presented at SURF, December 2025.

{[}20{]} S. S. McGaugh, F. Lelli, and J. M. Schombert, ``Radial
acceleration relation in rotationally supported galaxies,'' \emph{Phys.
Rev.~Lett.} \textbf{117}, 201101 (2016).

{[}21{]} R. Genzel et al., ``Strongly baryon-dominated disk galaxies at
the peak of galaxy formation ten billion years ago,'' \emph{Nature}
\textbf{543}, 397--401 (2017).

{[}22{]} R. Genzel et al., ``Rotation curves in z \textasciitilde{} 1--2
star-forming disks: evidence for cored dark matter distributions,''
\emph{Astrophys. J.} \textbf{902}, 98 (2020).

{[}23{]} Y. Jiao et al., ``Detection of the Keplerian decline in the
Milky Way rotation curve,'' \emph{Astronomy \& Astrophysics}
\textbf{678}, A208 (2023).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

{[}24{]} N. Bao, S. M. Carroll, and A. Singh, ``The Hilbert space of
quantum gravity is locally finite-dimensional,'' \emph{Int. J. Mod.
Phys. D} \textbf{26}, 1743013 (2017).

{[}25{]} H. C. Bäcker, S. Milz, and K. Modi, ``Operational
discrimination of classical and quantum memory in non-Markovian
processes,'' (2023).

{[}26{]} B. Yosifov et al., ``Distinguishing classical and quantum
memory in collision models with controllable correlations,'' (2025).

{[}27{]} S. Popescu, A. J. Short, and A. Winter, ``Entanglement and the
foundations of statistical mechanics,'' \emph{Nature Physics}
\textbf{2}, 754--758 (2006).

{[}28{]} W. F. Stinespring, ``Positive functions on C\emph{-algebras,''
}Proc. Amer. Math. Soc.* \textbf{6}, 211--216 (1955).

{[}29{]} M. A. Nielsen and I. L. Chuang, \emph{Quantum Computation and
Quantum Information} (Cambridge University Press, 10th anniversary
edition, 2010).

{[}30{]} D. N. Page, ``Average entropy of a subsystem,'' \emph{Phys.
Rev.~Lett.} \textbf{71}, 1291 (1993).

\end{document}
