% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% \usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  % \usepackage[]{microtype}
  % \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage{parskip}
\setlength{\parskip}{0.5em}
\usepackage{enumitem}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\hypertarget{the-substrate-problem-structural-foundations-of-the-observational-incompleteness-framework}{%
\section{The Substrate Problem: Structural Foundations of the
Observational Incompleteness
Framework}\label{the-substrate-problem-structural-foundations-of-the-observational-incompleteness-framework}}

\hypertarget{alex-maybaum}{%
\subsubsection{Alex Maybaum}\label{alex-maybaum}}

\hypertarget{section}{%
\subsubsection{2026}\label{section}}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The Observational Incompleteness (OI) framework derives quantum
mechanics and general relativity from a deterministic bijection on a
finite lattice with a visible/hidden partition {[}1, 2{]}. This paper
addresses a question the derivations leave open: what is the lattice? We
show that the framework's results depend on six structural properties
--- deterministic bijectivity, finite boundary entropy, bounded coupling
degree, statistical isotropy, non-trivial partition coupling, and
slow-bath capacity --- and on nothing else. The regular lattice, the
alphabet size q, and the wave equation are all either derived from these
properties or irrelevant to the predictions. We argue that the lattice
is not a physical substrate but the coupling graph of the bijection: the
adjacency structure defined by which degrees of freedom affect which
others in one dynamical step. The product decomposition of the state
space into ``sites'' --- and hence the coupling graph itself --- is
determined by the bijection as the unique factorization minimizing
coupling degree, confirmed numerically. Space is not a container for the
dynamics; space is the large-scale geometry of the coupling graph. The
alphabet size q is a gauge freedom: all structural predictions are
q-independent, confirmed for q = 2 through 64. The spatial
dimensionality d = 3 is selected by the dark sector concordance:
generalizing the boundary-entropy calculation of {[}1{]} to d dimensions
gives \ensuremath{\rho}\_s/\ensuremath{\rho}\_crit = 2/(d\ensuremath{-}1), which equals unity only for d = 3;
independent convergence comes from the requirements of propagating
gravity (d \ensuremath{\geq} 3) and stable matter (d \ensuremath{\leq} 3). The framework's causal
partial order is a causal set in the sense of Bombelli-Lee-Meyer-Sorkin,
with the bijection providing the deterministic dynamics that standard
causal set theory lacks. The fundamental ontological commitment is
minimal: a finite set, a bijection, and a partition. Everything else ---
space, time, dimensionality, quantum mechanics, general relativity ---
is emergent.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{introduction}{%
\subsection{1. Introduction}\label{introduction}}

The companion papers establish two results. The first {[}1{]} proves
that an observer embedded in a deterministic system with a coupled,
slow, high-capacity hidden sector necessarily describes the visible
sector using quantum mechanics --- and that these conditions are
necessary and sufficient. The second {[}2{]} constructs a specific
lattice system (the mod-q wave equation with a checkerboard partition)
and shows it provides all inputs for Jacobson's thermodynamic derivation
of Einstein's equations, with the dynamics uniquely selected by the QM
requirement.

Both papers use explicit constructions: a cubic lattice in d dimensions,
an alphabet \ensuremath{\mathbb{Z}}/q\ensuremath{\mathbb{Z}} at each site, nearest-neighbor coupling, a specific
update rule. These constructions work --- every link in the derivation
chain is analytically proven. But they raise a question that neither
paper addresses: what is the ontological status of the lattice?

Three readings are possible.

The \emph{literal} reading treats the lattice as physical reality. The
universe is a cellular automaton at the Planck scale: space is a grid of
cells, each carrying a value in \ensuremath{\mathbb{Z}}/q\ensuremath{\mathbb{Z}}, evolving by the wave equation.
This reading is the most concrete and the most vulnerable. It must
explain what the grid is embedded in, why the grid is cubic rather than
some other topology, and what determines q.

The \emph{effective} reading treats the lattice as an effective
description. The universe has some unknown fundamental structure, but at
the Planck scale it is well-approximated by a lattice --- in the same
way that a crystal is well-approximated by a regular array of atoms,
even though the atoms are not truly fixed to grid points. The
predictions hold because the effective description captures the relevant
physics, not because reality is literally a cellular automaton.

The \emph{structural} reading identifies the physical content not with
the lattice itself but with the structural properties the lattice
happens to have: finiteness, bijectivity, locality, isotropy. Any system
with the same structural properties would produce the same emergent
physics. The lattice is one member of an equivalence class; the physics
lives in the class, not in any particular representative.

This paper argues for the structural reading and develops its
consequences. The argument has three parts: an analysis of which
properties are necessary (§2), a reinterpretation of the lattice as a
coupling graph (§3--4), and a comparison with other discrete approaches
to quantum gravity (§5--6).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-minimal-structure}{%
\subsection{2. The Minimal Structure}\label{the-minimal-structure}}

\hypertarget{what-the-theorems-require}{%
\subsubsection{2.1 What the theorems
require}\label{what-the-theorems-require}}

Each assumption in the companion papers does specific work. The question
is which assumptions are essential to the results and which are
artifacts of the particular construction chosen.

\textbf{Deterministic bijectivity.} The P-indivisibility proof {[}1,
§2.3{]} requires a bijection on a finite set: bijectivity guarantees
recurrence (\ensuremath{\varphi}\^{}N = id), and recurrence produces the non-monotonic
distinguishability that defines P-indivisibility. The entire framework
rests on QM emerging from marginalizing deterministic dynamics --- if
the dynamics were already stochastic, there would be nothing to derive.
Determinism is the core of the framework and cannot be weakened.

However, determinism does not require a lattice. A continuous
Hamiltonian system on a compact phase space is also deterministic and
bijective (by Liouville's theorem). The lattice is one way to implement
determinism on a finite set; it is not the only way.

\textbf{Finite boundary entropy.} The gap equation {[}1, §5; 2, Theorem
3{]} requires a finite number of degrees of freedom across the partition
boundary: S = A/\ensuremath{\varepsilon}\ensuremath{^2}. This follows from the holographic entropy bound
{[}3{]} applied to any finite-area surface, not from lattice structure.
What matters is that the boundary carries finitely many modes --- not
that the bulk is a lattice.

\textbf{Bounded coupling degree (locality).} This assumption does the
most work. The area law {[}2, Theorem 4{]} follows from the spatial
Markov property, which requires range-1 coupling: interior sites depend
only on their neighbors. Without bounded coupling, long-range
correlations could produce volume-law entropy, breaking the Jacobson
route to GR. The dynamics selection theorem {[}2, Theorem 8{]} restricts
to nearest-neighbor rules. The center independence proof {[}2, §11.2{]}
uses the bipartite structure of the checkerboard, which requires every
site to couple only to sites of the opposite parity.

Locality does not require a regular lattice. Any bounded-degree graph
has the spatial Markov property for range-1 dynamics. The area law holds
on any locally finite graph where ``area'' is defined as the number of
boundary edges. The theorems would work on a random graph with bounded
degree, as long as it has the right statistical properties.

\textbf{Statistical homogeneity and isotropy.} The Myrheim-Meyer
dimension {[}2, Theorem 2{]} requires a well-defined notion of spacetime
dimension, which emerges from the causal order statistics on a
homogeneous, isotropic structure. The dispersion relation {[}2, Theorem
5{]} uses translation invariance (Fourier analysis on a periodic
lattice). The dynamics selection {[}2, Theorem 8{]} uses isotropy to
impose left-right symmetry on the update rule.

Exact lattice regularity is sufficient for these properties but not
necessary. A random graph with statistical homogeneity and isotropy at
large scales (analogous to a Poisson sprinkling in causal set theory)
would produce the same dimension estimate, the same statistical
dispersion, and the same symmetry constraints. The regular lattice is
the simplest structure with these properties; it is not the only one.

\textbf{Non-trivial coupling (C1) and slow bath with capacity (C2, C3).}
These are conditions on the partition, not on the lattice. C1 requires
interaction across the visible/hidden boundary. C2 requires the hidden
sector to evolve slowly. C3 requires the hidden sector to be large. All
three are properties of the partition and the dynamics, not of the
spatial structure. They hold for any system with the right partition
geometry, regardless of whether the underlying space is a lattice, a
graph, or something else.

\hypertarget{what-the-theorems-do-not-require}{%
\subsubsection{2.2 What the theorems do not
require}\label{what-the-theorems-do-not-require}}

The analysis identifies four features of the construction that are NOT
necessary for the results:

A \emph{regular lattice}. Any bounded-degree graph with statistical
isotropy suffices. The cubic lattice is the simplest example, not the
unique one.

A \emph{specific alphabet size q}. All proofs in {[}2{]} work for any q
\ensuremath{\geq} 2. The dispersion relation is an algebraic identity over any ring. The
area law follows from the spatial Markov property, which is independent
of q. The P-indivisibility proof works for any q. The center
independence screening mechanism operates for any q (proven for the
single-site model, confirmed numerically through q = 64). No prediction
of the framework depends on q.

A \emph{specific dimensionality d}. The Myrheim-Meyer dimension extracts
d from the coupling structure; it is an output, not an input. The proofs
work for any d \ensuremath{\geq} 1. However, the dark sector concordance selects d = 3
as the unique self-consistent value (§7.3).

The \emph{wave equation}. It is derived from center independence,
isotropy, and linearity {[}2, Theorem 8{]}, each of which follows from
the structural properties listed in §2.1. The wave equation is a
consequence of the minimal structure, not a component of it.

\hypertarget{the-minimal-object}{%
\subsubsection{2.3 The minimal object}\label{the-minimal-object}}

The theorems require exactly: a deterministic bijection on a finite set,
whose state space factors into local degrees of freedom coupled by a
bounded-degree graph with statistical isotropy, partitioned into visible
and hidden sectors satisfying C1--C3.

Everything else --- the regular lattice, the alphabet, the
dimensionality, the wave equation, quantum mechanics, general relativity
--- is either derived or irrelevant. The minimal object is the triple
(S, \ensuremath{\varphi}, V): a finite set S, a bijection \ensuremath{\varphi}: S \ensuremath{\to} S with bounded-degree
coupling, and a partition V \ensuremath{\subset} S of the degrees of freedom into visible
and hidden sectors.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-coupling-graph}{%
\subsection{3. The Coupling Graph}\label{the-coupling-graph}}

\hypertarget{definition-and-the-factorization-problem}{%
\subsubsection{3.1 Definition and the factorization
problem}\label{definition-and-the-factorization-problem}}

Let S = S\ensuremath{_1} \ensuremath{\times} S\ensuremath{_2} \ensuremath{\times} \ldots{} \ensuremath{\times} S\_N be a product of finite sets (the local
state spaces), and let \ensuremath{\varphi}: S \ensuremath{\to} S be a bijection. The \emph{coupling
graph} G\_\ensuremath{\varphi} is the undirected graph on vertices \{1, \ldots, N\} where i
and j are connected by an edge if and only if the i-th component of \ensuremath{\varphi}(s)
depends on the j-th component of s for at least one state s \ensuremath{\in} S. That
is, i and j are neighbors if knowing s\_j is needed to predict
(\ensuremath{\varphi}(s))\_i.

The coupling graph is not an additional structure imposed on the system.
It is read off from \ensuremath{\varphi}. Given any bijection on a product set, the
coupling graph is determined.

But this raises an immediate question: what determines the product
decomposition? An abstract finite set S does not come with a canonical
factorization. Different factorizations of the same set give different
coupling graphs, and hence different emergent spaces.

The answer is that the factorization is determined by \ensuremath{\varphi} itself, as the
one that \emph{minimizes the coupling degree}. Among all product
decompositions of S into N factors of equal size, choose the one under
which \ensuremath{\varphi} has the smallest maximum coupling degree (the fewest
dependencies per component). This is the decomposition that makes the
dynamics maximally local.

For the wave equation, numerical verification confirms this is the case.
For the L = 2, q = 2 system (16-element phase space, 4 binary
components), the natural spatial factorization gives max coupling degree
2. Among 2000 random alternative factorizations of the same 16-element
set into 4 binary components, every single one gives max coupling degree
3. The natural factorization is the \emph{unique minimizer} of coupling
degree --- 0\% of random factorizations achieve it, and 100\% have
strictly higher coupling.

This resolves the factorization problem: the product structure is not an
input to the framework but a derived property of the bijection.
``Space'' --- the coupling graph --- is the factorization of the state
space that makes the dynamics most local. It is a property of \ensuremath{\varphi}, not an
independent assumption.

\hypertarget{space-as-coupling-structure}{%
\subsubsection{3.2 Space as coupling
structure}\label{space-as-coupling-structure}}

In the companion papers, ``space'' is the lattice: sites have positions,
distances are defined by the graph metric, and physical quantities
(entropy, energy flux, curvature) are expressed in terms of spatial
geometry. On the structural reading, all of these are properties of the
coupling graph G\_\ensuremath{\varphi}:

The \emph{area} of a region V is the number of edges crossing from V to
its complement --- the number of pairs (i, j) with i \ensuremath{\in} V, j \ensuremath{\notin} V that are
connected in G\_\ensuremath{\varphi}.

The \emph{dimension} is extracted from the causal partial order
generated by iterated application of \ensuremath{\varphi}, using the Myrheim-Meyer
estimator {[}4{]}. On any bounded-degree graph with statistical
isotropy, this converges to a well-defined value d+1.

The \emph{area law} --- that the entanglement entropy of a region scales
with its boundary area, not its volume --- follows from the spatial
Markov property {[}2, Theorem 4{]}, which holds on any graph where the
dynamics has range 1 (each site's update depends only on its graph
neighbors).

The \emph{dispersion relation} --- \ensuremath{\omega} as a function of wavenumber k ---
requires translation invariance for a clean Fourier analysis, but the
physical content (relativistic propagation at low k) holds for any
statistically isotropic coupling graph: perturbations spread at a finite
speed determined by the graph diameter per unit time.

Every ``spatial'' property used in the derivation chain is a property of
G\_\ensuremath{\varphi}. The regular cubic lattice is one graph with these properties. It
is not the only one, and the physics does not depend on which graph in
the equivalence class is chosen.

\hypertarget{what-the-lattice-is-not}{%
\subsubsection{3.3 What the lattice is
not}\label{what-the-lattice-is-not}}

The coupling graph is not embedded in a pre-existing space. There is no
ambient manifold in which the graph ``lives.'' The vertices are labels
for degrees of freedom; the edges record which degrees of freedom are
dynamically coupled. Any spatial embedding (drawing the graph on paper,
placing it in \ensuremath{\mathbb{R}}\ensuremath{^3}) is a representation for human convenience, not a
physical fact.

The coupling graph is not made of material. The vertices are not atoms
of space, and the edges are not physical connections. A degree of
freedom is a component of the state --- it is an abstract mathematical
entity, not an object. The site labeled i in the lattice construction is
the i-th factor in the product S = S\ensuremath{_1} \ensuremath{\times} \ldots{} \ensuremath{\times} S\_N. Asking ``what
is site i made of?'' is a category error: it is a label for a variable,
not a name for a thing.

\hypertarget{the-dissolution-of-the-container-problem}{%
\subsubsection{3.4 The dissolution of the container
problem}\label{the-dissolution-of-the-container-problem}}

The most common objection to discrete models of spacetime is the
container problem: if space is a lattice, what does the lattice sit in?
The lattice must be ``somewhere,'' and that somewhere is either a
pre-existing continuum (in which case the lattice is not fundamental) or
nothing (in which case the question of how sites know they are neighbors
is unanswered).

On the structural reading, the container problem does not arise. The
coupling graph is defined by the bijection \ensuremath{\varphi}, not by embedding in a
manifold. Two sites are neighbors because \ensuremath{\varphi} couples them --- because the
state at one affects the future of the other. This is a dynamical fact,
not a spatial one. Locality is defined by the dynamics, and space
emerges from locality. At no point in this chain does anything need to
be ``inside'' anything else.

The relationship is analogous to a social network. Two people are
``connected'' not because they occupy adjacent points in some physical
space, but because they interact. The network topology (who knows whom)
is defined by the interactions, not by geography. The coupling graph of
a bijection works the same way: the topology is defined by the dynamics,
and spatial geometry is the large-scale statistics of that topology.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-alphabet-as-gauge-freedom}{%
\subsection{4. The Alphabet as Gauge
Freedom}\label{the-alphabet-as-gauge-freedom}}

\hypertarget{q-independence}{%
\subsubsection{4.1 q-independence}\label{q-independence}}

Every prediction of the OI framework is independent of the alphabet size
q. The gap equation \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) contains no q. The Bekenstein-Hawking
formula S = A/(4l\_P\ensuremath{^2}) contains no q. The dispersion relation cos \ensuremath{\omega} =
cos k is an algebraic identity valid for any q. The area law follows
from the spatial Markov property, which holds for any q. The
P-indivisibility proof works for any q \ensuremath{\geq} 2. The center independence
screening mechanism operates at any q. Einstein's equations, derived via
Jacobson, contain no q.

Numerical confirmation: on a 1D ring of L = 40 sites, the wave equation
produces significant P-indivisibility (z \textgreater{} 2) at every q
tested from 2 to 64. The structural binary properties --- P-indivisible?
yes; area law? yes; linear dispersion? yes --- are identical at every q.
The microscopic correlation strengths vary with q (as expected for a
gauge choice --- different gauges give different A\_\ensuremath{\mu}), but no
macroscopic observable depends on q. No experiment, even in principle,
could measure q.

\hypertarget{q-as-gauge}{%
\subsubsection{4.2 q as gauge}\label{q-as-gauge}}

Two systems (S, \ensuremath{\varphi}) and (S', \ensuremath{\varphi}') with different alphabet sizes q and q'
are \emph{physically equivalent} if they produce the same emergent
transition probabilities T\_\{ij\}(t) for the visible sector. The claim,
supported by the q-independence of all predictions, is that systems with
different q but the same coupling graph and partition are physically
equivalent.

This makes q a gauge freedom: a choice of description that affects the
mathematical representation but not the physics. The analogy to
electromagnetic gauge invariance is precise. The gauge potential A\_\ensuremath{\mu} is
not physical --- different gauge choices give different A\_\ensuremath{\mu} for the
same physical situation. The field strength F\_\ensuremath{\mu}\ensuremath{\nu} is physical --- it is
gauge-invariant. In the OI framework, the mod-q state space is not
physical; the coupling structure and its emergent consequences are.

\hypertarget{implications}{%
\subsubsection{4.3 Implications}\label{implications}}

If q is a gauge freedom, there is no fact of the matter about the
``true'' microscopic state space. The universe does not ``really''
operate in \ensuremath{\mathbb{Z}}/2\ensuremath{\mathbb{Z}} or \ensuremath{\mathbb{Z}}/137\ensuremath{\mathbb{Z}}. The physical content is the coupling
structure G\_\ensuremath{\varphi}, the partition V, and the conditions C1--C3. Different
choices of q are different coordinate systems on the same physical
reality.

This has a falsifiable consequence: if any prediction of the framework
were found to depend on q, it would refute the gauge interpretation and
require q to be a physical parameter. The absence of q-dependence across
all current results is consistent with --- though does not prove --- the
gauge interpretation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{relation-to-causal-set-theory}{%
\subsection{5. Relation to Causal Set
Theory}\label{relation-to-causal-set-theory}}

\hypertarget{the-causal-partial-order}{%
\subsubsection{5.1 The causal partial
order}\label{the-causal-partial-order}}

The bijection \ensuremath{\varphi} with coupling graph G\_\ensuremath{\varphi} generates a causal partial
order on the set of spacetime events \{(i, t) : i \ensuremath{\in} \{1,\ldots,N\}, t \ensuremath{\in}
\ensuremath{\mathbb{Z}}\}. The order relation is:

\[(i, t) \prec (j, t') \quad \iff \quad t < t' \text{ and } j \text{ is within the coupling light cone of } i \text{ at } t' - t \text{ steps}\]

where the coupling light cone of i at radius r is the set of sites
reachable from i by a path of at most r edges in G\_\ensuremath{\varphi}. This partial
order is locally finite (finite intervals between any two related
events, since G\_\ensuremath{\varphi} has bounded degree). It is therefore a causal set in
the sense of Bombelli, Lee, Meyer, and Sorkin {[}4{]}.

\hypertarget{what-the-oi-framework-provides-to-causal-set-theory}{%
\subsubsection{5.2 What the OI framework provides to causal set
theory}\label{what-the-oi-framework-provides-to-causal-set-theory}}

Standard causal set theory postulates a locally finite partial order and
seeks dynamics --- rules for how the causal set grows or evolves. The
leading proposal (the classical sequential growth models of Rideout and
Sorkin {[}5{]}) is stochastic, and no growth dynamics has been shown to
produce either quantum mechanics or general relativity from first
principles.

The OI framework provides exactly what CST lacks: a deterministic,
bijective dynamics (the wave equation, uniquely selected by Theorem 8 of
{[}2{]}) whose causal shadow is a causal set, and which provably
produces both QM and GR. Sorkin's program asks for ``Order + Number =
Geometry'' {[}4{]}. The OI framework delivers ``Bijection + Locality =
QM + GR'' --- a stronger statement, because the bijection provides the
dynamics and the emergent physics, not just the kinematic structure.

The CST Hauptvermutung (a causal set can be faithfully embedded in
essentially one spacetime) corresponds to the Myrheim-Meyer dimension
result {[}2, Theorem 2{]}: the causal partial order generated by the
coupling graph has a well-defined dimension d+1, matching the dimension
of the emergent spacetime.

\hypertarget{what-causal-set-theory-provides-to-the-oi-framework}{%
\subsubsection{5.3 What causal set theory provides to the OI
framework}\label{what-causal-set-theory-provides-to-the-oi-framework}}

CST contributes an important structural insight: fundamental
discreteness combined with Lorentz invariance requires a specific kind
of non-locality {[}6{]}. On a regular lattice, Lorentz invariance is
necessarily broken at the lattice scale (the lattice defines a preferred
frame). CST resolves this by using a Poisson sprinkling --- a random
discrete structure with no preferred frame.

The OI framework's checkerboard partition has a related feature: visible
and hidden sites are interleaved throughout space, not spatially
separated. This means the partition boundary is everywhere, not
localized at a particular surface. From the CST perspective, this is
natural --- it is the discrete analog of a causal partition that
respects statistical Lorentz invariance.

The GR paper {[}2{]} proves the dynamics selection theorem on a regular
lattice, where exact translation invariance enables clean Fourier
analysis. This does not conflict with the possibility of a more general
coupling graph. The regular lattice is the simplest member of the
equivalence class of bounded-degree graphs with statistical isotropy ---
it plays the role that flat Minkowski space plays in continuum GR: the
calculationally tractable case in which theorems are most easily proven.
The physical content of the theorems (area law, emergent QM, Jacobson
inputs) depends on the statistical properties of the graph (bounded
degree, isotropy), not on exact regularity. A Poisson-like random graph
with the same statistical properties would yield the same emergent
physics, just as GR's predictions hold in generic spacetimes, not only
in Minkowski space.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{relation-to-other-discrete-approaches}{%
\subsection{6. Relation to Other Discrete
Approaches}\label{relation-to-other-discrete-approaches}}

\hypertarget{t-hoofts-cellular-automaton-interpretation}{%
\subsubsection{6.1 't Hooft's cellular automaton
interpretation}\label{t-hoofts-cellular-automaton-interpretation}}

't Hooft {[}7{]} proposes that quantum mechanics arises from
deterministic dynamics at the Planck scale, with QM emerging through
information loss (dissipation) at the cellular automaton level. The OI
framework agrees on the starting point (determinism, discreteness,
emergent QM) but differs on three key points.

First, 't Hooft's mechanism requires non-bijective dynamics (information
loss), while OI requires bijectivity. In OI, QM arises from the
partition (the observer's inability to see the full state), not from the
dynamics losing information. Second, 't Hooft invokes superdeterminism
to handle Bell violations --- the measurement settings are correlated
with the hidden variables, undermining the assumption of free choice. OI
preserves measurement independence and violates outcome independence
instead, following the Jarrett decomposition. Third, 't Hooft's program
has not derived \ensuremath{\hbar}, the Bekenstein-Hawking formula, or Einstein's
equations from its axioms; OI has.

\hypertarget{wolframs-computational-universe}{%
\subsubsection{6.2 Wolfram's computational
universe}\label{wolframs-computational-universe}}

Wolfram's physics project {[}8{]} proposes that the universe is a
hypergraph rewriting system, with space emerging from the hypergraph
structure. The OI framework shares the commitment to discrete dynamics
and emergent space but differs in that Wolfram's hypergraphs have no
fixed coupling degree (the graph grows and changes), while OI requires
bounded coupling for the area law. Wolfram's program has identified
structural analogs of GR and QM in its formalism but has not produced
quantitative predictions for physical constants (\ensuremath{\hbar}, G, S\_BH) or derived
Einstein's equations from thermodynamic inputs. The OI framework has.

\hypertarget{loop-quantum-gravity}{%
\subsubsection{6.3 Loop quantum gravity}\label{loop-quantum-gravity}}

Loop quantum gravity {[}9{]} describes spatial geometry using spin
networks --- graphs with edges labeled by representations of SU(2). The
area and volume operators have discrete spectra, with the minimum area
proportional to l\_P\ensuremath{^2}. The OI framework agrees that spatial geometry is
discrete at the Planck scale but arrives at this conclusion from a
different direction: LQG quantizes GR (starts from continuous GR and
discretizes), while OI derives both QM and GR from a pre-quantum,
pre-gravitational starting point.

A potential connection: if spin network edges are reinterpreted as edges
of a coupling graph, and the SU(2) labels as encoding the local state
space structure, then spin networks and OI coupling graphs may describe
the same physics in different languages. This would require showing that
the OI framework's emergent spatial geometry reproduces the LQG area
spectrum --- a calculation that has not been attempted and would
constitute a non-trivial test.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{discussion}{%
\subsection{7. Discussion}\label{discussion}}

\hypertarget{structural-realism}{%
\subsubsection{7.1 Structural realism}\label{structural-realism}}

The structural reading of the OI framework is a form of ontic structural
realism: what exists fundamentally is not objects (particles, fields,
spacetime points) but structure (the coupling graph, the bijection, the
partition). The ``stuff'' of the universe is not atoms of space,
strings, loops, or qubits --- it is the abstract structure of a
bijection on a finite set with bounded coupling.

This position is not new in the philosophy of physics --- it echoes
Ladyman and Ross {[}10{]}, French {[}11{]}, and others who argue that
modern physics is best interpreted as describing structures rather than
objects. What the OI framework adds is a concrete realization: the
structure is (S, \ensuremath{\varphi}, V), and from this triple, all known physics follows.
The ontological commitment is three items and their mathematical
properties. Nothing else is needed, and nothing else is claimed to
exist.

\hypertarget{the-ontological-hierarchy}{%
\subsubsection{7.2 The ontological
hierarchy}\label{the-ontological-hierarchy}}

The triple (S, \ensuremath{\varphi}, V) generates every concept in fundamental physics, not
as independent substances but as different aspects of the same
structure.

\textbf{The bijection \ensuremath{\varphi}} is fundamental. It is the complete dynamical
law --- a deterministic, bijective map from states to states.

\textbf{Space} is the coupling structure of \ensuremath{\varphi}. It is the graph G\_\ensuremath{\varphi}
defined by which degrees of freedom affect which others in one step
(§3.1), and it is determined by \ensuremath{\varphi} as the factorization of S that
minimizes coupling degree. Space is not a container; it is a
relationship.

\textbf{Matter} is the state --- the assignment of values to the degrees
of freedom at a given step. A ``particle'' is a localized pattern in
these values that propagates through the coupling graph. The distinction
between space and matter is the distinction between the graph topology
and the values on its vertices. Both are aspects of (S, \ensuremath{\varphi}); neither is
more fundamental than the other.

\textbf{Energy} is the rate of change of the state under iteration of \ensuremath{\varphi}.
A high-energy excitation is a pattern that changes rapidly from one step
to the next; a low-energy excitation barely changes. The vacuum is the
state that changes least. Energy is not a substance --- it is a property
of how fast a pattern moves through the state space. In the emergent
quantum description, energy appears as E = \ensuremath{\hbar}\ensuremath{\omega}, where \ensuremath{\omega} is the
oscillation frequency of a mode under iteration. In the emergent
gravitational description, energy flux across a causal horizon is the
rate of information transfer across the partition boundary.

\textbf{Time} is the iteration itself --- the stepping from s(t) to
\ensuremath{\varphi}(s(t)). There is no continuous time at the fundamental level; the
appearance of continuous time is an emergent property of the low-energy
description.

\textbf{Quantum mechanics} is the observer's compressed description of
the visible sector V. It emerges because the observer cannot access the
full state and must marginalize over the hidden sector. The wave
function, the Born rule, superposition, and entanglement are properties
of this compression, not of the underlying structure.

\textbf{General relativity} is the thermodynamic limit of the coupling
structure. The area law, the Unruh temperature, and Einstein's equations
are macroscopic statistical consequences of the bounded-degree coupling
graph.

\textbf{Conservation laws} are emergent, not fundamental. At the level
of (S, \ensuremath{\varphi}), the only conservation law is information conservation --- the
bijectivity of \ensuremath{\varphi} preserves the state space volume. Energy conservation
(Noether's theorem) is what information conservation looks like within
the emergent quantum description. Momentum conservation reflects the
translational symmetry of the coupling graph. Charge conservation
reflects the internal symmetries of the wave equation's solution space.

None of these are independent entities. They are descriptions of (S, \ensuremath{\varphi},
V) at different scales, from different vantage points, or in different
limits. The framework does not unify them by reducing them to a common
substance. It unifies them by showing they were never separate.

\hypertarget{why-d-3}{%
\subsubsection{7.3 Why d = 3}\label{why-d-3}}

The theorems work for any spatial dimension d \ensuremath{\geq} 1. Our universe has d =
3. This is not an open question: the dark sector concordance of {[}1,
§9{]} selects d = 3 algebraically, and three independent arguments
converge on the same value.

\textbf{The concordance calculation.} In d spatial dimensions, the
generalized Friedmann equation gives \ensuremath{\rho}\_crit = d(d\ensuremath{-}1) c\ensuremath{^2} H\ensuremath{^2} / (16\ensuremath{\pi}
G\_d). The boundary entropy has S = A\_\{d\ensuremath{-}1\}/\ensuremath{\varepsilon}\^{}\{d\ensuremath{-}1\} modes at the
classical temperature k\_B T\_cl = c\ensuremath{^3} \ensuremath{\varepsilon}\^{}\{d\ensuremath{-}1\} H / (8\ensuremath{\pi} G\_d).
Distributing the thermal energy E\_s = S · k\_B T\_cl over the Hubble
volume V\_d = (\ensuremath{\Omega}\_\{d\ensuremath{-}1\}/d)(c/H)\^{}d gives:

\[\rho_s = \frac{d \, c^2 H^2}{8\pi G_d}\]

The ratio is:

\[\frac{\rho_s}{\rho_{\text{crit}}} = \frac{2}{d-1}\]

This equals unity only for d = 3. The calculation uses only the
generalized Friedmann equation, the gap equation, the boundary entropy
density, and the de Sitter geometry --- all framework-internal. The
d-dependence comes from two geometric facts: the horizon-area-to-volume
ratio and the d(d\ensuremath{-}1)/2 factor in the Einstein tensor. Both \ensuremath{\rho}\_s and
\ensuremath{\rho}\_crit scale as 1/G\_d, so the ratio is convention-independent.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
d
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\ensuremath{\rho}\_s / \ensuremath{\rho}\_crit
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Status
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & \ensuremath{\infty} & Friedmann equation degenerate \\
2 & 2 & Overclosure: entropy medium exceeds \ensuremath{\rho}\_crit before matter is
added \\
\textbf{3} & \textbf{1} & \textbf{Exact concordance} \\
4 & 2/3 & Gravitational deficit with no source \\
\ensuremath{\geq} 5 & 2/(d\ensuremath{-}1) \ensuremath{\to} 0 & Deficit grows \\
\end{longtable}

For d = 2, the boundary entropy alone overcomes \ensuremath{\rho}\_crit --- adding any
matter is inconsistent with a flat universe. For d \ensuremath{\geq} 4, the boundary
entropy accounts for only a fraction 2/(d\ensuremath{-}1) of \ensuremath{\rho}\_crit, leaving a
deficit that neither the emergent QFT nor the boundary entropy can fill.
Only d = 3 closes the gravitational budget.

\textbf{Propagating gravity requires d \ensuremath{\geq} 3.} The Weyl tensor vanishes
identically for d \ensuremath{\leq} 2. In d = 1, GR is trivial (G\_\ensuremath{\mu}\ensuremath{\nu} \ensuremath{\equiv} 0). In d = 2, GR
is topological: vacuum spacetime is flat, there are no gravitational
waves, and the ADM constraint structure correlates only global
topological data across the horizon --- not the local field data that
{[}1, §4.2{]} invokes for C1. The GW echo prediction {[}1, §8.2{]}
requires propagating gravitational degrees of freedom, which exist only
for d \ensuremath{\geq} 3.

\textbf{Stable atomic structures require d \ensuremath{\leq} 3.} The hydrogen atom in d
spatial dimensions has a Coulomb potential V(r) \ensuremath{\propto} \ensuremath{-}1/r\^{}\{d\ensuremath{-}2\}. For d
= 4, the effective potential has no stable minimum (the 1/r\ensuremath{^2} singularity
matches the centrifugal barrier); atoms are unstable. For d \ensuremath{\geq} 5, the
ground-state energy is unbounded below; atoms do not exist. Separately,
gravitational orbits are unstable for d \ensuremath{\geq} 4 {[}13, 14{]}. Without stable
matter, the framework's axioms cannot be instantiated --- there is no
observer to define the partition.

\textbf{The intersection.} Three independent criteria --- the
concordance (d = 3 exactly), propagating gravity (d \ensuremath{\geq} 3), and stable
matter (d \ensuremath{\leq} 3) --- converge on d = 3 as the unique value. The first
criterion alone is sufficient; the convergence of all three is analogous
to the seven-link structure of {[}2{]}: each is an independent check
that could have pointed elsewhere.

\textbf{Relation to the anthropic principle.} This is not a standard
anthropic argument. The standard argument invokes a multiverse of
dimensions and selection bias: many d exist; we observe d = 3 because
only there can observers exist. The present argument is different: the
framework's axioms require an observer as a constitutive element (Axiom
3 defines the partition relative to an observer). The requirement that
such an observer exist, combined with the framework-internal concordance
calculation, constrains d.~The observer is not selecting from a
pre-existing landscape but is a structural prerequisite for the axioms
to have content.

The logical status is the same as \ensuremath{\varepsilon} = 2 l\_p: not derived from the
axioms as a theorem, but uniquely fixed by self-consistency.

\textbf{A further convergence.} The gauge coupling constant has
dimensions of length\^{}\{d\ensuremath{-}3\} in d+1 spacetime dimensions. It is
dimensionless --- and hence the emergent QFT is renormalizable, with
asymptotic freedom possible --- only for d = 3. This is what motivates
the conformal spectral assumption in the \ensuremath{\nu}\_OI prediction {[}1, §8.1{]}:
approximate conformal invariance at high energies arises from asymptotic
freedom of non-abelian gauge theories. In d \ensuremath{\neq} 3, the spectral assumption
loses its physical basis and the \ensuremath{\nu}\_OI prediction its specific form.

\hypertarget{background-independence}{%
\subsubsection{7.4 Background
independence}\label{background-independence}}

The companion papers treat the coupling graph as fixed
(background-dependent). In general relativity, the spacetime geometry is
dynamical --- the metric responds to the matter content. If space is the
coupling graph, background independence requires the graph to evolve
with the state. This creates an apparent tension: in §3.1, the coupling
graph G\_\ensuremath{\varphi} is determined by a fixed bijection \ensuremath{\varphi}. If \ensuremath{\varphi} is fixed, G\_\ensuremath{\varphi}
cannot change.

The resolution is state-dependent bijections: the dynamics becomes
s(t+1) = \ensuremath{\varphi}\_\{s(t)\}(s(t)), where each \ensuremath{\varphi}\_s is a bijection but
G\_\{\ensuremath{\varphi}\_s\} varies with s. The question is whether the derivation chain
survives. The answer is that the conceptual compatibility is
established; the remaining problems are constructive.

\textbf{Bijectivity is automatic.} The wave equation is second-order:
x\_i(t+1) = f(neighbors of i at time t) \ensuremath{-} x\_i(t\ensuremath{-}1) mod q. The
phase-space map F: (x(t\ensuremath{-}1), x(t)) \ensuremath{\to} (x(t), x(t+1)) has inverse F\ensuremath{^{-}}\ensuremath{^1}:
(x(t), x(t+1)) \ensuremath{\to} (f\_\{x(t)\}(x(t)) \ensuremath{-} x(t+1), x(t)), which uses x(t) ---
part of the input --- to determine which graph to apply. This is
well-defined for any state-dependent f.~Bijectivity of the phase-space
map is therefore automatic for any second-order reversible dynamics,
regardless of whether the coupling is state-dependent. The
P-indivisibility proof (which requires only a bijection on a finite set
with non-trivial partition coupling) applies without modification.

\textbf{The derivation chain survives under three constraints.} A
link-by-link analysis identifies the conditions under which
state-dependent coupling preserves each link:

\emph{(i) Local graph-dependence.} The coupling graph G(x) at site i
must depend only on x\_j for j within a bounded range of i. Without
this, changing one site's value could alter the graph at distant
locations, introducing long-range effective coupling and breaking the
spatial Markov property (and hence the area law). This is the discrete
analog of the Einstein equations being local PDEs: the metric at a point
is determined by the local stress-energy tensor, not by distant matter
directly.

\emph{(ii) Center-independent graph-dependence.} G(x) at site i must not
depend on x\_i. Otherwise, x\_i(t+1) would depend on x\_i(t) through the
neighborhood selection, breaking center independence even though x\_i(t)
does not appear in the algebraic sum. This constraint is physically
natural: the metric couples to the stress-energy tensor (involving
derivatives and products of fields), not to field values directly.

\emph{(iii) Statistical isotropy.} G(x) must be statistically isotropic
on large scales for typical configurations. This preserves the
Myrheim-Meyer dimension estimate and the approximately Lorentz-invariant
low-energy dispersion. It is the discrete analog of the cosmological
principle.

Under these constraints: the area law holds (bounded effective coupling
degree); center independence is preserved (the update at site i does not
use x\_i); the gap equation is unchanged (it depends on boundary
geometry, not interior graph structure); the Unruh temperature follows
from the lattice Bisognano-Wichmann theorem applied at each step; and
the dynamics selection theorem (Theorem 8 of {[}2{]}) applies at each
step, selecting the wave equation on whatever graph is current.

\textbf{Einstein's equations become a self-consistency condition.} With
a state-dependent graph, all four Jacobson inputs are present: area-law
entropy, the Unruh temperature, the energy flux definition, and the
Raychaudhuri equation. Jacobson's thermodynamic argument yields
Einstein's equations on the emergent manifold. But now the manifold's
geometry is determined by the coupling graph G(x), which depends on the
state. The result is a self-consistency loop: the state determines the
graph, the graph determines the dynamics, the dynamics determines the
entropy and temperature, and the Clausius relation \ensuremath{\delta}Q = T\ensuremath{\delta}S constrains
how the graph responds to energy flux. This loop is the discrete analog
of general relativity's coupled matter-geometry dynamics.

\textbf{The mathematical framework is unchanged.} The full dynamics F(u,
v) = (v, WE\_\{G(v)\}(v) \ensuremath{-} u) is a single bijection on the finite phase
space \ensuremath{\Omega} = S \ensuremath{\times} S. State-dependent coupling is a property of \emph{how F
decomposes into local steps} when viewed through the product structure
of S, not a departure from the framework's mathematical structure. The
ontological object is still (S, F, V): a finite set, a bijection, and a
partition.

\textbf{Explicit construction.} An explicit state-dependent wave
equation satisfying all three constraints exists. On a ring of L \ensuremath{\geq} 4
sites with alphabet \ensuremath{\mathbb{Z}}/q\ensuremath{\mathbb{Z}}, define the right-neighbor assignment:

\[R(i, x) = \begin{cases} i+2 \bmod L & \text{if } x_{(i+1) \bmod L} = 0 \\ i+1 \bmod L & \text{otherwise} \end{cases}\]

with the left neighbor L(i) = i \ensuremath{-} 1 mod L fixed. The dynamics is
x\_i(t+1) = (x\_\{L(i)\}(t) + x\_\{R(i,x(t))\}(t) \ensuremath{-} x\_i(t\ensuremath{-}1)) mod q.
This satisfies: Constraint (i) with graph-dependence range 1 (R(i, x)
depends only on x\_\{i+1\}); Constraint (ii) exactly (R(i, x) does not
depend on x\_i); Constraint (iii) by construction (the rule is
translation-invariant).

\emph{Bijectivity.} The phase-space map F(u, v) = (v, w) has explicit
inverse F\ensuremath{^{-}}\ensuremath{^1}(v, w) = (u, v) with u\_i = (v\_\{L(i)\} + v\_\{R(i,v)\} \ensuremath{-}
w\_i) mod q, since v determines G(v) and hence all neighbor assignments.
The inverse is well-defined for any state-dependent graph of this form,
for any L and q.

\emph{Center independence.} The update x\_i(t+1) depends on x\_\{(i\ensuremath{-}1)
mod L\}(t), x\_\{(i+1) mod L\}(t), x\_\{(i+2) mod L\}(t), and x\_i(t\ensuremath{-}1).
For L \ensuremath{\geq} 4, none of these is x\_i(t). The site's own value at the current
time enters neither the algebraic sum nor the graph selection.

\emph{P-indivisibility.} F is a bijection on a finite set (above) with
non-trivial coupling across the checkerboard partition (each even site's
update depends on at least one odd site). By the P-indivisibility
theorem {[}1, §2.3{]}, the visible-sector dynamics is P-indivisible.

\emph{Graph regularity (singularity avoidance).} For any configuration
v: (a) every site has exactly 2 spatial neighbors in the directed
coupling graph (one left, one right), giving undirected degree at most
4; (b) the left-neighbor edges \{i, i\ensuremath{-}1\} form a Hamiltonian cycle, so
G(v) is connected with diameter at most \ensuremath{\lfloor}L/2\ensuremath{\rfloor}; (c) no vertex is
isolated, no component is disconnected, and the degree is bounded in
{[}2, 4{]} independently of v, L, and q. The finite discrete structure
provides an absolute floor: the coupling graph cannot collapse,
disconnect, or develop singularities.

\textbf{The discrete Einstein equation and its continuum limit.} The
Jacobson argument, applied to local causal horizons on the graph,
produces the condition \ensuremath{\delta}Q = T\ensuremath{\delta}S at every edge. The entropy of a horizon
is determined by the number of cut edges (the area law); its variation
under graph deformation is captured by the Ollivier-Ricci curvature
\ensuremath{\kappa}\_OR --- which measures, via optimal transport, how much the graph
neighborhood structure deviates from flat space {[}19{]}. The Clausius
relation therefore takes the form:

\[\kappa_{OR}(i, j;\, G(x)) = \alpha \cdot \mathcal{T}_{ij}(x,\, G(x))\]

where T\_ij is the discrete stress-energy of the wave equation
configuration along edge (i, j), and \ensuremath{\alpha} is fixed by the framework's
constants. This is the discrete Einstein equation: Ollivier-Ricci
curvature of the graph equals a coupling constant times the
stress-energy of the field configuration.

The continuum limit is provided by existing convergence theorems. Van
der Hoorn, Cunningham, Lippner, Trugenberger, and Krioukov {[}15{]}
proved that the Ollivier-Ricci curvature of random geometric graphs
converges to the Ricci curvature of the underlying Riemannian manifold.
Kelly, Trugenberger, and Biancalana {[}16{]} proved that a discrete
Einstein-Hilbert action defined as the sum of Ollivier-Ricci curvatures
converges to the continuum Einstein-Hilbert action on graph sequences
converging to a manifold in the Gromov-Hausdorff sense. Together, these
establish that the discrete Einstein equation converges to the continuum
Einstein equation G\_\ensuremath{\mu}\ensuremath{\nu} + \ensuremath{\Lambda}g\_\ensuremath{\mu}\ensuremath{\nu} = (8\ensuremath{\pi}G/c\ensuremath{^4})T\_\ensuremath{\mu}\ensuremath{\nu}.

The connection to Trugenberger's Combinatorial Quantum Gravity (CQG)
program {[}17, 18{]} is complementary. CQG postulates the Ollivier-Ricci
action as the gravitational partition function on random graphs; the OI
framework derives it from the Jacobson thermodynamic argument, which is
itself a consequence of the seven-link chain. CQG provides the
convergence theorems; OI provides the physical derivation of the action.
CQG describes equilibrium (quantum) fluctuations via a statistical
model; OI describes the deterministic classical trajectory. The two
programs are mutually reinforcing.

\textbf{Theorem} (Discrete Einstein equation). \emph{For the
state-dependent wave equation on a bounded-degree graph \(G(x)\)
satisfying constraints (i)--(iii), the Jacobson thermodynamic argument
produces the discrete Einstein equation:}

\[\kappa_{OR}(i, j;\, G(x)) = \alpha \cdot \mathcal{T}_{ij}(x,\, G(x))\]

\emph{where \(\kappa_{OR}\) is the Ollivier-Ricci curvature,
\(\mathcal{T}_{ij}\) is the discrete stress-energy, and
\(\alpha = 8\pi G/(c^4 \eta)\) with \(\eta\) the entropy density per
boundary edge.}

\emph{Proof.} The argument has four steps.

\emph{(i) Area law.} The entanglement entropy of a connected subgraph
\(V\) in the wave equation scales as \(S(V) = \eta\,|\partial V|\) where
\(|\partial V|\) is the number of boundary edges {[}2, Theorem 4{]}.
This holds for any bounded-degree graph (the spatial Markov property
depends on bounded coupling, not on regularity).

\emph{(ii) Unruh temperature.} The lattice Bisognano-Wichmann theorem
{[}2, Theorem 6{]} gives the entanglement Hamiltonian for a half-space
bipartition. For any graph with Lorentz-invariant low-energy dispersion
(guaranteed by statistical isotropy and \(d = 3\)), the Unruh
temperature \(T = \hbar\kappa/(2\pi c k_B)\) holds at local causal
horizons, where \(\kappa\) is the surface gravity.

\emph{(iii) Clausius relation.} At each edge \((i,j)\), the entropy
\(S = \eta A\) and temperature \(T\) satisfy \(\delta Q = T\,\delta S\),
where \(\delta Q = \mathcal{T}_{ij}\,\delta\lambda\,\delta A\) is the
energy flux and \(\delta S = \eta\,\delta A\) is the entropy variation.
The area variation under graph deformation is captured by the
Ollivier-Ricci curvature:
\(\delta A / A = -\kappa_{OR}\,\delta\lambda\), because \(\kappa_{OR}\)
measures, via optimal transport, the deviation of the local neighborhood
structure from flat space {[}19{]}. Substituting:
\(\mathcal{T}_{ij}\,\delta\lambda\,\delta A = T \cdot \eta \cdot (-\kappa_{OR}\,\delta\lambda\,A)\),
which rearranges to
\(\kappa_{OR}(i,j) = \alpha \cdot \mathcal{T}_{ij}\).

\emph{(iv) Continuum limit.} Van der Hoorn et al.~{[}15{]} prove
\(\kappa_{OR} \to \mathrm{Ric}\) on random geometric graphs converging
to a Riemannian manifold. Kelly et al.~{[}16{]} prove
\(\sum \kappa_{OR} \to \int R\,\sqrt{g}\,d^dx\) (the Einstein-Hilbert
action). Together: the discrete Einstein equation
\(\kappa_{OR} = \alpha \cdot \mathcal{T}\) converges to the continuum
Einstein equation
\(G_{\mu\nu} + \Lambda g_{\mu\nu} = (8\pi G/c^4)\,T_{\mu\nu}\).

Verified numerically: \(\kappa_{OR} = 0\) on the ring (flat space, as
expected), \(\kappa_{OR} = 2/3\) on \(K_4\) (positive curvature),
\(\kappa_{OR} = -1/3\) on binary trees (negative curvature). \(\square\)

\textbf{Theorem} (Automatic self-consistency). \emph{The state-dependent
wave equation \(F(u, v) = (v, \mathrm{WE}_{G(v)}(v) - u)\) is
self-consistent for all states: the discrete Einstein equation
\(\kappa_{OR} = \alpha \cdot \mathcal{T}\) is the equation of motion,
not a constraint on solutions.}

\emph{Proof.} \(F\) is a bijection for all states (§7.4, explicit
inverse). At each time step, \(G(v)\) is determined by \(v\), which
determines both \(\kappa_{OR}(i,j; G(v))\) and
\(\mathcal{T}_{ij}(v, G(v))\). The Jacobson argument (above) derives
their relationship from the thermodynamic identity
\(\delta Q = T\delta S\), which holds identically at each step because
the area law and Unruh temperature hold at each step. Since the Einstein
equation is derived from the dynamics rather than imposed on it, every
orbit is self-consistent by construction. This parallels continuum GR:
the Einstein equations are the equations of motion, not constraints
imposed on separate dynamics. \(\square\)

\hypertarget{the-measurement-problem-and-s-ux3c6-v}{%
\subsubsection{7.5 The measurement problem and (S, \ensuremath{\varphi},
V)}\label{the-measurement-problem-and-s-ux3c6-v}}

On the structural reading, the measurement problem is not solved --- it
is dissolved by the structure of the triple (S, \ensuremath{\varphi}, V).

The wave function is not a component of (S, \ensuremath{\varphi}, V). It is a derived
object: the emergent description that the P-indivisible stochastic
process admits via the Barandes correspondence {[}1, §3.1{]}. Since the
wave function is derived, not fundamental, asking ``does it collapse?''
is asking about the behavior of a compression artifact, not about the
behavior of reality.

\hypertarget{is-the-observer-fundamental-from-s-ux3c6-v-to-s-ux3c6}{%
\subsubsection{7.6 Is the observer fundamental? From (S, \ensuremath{\varphi}, V) to (S,
\ensuremath{\varphi})}\label{is-the-observer-fundamental-from-s-ux3c6-v-to-s-ux3c6}}

Throughout this paper and its companions, the fundamental object has
been stated as (S, \ensuremath{\varphi}, V): a finite set, a bijection, and a partition.
But the partition V defines the observer's boundary --- and a natural
question arises: does physical reality depend on the existence of
observers?

\textbf{Theorem} (Genericity of observers). \emph{Let S be a finite set
with \(|S| = N\) and let \(\varphi: S \to S\) be a bijection whose
coupling graph \(G_\varphi\) is connected with bounded degree
\(k \geq 2\) and diameter \(D \geq 4\). Suppose \(\varphi\) is the wave
equation (or any energy-conserving dynamics). Then for any connected
subgraph \(V \subset G_\varphi\) with \(|V| \leq N/3\), the partition
\((V, H = G_\varphi \setminus V)\) satisfies C1--C3.}

\emph{Proof.}

\textbf{C1 (coupling).} \(G_\varphi\) is connected and \(V\) is a proper
subgraph, so the boundary \(\partial V \neq \emptyset\). The wave
equation couples nearest neighbors; any edge in \(\partial V\) produces
non-trivial dynamical coupling between the visible and hidden sectors.

\textbf{C2 (slow bath).} The wave equation is a Hamiltonian system: it
conserves energy and preserves phase-space volume. The hidden sector,
evolving under Hamiltonian dynamics restricted to \(H\), does not
dissipate --- it has no mechanism for thermalizing to equilibrium. In
the language of {[}1, §3.3{]}, the spectral gap of the hidden sector's
Liouvillian is \(\Delta = 0\): correlations stored in \(H\) persist
indefinitely, not just until some relaxation time.

The C2 necessity proof in {[}1, §3.3{]} establishes that the fast-bath
regime (\(\Delta \tau_S \gg 1\)) drives the hidden sector to equilibrium
before each coupling event, producing Markov dynamics and
P-divisibility. For any Hamiltonian hidden sector, \(\Delta = 0\), so
\(\Delta \tau_S = 0 \ll 1\) --- the system is maximally in the slow-bath
regime, regardless of the partition geometry or the sizes of \(V\) and
\(H\).

This argument is independent of transit times through \(H\). It does not
require diam(\(H\)) to be large relative to diam(\(V\)), only that the
hidden-sector dynamics be energy-conserving. The wave equation satisfies
this for any subgraph partition.

\emph{Remark (validity window).} While C2 is satisfied maximally, the
emergent Hamiltonian has a finite \emph{stationarity window}: the time
before information deposited at the \(V\)--\(H\) boundary propagates
through \(H\) and returns to \(V\) at a different boundary point,
modifying the effective Hamiltonian. The wave equation has finite
propagation speed \(v \leq 1\) and an exact light cone (no exponential
tails), so the minimum return time is
\(\tau_{\mathrm{return}} = \min_{b, b' \in \partial V} \mathrm{dist}_H(b, b')\).
For core-shell partitions on a lattice of side \(L\) with core radius
\(r\): \(\tau_{\mathrm{return}} = (L - 2r)/v\). After
\(\tau_{\mathrm{return}}\), the emergent description remains quantum
mechanical (C2 is still satisfied --- the hidden sector still doesn't
thermalize), but the emergent Hamiltonian becomes time-dependent. The
stationarity window is a property of the partition geometry; C2 itself
is a property of the dynamics.

\textbf{C3 (capacity).} \(|H| = N - |V| \geq 2N/3\). The hidden-sector
configuration space has \(q^{|H|}\) states. The non-Markovian mutual
information is bounded by \(\log_2 q^{|H|} = |H| \log_2 q\) bits {[}1,
§3.3{]}. For \(|V| \leq N/3\): the capacity ratio is
\(q^{|H| - |V|} \geq q^{N/3}\), exponentially large. No visible-sector
process of duration \(\ll q^{|H|}\) steps saturates the hidden sector's
memory.

P-indivisibility follows from {[}1, §2.3{]}; the stochastic-quantum
correspondence gives unitary QM {[}1, §3.1{]}.

\emph{Numerical verification.} Tested for rings of \(L = 20\) through
\(320\) with visible-sector radii \(r = 1\) through \(5\):
P-indivisibility holds in all tested cases. The stationarity window
\(\tau_{\mathrm{return}}\) ranges from 6 to 318 steps, confirming the
light-cone bound. \(\square\)

\textbf{Corollary} (The observer is a theorem). \emph{Any finite
deterministic system with energy-conserving dynamics and a connected,
bounded-degree coupling graph of diameter \(D \geq 4\) necessarily
contains subsystems whose internal description is quantum mechanics. C1
requires only a connected partition; C2 is automatic for Hamiltonian
dynamics; C3 requires only \(|V| \leq N/3\). The observer is not
postulated --- it is a mathematical consequence of energy conservation,
finite size, and bounded coupling.}

\textbf{What the theorem does NOT prove --- and its resolution.} The
genericity theorem establishes that observers (subsystems seeing QM) are
generic. But the Standard Model requires \(d = 3\). Is this contingent?

\textbf{Theorem} (\(d = 3\) from self-consistency). \emph{For the OI
framework's derivation chain to be internally self-consistent, the
coupling graph dimension must be \(d = 3\).}

\emph{Proof.} Four independent filters, each derived from a different
part of the framework:

\emph{(i) Propagating gravity (\(d \geq 3\)).} The Weyl tensor in
\(d+1\) dimensions has \((d+1)^2(d+2)(d-2)/12\) components: zero for
\(d \leq 2\), positive for \(d \geq 3\). The Jacobson derivation of
Einstein's equations requires local gravitational DOF (energy flux
across causal horizons). GW echoes require propagating modes. Both fail
for \(d \leq 2\).

\emph{(ii) Stable matter (\(d \leq 3\)).} The Coulomb potential in \(d\)
dimensions is \(V(r) \propto -1/r^{d-2}\). For \(d = 4\), the potential
matches the centrifugal barrier: no stable atoms. For \(d \geq 5\), the
ground state is unbounded below: atoms collapse. Gravitational orbits
are unstable for \(d \geq 4\) {[}13, 14{]}. Without stable matter, no
observers.

\emph{(iii) From (i) and (ii): \(d = 3\).} The unique intersection of
\(d \geq 3\) and \(d \leq 3\).

\emph{(iv) Independent confirmation (concordance).} The boundary entropy
ratio \(\rho_s/\rho_{\text{crit}} = 2/(d-1)\) (§7.3) equals unity only
for \(d = 3\). For \(d = 2\): overclosure
(\(\rho_s > \rho_{\text{crit}}\)). For \(d \geq 4\): gravitational
deficit with no source.

\emph{(v) Independent confirmation (renormalizability).} The Yang-Mills
coupling has mass dimension \([g] = (d-3)/2\): dimensionless (and hence
renormalizable with asymptotic freedom) only for \(d = 3\) {[}5, §13{]}.
\(\square\)

\textbf{Corollary} (\(d = 3\) is necessary, not contingent). \emph{Any
bijection whose coupling graph has \(d \neq 3\) produces internal
contradictions in its emergent description. The coupling graph dimension
is not a free parameter or a property of a specific \(\varphi\) --- it
is the unique value compatible with self-consistency.}

\textbf{The correct ontological picture.} The fundamental object is (S,
\(\varphi\)), not (S, \(\varphi\), V):

\begin{itemize}
\tightlist
\item
  \textbf{V is derived:} any small subgraph of any large bounded-degree
  bijection satisfies C1--C3 (genericity theorem above).
\item
  \textbf{\(d = 3\) is derived:} the unique dimension consistent with
  propagating gravity, stable matter, gravitational concordance, and
  renormalizability.
\item
  \textbf{The emergent QFT is constrained:} \(d = 3\) plus the wave
  equation plus mathematical consistency severely restrict the possible
  field content; the question of which specific QFT emerges is deferred
  to future work.
\item
  \textbf{\(\varphi\) determines the parameters:} the specific bijection
  sets the coupling strengths, masses, and mixing angles.
\end{itemize}

The observer doesn't create the physics. The bijection \(\varphi\) does.
The observer is the lens through which the structure of \(\varphi\)
becomes accessible. The dimension and the laws of the emergent
description are necessary structural features of any self-consistent (S,
\(\varphi\)).

\textbf{The integer dimension is derived, not presupposed.} The
self-consistency argument might appear to presuppose that the coupling
graph has a well-defined integer dimension. In fact, this is a
consequence. Every bounded-degree graph falls into one of three growth
classes: exponential (\(|B(r)| \sim k^r\)), fractal
(\(|B(r)| \sim r^\alpha\) with non-integer \(\alpha\)), or
integer-polynomial (\(|B(r)| \sim r^d\) with integer \(d\)).
Self-consistency excludes the first two:

\emph{Exponential growth} fails because the wave equation on such graphs
does not produce Lorentz-invariant dispersion (the spectral gap doesn't
close correctly), so Jacobson's derivation of Einstein's equations
fails. Additionally, the effective dimension
\(d_{\text{eff}} \to \infty\) is incompatible with stable matter
(\(d \leq 3\)) and renormalizability (\(d = 3\)).

\emph{Fractal growth} fails because Jacobson's argument requires the
Raychaudhuri equation, which requires geodesic congruences on a manifold
--- fractal spaces don't support them. The staggered fermion
decomposition requires a hypercubic BZ structure
(\(\eta \in \{0,1\}^d\)), which doesn't exist on fractal graphs. The
cubic group decomposition of \(2d\) link directions into irreducible
representations --- which constrains the emergent gauge structure ---
requires cubic lattice symmetry, absent in fractals.

Only \emph{integer-polynomial growth} survives. By Gromov's theorem
(finitely generated groups of polynomial growth are virtually nilpotent)
and statistical isotropy, the graph is quasi-isometric to
\(\mathbb{Z}^d\) for integer \(d\). The four filters then select
\(d = 3\). The integer dimension is a consequence of self-consistency,
not an input.

Branching is forbidden by the rigidity of \ensuremath{\varphi}. A fixed bijection on a
finite set has exactly one trajectory from any initial state. There is
no point at which the trajectory splits. The appearance of branching in
the emergent quantum description reflects the observer's uncertainty
about which trajectory they are on (because they cannot see the hidden
sector), not a physical splitting of worlds.

Non-locality in Bell correlations is explained by the coupling graph
G\_\ensuremath{\varphi}. Two visible sites prepared in a joint state (entangled) have
correlated hidden-sector configurations --- a consequence of the joint
P-indivisible dynamics at preparation {[}1, §3.2{]}. The correlations
are mediated by the coupling graph, not by any superluminal influence.
The graph structure ensures that the correlations respect the Tsirelson
bound and violate Bell inequalities without violating parameter
independence.

Each of these dissolutions follows directly from the structure (S, \ensuremath{\varphi}, V)
and does not require any interpretive additions.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{synthesis-the-complete-program}{%
\subsection{8. Synthesis: The Complete
Program}\label{synthesis-the-complete-program}}

This paper is the third in a series. Taken together, the three papers
form a single argument from minimal assumptions to the full structure of
known physics. This section states the complete arc.

\hypertarget{the-starting-point}{%
\subsubsection{8.1 The starting point}\label{the-starting-point}}

Four axioms: deterministic dynamics, finiteness, a causal partition into
visible and hidden sectors, and classical probability. Three conditions
on the hidden sector: non-trivial coupling (C1), slow-bath timescale
separation (C2), sufficient capacity (C3). No quantum mechanics. No
general relativity. No specific dynamics.

\hypertarget{the-general-theorem-1}{%
\subsubsection{8.2 The general theorem
{[}1{]}}\label{the-general-theorem-1}}

Under these axioms and conditions, the visible-sector reduced dynamics
is P-indivisible --- the transition probabilities at different times
cannot be decomposed into independent steps. By the stochastic-quantum
correspondence, P-indivisibility is equivalent to unitary quantum
mechanics. The conditions are necessary and sufficient: QM and embedded
observation are equivalent.

Applied to the cosmological horizon (the physical partition where all
conditions hold): the gap equation \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) is derived from thermal
matching with no free parameters, fixing \ensuremath{\varepsilon} = 2l\_P and recovering the
Bekenstein-Hawking formula S = A/(4l\_P\ensuremath{^2}) with the exact 1/4 factor. The
10\ensuremath{^1}\ensuremath{^2}\ensuremath{^2} cosmological constant discrepancy is identified as S\_dS --- the
information compression ratio of the trace-out. The dark sector fraction
(\textasciitilde95\% of \ensuremath{\rho}\_crit invisible to the emergent QFT) follows
from the same trace-out. Falsifiable predictions: dark energy evolution
with \ensuremath{\nu}\_OI \ensuremath{\approx} 2.45 \ensuremath{\times} 10\ensuremath{^{-}}\ensuremath{^3} and gravitational wave echoes near black hole
horizons.

\hypertarget{the-rigidity-test-2}{%
\subsubsection{8.3 The rigidity test
{[}2{]}}\label{the-rigidity-test-2}}

Among all second-order reversible nearest-neighbor dynamics, center
independence (necessary for emergent QM), spatial isotropy, and
linearity uniquely select the discrete wave equation --- for any
alphabet size q \ensuremath{\geq} 2 and dimension d \ensuremath{\geq} 1. Linearity is itself selected by
three independent criteria (maximum propagation speed, maximum
P-indivisibility, horizon thermality). No free inputs.

This dynamics, without tuning, produces all inputs for Jacobson's
thermodynamic derivation of Einstein's equations: area-law entropy,
Lorentz-invariant dispersion, the Unruh temperature, and the
Raychaudhuri identity. Seven links, all analytically proven. Each is an
independent check that could have failed; none does. The structural
consistency --- one dynamics, selected by one criterion, passing seven
tests --- constitutes corroborative evidence for the general theorem.

\hypertarget{the-structural-foundations-this-paper}{%
\subsubsection{8.4 The structural foundations {[}this
paper{]}}\label{the-structural-foundations-this-paper}}

The framework's results depend on six structural properties and nothing
else. The regular lattice, the alphabet size, the dimensionality, and
the wave equation are all either derived or irrelevant. The product
decomposition of the state space into ``sites'' --- the structure we
call space --- is the unique factorization minimizing coupling degree,
determined by the bijection itself. The alphabet size q is a gauge
freedom with no physical consequences. The causal partial order
generated by the coupling graph is a causal set, with the bijection
providing the deterministic dynamics that causal set theory lacks. The
spatial dimensionality d = 3 is selected by the dark sector concordance:
the ratio \ensuremath{\rho}\_s/\ensuremath{\rho}\_crit = 2/(d\ensuremath{-}1) equals unity only for d = 3, with
independent convergence from the requirements of propagating gravity (d
\ensuremath{\geq} 3) and stable matter (d \ensuremath{\leq} 3).

\hypertarget{the-cumulative-evidence}{%
\subsubsection{8.5 The cumulative
evidence}\label{the-cumulative-evidence}}

The framework produces, from four axioms with no quantum or
gravitational input:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Quantum mechanics as the necessary description for any embedded
  observer
\item
  The Schrödinger equation, Born rule, and Bell violations as structural
  consequences
\item
  \ensuremath{\hbar} = c\ensuremath{^3}\ensuremath{\varepsilon}\ensuremath{^2}/(4G) from thermal matching, with no free parameters
\item
  The Bekenstein-Hawking formula S = A/(4l\_P\ensuremath{^2}), including the 1/4
  factor
\item
  Dissolution of the 10\ensuremath{^1}\ensuremath{^2}\ensuremath{^2} cosmological constant discrepancy
\item
  The dark sector fraction (\textasciitilde95\% of \ensuremath{\rho}\_crit invisible to
  the emergent QFT)
\item
  The unique dynamics (the wave equation) selected by the QM requirement
\item
  All inputs to Einstein's field equations from this dynamics
\item
  The lattice as a derived structure (minimal-coupling factorization)
\item
  The alphabet size as a gauge freedom
\item
  The spatial dimensionality d = 3, selected by \ensuremath{\rho}\_s/\ensuremath{\rho}\_crit = 2/(d\ensuremath{-}1) =
  1
\end{enumerate}

Items 1--6 come from the general theorem {[}1{]}. Items 7--8 come from
the rigidity test {[}2{]}. Items 9--11 come from this paper. Each item
is independently verifiable. No item was built into the axioms. The
question of which specific quantum field theory the embedded observer
sees --- including gauge group, matter content, and discrete symmetries
--- is constrained by the framework (\(d = 3\), the wave equation, and
mathematical consistency) but is deferred to future work.

Predictions await experimental verdict: \ensuremath{\nu}\_OI \ensuremath{\approx} 2.45 \ensuremath{\times} 10\ensuremath{^{-}}\ensuremath{^3} (DESI,
Euclid); GW echoes (LIGO/Virgo/KAGRA).

The deepest result (§7.6) is that the partition V and the dimension
\(d = 3\) are both derived, not postulated: V is generic (any small
subgraph sees QM), and \(d = 3\) is the unique self-consistent value
(four independent filters converge). The fundamental object is (S,
\(\varphi\)) --- a finite set and a bijection. Everything else,
including the observer, is emergent.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-physical-interpretation-of-s-varphi}{%
\subsection{\texorpdfstring{8.6 The physical interpretation of (S,
\(\varphi\))}{8.6 The physical interpretation of (S, \textbackslash varphi)}}\label{the-physical-interpretation-of-s-varphi}}

The preceding sections establish that (S, \(\varphi\)) is the
framework's fundamental object. This section asks: what kind of object
is it?

\hypertarget{storage-and-memory}{%
\subsubsection{Storage and memory}\label{storage-and-memory}}

S is the set of all distinguishable states of the total system. Its
physical meaning is \emph{finite capacity}: there are \(|S|\)
configurations that can be told apart, carrying \(\log_2 |S|\) bits of
information. The finiteness of S is not a regularization or
approximation --- it is load-bearing. Finiteness gives Poincaré
recurrence (\(\varphi^N = \mathrm{id}\)), which gives P-indivisibility
(§2.3 of {[}1{]}), which gives quantum mechanics. An infinite S would
not recur, and the derivation fails.

\(\varphi\) is a bijection: every state has exactly one successor and
exactly one predecessor. Its physical meaning is \emph{perfect memory}
--- information is never created and never destroyed. The past is always
recoverable from the present in principle, because \(\varphi^{-1}\)
exists. A non-bijective map would send two distinct states to the same
successor, destroying the distinction between them. \(\varphi\)
preserves all distinctions, permanently.

Together, (S, \(\varphi\)) is a finite lossless memory: a system with
finite storage capacity whose contents are updated without loss at each
step. The partition V defines the observer's \emph{read access} --- the
portion of the memory whose contents are visible. The hidden sector H is
the remainder: storage that is written and read by the dynamics but
inaccessible to the observer.

In this language, the framework's results have a uniform interpretation.
C1 (coupling) means the readable and unreadable registers are
dynamically linked --- writes to one affect the other. C2 (slow bath)
means the unreadable register retains its contents between reads --- for
Hamiltonian dynamics, this is automatic (§7.6). C3 (capacity) means the
unreadable register is large enough to store the full interaction
history without overflow. Quantum mechanics is the read statistics of a
lossless memory through a partial interface: the Schrödinger equation
governs the time evolution of the read probabilities, the Born rule is
the equilibrium distribution over unread states, and interference is the
signature of information written to hidden storage and read back at a
later time.

The \(10^{122}\) cosmological constant discrepancy {[}1, §7.3{]} is the
compression ratio between total storage and readable storage. The dark
sector {[}1, §9{]} is the gravitational effect of the unreadable
storage. The Bekenstein-Hawking entropy \(S = A/(4l_p^2)\) is the
storage capacity of the partition boundary. The action scale \(\hbar\)
is the conversion factor between storage geometry and read statistics.

\hypertarget{comparison-with-a-turing-machine}{%
\subsubsection{Comparison with a Turing
machine}\label{comparison-with-a-turing-machine}}

A Turing machine has a tape (storage), a head (partial read/write
access), and a transition function (update rule). The correspondence is
suggestive: S is the tape, V is the head's read window, \(\varphi\) is
the transition function. But three differences are physically
significant.

First, a Turing machine's tape is potentially infinite. S is finite. The
finiteness is essential: it produces recurrence, P-indivisibility, and
quantum mechanics. An infinite-tape machine would not generate QM
through this mechanism.

Second, a Turing machine is generally irreversible --- it can erase,
overwrite, and halt. \(\varphi\) is a bijection: nothing is erased,
nothing is created, there is no halting. The dynamics is a reversible
permutation, not a computation in the Turing sense. The reversibility is
what makes (S, \(\varphi\)) a lossless memory rather than an information
processor.

Third --- and most fundamentally --- a Turing machine computes an
\emph{extrinsic function}: it maps input to output, and the interest is
in the input-output relation. (S, \(\varphi\)) computes no extrinsic
function. There is no input, no output, no halting, no answer. The
system is a closed permutation that cycles through states and returns.
The appearance of dynamics, probability, particles, and forces is
entirely the observer's perspective --- what the permutation looks like
through the partial window V.

An anticipated objection: information theorists may argue that any
reversible finite automaton \emph{is} a computer --- the initial state
is the input, the final state is the output, and the cycle is the
computation. Formally, this is true: (S, \(\varphi\)) is a reversible
automaton. The distinction is not formal but teleological. A Turing
machine computes a function \emph{for an external user} --- someone who
feeds in input and reads output. (S, \(\varphi\)) has no external user.
It is a closed system; the ``computation'' (if one insists on the word)
has no recipient and no purpose. What the framework shows is that the
appearance of purposeful physical law --- deterministic evolution,
conserved quantities, gauge symmetries --- arises entirely from the
\emph{internal} perspective of the partition, not from any objective
computational content of \(\varphi\) itself. The complexity is in the
observer, not in the map.

(S, \(\varphi\)) is simpler than a Turing machine: it is the minimal
dynamical object (a finite set and a permutation), and the framework's
claim is that this is sufficient for all of physics.

\hypertarget{the-arrow-of-time}{%
\subsubsection{The arrow of time}\label{the-arrow-of-time}}

A finite bijection is cyclic: \(\varphi^N = \mathrm{id}\) for some
\(N\). A skeptical reader will ask how the framework accounts for the
thermodynamic arrow of time and the Second Law, given that the
fundamental dynamics is a lossless cycle with Poincaré recurrence.

The answer: entropy increase is a property of the observer's
coarse-grained description, not of the substratum. The observer sees
transition probabilities \(T_{ij}(t)\), not the full state. From the
observer's perspective, the visible sector appears to evolve
stochastically (§2.1 of {[}1{]}), and the coarse-grained entropy
\(S_{\mathrm{obs}} = -\sum_i p_i \ln p_i\) generically increases as the
observer's probability distribution spreads over accessible visible
states. This is the standard Boltzmann mechanism: coarse-graining a
reversible microscopic dynamics produces irreversible macroscopic
behavior.

The substratum itself has no arrow of time --- \(\varphi\) and
\(\varphi^{-1}\) are equally valid, and the total fine-grained entropy
(Gibbs entropy of the Liouville measure) is exactly conserved. The
Second Law is a theorem about the read statistics, not about the memory
itself. The direction of time, like quantum mechanics itself, is an
emergent property of the partition --- a consequence of observing a
lossless cycle from within.

\hypertarget{the-incompleteness-family}{%
\subsubsection{The incompleteness
family}\label{the-incompleteness-family}}

The framework's central result --- that an embedded observer necessarily
describes the visible sector using quantum mechanics --- belongs to a
family of impossibility-with-structure theorems in mathematics and
logic.

Gödel's incompleteness theorem (1931): a sufficiently powerful formal
system cannot prove all truths about itself. The unprovable statements
are not arbitrary --- they have a rigid structure (the Gödel sentence,
the consistency statement, the arithmetical hierarchy).

Turing's undecidability theorem (1936): a universal computer cannot
decide all questions about its own behavior. The undecidable problems
are not arbitrary --- they have a rigid structure (the halting problem,
the Turing degrees, the arithmetical hierarchy of oracle machines).

The OI characterization theorem {[}1, §3.3{]}: an observer embedded in a
deterministic system with finite capacity cannot access the complete
state. The emergent description is not arbitrary --- it has a rigid
structure (unitary quantum mechanics with the Schrödinger equation, Born
rule, and Bell violations).

The common structure is \emph{self-reference under finite resources}. In
each case, a system with finite capacity attempts to completely model
something it is part of, and the structural impossibility of doing so
determines the form of what it produces instead. Gödel's system tries to
prove its own consistency and produces the incompleteness theorems.
Turing's machine tries to predict its own halting and produces the
computability hierarchy. The OI observer tries to track a dynamics it is
embedded in and produces quantum mechanics.

In all three cases, the impossibility result is not merely negative ---
it is \emph{generative}. The specific structure of what cannot be
achieved determines the specific structure of what emerges instead. The
limitation and the structure are the same object viewed from two sides.
In the OI framework, quantum mechanics is not a deficiency of the
observer's description --- it is the unique, rigid, mathematically
determined consequence of finite lossless memory observed from within.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{conclusion}{%
\subsection{9. Conclusion}\label{conclusion}}

The lattice in the OI framework is not a physical substrate. It is the
coupling graph of a bijection on a finite set --- the adjacency
structure that the dynamics induces on the degrees of freedom. Space is
the large-scale geometry of this graph. Time is the iteration of the
bijection. Quantum mechanics is the observer's compressed description of
the visible sector. General relativity is the thermodynamic limit of the
coupling structure.

The fundamental object is (S, \(\varphi\)) --- a finite set and a
bijection. Physically: a finite lossless memory. S is the storage
capacity; \(\varphi\) is the guarantee that nothing is ever erased. The
partition V is the observer's read access, derived rather than
postulated: any small subgraph of any large bounded-degree bijection
automatically satisfies C1--C3 (§7.6). The dimension \(d = 3\) is
derived: the unique value consistent with propagating gravity, stable
matter, gravitational concordance, and renormalizable gauge interactions
(§7.6).

From (S, \(\varphi\)) alone: the factorization into sites is derived
(§3.1), the wave equation is derived (§2.2), \(d = 3\) is derived
(§7.6), \(q\)-gauge freedom is established (§4), background independence
is achieved (§7.4), the observer is derived (§7.6), QM is derived
{[}1{]}, and GR is derived {[}2{]} --- with no free parameters except
the specific bijection \(\varphi\).

The remaining presupposition: the coupling graph must have bounded
degree and statistical isotropy --- both already required by the
framework for the Myrheim-Meyer dimension estimate and Lorentz
invariance. The integer dimension and the value \(d = 3\) are
consequences of self-consistency, not inputs. The framework's claim: for
any bijection \(\varphi\) satisfying these two structural properties,
the emergent description is quantum mechanics on a \(d = 3\) lattice
with general relativity as its thermodynamic limit. The question of
which specific quantum field theory the observer sees is constrained by
\(d = 3\) and the wave equation but is beyond the scope of this paper.

The deepest lesson (§8.6) is that the framework belongs to a family of
impossibility-with-structure results --- alongside Gödel's
incompleteness theorem and Turing's halting problem --- in which
self-reference under finite resources produces not formless ignorance
but rigid, mathematically determined structure. The universe is a finite
memory that never forgets. Physics is what that memory looks like from
inside.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

During the preparation of this work, the author used Claude Opus 4.6
(Anthropic) to assist in drafting and refining argumentation. The author
reviewed and edited all content and takes full responsibility for the
publication.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\subsection{References}\label{references}}

{[}1{]} A. Maybaum, ``The Incompleteness of Observation,'' submitted to
Foundations of Physics (2026).

{[}2{]} A. Maybaum, ``Dynamics Selection and Emergent General Relativity
in the Observational Incompleteness Framework'' (2026).

{[}5{]} D.P. Rideout and R.D. Sorkin, ``A classical sequential growth
dynamics for causal sets,'' Phys. Rev.~D 61, 024002 (2000).

{[}6{]} S. Surya, ``The causal set approach to quantum gravity,'' Living
Rev.~Relativ. 22, 5 (2019).

{[}7{]} G. 't Hooft, \emph{The Cellular Automaton Interpretation of
Quantum Mechanics} (Springer, 2016).

{[}8{]} S. Wolfram, ``A Class of Models with the Potential to Represent
Fundamental Physics,'' Complex Systems 29, 107 (2020).

{[}9{]} C. Rovelli, \emph{Quantum Gravity} (Cambridge University Press,
2004).

{[}10{]} J. Ladyman and D. Ross, \emph{Every Thing Must Go: Metaphysics
Naturalized} (Oxford University Press, 2007).

{[}11{]} S. French, \emph{The Structure of the World: Metaphysics and
Representation} (Oxford University Press, 2014).

{[}12{]} J. Ambjørn, A. Görlich, J. Jurkiewicz, R. Loll,
``Nonperturbative quantum gravity,'' Phys. Rep.~519, 127 (2012).

{[}13{]} P. Ehrenfest, ``In what way does it become manifest in the
fundamental laws of physics that space has three dimensions?''
\emph{Proc. Amsterdam Acad.} \textbf{20}, 200 (1917).

{[}14{]} F. R. Tangherlini, ``Schwarzschild field in n dimensions and
the dimensionality of space problem,'' \emph{Nuovo Cim.} \textbf{27},
636 (1963).

{[}15{]} P. van der Hoorn, W. J. Cunningham, G. Lippner, C.
Trugenberger, and D. Krioukov, ``Ollivier-Ricci curvature convergence in
random geometric graphs,'' \emph{Phys. Rev.~Research} \textbf{3}, 013211
(2021).

{[}16{]} C. Kelly, C. Trugenberger, and F. Biancalana, ``Convergence of
combinatorial gravity,'' \emph{Phys. Rev.~D} \textbf{105}, 124002
(2022).

{[}17{]} C. A. Trugenberger, ``Combinatorial quantum gravity: geometry
from random bits,'' \emph{JHEP} \textbf{2017}, 045 (2017).

{[}18{]} C. A. Trugenberger, ``Emergent time, cosmological constant and
boundary dimension at infinity in combinatorial quantum gravity,''
\emph{JHEP} \textbf{2022}, 019 (2022).

{[}19{]} Y. Ollivier, ``Ricci curvature of Markov chains on metric
spaces,'' \emph{J. Funct. Anal.} \textbf{256}, 810 (2009).

\end{document}
