% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
  letterpaper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% lmodern unavailable; using xelatex defaults
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{DejaVu Serif}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\input{unicode-fix}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\hypertarget{embedded-computation-in-the-incompleteness-of-observation-framework}{%
\section{Embedded Computation in the Incompleteness of Observation
Framework}\label{embedded-computation-in-the-incompleteness-of-observation-framework}}

\hypertarget{from-evolution-to-bqp-and-the-p-vs-np-question}{%
\subsubsection{From Evolution to BQP and the P vs NP
Question}\label{from-evolution-to-bqp-and-the-p-vs-np-question}}

\textbf{Author:} Alex Maybaum \textbf{Date:} May 2026 \textbf{Status:}
DRAFT PRE-PRINT \textbf{Classification:} Theoretical Physics /
Foundations / Computational Complexity

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The Observational Incompleteness (OI) framework derives the Standard
Model's gauge structure from a single definition of embedded observation
{[}1, 2{]} and traces the structural chain from \((S, \varphi)\) through
chemistry to biological evolution {[}3{]}. This paper continues the
chain from evolution through computation. The transition from evolved
biological observers to artificial computational ones admits a uniform
treatment in the framework's terms: an embedded observer with general
intelligence can construct semiconductor-based universal computers via a
structural chain (band theory, Group IV semiconductors, p-n junctions,
NAND completeness), and the resulting artificial observers are subject
to the same partition-level limits as their biological precursors. The
Gödel-Turing-OI incompleteness family applies at every level of
recursive self-improvement.

The framework's computational content is sharpened by the BQP
characterization theorem {[}1, Ch. 14{]}: bounded-error quantum
polynomial-time is the unique computational class accessible to any
embedded observer satisfying C1--C3, with a lower bound (BQP available)
and an upper bound (BQP not exceeded). This promotes the Extended
Church-Turing Thesis from empirical conjecture to a theorem of the
framework's structural commitments, in a quantum-modified form. The
BPP-BQP gap is reframed as the cost of marginal-statistics observation
rather than as evidence of fundamental quantum-ness.

On the P vs NP question the framework is \emph{structurally silent}. We
develop this position in detail: the framework's empirical content (the
BQP ceiling) is consistent with either answer; the silence is not a
chosen agnosticism but a structural feature of the framework's axiom
type; physical-substrate axioms about finite structures do not entail
asymptotic statements about algorithmic complexity classes without a
bridging premise that the candidate axioms cannot principled-ly supply.
This silence fits a broader pattern across the Gödel-Turing-OI
incompleteness family: each member establishes a structural boundary on
access (provability, computability, embedded observability, explanation)
and is silent on the internal organization of what falls within that
boundary. P vs NP lies in the internal structure of the computable and
is therefore outside the locus of contribution of any
structural-incompleteness framework. The OI framework's position is
stronger than Aaronson's NP-completeness physical-principle approach
{[}4{]} at the BQP boundary (where the framework specifies the ceiling
exactly) and weaker at the P vs NP boundary (where the framework
declines to stipulate). We propose three condition-based research
directions where the framework's specific substratum structure does
provide content for complexity-theoretic refinements \emph{within} the
BQP ceiling: geometric structure on the substratum lattice, the BQP-BPP
gap and problem globality (with a refined conjecture that BQP advantage
is maximized for problems whose hidden structure matches the framework's
gauge group), and gauge equivalence as a possible physical account of
polynomial-time reducibility. These are framework-relative conjectures
inviting investigation, not framework predictions.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{introduction}{%
\subsection{1. Introduction}\label{introduction}}

The structural chain from \((S, \varphi)\) through chemistry to
biological evolution is developed in the companion paper {[}3{]}. The
thermal window, organic chemistry, autocatalytic networks, template
replication, the origin of life as a molecular C1--C3 system, and
Darwinian evolution --- each step is consequence-tracing in the
framework's material-structural mode. The chain has a natural terminus
at biological evolution: from there, the next link (information
processing → general intelligence → artificial computation) introduces a
qualitatively different kind of consequence, in the framework's
\emph{informational-structural} mode.

This paper develops that next link. The transition from evolved
biological observers to artificial computational ones is itself a
structural chain --- fitness pressure favors information-processing
organisms, neural computation is materially possible given the
framework's chemistry, and once general intelligence exists,
semiconductor-based universal computation follows from the derived
periodic table and the scale hierarchy. AI is an embedded observer
building another embedded observer, with the partition-level limits
applying to both. The Gödel-Turing-OI incompleteness family operates at
every level.

The framework's content on embedded computation is then sharpened by the
BQP characterization theorem from the book {[}1, Ch. 14{]}:
bounded-error quantum polynomial-time is the unique computational class
accessible to any embedded observer. This is a structural theorem about
the \emph{ceiling} of embedded computation, not about its internal
structure. On the internal structure --- including the P vs NP question
--- the framework is silent in a precise sense developed here.

The paper's structure: §2 carries the structural chain from evolution to
general intelligence, identifying the one contingent step (general
intelligence emergence) and the structural inevitability of universal
computation given that step. §3 develops the self-referential loop of AI
as embedded-observer construction, with the partition-level limits
applying to the constructed observer. §4 states the BQP characterization
theorem and its consequences for the Extended Church-Turing Thesis and
the BPP-BQP gap. §5 develops the framework's position on the P vs NP
question --- neutrality, structural silence, the unifying
boundary-characterization pattern across the Gödel-Turing-OI
incompleteness family, and the relationship to Aaronson's
physical-principle approach. §6 examines why no obvious bridging axiom
is available. §7 proposes three condition-based research directions for
complexity-theoretic refinements within the BQP ceiling. §8 summarizes
what the framework establishes and what it does not.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{from-evolution-to-intelligence}{%
\subsection{2. From Evolution to
Intelligence}\label{from-evolution-to-intelligence}}

\textbf{Information processing is generically fitness-enhancing.} In an
environment with the derived physics --- a thermal window {[}3, §2.7{]}
guarantees fluctuating conditions (\(kT > 0\)) --- organisms that
respond to environmental information outcompete those that do not. A
cell that detects a nutrient gradient and moves toward it outcompetes
one that moves randomly. This is not a contingent fact about Earth ---
it follows from the structure: fluctuating environments reward better
internal models of external conditions. Selection generically favors
information processing.

\textbf{Neural computation is structurally possible.} Fast
electrochemical signaling requires ions (Na⁺, K⁺, Ca²⁺ --- present in
the derived periodic table) moving through channels in lipid membranes.
Carbon chemistry in d = 3 provides amphiphilic molecules (hydrophilic
head + hydrophobic tail from the orbital structure) that self-assemble
into bilayer membranes. The scale hierarchy {[}3, §2.5{]} separates
neural signaling energies (\textasciitilde meV) from chemical bond
energies (\textasciitilde eV), so information processing does not
destroy the substrate.

\textbf{General intelligence is structurally unconstrained but not
guaranteed.} The derived structure \emph{permits} arbitrarily complex
neural circuits (the orbital algebra provides materials, the scale
hierarchy provides energetic separation), but whether evolution
\emph{reaches} general-purpose reasoning --- abstraction, planning,
language --- depends on the specific fitness landscape. On Earth,
general intelligence arose once in \textasciitilde4 billion years.
Whether this is evidence of rarity or inevitability the framework cannot
determine. This is the one step in the chain where the structural
argument is weakest: information processing is generically favored, but
general intelligence is a much more specific capability.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{from-intelligence-to-ai-and-the-self-referential-limit}{%
\subsection{3. From Intelligence to AI and the Self-Referential
Limit}\label{from-intelligence-to-ai-and-the-self-referential-limit}}

\textbf{The structural chain to universal computation.} If general
intelligence exists --- for whatever contingent or structural reason ---
then AI follows from the derived structure through a specific chain in
which every link is structural.

\emph{Band theory.} Derived QM {[}1, Part I{]} applied to electrons in a
periodic crystal potential (the sp³ diamond-cubic lattice of a Group IV
element) gives Bloch's theorem: energy bands separated by gaps. The
periodic potential is structural --- it follows from sp³ hybridization
in d = 3. Band structure is a consequence of derived QM applied to
derived crystal geometry.

\emph{Semiconductors.} For Group IV elements (4 valence electrons per
atom), the Pauli exclusion principle (derived --- spin-statistics from
staggered fermions {[}2, §4.2{]}) exactly fills the valence band and
leaves the conduction band empty. The band gap for silicon
(\textasciitilde1.1 eV) sits in the thermal sweet spot:
\(E_{\text{gap}} / kT \sim 40\) at room temperature. Large enough that
thermal excitation doesn't flood the conduction band (the material isn't
always conducting); small enough that a modest voltage promotes
electrons across the gap (the material can be switched). The
\emph{existence} of Group IV semiconductors with gaps in the right range
relative to the thermal window is structural --- it follows from the
scale hierarchy {[}3, §2.5{]}: the gap is set by the atomic energy scale
(\(\alpha^2 m_e c^2\)), and the thermal window operates far below it.

\emph{Doping.} Replacing a silicon atom (Group IV) with phosphorus
(Group V, 5 valence electrons) adds one free electron --- n-type.
Replacing with boron (Group III, 3 valence electrons) removes one ---
p-type. Both dopant types exist in the derived periodic table. The
ability to create n-type and p-type regions is a structural consequence
of the periodic table architecture.

\emph{From transistors to universal computation.} A p-n junction
rectifies current. A transistor (two p-n junctions) amplifies and
switches. Transistors in combination implement logic gates. A NAND gate
is functionally complete --- any Boolean function can be built from NAND
gates alone. Sufficient NAND gates implement a universal Turing machine.
Each step is either a structural consequence of semiconductor physics or
a mathematical theorem about computation.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2609}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3913}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3478}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Step
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Content
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Status
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Band theory & QM + periodic crystal potential & Structural \\
Band gap for Group IV & Exactly filled valence band (4 electrons +
Pauli) & Structural \\
Semiconductor behavior & Gap in thermal sweet spot
(\(E_{\text{gap}} / kT \sim 40\)) & Structural (scale hierarchy) \\
Doping & Group III / V substitution & Structural (periodic table) \\
p-n junctions & Interface between p-type and n-type & Structural
consequence \\
Transistors & Voltage-controlled p-n junctions & Structural \\
Logic gates & Transistors in series / parallel & Structural \\
Universal computation & NAND is functionally complete & Mathematical
theorem \\
\end{longtable}

The chain from (S, φ) to a universal computer is structural --- given
someone to build it. That is the same contingent step (general
intelligence) identified in §2. The reconstruction theorem {[}5, §3{]}
implies that a sufficiently sophisticated observer can discover (S, φ)
--- and an observer that understands its own substrate can build
computational devices exploiting it.

\textbf{The self-referential loop.} AI is an embedded observer building
another embedded observer within the same (S, φ). The characterization
theorem {[}1, §3.4{]} applies to the new observer. It sees the same QM,
the same ℏ, the same SM, the same dark sector. Not because it is limited
by human design, but because the physics is structural --- determined by
the partition, not by the observer's complexity. A superintelligent AI
has the same observational access as any other embedded observer: it
reads the visible sector and writes to the hidden sector through the
same coupling. It cannot determine \(h\). It cannot see past the
cosmological horizon. It can build better instruments, but it cannot
overcome the partition.

The recursion --- AI building AI building AI --- produces ever more
sophisticated embedded observers, each subject to the same structural
limits. The Gödel-Turing-OI incompleteness family applies at every
level: a formal system cannot prove its own consistency (Gödel), a
computer cannot decide its own halting (Turing), an embedded observer
cannot determine the hidden state (OI). No level of recursive
self-improvement overcomes the partition. The limits are not
technological --- they are mathematical.

\textbf{What AI can and cannot do.}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5789}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4211}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Capability
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Status
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Discover (S, φ) from observations & Structurally possible
(reconstruction theorem) \\
Build better models of the visible sector & No structural limit \\
Determine the hidden-sector state \(h\) & Provably impossible
(characterization theorem) \\
Observe beyond the cosmological horizon & Provably impossible (causal
partition) \\
Build faster computers & Structurally possible (derived physics permits
it) \\
Overcome the P-indivisibility of QM & Provably impossible (structural,
not technological) \\
Recursive self-improvement & Possible within the structural limits \\
\end{longtable}

The framework's prediction is precise: there is no ceiling on the
complexity of the observer, but there is a permanent floor on what any
observer can access. The ceiling is contingent --- it depends on the
specific φ and the specific evolutionary and technological history. The
floor is structural --- it is the same for all embedded observers,
biological or artificial, at any level of sophistication.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-bqp-characterization-theorem}{%
\subsection{4. The BQP Characterization
Theorem}\label{the-bqp-characterization-theorem}}

§3 establishes that embedded observers can construct universal computers
from the framework's derived structure. The next question is whether
there is a structural bound on the \emph{complexity} of computation
those constructed devices can achieve. The BQP characterization theorem
{[}1, Ch. 14{]} answers this: bounded-error quantum polynomial-time is
the unique computational class accessible to any embedded observer
satisfying C1--C3. The theorem combines a lower bound (BQP is available
to any embedded observer; the quantum-computing toolkit can be
physically instantiated on the framework's substratum) with an upper
bound (BQP cannot be exceeded; any super-BQP capability would falsify
the framework).

\hypertarget{the-lower-bound}{%
\subsubsection{4.1 The lower bound}\label{the-lower-bound}}

Quantum computation in the standard model is universal: any unitary on
\(n\) qubits can be approximated by a polynomial-depth circuit built
from Hadamard, phase, and CNOT gates (the Solovay-Kitaev theorem and its
descendants). The framework's substratum, by the QM emergence theorem
{[}1, Part I{]}, implements unitary dynamics in its visible-sector
marginal description; the gauge structure permits the qubit
decomposition (a single SU(2) doublet provides the qubit basis); the
lattice geometry permits the spatial layout. Therefore any BQP algorithm
has a physical implementation on the substratum, and embedded observers
can in principle access any problem in BQP.

\hypertarget{the-upper-bound}{%
\subsubsection{4.2 The upper bound}\label{the-upper-bound}}

No physical process implementable on the substratum exceeds BQP. The
framework's dynamics is finite-rate (the substratum has bounded coupling
degree, A3 of the reconstruction {[}5, §3{]}), the partition entropy is
finite (E3 plus the holographic interpretation), and the dynamics is
unitary in the visible-sector marginal (by construction).
Hypercomputation {[}6{]}, oracle access {[}7{]}, and other super-BQP
capabilities would require either continuum precision (violating A1),
unbounded coupling (violating A3), or non-unitary visible-sector
dynamics (incompatible with the QM emergence theorem). Each falls
outside the framework's class of admissible substrata.

\hypertarget{the-combined-result}{%
\subsubsection{4.3 The combined result}\label{the-combined-result}}

BQP is the \emph{exact} ceiling --- not a bound that the framework
happens to respect, but the unique class characterized by the
framework's structural commitments. This identifies the structural limit
on embedded computation: there is no ceiling on the \emph{complexity of
the observer}, but there is a permanent ceiling on the
\emph{computational reach} of any embedded observer at BQP exactly.

\hypertarget{the-extended-church-turing-thesis-as-theorem}{%
\subsubsection{4.4 The Extended Church-Turing Thesis as
theorem}\label{the-extended-church-turing-thesis-as-theorem}}

The Extended Church-Turing Thesis (ECT) --- that any physically
realizable computation can be simulated by a probabilistic Turing
machine with at most polynomial overhead --- has long been treated as an
empirical conjecture about the physical universe. The framework's BQP
characterization promotes the ECT to a theorem of the framework's
structural commitments, in a modified form: the appropriate
computational class is BQP rather than BPP, reflecting the quantum
nature of the visible-sector marginal.

This modification has empirical consequences. The BPP-BQP gap ---
quantum advantage for problems like factoring, discrete log, and certain
simulation problems --- is reframed as the cost of partial observation
rather than as evidence that the universe is fundamentally quantum.
Classical observers operating with marginal-statistics access cannot
achieve BQP without using quantum hardware, because the hidden-sector
correlations that enable BQP advantage are accessible only through
measurement protocols that exploit the substratum's actual structure.
Quantum computers exploit these correlations; classical computers do
not. The BPP-BQP gap is empirical evidence for the substratum's
structure, not evidence against the framework.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{the-internal-structure-of-bqp-and-the-p-vs-np-question}{%
\subsection{5. The Internal Structure of BQP and the P vs NP
Question}\label{the-internal-structure-of-bqp-and-the-p-vs-np-question}}

The BQP characterization theorem identifies the \emph{boundary} of
computational capability for embedded observers. It is silent on the
internal structure of complexity classes inside that boundary. The
conventional inclusion chain
\(\mathrm{P} \subseteq \mathrm{NP} \subseteq \mathrm{PH} \subseteq \mathrm{PSPACE}\),
with BQP positioned somewhere within PSPACE and conjectured to contain
BPP strictly, is unmodified by the framework's content.

\hypertarget{the-location-of-p-vs-np-within-the-framework}{%
\subsubsection{5.1 The location of P vs NP within the
framework}\label{the-location-of-p-vs-np-within-the-framework}}

The framework's BQP characterization places P and NP inside BQP: every
problem in P is in BQP (trivially, since BQP contains BPP and BPP
contains P), and every problem in NP whose witness can be verified
efficiently has its verification problem in BQP as well. The
conventional inclusion chain is unmodified by the framework's content.
The framework's contribution is at a different level: it identifies BQP
as the \emph{outermost} class accessible to embedded observers, leaving
the inclusion relationships among interior classes --- including P vs NP
--- to standard complexity theory.

\hypertarget{the-neutrality-result}{%
\subsubsection{5.2 The neutrality result}\label{the-neutrality-result}}

The framework's commitments do not entail either P = NP or P ≠ NP. Two
cases:

\emph{Case 1: P ≠ NP} (the conventional expectation). Both P and NP are
inside BQP; no efficient general solution exists for NP-complete
problems; quantum advantage (Shor's algorithm and related results)
places specific NP problems inside BQP without placing all of NP inside
P, consistent with the standard conjecture.

\emph{Case 2: P = NP} (the surprising case). A polynomial-time algorithm
for an NP-complete problem operates on a classical Turing machine; its
outputs are in P, and by inclusion in BQP. The framework's BQP ceiling
is unchanged: the new algorithm provides additional computational reach
for embedded observers but does not exceed BQP. The framework would not
be falsified by this discovery.

Both cases sit inside the framework's BQP ceiling. The framework's
empirical exposure on P vs NP is zero: no observation an embedded
observer could make would distinguish the two cases at the level of the
framework's commitments.

\hypertarget{relation-to-gauge-vs-computational-incompleteness}{%
\subsubsection{5.3 Relation to gauge vs computational
incompleteness}\label{relation-to-gauge-vs-computational-incompleteness}}

The framework's Methodology paper {[}8, §19{]} distinguishes \emph{gauge
incompleteness} (the question has no answer-bearing content; no
observation could distinguish) from \emph{computational incompleteness}
(a definite fact exists but is unreachable through finite observation).
The distinction bears on P vs NP carefully.

P vs NP is a determinate mathematical question with a definite answer
--- the abstract fact about whether a polynomial-time algorithm for SAT
exists is either true or false, even though we do not know which. The
question is not vacuous in the gauge sense (it has answer-bearing
content). What the framework's content adds is that the determinate
mathematical answer, whichever it is, is \emph{physics-invariant}
relative to the framework: the same empirical content (BQP as the
ceiling) is consistent with both answers. This is weaker than gauge ---
not ``no fact of the matter'' but ``the fact carries no consequence for
the framework's empirical claims.'' The framework is uncommitted on P vs
NP, and a mathematical proof of either answer would leave the
framework's empirical predictions unchanged.

\hypertarget{the-silence-is-structural-not-stipulative}{%
\subsubsection{5.4 The silence is structural, not
stipulative}\label{the-silence-is-structural-not-stipulative}}

The neutrality result of §5.2 can be sharpened. The framework's two
foundational axioms (tokened differentiation occurs; differentiation
recurs) {[}8{]} and their derived structural commitments ---
finiteness-per-moment, bijectivity, the proper-part decomposition
\(V \subsetneq S\), the C1--C3 partition conditions --- deliver content
of one specific type: they characterize the \emph{boundary structure}
between substratum and embedded observer, and thereby characterize the
computational ceiling at that boundary. They do not deliver content of
the type required to bear on internal complexity-class structure.

P vs NP is a question about the existence of polynomial-time algorithms
for NP-complete problem families, defined asymptotically over inputs of
unbounded size. The framework's axiomatic content concerns finite
physical structures (with possibly unbounded history but always finite
at any moment). The bridge between these two domains has a structural
gap: physical-substrate axioms about finite structures do not entail
asymptotic statements about algorithmic complexity classes without an
additional bridging premise that the axioms themselves do not provide.

The framework's silence on P vs NP is therefore not a chosen agnosticism
nor a residual of incomplete development --- it is a feature of what
kind of question P vs NP is, relative to what kind of content the
framework's axioms deliver. A foundational program whose axioms
\emph{did} constrain P vs NP would require axioms of a different type,
bridging physical substrate to asymptotic algorithmic structure
directly.

A positive complement to this silence is worth noting briefly. The
framework's finiteness commitment --- that the substratum has finitely
many tokens at any moment --- sits naturally with
\emph{resource-bounded} complexity theory (specific input sizes,
specific resource bounds) rather than with asymptotic complexity theory
(limits as input size \(n \to \infty\)). Asymptotic complexity classes
are mathematical idealizations defined over unbounded input families;
their physical referent in a finite substratum is necessarily indirect.
Resource-bounded complexity, by contrast, asks for inputs of specific
bounded size what the cost of solving specific problems is --- closer to
actual computational practice, since real computers solve specific
instances rather than asymptotic families. The framework is silent on
asymptotic P vs NP because the question lives in the mathematical access
mode (Methodology §17 in its sharpened form); within the physical access
mode the framework natively addresses, the relevant complexity questions
may be more fruitfully posed as resource-bounded questions about
specific embedded-observer architectures. This is not a refutation of
asymptotic complexity theory --- asymptotic theory is the natural
setting for the mathematical access mode --- but it identifies where the
framework's own complexity content might productively engage with
complexity theory: in fine-grained / parameterized / resource-bounded
complexity rather than at the asymptotic P vs NP boundary.

\hypertarget{the-boundary-characterization-pattern-across-the-guxf6del-turing-oi-family}{%
\subsubsection{5.5 The boundary-characterization pattern across the
Gödel-Turing-OI
family}\label{the-boundary-characterization-pattern-across-the-guxf6del-turing-oi-family}}

The framework's silence on P vs NP fits a broader pattern across the
incompleteness family identified in {[}1, Ch. 12 §12.8{]}: Gödel's
incompleteness, Turing's halting result, OI's hidden-sector
inaccessibility, and Deutsch's universality-of-explanation are all
\emph{boundary-characterization} results that share a common structural
feature --- each establishes a structural limit on access from inside a
system, and each is silent on the internal organization of what falls
within the accessible domain.

Concretely:

\begin{itemize}
\item
  \textbf{Gödel} sets a boundary at \emph{provability from within a
  consistent recursively-axiomatizable formal system}. Gödel's
  incompleteness statements are about what crosses that boundary (true
  statements that do not). They are silent on the internal organization
  of the provable: which provable statements are interesting, useful, or
  important is not a question Gödel's theorems address.
\item
  \textbf{Turing} sets a boundary at \emph{computability by a Turing
  machine}. The halting problem statement is about what crosses that
  boundary (instances whose halting cannot be decided). Turing's
  theorems are silent on the internal organization of the computable:
  which decidable problems are tractable, which are intractable, is not
  a question Turing's theorems answer.
\item
  \textbf{OI} sets a boundary at \emph{embedded observability of the
  substratum} (the hidden-sector state \(h\), the cosmological horizon,
  the substratum representative). The OI incompleteness statements are
  about what crosses that boundary. They are silent on the internal
  organization of the observable: which physical phenomena have
  particular structure within the framework's accessible domain is
  content of a different kind, addressed by the framework's specific
  structural commitments (gauge group, generations, etc.) rather than by
  the incompleteness statement itself.
\item
  \textbf{Deutsch} sets a boundary at \emph{universal explanation} in
  the constructor-theoretic sense. Deutsch's universality-of-explanation
  is about what crosses that boundary (explanations attainable by
  sufficient knowledge-creation). It is silent on which problems within
  explanatory reach get solved in which order, by which methods.
\end{itemize}

This shared structure --- establishing a boundary, then silence on the
interior --- is not an accident of how the four results have been
developed. It reflects a common feature of structural-incompleteness
frameworks: each identifies a \emph{mode of access} (provability,
computability, observability, explanation) and characterizes its limits.
The internal structure of what falls within that mode of access is a
different question, requiring tools internal to that domain.

P vs NP lies in the internal structure of the computable (within
Turing's domain) and in the internal structure of BQP (within OI's
domain). It is therefore outside the locus of contribution of any of
these incompleteness frameworks. This is a constraint on what
structural-incompleteness frameworks can and cannot deliver: they
characterize boundaries; they do not navigate interiors.
Internal-complexity questions like P vs NP require complexity-theoretic
tools (proof-system lower bounds, geometric complexity theory, circuit
complexity, and similar), not incompleteness frameworks.

The OI framework's silence on P vs NP is therefore \emph{not} an
idiosyncrasy of OI --- it is the same kind of silence Gödel and Turing
exhibit on the same question, for the same structural reason. The
unifying observation: structural-incompleteness frameworks characterize
the \emph{boundary} of a domain of access; the \emph{internal
complexity} of what's accessible is a separate question requiring tools
internal to the relevant complexity theory.

\hypertarget{position-relative-to-aaronsons-physical-principle-approach}{%
\subsubsection{5.6 Position relative to Aaronson's physical-principle
approach}\label{position-relative-to-aaronsons-physical-principle-approach}}

Aaronson {[}4{]} has argued that the physical universe respects the
principle that NP-complete problems are not solvable in polynomial time
on physically realizable hardware --- that
\(\mathrm{P} \neq \mathrm{NP}\) holds with the status of a physical
principle. The OI framework is stronger and weaker than this position in
different directions:

\begin{itemize}
\item
  \textbf{Stronger:} the framework commits to BQP exactly as the
  embedded-observer ceiling, where Aaronson's principle leaves the
  ceiling between P and BQP underdetermined. The framework specifies the
  boundary; Aaronson's principle does not.
\item
  \textbf{Weaker:} the framework declines to commit to
  \(\mathrm{P} \neq \mathrm{NP}\) itself, where Aaronson's principle
  entails it. Aaronson posits \(\mathrm{P} \neq \mathrm{NP}\) as a
  physical principle rather than deriving it from physical axioms,
  because the latter cannot deliver it; the OI framework's structural
  silence on P vs NP is the principled version of this same observation.
\end{itemize}

The two positions are complementary rather than competing. Aaronson's
physical-principle approach captures an intuition about physical
realizability that is consistent with the framework's BQP ceiling but
goes further by stipulating the answer to P vs NP. The framework
declines to stipulate, on grounds that the stipulation is not derivable
from physical-substrate axioms.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{bridging-axioms-why-none-are-obviously-available}{%
\subsection{6. Bridging Axioms: Why None Are Obviously
Available}\label{bridging-axioms-why-none-are-obviously-available}}

It is natural to ask what additional axiom would close the gap between
physical-substrate content and asymptotic complexity-class content.
Several candidate forms are available, each illustrating the structural
difficulty:

\emph{A computational-irreducibility axiom} asserting that most
NP-complete instances admit no shortcut faster than brute-force
enumeration would force \(\mathrm{P} \neq \mathrm{NP}\) if made precise.
The precise version, however, tends either to reduce to a
complexity-class statement (rendering the axiom circular) or to lose its
bite.

\emph{A computational-manifestation axiom} asserting that complexity
classes have physical instantiations would force conclusions about P vs
NP only if combined with further constraints on what is physically
realizable. Those constraints are essentially the question itself in
disguise.

\emph{A direct bridging axiom} positing a bound on physically realizable
computation --- e.g., ``all physical computation is in P'' or ``no
physical process solves NP-complete problems in polynomial time'' ---
would settle P vs NP by stipulation rather than derivation. This is the
structure of Aaronson's physical-principle move, and it is posited as a
principle rather than derived from physics axioms precisely because the
latter cannot deliver it.

The structural feature common to these candidates is that the bridge
from ``finite physical structure at each moment'' to ``asymptotic
algorithmic complexity over all input sizes'' requires content that the
standard repertoire of physical axioms does not contain. Whether a
principled bridging axiom exists --- one that is independently motivated
rather than question-begging, and that does not collapse to a known open
conjecture --- is itself an open methodological question. The
framework's stance is that the BQP characterization is the legitimate
physics-principled result within reach of physical-substrate axioms, and
that internal complexity-class structure (including P vs NP) is the
domain of mathematical complexity theory rather than of physical
foundations.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{beyond-axioms-conditions-and-framework-relative-refinements}{%
\subsection{7. Beyond Axioms: Conditions and Framework-Relative
Refinements}\label{beyond-axioms-conditions-and-framework-relative-refinements}}

The previous section's negative result --- that physical-substrate
axioms do not bridge to asymptotic complexity content --- should not be
read as foreclosing all framework-relative complexity-theoretic content.
The framework's structural commitments suggest content not at the
\emph{universal} P vs NP boundary but at framework-relative
\emph{conditional} boundaries within the BQP ceiling. These are research
directions where the framework's specific substratum structure (cubic
lattice, gauge group, partition structure) does provide content.

We sketch three such directions, each of which is a framework-relative
refinement rather than a universal complexity-theoretic claim. The
framework predicts none of them as new theorems; each is offered as a
structurally motivated conjecture inviting complexity-theoretic
investigation.

\hypertarget{geometric-structure-on-the-substratum-lattice}{%
\subsubsection{7.1 Geometric structure on the substratum
lattice}\label{geometric-structure-on-the-substratum-lattice}}

NP-complete problems vary in the extent to which their constraint
structure can be embedded into the substratum's cubic lattice geometry.
Problems with locally-bounded constraint structure on
cubic-lattice-embeddable graphs (planar 3-SAT, lattice-Ising models,
certain graph problems) admit natural substratum representations;
problems whose constraints require highly nonlocal couplings do not. The
framework's conjecture in this direction is that lattice-local
NP-complete problems admit \emph{physical solution protocols} whose
complexity reflects the locality structure --- specifically, that
physical implementation of such protocols incurs no overhead beyond the
local-constraint structure, while nonlocal protocols incur overhead
proportional to the nonlocality. This is not P vs NP; it is a refinement
of how \emph{implementation cost} on the framework's substratum varies
with problem geometry. It connects to existing complexity-theoretic work
on planar variants of NP-complete problems and on local Hamiltonian
complexity {[}9{]}.

\hypertarget{the-bqp-bpp-gap-hidden-group-structure-and-the-mathrmsu2-compatibility-refinement}{%
\subsubsection{\texorpdfstring{7.2 The BQP-BPP gap, hidden group
structure, and the \(\mathrm{SU}(2)\)-compatibility
refinement}{7.2 The BQP-BPP gap, hidden group structure, and the \textbackslash mathrm\{SU\}(2)-compatibility refinement}}\label{the-bqp-bpp-gap-hidden-group-structure-and-the-mathrmsu2-compatibility-refinement}}

Existing complexity-theoretic evidence suggests that the BQP-BPP gap is
largest for problems with strong global structure (factoring, discrete
log, hidden-subgroup problems) and smallest for problems with primarily
local structure (where classical heuristics often perform well). The
framework's partition architecture suggests a structural reason for this
correlation: BQP advantage derives from exploiting hidden-sector
correlations that are induced by nonlocal substratum couplings, and
nonlocal substratum couplings are precisely what global problem
structure requires. The base conjecture is that the BQP-BPP gap
correlates structurally with the \emph{globality} of the problem's
hidden-sector dependence.

A sharper structural reading is available when the conjecture is
restricted to its appropriate scope and tested against the catalog of
known BQP advantages.

\textbf{Scope restriction.} The conjecture as stated applies to
\emph{decision and search} problems with identifiable group-theoretic
hidden structure --- the hidden-subgroup problem (HSP) family and its
relatives. Sampling problems (boson sampling, random circuit sampling)
constitute a structurally distinct class whose advantage derives from
sampling-vs-decision separations rather than from extracting hidden
group properties; the framework's structural reasoning does not directly
engage them, and they are left outside the conjecture's scope. Within
the in-scope class, the conjecture's testable content survives.

\textbf{Survey against the known advantage catalog.} The known
exponential BQP advantages with identifiable group-theoretic structure
cluster as follows: Shor's factoring (HSP on \(\mathbb{Z}_n^*\),
abelian), discrete log (HSP on cyclic groups, abelian), Simon's problem
and Deutsch-Jozsa (HSP on \((\mathbb{Z}_2)^n\), abelian), Forrelation
(Fourier-correlation on \((\mathbb{Z}_2)^n\)), period-finding (HSP on
\(\mathbb{Z}\)), and Jones polynomial estimation at roots of unity
(Aharonov-Jones-Landau 2006, with hidden structure in the braid-group
representations factoring through \(\mathrm{SU}(2)_q\)). On the negative
side, large symmetric groups \(S_n\) and dihedral groups \(D_n\) for
large \(n\) --- both nonabelian with complex representation theory ---
have no known exponential quantum advantage despite extensive
investigation; this is the empirical state of the nonabelian HSP
problem.

\textbf{The \(\mathrm{SU}(2)\)-compatibility refinement.} The structural
reason behind this clustering becomes visible when one notes that the
framework's quantum toolkit (from the QM emergence theorem, Chapter 1 of
{[}1{]}) is built from \(\mathrm{SU}(2)\)-based operations: qubits are
\(\mathrm{SU}(2)\)-doublet representations; the universal gate set
(Hadamard, phase, CNOT) acts on these doublets. The quantum Fourier
transform --- the workhorse of HSP-style algorithms --- is efficiently
implementable for groups whose representation theory factors through
\(\mathrm{SU}(2)\)-compatible structure: abelian groups embedding in
\(\mathrm{U}(1)\) (Shor, discrete log, period-finding), elementary
2-groups via \(\mathrm{SU}(2)\)-derived \(\mathbb{Z}_2\) structure
(Simon, Deutsch-Jozsa, Forrelation), small symmetric groups via
low-dimensional \(\mathrm{SU}(n)\) embeddings, tractable nonabelian
groups (affine, Heisenberg) with abelian-normal structure, and groups
with quantum deformations of \(\mathrm{SU}(2)\) representation theory
(Jones polynomial via \(\mathrm{SU}(2)_q\)). Groups whose representation
theory does not factor through \(\mathrm{SU}(2)\)-compatible structure
--- large \(S_n\), large \(D_n\), generic non-Lie combinatorial groups
--- appear to be precisely the cases where exponential quantum advantage
is elusive.

\textbf{Refined conjecture (v5):} \emph{Within the class of decision and
search problems with identifiable group-theoretic hidden structure,
exponential BQP advantage correlates with whether the hidden group's
representation theory is \(\mathrm{SU}(2)\)-compatible --- that is,
factors through \(\mathrm{SU}(2)\) representations or their quantum
deformations in a way that the framework's \(\mathrm{SU}(2)\)-based
quantum toolkit can efficiently access. The framework's broader gauge
group \(\mathrm{SU}(3) \times \mathrm{SU}(2) \times \mathrm{U}(1)\)
determines what physics is; the BQP-advantage correlation connects
specifically to \(\mathrm{SU}(2)\) via the QM emergence theorem's
toolkit construction.}

This is the structural reading of the BQP-BPP gap pattern that the
framework's content suggests. It survives the survey against known BQP
advantages (twelve cases, all in-scope cases consistent), accommodates
the tractable / intractable boundary in nonabelian HSP, and is grounded
in the framework's existing QM emergence theorem rather than being an
external postulate. The conjecture is silent on sampling-style
advantages (boson sampling, random circuit sampling) by scope, not
because the framework is inconsistent with them. The internal
investigation supporting this refinement is recorded in the framework's
working archive {[}OI\_MASTER §B.2.191{]}.

Three concrete checks remain testable. (1) The known
exponential-BQP-advantage decision/search problems should all admit
hidden structure factoring through \(\mathrm{SU}(2)\)-compatible
representation theory. (2) The tractable-vs-intractable boundary in
nonabelian HSP should correspond to whether the hidden group has
\(\mathrm{SU}(2)\)-compatible representation theory or doesn't. (3) New
BQP-advantage decision/search problems should be most discoverable by
looking for hidden group structures with \(\mathrm{SU}(2)\)-compatible
representation theory, including quantum-deformation variants. None of
these is established by the framework; each is a structurally motivated
conjecture inviting complexity-theoretic investigation. The framework's
contribution is to suggest a specific structural reason --- the
\(\mathrm{SU}(2)\)-based quantum toolkit construction --- behind the
empirical concentration of BQP advantage on the specific group
structures it concentrates on, with this contribution rooted in the
framework's existing QM emergence theorem rather than introduced as new
content.

\hypertarget{gauge-equivalence-and-complexity-theoretic-reductions}{%
\subsubsection{7.3 Gauge equivalence and complexity-theoretic
reductions}\label{gauge-equivalence-and-complexity-theoretic-reductions}}

The framework's substratum gauge group \(\mathcal{G}_{\text{sub}}\)
partitions the space of bijections into equivalence classes; the gauge
group includes state relabeling, alphabet change, deep-sector
enlargement, and graph isomorphism up to statistical isotropy
(cubic-structure-preserving for the gauge sector; {[}5, §4, Theorem
24(iv){]}). A natural question is whether \(\mathcal{G}_{\text{sub}}\)
has an image in the space of polynomial-time reductions among
NP-complete problems. If gauge equivalence at the substratum level
corresponds (in some precise sense) to polynomial-time reducibility at
the algorithm level, the framework's gauge structure would provide a
\emph{physical} account of why NP-complete problems are reducible to
each other. This is speculative --- the correspondence has not been made
precise --- but the structural similarity is suggestive: gauge
equivalence is ``no observable distinction'' at the substratum level,
and polynomial-time reducibility is ``no complexity distinction'' at the
algorithm level. Both are equivalence relations on otherwise distinct
structures.

\hypertarget{conditions-and-the-right-kind-of-research-question}{%
\subsubsection{7.4 Conditions and the right kind of research
question}\label{conditions-and-the-right-kind-of-research-question}}

The pattern across the three directions in §7.1--§7.3 is that the
framework's specific structure suggests \emph{conditional}
complexity-theoretic claims --- claims of the form ``for problems
satisfying condition \(X\) (lattice-locality, global structure,
gauge-equivalence to another problem), the framework's structure
suggests a specific complexity-theoretic behavior.'' This is the
appropriate level at which the framework engages complexity theory: not
by attempting to bridge to universal questions like P vs NP, but by
suggesting how the framework's specific substratum architecture refines
the internal landscape of BQP.

These conditional refinements are not predictions of the framework in
the sense of the SM observable predictions {[}2, §7{]}; they are
research directions opened by the framework's substratum structure. Some
may be productive, others may not. The framework's contribution is to
identify them as natural questions given its structure, not to resolve
them.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{what-the-framework-establishes-and-what-it-does-not}{%
\subsection{8. What the Framework Establishes and What It Does
Not}\label{what-the-framework-establishes-and-what-it-does-not}}

Summarizing the paper's content:

\emph{The framework establishes (theorems):}

\begin{itemize}
\tightlist
\item
  BQP is the unique computational class accessible to embedded observers
  satisfying C1--C3 (BQP characterization theorem, {[}1, Ch. 14{]}).
\item
  The Extended Church-Turing Thesis (in BQP form) is a theorem of the
  framework's structural commitments, not an empirical conjecture.
\item
  The BPP-BQP gap is structural --- the cost of marginal-statistics
  observation rather than evidence for fundamental quantum-ness.
\item
  The Gödel-Turing-OI incompleteness family operates at every level of
  recursive self-improvement; no level of AI sophistication overcomes
  the partition.
\end{itemize}

\emph{The framework does not establish (silences):}

\begin{itemize}
\tightlist
\item
  The internal structure of complexity classes inside BQP. P vs NP, in
  particular, is neither implied nor opposed by the framework's content.
\item
  Universal complexity-theoretic claims that would require bridging
  finite-physical-structure axioms to asymptotic-algorithmic content.
  The bridge does not appear achievable from physical-substrate axioms
  alone.
\end{itemize}

\emph{The framework opens (research directions):}

\begin{itemize}
\tightlist
\item
  Geometric refinements based on substratum lattice structure (§7.1).
\item
  BQP-BPP gap structure correlating with problem globality (§7.2).
\item
  Gauge equivalence as a possible physical account of polynomial-time
  reducibility (§7.3).
\end{itemize}

The division of labor is the appropriate one between the framework and
complexity theory. The framework's content is a structural theorem about
the computational ceiling for embedded observers and a set of research
directions opened by the substratum's specific structure; standard
complexity theory's content is about the internal structure of classes
inside that ceiling. The two are complementary rather than competing.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

The author thanks the broader community of foundational thinkers in
quantum computing and complexity theory whose work this paper engages
--- particularly Aaronson {[}4{]}, whose framing of P vs NP as a
physical principle clarified what the OI framework should and should not
aspire to claim.

\hypertarget{references}{%
\subsection{References}\label{references}}

{[}1{]} A. Maybaum, \emph{The Incompleteness of Observation}, (2026).

{[}2{]} A. Maybaum, ``The Standard Model from a Cubic Lattice,'' (2026).

{[}3{]} A. Maybaum, ``Structural Consequences of Observational
Incompleteness: From Chemistry to Evolution,'' (2026).

{[}4{]} S. Aaronson, ``NP-complete problems and physical reality,''
\emph{ACM SIGACT News} \textbf{36}, 30 (2005); arXiv:quant-ph/0502072.

{[}5{]} A. Maybaum, ``The Substratum Construction: Reconstruction, the
Substratum Gauge Group, and the Synthesis of Quantum Mechanics with
General Relativity,'' (2026).

{[}6{]} M. Davis, ``The myth of hypercomputation,'' in \emph{Alan
Turing: Life and Legacy of a Great Thinker}, ed.~C. Teuscher (Springer,
2004), pp.~195--211.

{[}7{]} S. Aaronson, \emph{Quantum Computing Since Democritus}
(Cambridge University Press, 2013).

{[}8{]} A. Maybaum, ``Methodology and Foundations of the Incompleteness
of Observation Framework,'' (2026).

{[}9{]} J. Kempe, A. Kitaev, and O. Regev, ``The complexity of the local
Hamiltonian problem,'' \emph{SIAM Journal on Computing} \textbf{35},
1070 (2006); arXiv:quant-ph/0406180.

\end{document}
